\section{Introduction}
\label{sec:intro}

Redistricting ensemble methods have emerged as a principled approach to
evaluating and generating congressional maps. By sampling from the space of
all legally-feasible plans, methods such as ReCom \citep{deford2021recombination}
and GerryChain \citep{duchin2019gerrychain} provide a reference distribution
against which any proposed map can be measured for partisan or racial bias.
The core primitive is a \emph{recombination step}: two adjacent districts are
merged and re-split along a random spanning tree, producing a new feasible plan
in $O(n)$ time per step.

\paragraph{The bipartition failure problem.}
This approach works well when the number of districts $k$ is small or when
$k$ decomposes into small factors. But many states have seat counts that are
large primes or contain large prime factors: Texas has $k = 38 = 2 \times 19$,
Pennsylvania has $k = 17$ (prime), New York has $k = 26 = 2 \times 13$,
and Michigan has $k = 13$ (prime). A direct $k$-way ReCom ensemble on such
states requires, at each step, merging two adjacent districts and re-splitting
the merged subgraph into a balanced bipartition. The correct failure mode is
geometric, not a balance-ratio issue: in a $k=38$ state, merging two adjacent
districts produces a region whose spanning tree must achieve a $1:1$ balance
\emph{within} the merged region, but the surrounding $k-2 = 36$ districts
constrain the boundary so severely that balanced spanning tree cuts become rare.
GerryChain \emph{without pair reselection} stalls in this regime; GerryChain
\emph{with pair reselection} can succeed by trying different adjacent pairs
until a balanced cut is found, at the cost of lower throughput in Python.

\paragraph{BisectionEnsemble: local 2-way ReCom at every node.}
We propose a structural fix: instead of running a $k$-way ensemble at the top
level, embed 2-way ensemble sampling \emph{inside} the bisection tree. The
bisection tree (see ApportionRegions \citep{dellaLibera2026apportion} and
GeoSection \citep{dellaLibera2026geosection}) already decomposes the redistricting
problem into a hierarchy of 2-way cuts. At each node of this tree---which
manages a subregion of $\approx n/\text{level}$ tracts and must split it into
2 balanced parts---we replace the single METIS call with a $T$-step local
ReCom chain seeded by the METIS bisection.

Because every local ensemble is always 2-way at bisection nodes, bipartition
failure at the local ReCom level is structurally impossible: we never ask for
a $1:37$ balance at the full-state level, only ever a $1:1$ balance within each
node's subregion (or near-$1:1$ for prime seat counts using the floor/ceil split).
The bipartition failure guarantee applies to the local ReCom steps at each
bisection node; METIS calls at $p'>2$ ApportionRegions nodes are not covered
by this guarantee. The bisection tree structure is fully preserved, guaranteeing
population balance and contiguity through the same mechanisms as the original
bisection algorithm.

\paragraph{Key advantages.}
\begin{enumerate}
  \item \textbf{Always tractable at bisection nodes}: No bipartition failures
        at the local ReCom level regardless of $k$. TX ($k=38$), PA ($k=17$),
        NY ($k=26$) all succeed at their bisection nodes.
  \item \textbf{Local feasibility sampling}: Each node samples the local
        feasible space of bisections of its subregion. At depth $\ell$, each
        node manages $\approx n/2^\ell$ tracts; the locality advantage grows
        with depth. The root node ($\ell=0$) manages half the state, while
        leaf nodes manage $\approx n/k$ tracts.
  \item \textbf{Parallelisable}: Tree nodes at the same depth are independent;
        each node's ensemble can run on a separate CPU core. For a balanced
        binary tree of depth $d = \log_2 k$, level $\ell$ has $2^\ell$ nodes
        running in parallel.
  \item \textbf{Structure-preserving}: The bisection tree guarantees population
        balance and contiguity at every level, regardless of which bisection cut
        is selected by the ensemble percentile.
\end{enumerate}

\paragraph{Percentile selection and legal posture.}
The percentile parameter $p \in [0, 1]$ selects which bisection cut to use from
the ensemble distribution. At $p = 0.5$ (median), we select the plan at the
median edge-cut rank---the most ``typical'' feasible bisection of the subregion.
At $p = 0.0$ (minimum), we select the minimum edge-cut plan---the most compact
bisection. The legal argument for percentile selection is identical to that of
PercentileSweep \citep{dellaLibera2026percentilesweep}: the choice of $p$ is
a transparent, pre-committed procedure that does not depend on partisan outcomes.

\paragraph{Contributions.}
\begin{enumerate}
  \item \textbf{Algorithm}: BisectionEnsemble---a node-local 2-way ReCom
        ensemble embedded in the bisection tree, with percentile plan selection.
  \item \textbf{Implementation}: \texttt{SeedCompositor::BisectionEnsemble\{p, ensemble\_steps\}}
        in \texttt{redist-cli}, using \texttt{split\_subgraph\_bisection\_ensemble()}.
  \item \textbf{Tractability theorem}: Proof that BisectionEnsemble is always
        tractable (no bipartition failures) for any $k$.
  \item \textbf{Empirical evaluation}: Comparison with standard METIS bisection
        and GerryChain on NC ($k=14$), WI ($k=8$), and TX ($k=38$).
\end{enumerate}
