\section{Algorithm}
\label{sec:algorithm}

\subsection{Notation and Setup}
\label{sec:alg:notation}

Let $G = (V, E, w)$ be the census-tract adjacency graph for a state (or
sub-region), where $V$ is the set of census tracts, $E$ the set of
shared-boundary edges, and $w: E \to \mathbb{R}_{>0}$ the TIGER boundary length.
Let $\text{pop}: V \to \mathbb{Z}_{>0}$ be census population.
Let $k \geq 2$ be the number of districts to produce.
The bisection tree for seat count $k$ is determined by the split prescription
of \citet{dellaLibera2026apportion}: at each node with $k'$ target districts,
the node applies a $p'$-way split where $p' = \text{largest prime factor of } k'$
(or $p' = 2$ for prime $k' > 3$, using floor/ceil targets).

BisectionEnsemble replaces the METIS call at each bisection node (where $p' = 2$)
with a local ReCom ensemble. It is defined by two hyperparameters:
\begin{itemize}
  \item $p \in [0, 1]$: the percentile of the edge-cut distribution to select.
  \item $T \in \mathbb{Z}_{>0}$: the number of ReCom steps to run.
\end{itemize}

\noindent\textbf{Remark on locality.}
The term ``local'' refers to the subgraph managed by a single bisection node,
but its practical meaning depends on tree depth. At depth $\ell$, each node
manages $\approx n/2^\ell$ tracts: the root node ($\ell=0$) manages half the
state, while leaf nodes manage $\approx n/k$ tracts. The locality advantage
--- the reduction in problem size relative to a full-state ensemble --- therefore
grows with depth. At the root, the local ensemble still spans $n/2$ tracts, which
may be a substantial fraction of the state; the full locality benefit is realised
at the leaf level.

\subsection{The BisectionEnsemble Primitive}
\label{sec:alg:primitive}

\begin{definition}[Local bisection ensemble]
  \label{def:be-primitive}
  Given a subgraph $H = (V_H, E_H, w)$ with populations $\text{pop}|_{V_H}$ and
  target district count $k' = 2$ (with population targets $t_1 = \lfloor k/2
  \rfloor$ and $t_2 = \lceil k/2 \rceil$ as fractions of the node's total
  population), \textup{BisectionEnsemble}$(H, p, T)$ proceeds as follows:
  \begin{enumerate}
    \item \textbf{Initialise}: Run METIS 2-way partitioning on $H$ with targets
          $(t_1, t_2)$ to obtain an initial bisection $\pi_0 = (A_0, B_0)$.
    \item \textbf{ReCom chain}: For $t = 1, \ldots, T$:
          \begin{enumerate}
            \item Find a random boundary edge $(u, v) \in E_H$ with
                  $u \in A_{t-1}$, $v \in B_{t-1}$.
            \item Form the merged subgraph $M = H[A_{t-1} \cup B_{t-1}] = H$.
            \item Sample a spanning tree $\mathcal{T} \sim \text{UST}(M)$ via
                  Wilson's algorithm.
            \item Find a cut edge $e^* \in \mathcal{T}$ such that removing $e^*$
                  produces two components $(A', B')$ with
                  $|\text{pop}(A') - t_1 \cdot \text{pop}(H)| \leq \varepsilon$.
            \item If no such $e^*$ exists, reject (keep $\pi_{t-1}$); else
                  accept: $\pi_t = (A', B')$.
          \end{enumerate}
    \item \textbf{Collect}: Let $\Pi = \{\pi_t : \pi_t \text{ was accepted}\}$
          with $|\Pi| = m$. For each $\pi \in \Pi$, compute the edge-cut
          weight $\text{EC}(\pi) = \sum_{(u,v) \in \delta(\pi)} w(u,v)$.
    \item \textbf{Select}: Sort $\Pi$ by $\text{EC}(\cdot)$. Return
          $\pi^* = \Pi[\lfloor p \cdot m \rfloor]$ (zero-indexed).
          If $m = 0$ (no plans accepted within $T$ steps), return the METIS
          initialization $\pi_0$ as a fallback. By Theorem~\ref{thm:tractable},
          $m = 0$ has probability zero in the limit; this fallback applies only
          for small $T$ where the chain may not yet have encountered a balanced cut.
  \end{enumerate}
\end{definition}

\noindent\textbf{Remark on step 2(c).} In BisectionEnsemble, the full merge
region $M$ is simply the subgraph $H$ itself (since we are splitting $H$ into
exactly 2 parts). The spanning tree is thus sampled over the complete local
subgraph at each step. This differs from a general $k$-way ReCom step where
only two of the $k$ districts are merged; here, the entire node region is
always the merge domain.

\subsection{Integration with the Bisection Tree}
\label{sec:alg:tree}

Algorithm~\ref{alg:bisection-ensemble} gives the full recursive procedure.
BisectionEnsemble replaces every 2-way METIS call in the bisection tree.
For $p'$-way splits with $p' > 2$ (ApportionRegions nodes), standard METIS
is used (BisectionEnsemble is not applicable to $p' > 2$ nodes since those
are not bisections). A pure binary tree (all seat counts $k = 2^d$) applies
BisectionEnsemble at every node.

\begin{algorithm}[h]
\caption{BisectionEnsemble redistricting}
\label{alg:bisection-ensemble}
\begin{algorithmic}[1]
\Require Subgraph $H$, seat count $k$, percentile $p$, steps $T$
\Ensure Set of $k$ district subgraphs
\If{$k = 1$}
  \Return $\{H\}$
\EndIf
\State $(p', k_1, \ldots, k_{p'}) \leftarrow \text{split}(k)$ \Comment{Per AR split prescription}
\If{$p' = 2$} \Comment{Bisection node: use local ensemble}
  \State $(A, B) \leftarrow \textup{BisectionEnsemble}(H, p, T)$
\Else \Comment{$p' > 2$ node: use METIS directly}
  \State $(R_1, \ldots, R_{p'}) \leftarrow \text{METIS}(H, p')$
\EndIf
\For{each sub-region $R_i$ with target $k_i$}
  \State $D_i \leftarrow \textup{BisectionEnsemble-Redist}(R_i, k_i, p, T)$
\EndFor
\Return $\bigcup_i D_i$
\end{algorithmic}
\end{algorithm}

\subsection{Key Theoretical Properties}
\label{sec:alg:theory}

\begin{theorem}[Tractability]
  \label{thm:tractable}
  BisectionEnsemble never encounters a bipartition failure. That is, for any
  local subgraph $H$ and any $T \geq 1$, the ReCom chain in Definition~\ref{def:be-primitive}
  is well-defined and will produce at least one accepted plan within
  $O(|V_H|)$ steps in expectation.
\end{theorem}

\begin{proof}
  We establish the result in three parts.

  \textbf{Part 1: Every spanning tree of $H$ contains a balanced cut edge.}
  Let $\mathcal{T}$ be any spanning tree of $H = (V_H, E_H, w)$.
  Root $\mathcal{T}$ at an arbitrary vertex $r$.
  For each edge $e \in \mathcal{T}$, let $A_e \subseteq V_H$ denote the
  component not containing $r$ when $e$ is removed, and let $B_e = V_H
  \setminus A_e$.  As $e$ ranges over the $|V_H| - 1$ tree edges ordered by
  a DFS traversal, $\text{pop}(A_e)$ increases from $\text{pop}(v_{\min})$
  (a single leaf) to $\text{pop}(H) - \text{pop}(v_r)$ (all but the
  root's component), with consecutive values differing by at most
  $\max_{v \in V_H} \text{pop}(v)$.
  By a discrete intermediate-value argument, there exists an edge $e^*$ such
  that $|\text{pop}(A_{e^*}) - t_1 \cdot \text{pop}(H)| \leq \max_{v}
  \text{pop}(v)$.
  Since individual census tracts represent $\ll 0.5\%$ of any node's
  population in all experimental states (NC, WI, TX), the tolerance
  $\varepsilon$ specified in Definition~\ref{def:be-primitive} is met.
  Hence \emph{every} spanning tree of $H$ has at least one balanced cut edge.

  \textbf{Part 2: The balanced spanning tree is sampled with positive probability.}
  Wilson's algorithm \citep{wilson1996generating} samples exactly from the
  uniform spanning tree (UST) distribution on $H$.  The UST assigns
  \emph{positive probability} to every spanning tree of a connected graph:
  for any spanning tree $\mathcal{T}^*$, $\Pr[\text{UST} = \mathcal{T}^*]
  \propto \prod_{e \in \mathcal{T}^*} w(e) > 0$ (since all edge weights are
  positive).  Therefore, the spanning tree established in Part~1 is sampled
  with positive probability at each ReCom step.

  \textbf{Part 3: The chain accepts in finite expected steps.}
  Because a balanced cut edge exists in every spanning tree (Part~1) and every
  spanning tree is sampled with positive probability (Part~2), each ReCom step
  produces a balanced bisection with probability bounded away from zero.
  Formally, letting $q > 0$ denote the minimum acceptance probability over all
  states of the chain, the number of steps until the first acceptance is
  stochastically dominated by a geometric random variable with success
  probability $q$, so the expected number of steps to first acceptance is
  $1/q < \infty$.  The expected number of steps to accumulate $m \geq 1$
  accepted plans is $m/q = O(|V_H|)$ under the average-case cover-time
  assumption for the UST sampler.

  \emph{Note on ``existence of acceptance'' vs.\ ``per-step acceptance
  probability.''} Part~1 establishes that acceptance is always possible
  (existence); Part~2--3 establish that it occurs in finite expected time
  (positive probability). These are logically distinct: it would be possible
  for a balanced cut to exist yet be sampled with probability zero if the
  UST distribution were not fully supported, but the full support of UST
  (Part~2) rules this out.
\end{proof}

\begin{proposition}[Contiguity preservation]
  \label{prop:contiguity}
  Every plan returned by BisectionEnsemble is contiguous. Specifically, every
  district subgraph produced at every tree node is a connected induced subgraph
  of the census-tract adjacency graph $G$.
\end{proposition}

\begin{proof}
  The spanning tree $\mathcal{T}$ of the merge domain $M = H$ is a connected tree.
  Removing any single edge $e^*$ from $\mathcal{T}$ produces exactly two
  connected subtrees $(A, B)$, both of which are connected. Since $A$ and $B$
  are subgraphs of $H \subseteq G$, they are connected induced subgraphs of $G$.
  Inductively, all sub-regions passed to recursive calls are connected, and the
  final districts are connected.
\end{proof}

\begin{proposition}[Complexity]
  \label{prop:complexity}
  For a bisection tree of depth $d = O(\log k)$ and a state graph with $n$ tracts,
  BisectionEnsemble runs in $O(T \cdot n \log n)$ total time across all tree nodes.
\end{proposition}

\begin{proof}
  At depth $\ell$ of a binary tree, there are $2^\ell$ nodes, each managing a
  subgraph of $\approx n / 2^\ell$ tracts. Each node runs $T$ ReCom steps.
  Wilson's algorithm samples a uniform spanning tree in $O(|V_H|)$ expected
  time for graphs with good expansion properties (e.g., constant spectral gap)
  \citep{wilson1996generating}. For near-path-like subgraphs (rural regions),
  worst-case expected time is $O(|V_H|^2)$; the bound $O(|V_H|)$ should be
  interpreted as an average-case statement over typical redistricting subgraphs.
  The cost per node is $O(T \cdot n / 2^\ell)$ under the average-case assumption.
  Summing over all $d$ levels:
  \[
    \sum_{\ell=0}^{d-1} 2^\ell \cdot O\!\left(\frac{Tn}{2^\ell}\right)
    = \sum_{\ell=0}^{d-1} O(Tn)
    = O(Tn \cdot d)
    = O(Tn \log k).
  \]
  Since $n \log k \leq n \log n$, the total time is $O(T n \log n)$.
\end{proof}

\subsection{Parallelism}
\label{sec:alg:parallel}

Tree nodes at the same depth are independent: each node's ensemble depends
only on the subgraph it inherits from its parent, which is fixed once the
parent's bisection is chosen. This gives two natural parallelism strategies:
\begin{itemize}
  \item \textbf{Level-parallel}: Execute all nodes at depth $\ell$ in parallel,
        then proceed to depth $\ell + 1$. Requires synchronisation at level
        boundaries but achieves $2^\ell$-fold parallelism at level $\ell$.
  \item \textbf{Asynchronous}: Begin all nodes immediately; sub-regions start
        as soon as their parent's bisection is selected. This is a pipeline
        over the tree depth with $d$ concurrent stages.
\end{itemize}
The \texttt{redist-cli} implementation uses Rayon's work-stealing thread pool,
which naturally implements the asynchronous strategy: nodes at deeper levels
are spawned as Rayon tasks as soon as their parent completes.
