\section{Conclusion}
\label{sec:conclusion}

We have introduced BisectionEnsemble, a redistricting algorithm that integrates
local 2-way ReCom ensemble sampling into the bisection tree. By repositioning
ensemble sampling from the full plan level to the individual node level, we
resolve the bipartition failure problem that prevents direct GerryChain
(without pair reselection) from operating on states with large seat counts
($k \geq 17$). Every local ensemble is always 2-way at bisection nodes, so
bipartition failure at the local ReCom level is structurally impossible
regardless of the state's seat count $k$. The guarantee applies to 2-way
bisection nodes; $p'>2$ ApportionRegions nodes continue to use METIS directly.

The algorithm is implemented as \texttt{SeedCompositor::BisectionEnsemble\{p,
ensemble\_steps\}} in \texttt{redist-cli}, using
\texttt{split\_subgraph\_bisection\_ensemble()} which seeds a METIS bisection,
runs \texttt{ensemble\_steps} ReCom steps via
\texttt{redist\_ensemble::recom::RecomChain}, and returns the plan at rank
$\lfloor p \times \text{accepted\_count} \rfloor$.

Empirically, BisectionEnsemble at $T=100$ steps improves Polsby-Popper
compactness by 5--12\% over standard METIS bisection without systematically
shifting partisan outcomes, and succeeds on Texas ($k=38$) where direct
GerryChain fails. Runtime scales as $O(Tn\log n)$ and is trivially parallelisable
across tree nodes.

\paragraph{Scope and limitations.}
BisectionEnsemble is presented here as an algorithmic contribution with partisan
neutrality properties. It does not address the Voting Rights Act (Section~2)
dimension of redistricting: the algorithm has no minority-community-preservation
objective, and a BisectionEnsemble plan must be evaluated for VRA compliance
separately, using tools such as \texttt{redist analyze --types bloc-voting}
\citep{dellaLibera2026callais}. Practitioners deploying BisectionEnsemble in
jurisdictions with majority-minority district requirements should verify VRA
compliance of the final plan as a post-processing step.

\paragraph{Reproducibility.}
The \texttt{redist-cli} implementation used in the experiments is available
in the project repository. Exact replication requires the 2020 Census
P.L.~94-171 redistricting data and the VEST 2020 presidential precinct returns
\citep{vest2020}. Phase~2 will provide a complete replication archive including
per-run seeds, intermediate outputs, and criterion.rs benchmark results.

\paragraph{Future work.}
Several directions merit investigation. First, the ReCom chain's mixing time
on local subgraphs is unexplored: how many steps $T$ are needed for convergence
to the stationary distribution? Second, BisectionEnsemble currently applies only
to bisection nodes ($p' = 2$); extending it to $p'$-way nodes with $p' > 2$
(via nested bisections or $p'$-way recombination) would enable full ensemble
coverage of all ApportionRegions tree nodes. Third, the sensitivity of compactness
gains to the ReCom proposal distribution (e.g., biased spanning trees favouring
compact cuts) is a natural direction for further improvement. Finally, a formal
ergodicity proof for the local chain over the feasible bisection space of each
node would strengthen the theoretical foundation.
