% Section 1: Introduction (700 words)

Congressional redistricting in the United States operates under a fundamental constraint: district boundaries cannot cross state lines. This requirement, rooted in Article I, Section 2 of the U.S. Constitution, mandates that ``Representatives...shall be apportioned among the several States,'' establishing states as the organizing unit for congressional representation. For over two centuries, this federalist structure has shaped the geographic configuration of all 435 House districts, with each state subdividing its allocated seats through independent redistricting processes.

While the prohibition on cross-state districts is typically understood as a legal and political necessity, its geometric consequences remain largely unexplored. Does constraining districts to respect state boundaries impose a meaningful cost on compactness---the widely accepted principle that districts should be reasonably compact rather than irregularly shaped? Or do state boundaries, by providing natural partitioning at appropriate population scales, actually facilitate more compact district configurations?

This paper addresses these questions through a computational experiment: we apply graph partitioning algorithms to redistrict all 435 congressional seats at the national level, treating the entire United States as a single graph of 84,414 census tracts with no regard for state boundaries. By comparing the resulting national optimization to state-based redistricting, we quantify the geometric trade-offs inherent in the federalist structure of American representation.

Our findings challenge the intuition that geographic constraints harm compactness. We discover that national optimization---which allows districts to freely cross state lines---produces districts that are \textbf{42.5\% less compact} than state-based redistricting (Polsby-Popper scores of 0.265 vs 0.461). This substantial difference suggests that state boundaries, far from being geometric obstacles, serve as beneficial constraints that improve overall district compactness. The result holds consistently across a boundary crossing penalty ablation study ($ \beta \in [0, 10] $), with all penalty values producing worse compactness than state-based redistricting.

The mechanism driving this counterintuitive finding relates to population distribution and geometric scale. States provide subdivision of the national population into units of manageable size, each containing between 1 and 52 districts. This multi-scale structure allows the partitioning algorithm to work at more appropriate population densities and geographic regions, producing more regular shapes. National optimization, by contrast, forces the algorithm to balance population constraints across vastly different geographic regions---urban coastal areas, rural plains, mountainous terrain---leading to elongated districts that stretch across diverse topographies to meet population targets.

A secondary finding emerges from our analysis of cross-state districts: 180 of 435 districts (41.4\%) cross state boundaries in the national optimization, with most districts drawing a mean of 74.4\% of their population from a single ``primary'' state. This pattern suggests the possibility of \textit{fractional representation}, where states would receive non-integer allocations of congressional seats based on population. While such an arrangement would require constitutional amendment (Article I mandates apportionment ``among the several States''), it represents a theoretically coherent alternative to the current system for addressing small-state inequities.

Importantly, the ablation study reveals that intermediate boundary crossing penalties do not improve outcomes. The relationship between penalty strength and cross-state districts is non-monotonic, peaking at $ \beta = 1.0 $ (187 cross-state districts) rather than decreasing monotonically. This result indicates that small penalties cannot selectively allow ``good'' boundary crossings while preventing ``bad'' ones; the optimization landscape contains multiple local optima that preclude simple threshold-based solutions.

This work contributes to three literatures. First, it extends research on redistricting algorithms by demonstrating that geographic constraints can improve rather than harm optimization objectives. Second, it provides empirical evidence relevant to normative debates about federalism and representation, quantifying the geometric costs of alternative institutional structures. Third, it introduces a methodological framework---national-scale graph partitioning with boundary crossing penalties---that can be applied to other multi-jurisdictional redistricting problems, from state legislative districts to international electoral boundary commissions.

The remainder of the paper proceeds as follows. Section 2 reviews the constitutional and historical context of state-based redistricting. Section 3 describes our methodology, including the graph partitioning algorithm, data sources, and experimental design. Section 4 presents results from the national optimization, compactness analysis, cross-state district characterization, and ablation study. Section 5 discusses implications for redistricting theory, federalism debates, and future research directions. Section 6 concludes.
