% Section 2: Constitutional and Historical Context (500 words)

The requirement that congressional districts respect state boundaries derives from Article I, Section 2 of the U.S. Constitution: ``Representatives and direct Taxes shall be apportioned among the several States which may be included within this Union, according to their respective Numbers.'' This language establishes states---not individuals, regions, or the nation as a whole---as the fundamental unit of congressional apportionment. Each state receives an integer number of seats based on its decennial census population, and districts within each state are then drawn through independent redistricting processes.

This state-based structure reflects the federalist compromise at the Constitutional Convention. The framers sought to balance representation of individuals (the House) with representation of states as sovereign political entities (the Senate). By apportioning House seats ``among the several States,'' they ensured that states would remain salient units of political organization even in the directly elected chamber. James Madison argued in Federalist No. 62 that this federal structure would ``protect the people against their rulers'' by maintaining multiple centers of political authority.

Historical practice reinforces the constitutional mandate. Since the first apportionment in 1790, all 435 congressional districts have respected state boundaries, producing 118 decennial redistricting cycles without exception. The Apportionment Act of 1911 fixed the House size at 435 members and codified the state-based allocation method, while the 1967 Voting Rights Act reinforced single-member districts but maintained state-level control of redistricting. Even in states with only one at-large representative, the state boundary serves as the district boundary.

The prohibition on cross-state districts has occasionally generated controversy in cases where population distribution creates inequities. Vermont and Wyoming, for instance, each receive one representative despite populations well below the national average district size (143,000 in Wyoming vs 761,000 nationally in 2020). Proposals for fractional representation---where states could share districts or receive non-integer allocations---have been discussed since the 19th century but never implemented. Legal scholars generally agree such arrangements would require constitutional amendment, as the text explicitly requires apportionment ``among'' rather than ``across'' or ``without regard to'' state boundaries.

The federalist structure imposes geometric constraints on redistricting. Each state must partition its territory into its allocated number of districts, forcing some states (California with 52 seats, Texas with 38) to subdivide dense populations while others (Alaska, Delaware) consist of single at-large districts. State boundaries create ``hard'' partition boundaries that algorithms cannot cross, potentially limiting the compactness and population balance achievable through optimization.

Yet the counterfactual remains unexplored: What would congressional districts look like if we removed the state boundary constraint? Would national optimization produce meaningfully more compact districts, or do state boundaries actually facilitate better geometric outcomes? This paper addresses these questions through computational experimentation, comparing national and state-based redistricting to quantify the geometric trade-offs of federalism.

Understanding these trade-offs matters for normative debates about representation and reform. If state boundaries substantially harm geometric compactness, this cost should factor into discussions of constitutional amendment. Conversely, if state boundaries improve compactness, this finding strengthens the geometric case for federalist redistricting structures. More broadly, the results inform our understanding of how institutional constraints interact with optimization objectives in multi-jurisdictional redistricting problems.
