Graph partitioning algorithms for congressional redistricting face a fundamental architectural choice: should states be divided directly into $k$ districts (n-way partitioning) or hierarchically through repeated binary splits (recursive bisection)? While both approaches build on the same METIS optimization core \citep{karypis1998}, they differ in optimization structure, parameter sensitivity, and computational characteristics. This paper provides the first comprehensive empirical comparison of these methods for Voting Rights Act (VRA) compliance.

\subsection{The Architectural Choice}

\textbf{N-way partitioning} performs direct $k$-way division, partitioning the state graph into exactly $k$ congressional districts in a single optimization. Edge weights influence all $k$ districts simultaneously, creating a global optimization landscape where minority concentration in one district automatically affects all others. This "one-shot" approach offers computational efficiency but may struggle with complex constraint satisfaction.

\textbf{Recursive bisection} employs hierarchical decomposition, repeatedly splitting into two parts until reaching $k$ districts. Each binary split is a simpler 2-way optimization, potentially more stable than complex $k$-way problems. However, early splitting decisions cascade through the tree—a suboptimal early split propagates to all descendants. Edge weights influence only local binary decisions, not global $k$-district structure.

These architectural differences suggest potential performance trade-offs. N-way's global view might better preserve minority communities across all districts but require longer computation. Recursive's local optimizations might execute faster but fragment minorities through cascading errors. Prior work has assumed these trade-offs are substantial, but no systematic comparison exists.

\subsection{Voting Rights Act Compliance Challenge}

The Voting Rights Act prohibits states from diluting minority voting strength through redistricting \citep{thornburg1986}. Section 2 requires creation of majority-minority (MM) districts—where minority voters constitute $\geq 50\%$ of voting-age population—when feasible \citep{johnson2001}. Achieving VRA compliance algorithmically requires preserving geographically dispersed minority communities without explicit racial gerrymandering.

Edge-weighted graph partitioning offers a solution: assign higher weights $w_{ij} = \alpha$ to edges connecting minority-dense census tracts (minority percentage $\geq \tau$), making such edges "expensive" to cut. METIS minimizes total edge cut weight, thereby concentrating minority populations into contiguous districts. Two parameters control performance:

\begin{itemize}
\item \textbf{Weight factor} ($\alpha$): Multiplier for minority-minority edges (tested: 1$\times$, 5$\times$, 10$\times$, 50$\times$, 100$\times$)
\item \textbf{Minority threshold} ($\tau$): Percentage defining "minority tract" (tested: 40\%, 45\%, 50\%, 55\%)
\end{itemize}

Edge weights are:
\begin{equation}
w_{ij} = \begin{cases}
\alpha & \text{if } m_i \geq \tau \text{ and } m_j \geq \tau \\
1 & \text{otherwise}
\end{cases}
\end{equation}
where $m_i$ is tract $i$'s minority percentage (total non-White population).

The question: do n-way and recursive bisection respond differently to edge-weighting? Does architecture affect VRA performance?

\subsection{Research Questions}

This paper addresses four questions through comprehensive 50-state empirical analysis:

\textbf{RQ1: Overall Performance} — Do n-way and recursive bisection achieve equivalent VRA compliance rates, or does one method systematically outperform?

\textbf{RQ2: Parameter Sensitivity} — Do methods exhibit different optimal configurations? Does n-way require different weight factors or thresholds than recursive?

\textbf{RQ3: State-Specific Behavior} — Do certain state characteristics (minority percentage, geographic clustering, district count) favor one method?

\textbf{RQ4: Speed-Quality Trade-offs} — Does faster method (if any) sacrifice VRA performance? What practical implications guide method selection?

\subsection{Preview of Key Findings}

Our comprehensive ablation study (1,760 redistricting runs across 44 states, 20 configurations per method per state) yields surprising results:

\textbf{Finding 1: Statistical Equivalence} — Both methods achieve nearly identical success rates (n-way: 47.5\%, recursive: 48.3\%, difference: 0.8 percentage points). Paired t-test shows no significant difference ($t=-0.48$, $p=0.634$), with negligible effect size (Cohen's $d=-0.018$). The architectural choice does not substantially impact VRA compliance.

\textbf{Finding 2: Speed vs Optimization Trade-off} — N-way executes 67.5\% faster (3.68s vs 11.33s average). Recursive achieves higher performance ceiling with optimal parameters (56.8\% vs 52.3\% best configuration). Choice depends on priority: speed (n-way) or maximum tuning potential (recursive).

\textbf{Finding 3: Different Parameter Landscapes} — Methods prefer different configurations. N-way optimizes at lower thresholds (40\%) with moderate weights (50$\times$). Recursive optimizes at higher thresholds (50\%) with heavy weights (100$\times$). This suggests different optimization mechanisms despite shared METIS core.

\textbf{Finding 4: State-Specific Advantages} — Neither method dominates state-by-state. N-way wins 5 states (largest: Virginia +35\%), recursive wins 8 states (largest: Connecticut +45\%), 31 states tie. State characteristics (minority dispersion, urban clustering) interact with method architecture.

\textbf{Finding 5: Bimodal Success Distribution} — Both methods show bimodal state-level performance: states either succeed consistently (100\% across configurations) or fail consistently (0\% across configurations). Intermediate success rates are rare. This indicates VRA compliance depends primarily on state demographics/geography, not method choice.

\subsection{Implications for Redistricting Practice}

These findings transform method selection guidance. Previous work assumed architectural differences create substantial performance gaps, leading to complex decision frameworks. Our equivalence result simplifies: \textit{both methods work equally well for VRA compliance when properly parameterized}.

Practical recommendation: Use n-way for rapid prototyping, ensemble generation, and exploratory analysis where speed matters. Use recursive for carefully tuned single plans where maximizing MM district count justifies longer computation. Both approaches are viable for production redistricting.

The different parameter sensitivities suggest practitioners should tune each method independently rather than assuming universal optimal configurations. N-way's preference for $\tau=0.40$ and recursive's preference for $\tau=0.50$ reflect underlying optimization dynamics—n-way benefits from broader minority definition, recursive from sharper minority boundaries.

\subsection{Contributions}

This paper makes four contributions:

\textbf{(1) First Comprehensive Comparison} — Systematic evaluation of n-way vs recursive bisection across all 50 states with extensive parameter ablation (1,760 runs). Establishes empirical equivalence with statistical rigor.

\textbf{(2) Parameter Sensitivity Analysis} — Identifies distinct optimal configurations for each method. Shows weight factor and threshold preferences differ by architecture, providing tuning guidance.

\textbf{(3) State-Level Characterization} — Maps which state characteristics predict method advantages. Identifies Virginia-type states (n-way better) and Connecticut-type states (recursive better).

\textbf{(4) Practical Decision Framework} — Provides speed-quality trade-off analysis guiding method selection based on computational constraints and performance requirements.

\subsection{Paper Organization}

Section 2 describes edge-weighted graph partitioning methodology, parameter configurations, and experimental design. Section 3 presents head-to-head comparison results: overall success rates, statistical significance, runtime analysis. Section 4 analyzes parameter sensitivity, identifying optimal configurations for each method. Section 5 examines state-by-state results, identifying geographic patterns and method-specific advantages. Section 6 discusses implications for redistricting practice and theoretical insights into optimization architectures. Section 7 concludes with recommendations for method selection.
