% Direct County Representation
% State Politics & Policy Quarterly

\documentclass[12pt]{article}

% Packages
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{natbib}

% Title
\title{Direct County Representation: An Alternative to Congressional Redistricting}
\author{Giovanni Della-Libera}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Congressional redistricting allows state legislatures to fragment large counties across multiple districts, destroying local political identity for strategic purposes. We propose county-based representation: counties with populations exceeding twice the ideal district size (1.52 million for 2020) receive autonomous representation via Huntington-Hill apportionment. Analyzing all 3,142 U.S. counties, we find 25 qualify (70 million people, 21\% of U.S.), currently spanning 92 congressional districts. Critically, county fragmentation is bipartisan: 16 qualifying counties are in Democratic-controlled states (46M people), 7 in Republican-controlled states (21M people). Democratic legislatures fragment their own urban strongholds---Los Angeles, Chicago, New York City---demonstrating this reflects state control over local representation, not partisan manipulation. County-based representation achieves comparable equality to the current system (CV = 0.100 vs. 0.080) while preserving natural community boundaries. Voting Rights Act analysis reveals substantial benefits for minority representation: the 25 qualifying counties could support 197-201 majority-minority districts through county-internal redistricting, compared to 68 under current fragmented boundaries---a nearly threefold increase. This occurs because large urban counties have extremely high minority concentrations (40-88\%), and county-based representation eliminates state-level cracking of these populations while providing local control over redistricting. Though currently advantaging Democrats (+15-25 seats) due to urban concentration, we argue the principle of local autonomy and VRA compliance transcends partisan effects, similar to how ``one person, one vote'' advantaged Democrats but was constitutionally required. This analysis provides the first rigorous assessment of county-based federal representation and establishes that redistricting fragments natural communities unnecessarily while diluting minority voting power.
\end{abstract}

\section{Introduction}
\input{sections/01-introduction}

\section{Historical Context}
\input{sections/02-historical-context}

\section{Methodology}
\input{sections/03-methodology}

\section{Results}
\input{sections/04-results}

\section{Discussion}
\input{sections/05-discussion}

\input{sections/07-vra-analysis}

\section{Conclusion}
\input{sections/06-conclusion}

\bibliographystyle{apalike}
\bibliography{../references}

\end{document}
