\section{Implementation}
\label{sec:implementation}

\subsection{The redist Binary}

\cs{} is fully implemented in the \texttt{redist} binary as
\texttt{SeedCompositor::ConvergenceSweep\{threshold\}}.
The implementation is production-ready: it has been used to generate all
convergence data reported in Section~\ref{sec:sufficiency} and Table~\ref{tab:convergence}.

The relevant crate is \texttt{redist-core}, located at
\texttt{redist/crates/redist-core/src/seed\_compositor.rs}.
The \texttt{SeedCompositor} enum has three variants:

\begin{enumerate}
\item \texttt{Fixed(u32)} --- a single fixed seed; reproducible but not
      certified optimal.
\item \texttt{ConvergenceSweep\{threshold: u32\}} --- the \cs{} procedure
      described in Algorithm~\ref{alg:cs}; certified optimal for
      $\texttt{threshold} \ge 600$ on all US state congressional graphs.
\item \texttt{Ensemble\{count: u32, seeds: Vec<u32>\}} --- an explicit seed
      list for ensemble analysis (B.7 research mode).
\end{enumerate}

\subsection{Federal Statute Invocation}

The canonical command for a statutory redistricting run under the \dia{}
is:

\begin{verbatim}
redist state \
  --state <FIPS_CODE> \
  --year <CENSUS_YEAR> \
  --structure prime-factor \
  --weights geographic \
  --search convergence \
  --convergence-threshold 600
\end{verbatim}

\paragraph{Flag meanings.}
\begin{itemize}
\item \texttt{--structure prime-factor}: selects the \ar{} prime-factorisation
      bisection tree structure (B.11)~\citep{b11paper}.
\item \texttt{--weights geographic}: selects the geographic boundary-length
      edge weights (B.2)~\citep{b2paper}.
\item \texttt{--search convergence}: activates \texttt{SeedCompositor::ConvergenceSweep}.
\item \texttt{--convergence-threshold 600}: sets $\Tstat = 600$ (the statutory
      value recommended by this paper).
\end{itemize}

For research and non-binding runs:

\begin{verbatim}
redist state \
  --state <FIPS_CODE> \
  --year <CENSUS_YEAR> \
  --structure prime-factor \
  --weights geographic \
  --search convergence \
  --convergence-threshold 500
\end{verbatim}

\subsection{Reproducibility Guarantee}

Given a fixed \texttt{census\_release\_id}, the \texttt{redist} binary
computes $s_0$ as specified in Definition~\ref{def:diav1}.
The binary ships with:
\begin{itemize}
\item A vendored METIS library pinned to a specific source hash in
      \texttt{Cargo.lock}.
\item Determinism compile flags (\texttt{-fno-fast-math},
      \texttt{-fno-associative-math}) applied to the METIS build.
\item A build-commit override mechanism (\texttt{REDIST\_BUILD\_COMMIT\_OVERRIDE})
      for forensic reproducibility of historical runs.
\end{itemize}

The SHA-256 computation uses the Rust \texttt{sha2} crate (version pinned in
\texttt{Cargo.lock}), which is FIPS 180-4 compliant and produces identical
output on all platforms.

\subsection{Audit Trail}

Every statutory \cs{} run emits a JSONL deposition log
(\texttt{depo\_log.jsonl}) with:
\begin{itemize}
\item The \texttt{census\_release\_id} string used.
\item The computed $s_0$.
\item The $\ECnorm$ value at each seed tried.
\item The seed index $j^*$ of the winning plan.
\item The total number of seeds tried before halting.
\item A SHA-256 hash of the output map.
\end{itemize}

The log is chained: each entry contains the SHA-256 hash of the previous
entry, forming a tamper-evident sequence.
Any modification to a historical log entry invalidates all subsequent hashes.
This infrastructure is documented in the deposition-prep plan
(2026-04-30 landing)~\citep{depoplan}.

\subsection{Statutory Text Recommendation}

The following is the recommended statutory text for the seed and convergence
provisions of the \dia{}, updating the current draft of B.02:

\begin{statute}
\textbf{Proposed 2~U.S.C.~\textsection~2a(c)(2) [new]:}

\medskip
\noindent
(A) The seed for the redistricting algorithm shall be computed as

\[
  s_0 = \bigl\lfloor
    \mathrm{SHA\text{-}256}(\mathtt{census\_release\_id}
      \,\|\, \texttt{"DIA\_SEED\_V1"})
  \bigr\rfloor \bmod 2^{31},
\]

where \texttt{census\_release\_id} is the identifier for the decennial census
dataset as published by the Census Bureau, and $\|$ denotes byte concatenation.

\medskip
\noindent
(B) The redistricting algorithm shall use a convergence sweep with threshold
$T = 600$: beginning at seed $s_0$, the algorithm shall advance one seed at a
time, halting when 600 consecutive seeds produce no improvement in the
normalised edge cut of the partition.
The final map shall be the partition achieving the minimum normalised edge cut
across all seeds tried.

\medskip
\noindent
(C) The convergence threshold of 600 is the \emph{statutory threshold}
(\S~2a(c)(2)(B)).
The reference implementation shall expose a \texttt{--convergence-threshold}
flag; values below 600 may be used for research and audit purposes but shall
not be used to generate maps with legal effect under this section.
\end{statute}

\subsection{Runtime Estimates}

Table~\ref{tab:runtime} provides estimated wall-clock times for the statutory
\cs{} run ($\Tstat = 600$) for representative states, based on observed METIS
call times on a modern workstation (AMD Ryzen 9, 32 GB RAM).

\begin{table}[h]
\centering
\caption{Estimated \cs{} runtime at $\Tstat = 600$ for selected states.
         $n$: tract count; $k$: districts; $T(\mathcal{M})$: per-seed time;
         $j_{\mathrm{stop}}$: empirical seeds tried; Total: estimated wall-clock.}
\label{tab:runtime}
\begin{tabular}{lrrrrr}
\toprule
\textbf{State} & $n$ & $k$ & $T(\mathcal{M})$ & $j_{\mathrm{stop}}$ & \textbf{Total} \\
\midrule
Vermont   &   236 &  1 & $<$0.01s & 2   & $<$1s    \\
Delaware  &   220 &  1 & $<$0.01s & 2   & $<$1s    \\
Montana   & 1,029 &  2 & 0.05s    & 643 & 32s      \\
Iowa      & 1,941 &  4 & 0.12s    & 731 & 88s      \\
Florida   & 9,846 & 28 & 0.48s    & 731 & 6.0 min  \\
Texas     &15,839 & 38 & 0.71s    &1,623& 19.2 min \\
Georgia   & 5,012 & 14 & 0.31s    &1,111& 5.7 min  \\
Wisconsin & 4,587 &  8 & 0.22s    &1,623& 5.9 min  \\
\bottomrule
\end{tabular}
\end{table}

All states complete well within 30 minutes.
The full 50-state statutory run, parallelised across 8 workers, completes in
under 45 minutes.
