\section{The Huntington-Hill Precedent}

Congressional redistricting is not the first time American democracy has confronted the need for algorithmic objectivity. For 150 years, Congress struggled with an analogous problem: apportioning House seats among states based on decennial census counts. From 1790 to 1941, every census triggered political battles over which mathematical method to use, with each formula favoring different states. Congress changed the apportionment method six times, always to advantage particular political coalitions~\cite{balinski1982fair}.

In 1941, Congress permanently adopted the Huntington-Hill method of equal proportions, a mathematical formula that has allocated House seats for eight decades without controversy~\cite{huntington1928methods}. The method succeeded not because it produces ``optimal'' outcomes---mathematicians identify five reasonable apportionment formulas, each defensible---but because it removes human judgment from the process. Once Congress set the House size at 435 members, the Huntington-Hill formula mechanically determined each state's allocation. No partisan actor could manipulate the result because no human discretion remained.

This precedent suggests a pathway for redistricting. Just as apportionment answered ``How many seats per state?'' with mathematical objectivity, redistricting might answer ``Where to draw district boundaries?'' through algorithmic methods. The parallel is instructive:

\begin{center}
\begin{tabular}{ll}
\textbf{Apportionment (solved)} & \textbf{Redistricting (unsolved)} \\
How many seats per state? & Where to draw boundaries? \\
Mathematical formula (1941) & Partisan negotiation \\
Universal acceptance & Persistent conflict \\
\end{tabular}
\end{center}

If mathematical governance resolved congressional apportionment, why should redistricting remain hostage to partisan manipulation? The question is whether algorithms can provide similar structural objectivity for boundary design.
