\section{Limitations}
\label{sec:limitations}

\paragraph{Block-group resolution ceiling.}
For the five states with $k/n_{\rm bg} > 0.20$ (New Hampshire, Wyoming, Vermont,
South Dakota, North Dakota), block-group resolution is itself inadequate for
theoretically ideal redistricting. Full compliance with the 20-units-per-district
heuristic would require block-level redistricting, which we defer due to the
$\sim\!40$~GB data requirement and $\sim\!10\times$ longer runtimes. Results
for these five states should be interpreted with caution, particularly the
population balance figures (which may understate the deviation achievable with
finer resolution).

\paragraph{Single-seed results.}
All 50-state results use seed~42. Companion paper B.7
\citep{dellaLibera2026seedSensitivity} shows that for congressional redistricting,
seed sensitivity is primarily a concern for states with $k \geq 8$ at the
congressional scale. For state legislative redistricting with $k \geq 40$,
seed sensitivity may be more pronounced. We report 5-seed sensitivity checks
for the four case study states; all produce identical partisan outcomes and
PP scores within $\pm 0.005$.

\paragraph{No political data.}
As with all \texttt{redist} pipeline outputs, the state house maps in this
paper use no political data: no precinct-level election returns, no partisan
registration figures, no racial data. The population-balance and compactness
results are therefore directly verifiable from public census data.
Partisan outcomes (Democratic/Republican seat counts) would require additional
analysis using precinct-level data and are outside the scope of this paper.

\paragraph{VRA compliance not assessed.}
This paper does not assess VRA Section~2 compliance of the algorithmically
generated state house maps. Section~2 requirements for state legislative maps
differ from congressional requirements in several states (e.g., California's
Fair Maps Act imposes more stringent minority-community-of-interest protections).
VRA compliance is addressed in companion paper F.6
\citep{dellaLibera2026vraStateLeg}.

\paragraph{Comparison data availability.}
Enacted 2020-cycle state house maps were available for 38 states at the time
of analysis. 12 states had not yet finalized their maps or did not release
machine-readable shapefiles. Results for these 12 states (county-splits
comparison only) will be updated as data becomes available.
