\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}[proposition]
\newtheorem{lemma}{Lemma}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}

\title{Prime-Factored Maps:\\
       Redistricting as the Geographic Completion of Huntington-Hill}

\author{Gio Della-Libera}

\date{Draft --- \today}

\begin{document}

\maketitle

\begin{abstract}
The U.S. Constitution allocates congressional seats to states via the
Huntington-Hill apportionment method (Article~I, \S2), but treats the drawing
of district boundaries as a separate ``Manner of election'' question (Article~I,
\S4). We propose collapsing this distinction. Given that a state receives $n$
seats, we construct its congressional map by factoring $n$ into primes and
applying those factors hierarchically as $k$-way geographic partitions of the
state's census-tract graph. The resulting \emph{Prime-Factored Redistricting}
(PFR) algorithm has two inputs: the prime factorization of the seat count
(arithmetic) and the minimum-cut partition of the geographic graph (TIGER
adjacency). All prime-$p$ splits of a given state are computed once and reused
across every seat count sharing that prime factor. When $n$ changes after
reapportionment, only the levels of the factorization tree affected by the new
prime structure require recomputation. The map is uniquely determined by
census data and number theory --- no political inputs, no interpretive degrees
of freedom. We prove that PFR is constitutional under Article~I~\S2 directly,
evaluate the algorithm on all 50 states, and characterize the striking
discontinuities that occur when consecutive seat counts have disjoint prime
factorizations (e.g., California's 51\,=\,3\,$\times$\,17 and
52\,=\,$2^2\times 13$ maps share no structural elements).
\end{abstract}

\input{sections/01-introduction}
\input{sections/02-related-work}
\input{sections/03-methodology}
\input{sections/04-evaluation}
\input{sections/05-discussion}
\input{sections/06-conclusion}

\bibliographystyle{plainnat}
\bibliography{references,../../references}

\end{document}
