\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{array}
\usepackage{setspace}

\onehalfspacing

\title{E.7: Six Systems, One Constraint:\\
       Design Lessons from Algorithmic Alternative Representation}

\author{Gio Della-Libera}

\date{Draft --- \today}

\begin{document}

\maketitle

\begin{abstract}
Algorithmic redistricting research has largely focused on optimizing the
drawing of single-member districts.  This paper reports what we learn
about representation design more broadly by comparing six algorithmically
implemented alternatives to the standard single-member geographic district:
multi-member districts (E.1), county-based representation (E.2), national
redistricting without state lines (E.3), partisan-similarity clustering
(E.4), partisan-fairness seat allocation (E.5), and international
applications (E.6).  Five design lessons emerge from systematic comparison.
First, geography is destiny for single-member districts: the Rodden sorting
gap produces systematic underrepresentation of Democrats regardless of the
drawing method, algorithmic or otherwise.  Second, multi-member districts
partially but not fully escape geographic sorting: three-member districts
reduce the gap by approximately 40\%, five-member by 65\%.  Third,
eliminating districts entirely (county representation) trades proportionality
in competitive states for a fractional-voting mechanism that is institutionally
unfamiliar and constitutionally problematic.  Fourth, crossing state lines
produces geographically compact but politically fragmented districts: the
national map achieves 12\% better mean Polsby--Popper compactness but
preserves 40\% fewer counties.  Fifth, the compactness--proportionality
Pareto frontier is steep: gaining one percentage point of proportionality
costs approximately 0.015 Polsby--Popper units, and no feasible
single-election geographic system escapes this constraint.  These lessons
imply a design hierarchy: for single-member systems, the constraint hierarchy
(population $\to$ contiguity $\to$ compactness) is optimal under geographic
sorting; multi-member districts are the only structural escape available
within current constitutional constraints.
\end{abstract}

\input{sections/01-introduction}
\input{sections/02-lesson1}
\input{sections/03-lesson2}
\input{sections/04-lesson3}
\input{sections/05-lesson4}
\input{sections/06-lesson5}
\input{sections/07-synthesis}
\input{sections/08-conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
