\section{Background}
\label{sec:background}

\subsection{Graph Minimum Bisection and Computational Hardness}

Let $G = (V, E, w)$ be a connected weighted graph with $n = |V|$ vertices and
edge weights $w : E \to \mathbb{R}_{>0}$. A \emph{$k$-partition} assigns every
vertex to one of $k$ parts; the \emph{edge-cut} of a partition is the total
weight of edges crossing part boundaries.

Graph minimum $k$-bisection (equal-weight, $k=2$) is NP-hard even on planar
graphs~\citep{garey1979computers}. METIS~\citep{karypis1998fast} approximates
the minimum using multilevel coarsening and local refinement. The random seed
controls the initial coarsened-graph bisection; refinement proceeds
deterministically given the initial partition, so different seeds yield
different local optima.

\subsection{Theoretical Lower Bounds on Minimum Edge-Cut}
\label{sec:background:bounds}

Two complementary bounds relate graph structure to minimum edge-cut:

\paragraph{Cheeger constant.}
The \emph{isoperimetric number} (Cheeger constant) of $G$ is:
\[
  h(G) = \min_{S \subseteq V,\, |S| \leq n/2}
           \frac{\sum_{(u,v)\in E,\, u\in S,\, v\notin S} w_{uv}}{|S|}
\]
Any bisection with $|S| = n/2$ satisfies
$\mathrm{EC}(S) \geq h(G) \cdot n/2$, giving a lower bound independent of the
algorithm.

\paragraph{Fiedler bound (algebraic connectivity).}
The second eigenvalue $\lambda_2$ of the weighted graph Laplacian $L = D - W$
(where $D_{ii} = \sum_j w_{ij}$ and $W_{ij} = w_{ij}$) is the \emph{algebraic
connectivity}~\citep{fiedler1973algebraic}. By the discrete Cheeger inequality:
\[
  h(G) \geq \lambda_2 / 2,
  \qquad\text{so}\qquad
  \mathrm{EC}_{\min} \geq \frac{\lambda_2 \cdot n}{4}.
\]
$\lambda_2$ can be computed in $O(n^2)$ or approximated in $O(n \log n)$ via
Lanczos iteration. For US census-tract graphs with $n \approx 200$--$4{,}000$
vertices, exact computation takes milliseconds.

\paragraph{Geometric scaling law.}
For geographically compact states, a dimensional-analysis argument predicts:
\[
  \mathrm{EC}_{\min}(k) \approx C_s \cdot \sqrt{k \cdot A_s}
\]
where $A_s$ is the state land area (m$^2$), $k$ is the district count, and
$C_s$ is a state-specific constant capturing geographic compactness.
Intuition: $k$ equal-area districts each have radius $r \sim \sqrt{A_s / (k\pi)}$
and perimeter $\sim 2\pi r$; internal boundaries are shared, so total internal
boundary $\sim k \cdot \pi r = \sqrt{\pi k A_s}$. The $\sqrt{k}$ scaling means
adding more districts increases edge-cut sublinearly. We test this empirically
in \S\ref{sec:eval:national}.

\subsection{Polsby-Popper Compactness}

The Polsby-Popper score~\citep{polsby1991third} of a region $R$ is:
\[
  \mathrm{PP}(R) = \frac{4\pi A(R)}{P(R)^2} \in (0, 1]
\]
where $A(R)$ is area and $P(R)$ is perimeter; PP = 1 iff $R$ is a circle.
Courts have accepted PP as a neutral compactness criterion for redistricting
(\emph{Karcher v. Daggett}, 462 U.S. 725 (1983)).

Polsby-Popper and edge-cut are related but not identical. The perimeter of a
district $d$ has two components:
\[
  P(d) = \underbrace{P_{\mathrm{ext}}(d)}_{\text{external (fixed)}}
        + \underbrace{P_{\mathrm{int}}(d)}_{\text{internal (controlled by partition)}}.
\]
Edge-cut minimisation controls only $P_{\mathrm{int}}$ (the shared boundaries
between districts). $P_{\mathrm{ext}}$ — coastlines, state borders, rivers —
is determined by geography and is independent of the partition. For coastal
states (Maryland, Florida, California), $P_{\mathrm{ext}}$ dominates and
edge-cut minimisation can diverge from PP maximisation. This motivates
\textsc{CompactBisect}'s use of the full PP formula rather than edge-cut as
the selection criterion.

\subsection{The Huntington-Hill Apportionment Method}

The Huntington-Hill equal-proportions method (2~U.S.C.~\S~2a) assigns seats to
states using the geometric mean $\sqrt{n(n+1)}$ as the critical divisor. A
state with $n$ seats receives the next seat if its \emph{priority value}
$P_i / \sqrt{n_i(n_i+1)}$ (where $P_i$ is population) is the largest among
all states. The geometric mean is used because it treats proportional
differences in representation equally: the proportional increase in
representation when going from $n$ to $n+1$ seats, $1/n - 1/(n+1) = 1/(n(n+1))$,
is equalised across all $n$ by the geometric-mean divisor.

\textsc{CompactBisect}'s selection criterion $\sqrt{\mathrm{PP}(L) \cdot
\mathrm{PP}(R)}$ applies the identical principle to bisection fairness: choose
the split where the proportional compactness difference between the two halves
is minimised. The geometric mean is maximised when $\mathrm{PP}(L) =
\mathrm{PP}(R)$, just as Huntington-Hill achieves equal proportional
representation. Congress adopted the geometric mean in 1941 as the uniquely
neutral apportionment standard; we apply the same criterion one level down to
the line-drawing problem.

\subsection{MCMC Ensemble Methods}

Ensemble redistricting methods~\citep{chikina2017assessing, herschlag2020quantifying}
generate large samples of neutral plans and use the resulting distribution as
a benchmark against which to evaluate enacted maps. \emph{GerryChain}
\citep{duchin2019gerrychain} implements a Markov-chain walk over the space of
valid partitions. These methods answer ``is the enacted map an outlier?'';
\textsc{CompactBisect} addresses a complementary question: ``which specific
neutral plan is geometrically optimal?'' The two approaches are compatible:
the ensemble establishes the distribution; CompactBisect identifies the plan
at the geometric optimum of that distribution.
