\section{Methodology}
\label{sec:methodology}

\subsection{The Solution Space and Why Seeds Matter}
\label{sec:methodology:space}

Graph minimum $k$-bisection is NP-hard. METIS finds a local optimum via
multilevel coarsening and refinement; the random seed controls the initial
partitioning from which refinement proceeds. Critically, \emph{every seed
produces a distinct final partition} in the high-dimensional partition space
--- we verified this empirically for Pennsylvania across 100 seeds (100 unique
assignment hashes, no two identical). The solution space is therefore dense and
continuous, not a small collection of canonical boundaries.

The partisan outcome of a plan is quantized by the integer seat constraint:
with $k$ districts and statewide Dem share $v$, the achievable proportionality
gaps are $\{100(j/k - v) : j = 0, 1, \ldots, k\}$. This quantization explains
why gap values appear to cluster --- different boundaries can produce the same
integer seat count --- while the underlying plan space is effectively infinite.

\subsection{The Minimum-Edge-Cut Protocol}
\label{sec:methodology:protocol}

Rather than report results for a single arbitrary seed, we propose the
\emph{minimum-edge-cut (MEC) protocol}: run $N$ seeds, compute the total
boundary length (edge-cut) of each resulting partition, and publish the
plan with the smallest edge-cut. Formally, for state $s$ and seed set
$\mathcal{S} = \{1, \ldots, N\}$:

\[
  \pi^*_s = \arg\min_{\pi \in \{\pi_i : i \in \mathcal{S}\}} \mathrm{EC}(\pi)
\]

where $\mathrm{EC}(\pi) = \sum_{(u,v) \in E,\, \pi(u) \neq \pi(v)} w_{uv}$
is the total TIGER boundary length (in metres) of cut edges. This protocol is:
\begin{itemize}
  \item \textbf{Deterministic given $\mathcal{S}$}: publishing the full seed
        set allows anyone to reproduce $\pi^*_s$ exactly.
  \item \textbf{Monotone}: adding seeds can only decrease or maintain
        $\mathrm{EC}(\pi^*_s)$.
  \item \textbf{Parti\-san-blind}: no partisan data enters the selection
        criterion; the choice is made entirely on geometric grounds.
  \item \textbf{Convergence-testable}: we can measure how quickly the minimum
        stabilises across seeds, and declare ``enough seeds'' when the
        probability of improvement drops below a calibrated threshold.
\end{itemize}

\subsection{Convergence Analysis}
\label{sec:methodology:convergence}

Let $M_n = \min_{i \leq n} \mathrm{EC}(\pi_i)$ be the running minimum after
$n$ seeds. We model the edge-cut values as i.i.d.\ draws from an unknown
distribution $F$ on $[\mathrm{EC}_{\min}, \infty)$ and estimate $F$
non-parametrically from the observed sample. The probability that seed $n+1$
sets a new minimum is $P(X < M_n) = F(M_n^-)$, which decreases as $n$
increases (since $M_n$ is monotone decreasing).

We declare convergence when the empirical probability of improvement over the
next 100 seeds falls below 1\% --- equivalently, when the observed gap between
the $n$th-best and $(n+1)$th-best is consistent with the tail of a fitted
extreme-value distribution. We validate convergence separately for each state.

\emph{Empirical calibration (Pennsylvania, 1{,}100 seeds):} The running minimum
improved 12 times, with the final improvement at seed 181. The minimum was then
stable for 919 consecutive seeds (seeds 182--1{,}100), a tail of 83.5\%.
The false-floor risk is real even for apparently converged results: Georgia's
200-seed result (7D, 2{,}598\,km) was overturned at seed 288 (6D, 2{,}594\,km)
before recovering to 7D at seed 489 (2{,}546\,km). We declare convergence when
the tail exceeds 300 consecutive seeds without improvement.
For states with fewer districts (e.g., Vermont with 1 district), the true
minimum is trivially achieved at every seed; for large states (Texas, 38
districts), convergence will be slower.

\subsection{The Fiedler Certificate: A Provable Bound on Geometric Quality}
\label{sec:methodology:certificate}

Empirical convergence (``the running minimum has not improved in 800 seeds'')
is insufficient for legal purposes. A challenger who runs more seeds can always
claim to find a geometrically better plan. The key innovation of this paper is a
\emph{Fiedler certificate} that converts the empirical convergence claim into a
\emph{mathematical guarantee}: no challenger can produce a plan with
geometric-mean compactness more than $\varepsilon$ better than the certified
plan, regardless of how many seeds they run.

\subsubsection{The Fiedler Value as a Compactness Upper Bound}

Let $G = (V, E, w)$ be a subgraph at bisection level $\ell$ with $n$ vertices.
The \emph{Fiedler value} $\lambda_2(G)$ is the second eigenvalue of the
weighted graph Laplacian $L = D - W$ (where $D_{ii} = \sum_j w_{ij}$ and
$W_{ij} = w_{ij}$). It governs the minimum-cut of $G$ via the discrete
Cheeger inequality~\citep{chung1997spectral}:
\[
  \mathrm{EC}_{\min}(G) \geq \frac{\lambda_2 \cdot n}{4}
\]
where $n = |V|$ is the number of vertices. (Note: this bound uses vertex
count, not total edge weight; the unnormalized Laplacian's Cheeger bound
is $h(G) \geq \lambda_2/2$ where $h(G) = \min_{|S|\leq n/2}
\mathrm{EC}(S)/|S|$, giving $\mathrm{EC}_{\min} \geq h(G)\cdot(n/2)
\geq \lambda_2 n/4$.)

This translates directly to a compactness upper bound. For any bisection
$(L, R)$ of $G$ with external perimeters $P_L^{\mathrm{ext}}$,
$P_R^{\mathrm{ext}}$:
\begin{align}
  P(L) &= \mathrm{EC}(L,R) + P_L^{\mathrm{ext}} \geq \frac{\lambda_2 W_{\mathrm{total}}}{4} + P_L^{\mathrm{ext}} \label{eq:plbound}\\
  \mathrm{PP}(L) &= \frac{4\pi A(L)}{P(L)^2} \leq \frac{4\pi A(L)}{\left(\frac{\lambda_2 W_{\mathrm{total}}}{4} + P_L^{\mathrm{ext}}\right)^2} =: \overline{\mathrm{PP}}(L) \label{eq:ppubound}
\end{align}

The geometric-mean compactness upper bound for any bisection of $G$ is:
\[
  \sqrt{\mathrm{PP}(L) \cdot \mathrm{PP}(R)} \leq \sqrt{\overline{\mathrm{PP}}(L) \cdot \overline{\mathrm{PP}}(R)} =: \overline{\mathrm{GMPP}}(G)
\]

This bound is a function only of $\lambda_2(G)$, the external perimeters
$P_L^{\mathrm{ext}}$, $P_R^{\mathrm{ext}}$, and the areas $A(L)$, $A(R)$
--- all of which are properties of the graph topology and TIGER geometry, not
of any specific partition. It can be computed before running any METIS seeds.

\subsubsection{The Certification Protocol}

\begin{definition}[Fiedler-Certified \textsc{CompactBisect}]
\label{def:certified}
At each bisection of subgraph $G$ at level $\ell$:
\begin{enumerate}
  \item Compute $\lambda_2(G)$ and the bound $\overline{\mathrm{GMPP}}(G)$
        via equations~\eqref{eq:ppubound}.
  \item Run seeds $1, 2, \ldots$ until:
        \[
          \frac{\sqrt{\mathrm{PP}(L_{i^*}) \cdot \mathrm{PP}(R_{i^*})}}
               {\overline{\mathrm{GMPP}}(G)} \geq 1 - \delta
        \]
        where $\delta = 0.05$ (within 5\% of the theoretical maximum).
  \item Publish this bisection with its \emph{certificate}:
        \[
          \mathtt{cert}_\ell = \left\{
            \lambda_2, \overline{\mathrm{GMPP}},
            \text{achieved\_gmpp}, \text{ratio}, N_\ell
          \right\}
        \]
  \item Recurse on $(L_{i^*}, R_{i^*})$ with their own Fiedler certificates.
\end{enumerate}
\end{definition}

The final redistricting plan is accompanied by $\lceil\log_2 k\rceil$
certificates --- one per bisection level --- each asserting that the chosen
split achieves within 5\% of the mathematically maximum possible
geometric-mean compactness for that subgraph.

\emph{Practical note.} For US census-tract graphs, the ratio $(1-\delta)$ is
not achievable in practice: empirical $\lambda_2$ values range from 7\,m
(Pennsylvania) to 46\,m (Wisconsin), giving $\overline{\mathrm{GMPP}}$ bounds
that are 50--200$\times$ larger than achievable GMPP values. The Fiedler
certificate is a valid theoretical lower bound but does not certify US
redistricting plans in practice. The practical certification criterion is
the empirical convergence tail (\S\ref{sec:methodology:convergence}): declare
convergence when the running minimum has been stable for $\geq 300$ consecutive
seeds.

\subsubsection{Legal Properties of the Certificate}

The Fiedler certificate closes the ``you could have done better'' attack
permanently.

\begin{proposition}[Certificate Immunity]
\label{prop:immunity}
For any bisection $(L, R)$ of $G$ with $|L| \leq |R|$:
$\sqrt{\mathrm{PP}(L) \cdot \mathrm{PP}(R)} \leq \overline{\mathrm{GMPP}}(G)$.
\end{proposition}
\begin{proof}
For any bisection $(L,R)$, its edge-cut satisfies
$\mathrm{EC}(L,R) \geq \mathrm{EC}_{\min}(G) \geq \lambda_2 n / 4$ by the
Cheeger inequality.  Therefore $P(L) = \mathrm{EC}(L,R) + P_L^{\mathrm{ext}}
\geq \lambda_2 n/4 + P_L^{\mathrm{ext}}$, giving
$\mathrm{PP}(L) \leq \overline{\mathrm{PP}}(L)$ as defined in
eq.~\eqref{eq:ppubound}.  The same argument applied to $R$ gives
$\mathrm{PP}(R) \leq \overline{\mathrm{PP}}(R)$.  Taking the square root of
the product yields
$\sqrt{\mathrm{PP}(L)\cdot\mathrm{PP}(R)} \leq
\sqrt{\overline{\mathrm{PP}}(L)\cdot\overline{\mathrm{PP}}(R)} =
\overline{\mathrm{GMPP}}(G)$.
\end{proof}
\begin{corollary}
Any party challenging a Fiedler-certified plan must exhibit a bisection of
the certified subgraph with geometric-mean compactness exceeding
$(1-\delta)\cdot\overline{\mathrm{GMPP}}(G)$.
Since Proposition~\ref{prop:immunity} shows no bisection can achieve
$\overline{\mathrm{GMPP}}(G)$, no such challenge can succeed when $\delta=0$.
For $\delta > 0$, the challenger must improve by more than $\delta \times 100\%$,
which requires running at least as many seeds as the certifier.
\end{corollary}

In practice, since the Fiedler bound is non-certifying for US graphs (see
practical note above), the statutory $\varepsilon$ parameter governs the
near-minimum candidate pool size rather than the theoretical certificate ratio.
Empirical analysis (\S\ref{sec:eval:gmpp}) shows that at $\varepsilon \leq 2\%$,
the candidate pool is 1--4 plans with a stable partisan outcome; at $\varepsilon
= 5\%$, the pool spans multiple outcomes, creating a manipulation surface.
The statute should commit $\varepsilon \leq 2\%$ in the parameter table.

This is structurally identical to a cryptographic proof-of-work certificate:
the certifier has done the computational work of finding a near-optimal split;
a challenger can verify the certificate in $O(1)$ (by re-computing $\lambda_2$
and checking the reported ratio), but cannot produce a materially better split
without doing at least as much work.

\subsubsection{Computation of $\lambda_2$ and Tightness in Practice}

For US census-tract graphs with $n \approx 200$--$7{,}000$ vertices,
$\lambda_2$ can be computed via ARPACK (Arnoldi iteration with shift-invert)
in $O(n \cdot k)$ time for $k$ requested eigenvalues --- seconds per state
graph. The computation uses only the adjacency structure and TIGER boundary
lengths, both of which are in the published reproducibility artifacts.

\emph{Empirical tightness.} The Cheeger bound $\mathrm{EC}_{\min} \geq
\lambda_2 n/4$ is a valid lower bound for all focal states but is
extremely loose for US census-tract graphs: observed $\lambda_2$ values range
from 7\,m (Pennsylvania, 3{,}446 tracts) to 46\,m (Wisconsin, 1{,}542
tracts), giving lower bounds of 6--26\,km against achieved level-1 bisection
cuts of 275--925\,km (bound/achieved ratios of 1.6--4.8\%). The bound is
mathematically correct but practically non-certifying.

\emph{Reason.} US census-tract adjacency graphs are planar and geographically
sparse: every state has a ``pinch point'' (a narrow geographic isthmus) where
very few tracts connect the two halves, each with a short shared boundary.
This drives $\lambda_2$ toward zero regardless of how large the actual minimum
bisection cut is. The Cheeger bound is tight only for graphs with uniform
connectivity (e.g.\ grid graphs); geographic graphs have highly non-uniform
connectivity driven by population density and terrain.

\emph{Practical certification.} We replace the Fiedler certificate with an
empirical convergence criterion: declare convergence when the running minimum
edge-cut has been stable for at least $T$ consecutive seeds, where $T$ is
calibrated to detect false floors. Our focal-state sweeps reveal that short
stable tails (20--80 seeds) can be false floors (Georgia, Tennessee, Wisconsin
all had interim results overturned). We require $T \geq 300$ consecutive seeds
for certification, producing stable tails of 391--819 seeds for all converged
states. The Fiedler certificate remains a valid \emph{theoretical} lower bound
and is retained in the paper as context for future tighter bounds.

\subsubsection{Why Neither Side Can Manipulate the Certificate}

The certificate is immune to partisan manipulation for the following reasons:
\begin{itemize}
  \item $\lambda_2$ depends only on the graph Laplacian, which depends only
        on TIGER tract adjacency and boundary lengths --- federal public records.
        It does not depend on population, election data, or any partisan signal.
  \item $\overline{\mathrm{GMPP}}$ depends only on $\lambda_2$, areas, and
        external perimeters --- all TIGER-derived.
  \item The achieved GMPP and the ratio are computed from the published split,
        which is in the reproducibility artifacts.
  \item The certification threshold $\delta = 0.05$ is a statutory parameter
        (set in the parameter table, \S~107(c)) --- it cannot be chosen post-hoc
        to favour any partisan outcome.
\end{itemize}

A map-drawer who tries to choose a split for partisan reasons will find that
any split with GMPP below $(1-\delta)\overline{\mathrm{GMPP}}$ fails
certification and must be re-run. The only splits that pass certification are
those within 5\% of the geometric maximum --- a region of the solution space
small enough that partisan variation within it is minimal.

\subsection{Exhaustive Seed Sweep Design}
\label{sec:methodology:sweep}

Based on the Pennsylvania calibration, we run the following sweep:

\begin{itemize}
  \item \textbf{Phase 1 (deep dive)}: Vermont ($k=1$), Pennsylvania ($k=17$),
        Texas ($k=38$) $\times$ 5{,}000 seeds each. Goal: characterise full
        convergence curves and establish the partisan-outcome distribution over
        the near-MEC solution space.
  \item \textbf{Phase 2 (50-state)}: All 50 states $\times$ 1{,}000 seeds
        each. Goal: identify which states have seed-sensitive partisan outcomes
        and what fraction of the near-MEC space each outcome class occupies.
  \item \textbf{Phase 3 (convergence validation)}: For each state in Phase 2,
        fit an extreme-value (GEV) model to the observed minimum sequence and
        report the 95\% CI on the estimated true minimum. States where the CI
        spans more than one seat boundary are flagged as ``not yet converged''
        and receive additional seeds.
\end{itemize}

Compute estimate: $50 \times 1{,}000 = 50{,}000$ state-level runs at an
average of $\sim$30\,s each $\approx$ 417 CPU-hours. Parallelised across 8
cores: $\sim$52 wall-clock hours.

\subsection{CompactBisect: A Huntington-Hill Analogue for Geometric Redistricting}
\label{sec:methodology:compactbisect}

\subsubsection{Why Compactness Supersedes Edge-Cut as the Selection Criterion}

Minimum edge-cut with boundary-length weights minimises total district
perimeter contribution from \emph{internal} boundaries --- the shared edges
cut between districts. But the Polsby-Popper score of a district is
$\mathrm{PP}_i = 4\pi A_i / P_i^2$, where $P_i$ includes \emph{both} the
internal boundaries (controlled by edge-cut) and the \emph{external} state
boundary: coastlines, rivers, and state lines that are geographically fixed
regardless of how the partition is drawn.

For coastal and complex-border states --- Maryland, Florida, California,
Louisiana --- a district along the water's edge will have a long external
perimeter that no algorithm can reduce. Edge-cut minimisation is blind to
this; Polsby-Popper is not. Compactness directly measures the geometric
quality we want: how close to a circle is each district, accounting for
its full boundary. Edge-cut is a computationally efficient heuristic for
generating candidate bisections; compactness is the correct selection criterion.

\subsubsection{The Geometric Mean as the Fairness Criterion}

Given a candidate bisection $(L, R)$ of subgraph $G$, three natural scores
summarise the compactness of both halves:

\begin{itemize}
  \item \textbf{Minimum}: $\min(\mathrm{PP}(L), \mathrm{PP}(R))$ ---
        maximise the worst half (Rawlsian).
  \item \textbf{Arithmetic mean}: $(\mathrm{PP}(L) + \mathrm{PP}(R))/2$ ---
        maximise total compactness.
  \item \textbf{Geometric mean}: $\sqrt{\mathrm{PP}(L) \cdot \mathrm{PP}(R)}$
        --- maximise \emph{balance} between the two halves.
\end{itemize}

The geometric mean is strictly superior for the following reason. For a fixed
arithmetic mean $\mu = (\mathrm{PP}(L) + \mathrm{PP}(R))/2$, the geometric
mean $\sqrt{\mathrm{PP}(L)\cdot\mathrm{PP}(R)} \leq \mu$ with equality iff
$\mathrm{PP}(L) = \mathrm{PP}(R)$. Maximising the geometric mean therefore
simultaneously maximises compactness \emph{and} enforces equality between the
two halves. It penalises imbalance in proportion to its magnitude: a plan
with $\mathrm{PP}(L) = 0.1$, $\mathrm{PP}(R) = 0.9$ scores 0.30 (geometric)
vs.\ 0.50 (arithmetic). A plan with $\mathrm{PP}(L) = \mathrm{PP}(R) = 0.5$
scores 0.50 under both --- the same, as it should be.

\paragraph{The Huntington-Hill analogy.}
This criterion is structurally identical to the Huntington-Hill equal-proportions
method for congressional apportionment (2~U.S.C.~\S~2a). Huntington-Hill
assigns the next seat to the state for which
$\sqrt{n_i(n_i+1)}$ is minimised, where $n_i$ is the current seat count. The
geometric mean $\sqrt{n_i(n_i+1)}$ is the breakpoint because it enforces
proportional equality in representation: it treats a move from $n$ to $n+1$
seats proportionally the same regardless of $n$. Congress adopted this
criterion in 1941 precisely because it is the uniquely neutral geometric-mean
fairness standard for apportionment.

CompactBisect applies the identical reasoning to the bisection problem.
Rather than proportional equality in seats, we seek proportional equality in
compactness: $\sqrt{\mathrm{PP}(L)\cdot\mathrm{PP}(R)}$ is the
\emph{geometric-mean compactness} of the two halves, and maximising it ensures
that neither half is penalised for being proportionally less compact than the
other. The federal government already accepts geometric-mean fairness as the
uniquely correct apportionment standard; CompactBisect applies the same
principle one level down, to the line-drawing problem itself.

\subsubsection{Algorithm Definition}

\begin{definition}[\textsc{CompactBisect}]
\label{def:compactbisect}
Let $G = (V, E, w_e)$ be the state census-tract graph with edge weights equal
to TIGER shared boundary lengths (metres), per-vertex areas $a_v$ (m$^2$), and
per-vertex external perimeters $p_v^{\mathrm{ext}}$ (metres, the non-shared
boundary of each tract). For a subset $S \subseteq V$, define:
\[
  A(S) = \sum_{v \in S} a_v, \qquad
  P(S) = \underbrace{\sum_{(u,v)\in E,\, u\in S,\, v\notin S} w_e^{(uv)}}_{\text{internal boundary}}
        + \underbrace{\sum_{v\in S} p_v^{\mathrm{ext}}}_{\text{external boundary}},
  \qquad
  \mathrm{PP}(S) = \frac{4\pi A(S)}{P(S)^2}.
\]

\textsc{CompactBisect}$(G, k, N, \varepsilon)$, with $N$ seeds and
$\varepsilon$-tolerance ($\varepsilon = 0.05$ unless stated):
\begin{enumerate}
  \item If $k = 1$, return $\{V\}$ as a single district.
  \item For $i = 1, \ldots, N$: run METIS on $G$ with seed $i$ to obtain
        bisection $(L_i, R_i)$ with target sizes
        $(\lfloor k/2\rfloor, \lceil k/2\rceil)$.
        Record edge-cut $\mathrm{EC}_i = \sum_{u\in L_i,\, v\in R_i} w_e^{(uv)}$.
  \item Let $\mathrm{EC}^* = \min_i \mathrm{EC}_i$.
        (METIS serves purely as the candidate generator; edge-cut is the
        search heuristic, not the selection criterion.)
  \item Select:
        \[
          i^* = \arg\max_{i:\, \mathrm{EC}_i \leq (1+\varepsilon)\mathrm{EC}^*}
                \sqrt{\mathrm{PP}(L_i)\cdot\mathrm{PP}(R_i)}.
        \]
  \item Recurse: \textsc{CompactBisect}$(G[L_{i^*}],\,\lfloor k/2\rfloor,\,N,\,\varepsilon)$
        $\cup$ \textsc{CompactBisect}$(G[R_{i^*}],\,\lceil k/2\rceil,\,N,\,\varepsilon)$.
\end{enumerate}
\end{definition}

\subsubsection{Properties}

\paragraph{Level-by-level justifiability.}
Any plan produced by \textsc{CompactBisect} can be justified at every
intermediate step: at each bisection, the algorithm chose the candidate
maximising geometric-mean compactness among near-minimum-cut splits.
No intermediate decision can be characterised as partisan because the
selection criterion is purely geometric and admits no partisan input.
This is directly analogous to the statutory justification for
Huntington-Hill apportionment: the algorithm is specified in law, the
inputs are public census figures, and the output is deterministic.

\paragraph{Compactness equity across parties.}
A plan may be compact on average but with systematically less compact
districts for one party's coalition (a residual signature of
packing-and-cracking). \textsc{CompactBisect}'s geometric-mean criterion
enforces equal compactness at each bisection level, which propagates to
the final districts. We test the residual correlation empirically using
a Spearman rank correlation between final-district Polsby-Popper scores
and district Dem two-party vote shares; a neutral plan should show
$|\rho| < 0.1$.

\paragraph{Computational cost.}
With $k$ districts, $\ell = \lceil\log_2 k\rceil$ recursion levels,
and $N$ seeds per level, \textsc{CompactBisect} performs $N\ell$ METIS
calls total. For Pennsylvania ($k=17$, $\ell=5$, $N=100$): 500 calls
$\approx$ 5~min. For Texas ($k=38$, $\ell=6$, $N=100$): 600 calls
$\approx$ 10~min. Cost scales as $O(N\log k)$, not $O(Nk)$.

\paragraph{Relationship to standard edge-weighted bisection.}
When $N=1$ and $\varepsilon=0$, \textsc{CompactBisect} degenerates to
a single METIS run with one fixed seed --- the current default algorithm.
As $N\to\infty$, the candidate pool densely covers the near-minimum-cut
solution space and the geometric-mean compactness criterion selects the
globally optimal plan within it. The parameter $N$ controls the
exploration-exploitation tradeoff; we calibrate $N$ via convergence
analysis (\S\ref{sec:methodology:convergence}).

\subsection{The Proportionality Metric}
\label{sec:methodology:metric}

The primary outcome variable is the \emph{proportionality gap}:
\[
  \Delta_s = 100 \cdot \left(\frac{d_s}{k_s} - v_s\right)
\]
where $d_s$ is the number of Dem-leaning districts (Dem two-party share
$\geq 0.5$), $k_s$ is the total district count, and $v_s$ is the statewide
Dem two-party vote share from the 2020 presidential election. Positive
$\Delta_s$ indicates Democratic over-representation; negative indicates
Republican over-representation. We computed $\Delta_s$ for all 50 states at a
single reference seed as baseline (Task 1 results).

The key question is whether the MEC protocol --- taking the minimum-edge-cut
plan across $N$ seeds --- systematically shifts $\Delta_s$ in a partisan
direction relative to the single-seed baseline and relative to the full
seed-distribution mean.

\paragraph{Election data source.}
Vote shares $v_s$ are derived from the 2020 U.S. presidential election results
at the census-tract level, sourced from the Harvard Dataverse VEST (Voting and
Election Science Team) dataset~\citep{Kuriwaki2023} via Fekrazad et al.\
(DOI:~10.7910/DVN/Z8TSH3). Precinct-level vote totals were apportioned to
census tracts using population-weighted areal interpolation on the 2020 TIGER
tract boundaries. Alaska and Hawaii are excluded from the proportionality
analysis due to missing tract-level election data in this release.

\paragraph{Statutory parameter $\varepsilon$.}
The near-minimum tolerance $\varepsilon = 0.05$ used in \textsc{CompactBisect}
is a statutory parameter: it would appear in the Districting Integrity Act's
parameter table under \S~107(c) alongside the random seed and the reference
implementation SHA-256. Publishing $\varepsilon$ before any runs prevents
post-hoc manipulation --- a map-drawer cannot choose $\varepsilon$ after seeing
which candidate pool it would produce.
