\section{Evaluation}
\label{sec:evaluation}

\subsection{Vermont: Zero-Variance Baseline}
\label{sec:eval:vt}

Vermont has one congressional district (the entire state). Every seed produces
the identical assignment --- the trivial partition assigning all 192 census
tracts to district 1. The edge-cut is zero, the proportionality gap is
$100(1.0 - 0.683) = +31.7$\,pp regardless of seed, and the assignment files
are byte-identical across all 20 tested seeds. This confirms that for
at-large states, the seed is \emph{mathematically irrelevant}: no bisection
is performed and no partitioning choice is made.

\subsection{Pennsylvania: The Dense Solution Space}
\label{sec:eval:pa}

Pennsylvania (17 districts, 50.6\% Dem statewide vote, 2020 presidential)
is the paper's central case. The achievable proportionality gaps are
$\{-15.3, -9.4, -3.5, +2.3\}$\,pp corresponding to 6D, 7D, 8D, and 9D
Republican/Democratic seat splits.

\paragraph{Solution-space density.}
Across 100 tested seeds, every seed produced a \emph{distinct} final
partition (100 unique SHA-256 fingerprints of the assignment files). The
solution space is dense: there is no small set of ``canonical'' boundaries
that seeds converge to. This is a direct consequence of the high dimensionality
of the partition space ($\sim$193 census tracts, $\binom{193}{k}$-scale
feasible-partition count).

\paragraph{Edge-cut distribution.}
The observed edge-cuts span 2{,}529--3{,}003\,km (18.7\% range). Edge-cut
distributions overlap substantially across seat-outcome classes:

\begin{table}[h]
\centering
\begin{tabular}{lrrrr}
\toprule
Outcome & $n$ (998 valid seeds) & EC min & EC mean & EC max \\
\midrule
6D/11R ($-15.3$\,pp) & 49  (5\%) & 2{,}561 & 2{,}717 & 2{,}986 \\
7D/10R ($-9.4$\,pp)  & 591 (59\%) & 2{,}466 & 2{,}708 & 3{,}110 \\
8D/9R  ($-3.5$\,pp)  & 315 (32\%) & 2{,}441 & 2{,}703 & 2{,}949 \\
9D/8R  ($+2.3$\,pp)  & 43  (4\%) & 2{,}455 & 2{,}690 & 2{,}995 \\
\bottomrule
\end{tabular}
\caption{Edge-cut (km) by seat outcome, Pennsylvania, 1{,}100 seeds (998 valid).
  All four outcome classes contain plans within 5\% of the global minimum.
  7D is the modal outcome (59\%); 8D is the MEC winner (min EC = 2{,}441\,km).}
\label{tab:pa-ec-by-outcome}
\end{table}

\paragraph{Convergence.}
The running minimum improved 12 times across 1{,}100 seeds, last at seed 181
(EC = 2{,}441\,km, 8D/9R, $-3.5$\,pp). The minimum was stable for 919
consecutive seeds afterward (seeds 182--1{,}100), a 83.5\% stable tail.
Table~\ref{tab:pa-convergence} shows the improvement trajectory.

\begin{table}[h]
\centering
\begin{tabular}{rrrl}
\toprule
Seed & New minimum (km) & Improvement & Outcome \\
\midrule
1   & 3{,}003 & --- (first) & 7D \\
2   & 2{,}822 & $-$181     & 9D \\
3   & 2{,}632 & $-$190     & 8D \\
5   & 2{,}606 & $-$26      & 8D \\
12  & 2{,}596 & $-$10      & 7D \\
18  & 2{,}578 & $-$18      & 8D \\
34  & 2{,}529 & $-$49      & 8D \\
101 & 2{,}520 & $-$9       & 8D \\
106 & 2{,}520$\to$2{,}509 & cont. & 8D \\
181 & 2{,}441 & $-$68      & 8D \\
\midrule
\multicolumn{4}{l}{\emph{Seeds 182--1{,}100: no improvement (919-seed stable tail)}} \\
\bottomrule
\end{tabular}
\caption{Running minimum edge-cut improvements, Pennsylvania, 1{,}100 seeds.
  12 improvements total; final minimum 2{,}441\,km at seed 181.
  Subsequent 919 seeds confirm convergence (83.5\% stable tail).}
\label{tab:pa-convergence}
\end{table}

\paragraph{Near-minimum partisan distribution.}
Of the 29 plans within 5\% of the current global minimum (EC $\leq$ 2{,}655\,km):
2 produce 6D, 15 produce 7D, 11 produce 8D, and 1 produces 9D. The near-minimum
region of the solution space spans all four outcome classes. This means:
\emph{the minimum edge-cut criterion does not deterministically select a
particular partisan outcome in Pennsylvania}. The outcome of the MEC plan
depends on which local optimum is the global minimum --- a question only
resolved by exhaustive (or near-exhaustive) seed sampling.

\subsection{\textsc{CompactBisect} vs.\ Global MEC: Empirical Comparison}
\label{sec:eval:compactbisect}

We ran \textsc{CompactBisect} with 50 seeds per bisection level on the three
focal states. Table~\ref{tab:compactbisect-vs-mec} compares the results to
the global MEC benchmark (best plan from 200 seeds for the full partition).

\begin{table}[h]
\centering
\begin{tabular}{llrrrrrr}
\toprule
State & & \multicolumn{3}{c}{Global MEC (200 seeds)} & \multicolumn{3}{c}{\textsc{CompactBisect} (50/level)} \\
\cmidrule(lr){3-5}\cmidrule(lr){6-8}
& Dem\% & D seats & $\Delta$ & EC (km) & D seats & $\Delta$ & EC (km) \\
\midrule
WI & 50.3\% & 3D/5R* & $-12.8$\,pp & 1{,}604 & 3D/5R & $-12.8$\,pp & 1{,}584 \\
NC & 49.3\% & 5D/9R & $-13.6$\,pp & 2{,}400 & 5D/9R & $-13.6$\,pp & 2{,}271 \\
PA & 50.6\% & 8D/9R & $-3.5$\,pp  & 2{,}441 & 7D/10R & $-9.4$\,pp & 2{,}392 \\
\bottomrule
\end{tabular}
\caption{\textsc{CompactBisect} (50 seeds/level, release binary) vs.\ global MEC
  (best of N seeds). $\Delta$ = proportionality gap in pp.
  *WI: 200-seed false floor was 2D ($-25.3$\,pp); 1{,}200-seed MEC is 3D ($-12.8$\,pp),
  matching \textsc{CompactBisect}. NC: same outcome, lower edge-cut.
  PA: \textsc{CompactBisect} slightly less proportional. Results confirmed with release binary.}
\label{tab:compactbisect-vs-mec}
\end{table}

\paragraph{Wisconsin (updated).}
The initial 200-seed MEC result for Wisconsin was 2D/6R ($-25.3$\,pp) at
1{,}689\,km. This is the third false floor in our focal analysis (after Georgia
and Tennessee). The 1{,}000-seed sweep yields:

\begin{center}
\begin{tabular}{rrrl}
\toprule
Seed & EC (km) & D seats & Gap \\
\midrule
113 & 1{,}689 & 2D/6R & $-25.3$\,pp \\
318 & 1{,}642 & 3D/5R & $-12.8$\,pp \\
891 & 1{,}615 & 3D/5R & $-12.8$\,pp \\
\bottomrule
\end{tabular}
\end{center}

The partisan outcome stabilised at 3D/5R at seed 318 and did not change through
seed 1{,}000, despite the EC improving by a further 27\,km at seed 891. The
3D outcome has persisted through 573 consecutive seeds. The 109-seed tail at
seed 1{,}000 (Fiedler ratio 3.06×) indicates the EC minimum is still
converging, but the partisan outcome is stable.

\textsc{CompactBisect} independently produces 3D/5R ($-12.8$\,pp) at
1{,}584\,km --- 31\,km more compact than the 1{,}000-seed MEC. For Wisconsin,
both geometric criteria agree on 3D when given sufficient computation.

\paragraph{North Carolina.}
Both algorithms produce 5D/9R. However, \textsc{CompactBisect} achieves this
at a lower edge-cut (2{,}271\,km vs.\ 2{,}400\,km). The level-by-level
compactness criterion finds a more geometrically efficient partition that
happens to produce the same partisan outcome. NC's strong geographic sorting
(Charlotte, Research Triangle, Greensboro) constrains any compact algorithm
to produce a Republican-favoring outcome.

\paragraph{Pennsylvania.}
\textsc{CompactBisect} produces 7D/10R ($-9.4$\,pp) vs.\ the global MEC's
8D/9R ($-3.5$\,pp) --- 5.9\,pp \emph{worse}. This is an important finding:
the greedy level-by-level approach does not always outperform global
optimisation. PA's mixed geography (Philadelphia suburbs, Pittsburgh, Scranton)
means that locally optimal bisection choices at early levels can constrain
later levels in ways that reduce the total Dem seat count.

\paragraph{Synthesis.}
\textsc{CompactBisect} outperforms global MEC on proportionality in geographically
sorted states (WI: +12.5\,pp) and matches it in moderately sorted states (NC:
0\,pp). In mixed-geography states where global MEC is already near-proportional
(PA: $-3.5$\,pp), \textsc{CompactBisect}'s greedy structure can produce a worse
outcome. The appropriate selection criterion depends on the state's geographic
structure, which is itself a property of the census data and not a partisan choice.

\subsection{GMPP Convergence vs.\ MEC Convergence}
\label{sec:eval:gmpp}

We tracked the running maximum geometric-mean Polsby-Popper of the level-1
bisection (GMPP = $\sqrt{\mathrm{PP}(L)\cdot\mathrm{PP}(R)}$) across seeds
in parallel with the running minimum edge-cut, for Wisconsin, Pennsylvania,
and North Carolina at 200--1{,}000 seeds.

\paragraph{Convergence rates differ.}
For Wisconsin (1{,}000 seeds), GMPP last improved at seed 109 (890-seed
tail, 89\%); MEC last improved at seed 891 (108-seed tail, 11\%). GMPP
converges roughly 8× faster than MEC for this state.

\paragraph{Criteria are aligned but not identical.}
The Pearson correlation between EC and GMPP across 1{,}000 Wisconsin seeds is
$r = -0.354$: lower EC tends to produce higher GMPP. This confirms that
edge-cut minimisation and geometric-mean PP maximisation point in the same
direction. However, the extreme cases diverge: the MEC-minimising plan
(seed 891, 1{,}615\,km, GMPP\,=\,0.358, PP\,=\,(0.40, 0.32)) is not the same
plan as the GMPP-maximising plan (seed 930, 1{,}880\,km, GMPP\,=\,0.397,
PP\,=\,(0.54, 0.29)).

\paragraph{Partisan outcomes agree for Wisconsin.}
For Wisconsin at 1{,}000 seeds, both the MEC-minimising plan (seed 891) and the
GMPP-maximising plan (seed 930) produce 3D/5R. \textsc{CompactBisect} also
produces 3D/5R. All three geometric criteria agree on Wisconsin's partisan
outcome when given sufficient seeds.

For Pennsylvania (200 seeds), MEC gives 8D while GMPP-max gives 7D --- the
two criteria disagree. This reflects Pennsylvania's mixed geographic structure,
where the near-minimum solution space spans both 7D and 8D outcomes and the
MEC and GMPP criteria select different plans from that space.

For North Carolina, both MEC (seed 41) and GMPP-max (seed 69) give 5D/9R.
Correlation $r = -0.223$: moderate alignment.

\paragraph{Wisconsin: partisan outcome is certified, EC minimum is not.}
At 1{,}200 seeds, the EC minimum last improved at seed 1{,}023 (1{,}604\,km,
177-seed tail) --- not EC-certified. However, the \emph{partisan outcome}
(3D/5R) has been stable since seed 318 for \textbf{882 consecutive seeds}.
For legal purposes, the outcome is certified: no redistricting challenge could
require a materially more compact 4D plan without demonstrating one.

Wisconsin GMPP (exact TIGER perimeters) last improved at seed 930 with a
69-seed tail --- not GMPP-certified. Both MEC and GMPP give 3D.

\paragraph{Epsilon sensitivity.}
The near-minimum tolerance $\varepsilon$ must be small and pre-committed to
the parameter table. Table~\ref{tab:epsilon} shows the partisan outcomes in
the candidate pool at different $\varepsilon$ values.

\begin{table}[h]
\centering
\small
\begin{tabular}{lrrrr}
\toprule
State & $\varepsilon$=1\% & $\varepsilon$=2\% & $\varepsilon$=5\% & $\varepsilon$=10\% \\
\midrule
WI (8D)  & 3D only & 3D only & 2D--3D & 2D--4D \\
PA (17D) & 8D--9D  & 7D--9D  & 6D--9D & 6D--9D \\
NC (14D) & 5D only & 5D--6D  & 4D--8D & 4D--8D \\
\bottomrule
\end{tabular}
\caption{Partisan outcomes in the near-minimum candidate pool. At
  $\varepsilon \leq 2\%$ the pool is 1--4 plans with stable outcomes.
  At $\varepsilon = 5\%$ (prior default) the pool spans multiple outcomes,
  creating a manipulation surface. Statutory recommendation: $\varepsilon \leq 2\%$.}
\label{tab:epsilon}
\end{table}

\paragraph{Criterion choice.}
The paper recommends MEC as primary: no degree-of-freedom choices (no area
measure, no shape metric, no approximation), and independently verifiable
from TIGER boundary lengths alone. GMPP (with exact perimeters) is
verification: when MEC and GMPP agree, confidence is high.

\subsection{Texas: Large-State Failures and Convergence Rate}
\label{sec:eval:tx}

Texas (38 districts) exhibited a 4\% seed-failure rate (seeds 1, 2, 6, 16
out of 100 produced population-balance violations exceeding the 0.5\%
tolerance). These failures are expected: with 38 districts and $\lceil
\log_2 38 \rceil = 6$ levels of recursion, small per-level imbalances
compound. The statute's tiered tolerance schedule (\S~104(b)(4)) addresses
this; our implementation applies a flat 0.5\% tolerance. % TODO: implement
% tiered tolerance and re-run.

Among 96 successful seeds, edge-cuts ranged from [TODO] to [TODO]\,km with
[TODO] distinct seat outcomes observed. Convergence is expected to be slower
than Pennsylvania due to the larger partition space.

\subsection{Georgia: Why Convergence Depth Matters}
\label{sec:eval:ga}

Georgia (14 districts, 50.1\% Dem statewide vote) illustrates why the 200-seed
convergence threshold from the initial sweep is insufficient for legal-quality
claims.

\begin{center}
\begin{tabular}{rrrl}
\toprule
Seed & EC (km) & D seats & Gap \\
\midrule
190 & 2{,}598 & 7D/7R & $-0.1$\,pp \\
288 & 2{,}594 & 6D/8R & $-7.3$\,pp \\
489 & 2{,}546 & 7D/7R & $-0.1$\,pp \\
\bottomrule
\end{tabular}
\end{center}

At seed 190 (within the initial 200-seed window), the running minimum appeared to
be 2{,}598\,km producing a near-perfectly proportional 7D/7R plan ($-0.1$\,pp
gap). A na{\"i}ve convergence declaration at seed 200 would have reported this
as the MEC result.

At seed 288, a new minimum at 2{,}594\,km overturned the result: the better
plan produced 6D/8R ($-7.3$\,pp). But at seed 489, a still-better minimum at
2{,}546\,km recovered the 7D/7R outcome. Seeds 490--1{,}000 produced no
further improvement: a 511-seed stable tail (51\%) confirms convergence.

The final MEC result for Georgia is 2{,}546\,km, 7D/7R, $-0.1$\,pp. The key
lesson: the initial 200-seed result was misleading not because it was wrong
(7D/7R is the correct answer) but because an intermediate false floor at
seed 288 produced an incorrect 6D result that would have been reported as the
MEC if the sweep had stopped there. \emph{Convergence certification --- not
just a fixed seed budget --- is the correct protocol.}

This contrasts sharply with Wisconsin (\S\ref{sec:eval:pa}), where the MEC
plan consistently produces 2D/6R despite 50.3\% Dem vote. The difference is
geographic sorting: Wisconsin's Democratic voters concentrate in Milwaukee and
Madison; Georgia's Democratic coalition spans Atlanta, Augusta, Savannah, and
diverse suburban counties.

\subsection{50-State Summary}
\label{sec:eval:national}

Table~\ref{tab:national} presents the MEC plan's proportionality gap for
all 50 states at 200 seeds (Texas excluded: 22\% seed failure rate with flat
tolerance; see below). Figure~\ref{fig:national-scatter} plots $\Delta_s$
against Dem vote share $v_s$, revealing the paper's central pattern.

\paragraph{Near-proportional states (MEC within $\pm$10\,pp).}
States with geographically mixed partisan electorates produce near-proportional
MEC results: Georgia ($-0.1$\,pp), Nevada ($-1.2$\,pp), South Carolina
($-1.2$\,pp), Pennsylvania ($-3.5$\,pp), Arizona ($-5.7$\,pp), Florida
($-5.5$\,pp), Ohio ($-5.9$\,pp). These are states where Democratic voters
are distributed across multiple urban and suburban centers rather than
concentrated in one metropolitan area.

\paragraph{Sorted Republican-favoring states ($\Delta < -15$\,pp).}
Deep-red states and states with strong geographic sorting produce large
negative gaps: Montana ($-41.6$\,pp), Iowa ($-45.8$\,pp), Mississippi
($-41.6$\,pp), Wisconsin ($-25.3$\,pp), North Carolina ($-13.6$\,pp). In
these states, Democratic voters concentrate in 1--2 urban centers; the MEC
criterion creates compact, safe Democratic districts and leaves the remaining
geography safely Republican.

\paragraph{Sorted Democratic-favoring states ($\Delta > +15$\,pp).}
New England and coastal states produce large positive gaps: Connecticut
($+39.8$\,pp), Rhode Island ($+39.4$\,pp), Delaware ($+40.4$\,pp),
Vermont ($+31.7$\,pp), Massachusetts ($+32.9$\,pp). In these states,
Democratic majorities are large enough that compact partisan-blind
partitioning produces Dem sweeps of all districts.

\paragraph{Structural vs.\ intentional disproportion.}
The key distinction for legal purposes: large $|\Delta_s|$ in states with
extreme geographic sorting (Montana: 41.6\% Dem vote, 0 Dem seats) reflects
geography, not algorithmic bias. Any compact algorithm --- including a
hand-drawn map designed to be neutral --- produces the same outcome in Montana
because no compact district in Montana can produce a Dem majority. This is
the ``geometric bias theorem'' (\S\ref{sec:discussion}): geographic sorting
makes proportional representation geometrically unachievable without
explicitly importing partisan data.

\paragraph{National seat totals.}
Summing the 48 states with available election data: the MEC protocol produces
228D/204R (52.8\% Democratic seats). Alaska (1 seat) and Hawaii (2 seats) are
excluded due to missing tract-level election data in our source. Best-guess
inclusion: Alaska's 2020 at-large seat went Republican (Don Young won with
54.1\%); Hawaii's 2 seats went Democratic (Kai Kahele and Ed Case). Including these gives \textbf{229D/206R} as the
national estimate. Actual 2020 House outcome: 222D/213R (51.0\% D). The
algorithmic maps give Democrats roughly 7 more seats nationally than the
enacted plans.

\paragraph{Texas and California: tiered tolerance needed.}
Texas (38 districts) and California (52 districts) exhibit high seed failure
rates under the flat 0.5\% tolerance: TX 22\%, CA 64\%. With 5--6 recursion
levels, cumulative balance errors compound. The Districting Integrity Act's
statutory tiered tolerance schedule (\S~104(b)(4): 0.5\% at levels 1--2,
1.0\% at level 3, 1.5\% at level 4, $\geq$2.0\% at level 5+) is necessary
for large states to achieve a high success rate. Implementation of the
tiered schedule is deferred; results for TX and CA are reported at 200 seeds
with the flat tolerance.

\begin{table}[h]
\centering\small
\begin{tabular}{lrrrr|lrrrr}
\toprule
State & $k$ & $v_s$\% & MEC & $\Delta_s$\,pp
& State & $k$ & $v_s$\% & MEC & $\Delta_s$\,pp \\
\midrule
MT & 2 & 41.6 & 0D & $-41.6$ & GA & 14 & 50.1 & 7D & $-0.1$  \\
AL & 7 & 37.1 & 0D & $-37.1$ & MI & 13 & 51.4 & 7D & $+2.4$  \\
IA & 4 & 45.8 & 1D & $-20.8$ & CO &  8 & 56.9 & 5D & $+5.6$  \\
WI & 8 & 50.3 & 2D & $-25.3$ & IL & 17 & 58.7 &11D & $+6.1$  \\
NC &14 & 49.3 & 5D & $-13.6$ & OR &  6 & 58.3 & 4D & $+8.4$  \\
PA &17 & 50.6 & 8D & $-3.5$  & VA & 11 & 55.2 & 8D & $+17.6$ \\
AZ &  9 & 50.2 & 4D & $-5.7$  & NY & 26 & 61.7 &21D & $+19.0$ \\
FL &28 & 48.3 &12D & $-5.5$  & CA & 52 & 64.9 &45D & $+21.6$ \\
NV &  4 & 51.2 & 2D & $-1.2$  & VT &  1 & 68.3 & 1D & $+31.7$ \\
\bottomrule
\end{tabular}
\caption{Selected MEC plan results. Full 50-state table in supplementary material.
  $k$ = districts; $v_s$ = statewide Dem two-party vote share; MEC = minimum
  edge-cut plan; $\Delta_s$ = proportionality gap ($+$ = Dem over-represented).
  200-seed results; convergence certified separately for focal states.}
\label{tab:national}
\end{table}

Key hypotheses to evaluate:
\begin{itemize}
  \item \textbf{H1 (at-large states)}: States with $k=1$ (VT, AK, WY, DE, ND,
        SD, MT) have zero seed variance. \emph{Confirmed for Vermont.}
  \item \textbf{H2 (competitive states)}: States with 10--20 competitive
        districts show the largest seed variance in partisan outcomes.
        \emph{Pennsylvania preliminary data: 3-seat swing across 100 seeds.}
  \item \textbf{H3 (sorted states)}: Deep-red and deep-blue states show low
        seed variance because geography dominates. \emph{TBD.}
  \item \textbf{H4 (MEC stability)}: The minimum-edge-cut plan is a typical
        draw from the near-minimum distribution, not a partisan outlier.
        \emph{Pennsylvania: MEC plan (8D) sits in the interior of the
        distribution, not at the partisan extreme.}
\end{itemize}
