\section{Discussion}
\label{sec:discussion}

\subsection{The Fundamental Theorem: Proportional Outcomes Require Partisan Input}

Our empirical results across [N] states establish a striking pattern. In states
with geographically sorted partisan electorates --- where one party's voters
cluster in urban cores while the other's spread across rural areas --- the
minimum-edge-cut criterion systematically produces fewer seats for the
concentrated party than their statewide vote share would predict.

Wisconsin is the paradigmatic case. With 8 congressional districts and a 50.3\%
Democratic vote share, the proportional allocation is exactly 4D/4R. The
200-seed MEC plan yielded 2D/6R ($-25.3$\,pp) --- but this was a false floor.
The 1{,}000-seed deep sweep converged to \textbf{3D/5R ($-12.8$\,pp)} at
1{,}642\,km (seed 318), with \textsc{CompactBisect} independently agreeing on
3D/5R at 1{,}584\,km. The maximum edge-cut plan yields 4D/4R ($-0.3$\,pp).
The Pearson correlation between edge-cut and $|\Delta|$ across 200 seeds is
$r = -0.166$: lower edge-cut correlates with worse proportionality.

This is not an artefact of the algorithm; it is a property of the geography.
Wisconsin Democrats concentrate in Milwaukee, Madison, and Green Bay. A
compact partition --- one that minimises the interface between partisan regions
--- creates compact urban districts (efficiently Democratic) and compact rural
districts (efficiently Republican). The result is the same as deliberate
partisan packing, achieved without any partisan input.

\begin{theorem}[Informal]
\label{thm:proportionality}
In any state with geographically sorted partisan electorates, no purely
geometric criterion can guarantee proportional seat allocation. Specifically,
any criterion that rewards geographic compactness will tend to produce
\emph{less} proportional outcomes in sorted states, because compactness
reinforces the natural clustering of voters.
\end{theorem}

The proof sketch: let $D$ and $R$ be the geographic regions where Democrats
and Republicans are the majority. If $|D| < |R|$ in area (urban concentration)
but $|D| \approx |R|$ in population (near parity), then any algorithm that
minimises boundary length will draw $D$ as a small number of dense districts
and $R$ as a larger number of sparse districts. The resulting seat share for
the concentrated party falls below their vote share. No geometric criterion
can correct this without knowledge of which regions are $D$ and which are $R$
--- that is, without partisan data.

\subsection{Proportional Bisection: The Minimum Partisan-Input Algorithm}

We propose \emph{Proportional Bisection} as the minimum-partisan-input
algorithm that achieves near-proportional outcomes. At each bisection level,
the algorithm computes the Dem vote share $v_s$ within the current subgraph
$G_s$ and sets the district allocation ratio:
\[
  k_{\mathrm{left}} = \lfloor v_s \cdot k \rceil, \qquad
  k_{\mathrm{right}} = k - k_{\mathrm{left}}.
\]
METIS then minimises edge-cut subject to the population constraint implied by
this district count ratio. The boundary location is entirely geometric; only
the allocation ratio uses partisan data.

Proportional Bisection is ``minimum partisan input'' in the following sense:
it uses no more partisan information than is needed to state the proportionality
target (the vote share $v_s$). It does not use:
\begin{itemize}
  \item Which specific tracts belong to which party (it treats vote share as a
        scalar property of the subgraph, not an assignment to individual tracts)
  \item Which \emph{side} of the boundary is Democrat-leaning (METIS draws the
        boundary; which side is left vs. right is not specified)
  \item Any race, partisan registration, or candidate identity data
\end{itemize}

In practice, Proportional Bisection is implemented by feeding the per-vertex
Democratic vote share into the METIS target-weight parameter, setting
$\mathrm{tpwgts} = (v_s, 1-v_s)$ rather than the standard $(0.5, 0.5)$.

\paragraph{Legal status.}
The Districting Integrity Act \S~104(e) prohibits partisan data as algorithm
input for federal congressional districts. Proportional Bisection falls within
this prohibition: the vote share $v_s$ is derived from past election results.

For state legislative redistricting --- which the Act explicitly does not
govern (\S~101(d)) --- Proportional Bisection is available and appropriate.
States that wish to achieve proportional representation in their legislative
chambers without specifying partisan district assignments can mandate
Proportional Bisection.

\subsection{Compactness as an Intermediate Criterion}

Our Wisconsin analysis reveals a three-way ordering of criteria by
proportionality:

\begin{enumerate}
  \item \textbf{Proportional Bisection}: 4D/4R ($-0.3$\,pp). Requires
        partisan vote data; achieves near-perfect proportionality.
  \item \textbf{Max Compactness (\textsc{CompactBisect})}: 3D/5R ($-12.8$\,pp,
        1{,}584\,km). No partisan data; agrees with deep-sweep MEC.
  \item \textbf{Min Edge-Cut (1{,}000 seeds)}: 3D/5R ($-12.8$\,pp, 1{,}642\,km).
        Same partisan outcome as \textsc{CompactBisect}; slightly less compact.
        (200-seed false floor was 2D/6R at 1{,}689\,km.)
  \item \textbf{Max Edge-Cut}: 4D/4R ($-0.3$\,pp). Least compact; most proportional.
\end{enumerate}

The provisional convergence of criteria 2 and 3 on the same 3D/5R outcome
is an encouraging methodological result: given sufficient seeds, MEC and
\textsc{CompactBisect} appear to agree. The difference is the path:
\textsc{CompactBisect} finds 3D in 50 seeds per level; MEC requires 318
full-state seeds. However, MEC (seed 891) has not yet reached the 300-seed
certification tail, and neither has GMPP (seed 930, 69-seed tail with exact
perimeters). Both need further sweeps to be formally certified.

The key empirical finding: maximum Polsby-Popper selects \emph{3D}, not 2D.
With exact TIGER perimeters (EPSG:5070), the GMPP-maximising level-1 bisection
(seed 930) produces 3D/5R at GMPP\,=\,0.397; the MEC-minimising plan (seed 891)
produces 3D/5R at GMPP\,=\,0.358. Both criteria agree on outcome; the
MEC plan is 58\,km more compact in edge-cut, the GMPP plan 0.039 higher in
geometric-mean Polsby-Popper.

Why does PP perform better than edge-cut for proportionality?
Polsby-Popper is \emph{size-normalised}: $\mathrm{PP}(R) = 4\pi A / P^2$
scales with the ratio of area to perimeter squared, treating small urban
districts and large rural districts equivalently. A compact Milwaukee
district (small area, many people, round shape) and a compact rural district
(large area, few people, round shape) can achieve identical PP scores.
Edge-cut, by contrast, accumulates the absolute boundary lengths of all cut
edges. Urban tracts have many short adjacency edges; rural tracts have fewer
but longer edges. Minimising edge-cut therefore creates fewer rural districts
(respecting long natural boundaries) and fewer urban districts (packing dense
tracts together) --- the same outcome as deliberate packing.

By using PP rather than edge-cut as the selection criterion within the
near-minimum-cut candidate set, \textsc{CompactBisect} partially corrects
the geographic-sorting bias of pure edge-cut minimisation. It does not
eliminate the bias entirely: the most proportional Wisconsin plan (4D/4R)
has PP = 0.3825, \emph{lower} than the max-compactness plan's PP = 0.4750.
Achieving full proportionality requires drawing some non-compact districts
--- specifically, splitting Milwaukee in a way that creates less circular
but more electorally balanced sub-regions. Perfect proportionality and
maximum compactness are in tension in geographically sorted states.

The \emph{Pearson correlation} between mean PP and $|\Delta|$ across 97
Wisconsin plans is $r = 0.047$ --- essentially no linear relationship. The
relationship is non-monotone: 3D plans (the modal outcome) have the highest
average PP (0.4078), while 2D plans (0.3934) and 4D plans (0.3916) score
lower. This confirms that the solution space does not organise along a
single compactness-proportionality axis; rather, compactness and partisan
outcomes are jointly determined by the specific geographic partitioning at
each bisection level.

\subsection{Implications for the Districting Integrity Act}

\paragraph{The statute achieves procedural, not substantive, fairness.}
Our results confirm that the Districting Integrity Act's procedural guarantee
--- partisan-blind, reproducible, geometric partitioning --- does not guarantee
proportional representation. In geographically sorted states, the minimum-edge-cut
plan will systematically under-represent the geographically concentrated party.

This is consistent with the statute's design. The statute explicitly does not
claim to achieve proportional representation:

\begin{quote}
\emph{The proportionality metric does not appear in the bill's required
reproducibility artifacts --- and that's intentional. The bill's claim is
procedural: did the State run the prescribed algorithm? It is not substantive:
did the resulting plan achieve proportional representation?}\\
\hspace*{\fill}--- \texttt{PARTISAN\_OPTIONS.md} \S~8
\end{quote}

The statute survives \textit{Rucho v. Common Cause} precisely because it makes
no substantive claim about partisan outcomes. Our empirical results provide the
quantitative characterisation of what the statute's procedural guarantee
\emph{does} produce: near-proportional outcomes in geographically mixed states
(Pennsylvania, Michigan, Arizona) and systematic under-representation of the
concentrated party in sorted states (Wisconsin, North Carolina).

\paragraph{Proportionality as a transparency metric.}
We recommend that the Districting Integrity Act's required reproducibility
artifacts (\S~105) include the proportionality gap $\Delta_s$ as a mandatory
reported diagnostic. The gap is computed from election data \emph{after} the
algorithm runs --- not as an input, but as an output characterisation. This
is legally clean: the algorithm does not optimise $\Delta_s$; it reports it.

Publishing $\Delta_s$ alongside the plan gives citizens, legislators, and courts
a quantitative measure of the geographic sorting effect in their state. A state
with $|\Delta_s| > 20$\,pp is not gerrymandered; it is geographically sorted.
The statute's procedural guarantee is a ceiling on what any algorithm can
achieve without importing partisan data; $\Delta_s$ measures how far below the
proportionality ceiling the geographic sorting places the outcome.

\paragraph{Seed-choice neutrality and content-derived seeds.}
Our convergence analysis demonstrates that the minimum-edge-cut plan is
stable once certified: for Pennsylvania, the running minimum did not improve
for 819 consecutive seeds after seed 181 (out of 1{,}100 total).

However, a challenger could still argue that any specific seed was chosen
post-hoc after seeing its partisan consequences. The cleanest defence is to
derive the seed from the census data itself, precluding any post-hoc choice:

\begin{quote}
\emph{The random seed shall be}
$\mathrm{SHA256}(\textit{census\_release\_id} \mathbin\| \texttt{"DIA\_SEED\_V1"})$
\emph{truncated to 64 bits, where}
$\textit{census\_release\_id}$
\emph{is the identifier published under \S~107(c)(4).}
\end{quote}

This seed is deterministic given public data, not chosen by any person after
knowing the partisan consequences, and uniform across all states. We recommend
this as the statutory default in lieu of Director-discretion seed publication.
The full $N$-seed sweep (with convergence certification) is the reproducibility
protocol; the content-derived seed is the statutory shortcut that any party
can verify reproduces the certified minimum.

\subsection{Relationship to Prior Work}

\paragraph{Ensemble methods.}
Our convergence analysis is related to but distinct from MCMC ensemble
methods \citep{chikina2017assessing}. MCMC generates a distribution of neutral
plans to establish a baseline; our seed sweep generates a distribution of
local optima of the edge-cut objective. The key difference: MCMC samples
uniformly from the feasible partition space, while our sweep samples from the
attraction basins of the edge-cut objective's local optima. The seed sweep
is less diverse (it over-samples near-optimal plans) but more useful for
identifying the global minimum.

\paragraph{Efficiency gap.}
The efficiency gap \citep{stephanopoulos2015partisan} measures partisan bias
at a fixed vote share; our proportionality gap measures seat allocation at the
\emph{actual} vote share. The two metrics are complementary: efficiency gap
diagnoses whether the map systematically favours one party; proportionality
gap diagnoses whether the actual outcome is representative. A plan can have a
small efficiency gap (low systematic bias) and a large proportionality gap
(poor actual proportionality) --- which is exactly what we observe in Wisconsin
under the minimum-edge-cut criterion.
