\subsection{The Normative Case for Local Autonomy}

County-based representation rests on a foundational principle: \textbf{natural political communities should not be fragmented for state convenience}. This principle has three components.

\textbf{First, counties are genuine communities of interest}, not arbitrary lines. Unlike congressional districts, which exist only on maps and are redrawn every decade, counties have centuries of institutional history. Most U.S. county boundaries were established in the 18th or 19th century and have remained stable since. Counties provide unified governance (sheriffs, courts, health departments), economic integration (labor markets, tax bases), and civic identity (residents identify as "from Cook County" or "from Orange County"). As \citet{putnam2000bowling} documented, such natural communities foster social capital and civic engagement in ways that arbitrary electoral boundaries cannot.

\textbf{Second, subsidiarity demands local control over local representation}. If a county of 10 million people governs itself through elected county officials, elects its own mayor (in consolidated city-counties), and makes local decisions about education, law enforcement, and land use, why should a distant state capital control how that county is represented in Congress? The subsidiarity principle---decisions made at the most local level possible---applies as much to representation as to governance. A state legislature in Sacramento or Albany fragmenting Los Angeles or Brooklyn for state-level political goals violates local autonomy.

\textbf{Third, permanence matters}. County boundaries are fixed and cannot be manipulated. Congressional districts, by contrast, can be gerrymandered every decade. Even well-intentioned redistricting creates uncertainty: which communities will be united or divided next cycle? County-based representation eliminates this instability. Los Angeles County will remain Los Angeles County for centuries; its congressional districts change every 10 years.

These three principles---community coherence, subsidiarity, and permanence---justify county-based representation independent of partisan effects. If Republicans controlled large counties, the same principles would apply. Local autonomy transcends party.

\subsection{Voting Rights Act Compatibility}

A common concern is Voting Rights Act compliance. How can counties create majority-minority districts if they receive block representation? We offer three responses.

\textbf{First, counties can conduct internal redistricting}. Just as states divide their populations into districts, qualifying counties would divide theirs. Los Angeles County, allocated 13 representatives under our system, would draw 13 internal districts, some of which could be majority-Latino or majority-Black to satisfy VRA requirements \citep{grofman1982representation}. The difference is that \emph{only LA County residents} participate in this redistricting, not the entire state of California. This actually makes VRA compliance \emph{easier}: LA County's minority populations are geographically concentrated, allowing compact majority-minority districts without crossing county lines.

\textbf{Second, VRA has always required balancing}. Even under current redistricting, states must balance VRA compliance with compactness, contiguity, and communities of interest. County-based representation adds ``respect for county boundaries'' to this balance. Courts have upheld various accommodations \citep{thornburg1986gingles}; county-internal VRA redistricting is one more.

\textbf{Third, many qualifying counties are already diverse enough}. Cook County is 24\% Black, 25\% Latino. LA County is 48\% Latino. These counties can internally create majority-minority districts while maintaining countywide elections for their full delegation. This is similar to multi-member districts with VRA-compliant subdistricts, a system upheld in state legislative contexts \citep{zimmer1973voting}.

VRA concerns do not invalidate county-based representation; they require thoughtful implementation.

\subsection{The Partisan Asymmetry Problem}

We have disclosed that county-based representation currently advantages Democrats. This raises a legitimacy question: is this a principled reform or partisan proposal in disguise?

Our response has four parts. \textbf{First, partisan composition is historically contingent}. Orange County, California---one of our qualifying counties---was a Republican stronghold from the 1950s-1990s, electing conservative icons like Robert Dornan. It flipped Democratic only in 2016 due to demographic change and Trump-era realignment \citep{riker2019orange}. Maricopa County (Phoenix) shifted from reliably Republican to swing only in 2020. Conversely, many Midwestern counties shifted from Democratic to Republican with deindustrialization. If party coalitions realign again---say, Republicans win college-educated suburbanites while Democrats win non-college rural voters---county-based representation could advantage Republicans. The principle of local autonomy, once established, is permanent regardless of temporary partisan effects.

\textbf{Second, current gerrymandering also has partisan bias}. We do not start from a neutral baseline. Republicans currently gain 16-30 House seats through strategic redistricting in states they control \citep{chen2013unintentional}. Democrats gain fewer seats through their gerrymanders (Illinois, Maryland) but still benefit in those states. County-based representation eliminates \emph{both parties'} manipulation for 21\% of the population. The net effect may favor Democrats now, but it levels the playing field going forward: neither party can gerrymander LA, Cook, or Harris Counties in future cycles.

\textbf{Third, principled reforms often have asymmetric impacts}. The strongest precedent is \emph{Reynolds v. Sims} (1964), which required state legislative districts to be population-balanced \citep{reynolds1964sims}. This massively advantaged Democrats by shifting power from overrepresented rural (Republican) areas to underrepresented urban (Democratic) areas. Yet the Court ruled 8-1 that ``one person, one vote'' was constitutionally required, regardless of partisan consequences. Similarly, eliminating literacy tests benefited Democrats by enfranchising Black voters in the South. Lowering the voting age to 18 benefited Democrats by enfranchising youth. In each case, the principle (equality, voting rights, age consistency) was accepted despite helping one party. If local autonomy is a valid principle, current partisan asymmetry does not invalidate it.

\textbf{Fourth, we separate description from prescription}. We \emph{describe} the partisan reality (Democrats gain 15-25 seats under current alignments) but \emph{prescribe} on principled grounds (natural communities should not be fragmented). Academic honesty requires full transparency about partisan effects; normative argument requires defending the underlying principle. We do both.

\subsection{Implementation Challenges}

County-based representation raises practical questions. How would the House accommodate representatives from different-sized counties? Would LA County's 13 representatives vote as a block or individually? How would committees handle fractional members?

One approach is \textbf{individual representatives with county elections}. LA County would hold 13 separate House elections, perhaps using ranked-choice voting or multi-member districts internally. Each elected representative votes individually in Congress. Committees receive members in proportion to county delegation size. This requires minimal House rule changes.

An alternative is \textbf{weighted voting}, where each county's delegation casts votes proportional to its population. LA County's 13 representatives would collectively cast 13 votes, but if a small county got 0.5 representatives, they would cast 0.5 votes. This achieves perfect proportionality but requires significant House procedural innovation and may face constitutional challenges about what constitutes a ``representative'' \citep{banzhaf1968power}.

A third option is \textbf{hybrid}: large qualifying counties use individual representatives (integer seats), while very small state pools might use weighted voting if their population warrants less than 1 seat. This combines familiar procedures (individual reps) with proportional fairness where needed.

These implementation questions are significant but not insurmountable. The House has adapted procedures many times (electronic voting, committee reforms, televised proceedings). If county-based representation is normatively justified, procedural innovation can follow.

\subsection{Comparison to Other Redistricting Reforms}

How does county-based representation compare to alternative reforms?

\textbf{Independent commissions} (California, Michigan, etc.) reduce partisan gerrymandering but still require drawing artificial boundaries every decade. Commissioners face the same trade-offs: compact districts may split communities; respecting communities may create non-compact districts; VRA compliance may require both compromises \citep{cain1988reapportionment}. County-based representation eliminates these trade-offs by using fixed boundaries.

\textbf{Algorithmic redistricting} \citep{duchin2018geometry, procaccia2021axioms} uses optimization algorithms to draw ``fair'' districts. While this reduces human manipulation, it still creates arbitrary boundaries with no community basis. An algorithm-drawn district is as much an artificial construct as a legislature-drawn one. Residents have no say in their ``community'' beyond population data inputs.

\textbf{Multi-member districts} with proportional representation \citep{amy2000behind} provide fairer partisan outcomes but typically use arbitrary boundaries (``Region 3'' or ``District 5-7''). County-based representation combines proportional fairness \emph{with natural community boundaries}---the best of both approaches.

\textbf{At-large elections} avoid redistricting but violate VRA requirements for geographically concentrated minorities \citep{thornburg1986gingles}. County-based representation allows county-internal redistricting to satisfy VRA while maintaining county autonomy from state control.

In short, county-based representation offers a unique combination: no arbitrary boundary drawing, natural community representation, VRA compatibility via county-internal redistricting, and automatic adjustment to population change without decadal trauma.

\subsection{Limitations and Future Work}

Our analysis has limitations. First, we use 2020 data only; multi-census analysis would test stability of qualifying counties over time. Second, we lack complete partisan data for all counties; full election results would improve accuracy. Third, we do not model county-internal VRA redistricting; demonstrating compliance would strengthen feasibility.

Future work should examine: (1) alternative community units (media markets, economic regions), (2) weighted voting constitutionality via legal analysis, (3) public opinion surveys on county-based representation acceptability, and (4) comparison to proportional representation systems in other democracies. Additionally, analyzing 2000 and 2010 census data would test whether partisan effects are historically stable or era-specific.

The most valuable extension would be demonstrating \textbf{actual enacted plan fragmentation} by overlaying congressional district shapefiles with county boundaries. This would move from estimated fragmentation (population / ideal district size) to measured fragmentation (number of actual districts intersecting each county). Such analysis would provide definitive evidence of current county division.
