\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{algorithm}
\usepackage{algorithmic}

% Define theorem environments
\newtheorem{proposition}{Proposition}

\title{Cross-Census Temporal Stability:\\
Recursive Bisection vs N-Way Partitioning Over Decades}
\author{Giovanni Della-Libera}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Congressional redistricting occurs every decade following the census, raising critical questions about boundary stability over time. While graph partitioning methods optimize current-census districts, their temporal stability across census cycles remains unexplored. This paper presents the first systematic comparison of hierarchical recursive bisection versus flat n-way partitioning methods for maintaining district continuity across the 2010-2020 decade.

We analyze five southern states with significant minority populations, running both methods on 2010 and 2020 census data (20 total redistricting runs). Our key finding: \textbf{recursive bisection provides measurably better temporal stability}, with 71.6\% population disruption versus 72.4\% for n-way partitioning (0.8 percentage point improvement, representing 1.1\% less disruption). This small but consistent advantage appears in 4 out of 5 states tested, suggesting that hierarchical structure preserves coarse-level geographic divisions more effectively than direct k-way optimization.

The stability advantage is modest but meaningful. For a state with 5 million residents, recursive bisection results in approximately 40,000 fewer people reassigned to different districts across the decade. Both methods maintain comparable district quality on current-census metrics (population balance, minority representation, compactness), indicating that recursive bisection's temporal benefits come without sacrificing spatial optimization performance.

These findings suggest redistricting method choice should consider not only current optimization but also future continuity. The 1.1\% stability advantage, while small, is consistent across diverse demographic and geographic contexts. States adopting recursive bisection in 2020 should experience modestly easier 2030 transitions, a prediction testable with forthcoming census data.
\end{abstract}

\section{Introduction}
\input{sections/01_introduction}

\section{Background}
\input{sections/02_background}

\section{Methodology}
\input{sections/03_methodology}
\input{sections/03_section_3.5_hierarchical_validation}
\input{sections/03_section_3.6_theoretical_foundation}

\section{Results}
\input{sections/04_results}
\input{sections/04_section_4.4_vra_compliance}

\section{Discussion}
\input{sections/05_discussion}

\section{Limitations}
\input{sections/06_limitations}

\section{Future Work}
\input{sections/07_future_work}

\section{Conclusion}
\input{sections/08_conclusion}

\bibliographystyle{plain}
\bibliography{references}

\end{document}
