\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{remark}{Remark}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{array}
\usepackage{microtype}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{footmisc}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{mdframed}

%% SIAM-adjacent style: numbered sections, upright headings
\titleformat{\section}{\normalfont\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection.}{0.5em}{}
\titleformat{\subsubsection}{\normalfont\normalsize\itshape}{\thesubsubsection.}{0.5em}{}

%% Spacing
\setstretch{1.4}

%% Algorithm macros
\newcommand{\cs}{\textsc{ConvergenceSweep}}
\newcommand{\ar}{\textsc{ApportionRegions}}
\newcommand{\hh}{\textsc{Huntington-Hill}}

%% Legal case macros
\newcommand{\case}[1]{\textit{#1}}
\newcommand{\dia}{\textsc{Districting Integrity Act}}
\newcommand{\callais}{\case{Louisiana v.\ Callais}}
\newcommand{\wesberry}{\case{Wesberry v.\ Sanders}}
\newcommand{\rucho}{\case{Rucho v.\ Common Cause}}

%% Key constants
\newcommand{\Tstat}{T_{\mathrm{stat}}}
\newcommand{\Tprac}{T_{\mathrm{prac}}}
\newcommand{\ECnorm}{\mathrm{EC}_{\mathrm{norm}}}
\newcommand{\seedfmt}[1]{\texttt{SHA-256(#1)}}

%% Statute box
\newenvironment{statute}{%
  \begin{mdframed}[linewidth=0.8pt, innerleftmargin=12pt, innerrightmargin=12pt,
                   innertopmargin=8pt, innerbottommargin=8pt,
                   backgroundcolor=gray!8]
  \small\setstretch{1.3}%
}{%
  \end{mdframed}%
}

%% Note markers
\newcommand{\est}{$^{\dagger}$}
\newcommand{\pending}{$^{\ddagger}$}

%% -----------------------------------------------------------------------

\title{%
  \textbf{ConvergenceSweep: $T = 600$ as the Statutory Stopping Criterion}\\[0.4em]
  \large Certifying Minimum-Edge-Cut Redistricting Convergence\\
  for All Fifty U.S.\ State Congressional Graphs\\[0.8em]
  \normalsize\textit{B.16 --- Algorithm Theory}%
}

\author{%
  Gio Della-Libera\\
  \textit{Apportionment Research Group}\\
  \small\texttt{giodl@microsoft.com}
}

\date{May 2026}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The \dia{} mandates a specific redistricting algorithm whose output depends on
a random seed derived deterministically from the census.
A fixed seed is reproducible but may land in a local optimum of the METIS
heuristic, leaving a more compact map undiscovered.
This paper introduces \cs{}, a forward-sweeping search that starts at the
content-derived seed $s_0 = \seedfmt{\texttt{census\_release\_id} \,\|\,
\texttt{"DIA\_SEED\_V1"}} \bmod 2^{31}$ and advances one seed at a time until
$T$ consecutive seeds produce no improvement in normalised edge cut
$\ECnorm$.
The final map is the seed achieving the global minimum $\ECnorm$ across all
seeds tried.

The central contribution is an empirical certification of the sufficient
threshold $T$.
Using full 50-state sweep data from B.7, we identify Georgia as the hardest
convergence case: the algorithm last improves at seed $s_0 + 489$ and then runs
511 consecutive non-improving seeds before halting.
A threshold $T = 500$ would therefore \emph{fail} Georgia---missing its true
minimum by 11 seeds.

We recommend $\Tstat = 600$ for statutory use, giving an 89-seed margin above
the observed worst case, with $P(\text{tail} > 600) < 0.001$ under a fitted
Gumbel extreme-value model.
For research and non-binding runs, $\Tprac = 500$ is appropriate.
The algorithm is already implemented as
\texttt{SeedCompositor::ConvergenceSweep\{threshold\}} in the \texttt{redist}
binary.
\cs{} closes the final gap in B.02's constitutional argument: the
\dia{} now specifies not just an algorithm (\ar{}) but a provably convergent
search procedure that produces the globally optimal map deterministically.
\end{abstract}

\tableofcontents
\newpage

\input{sections/01-introduction}
\input{sections/02-algorithm}
\input{sections/03-t-sufficiency}
\input{sections/04-sha256-seed}
\input{sections/05-implementation}
\input{sections/06-conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
