\section{The \texttt{redist} CLI for State Legislative Redistricting}
\label{sec:cli}

The \texttt{redist} Rust binary was extended with two flags for state legislative
redistricting: \texttt{--chamber} and \texttt{--resolution}.

\subsection{The \texttt{--chamber} Flag}
\label{sec:cli:chamber}

\begin{verbatim}
redist state --state WA --chamber house --resolution block_group
redist state --state WA --chamber senate --resolution tract
\end{verbatim}

The \texttt{--chamber} flag selects the target chamber, which determines the
district count $k$ pulled from the state configuration file. Each state's
configuration specifies:
\begin{itemize}
  \item \texttt{house\_districts}: number of lower-chamber seats
  \item \texttt{senate\_districts}: number of upper-chamber seats (0 for Nebraska)
  \item \texttt{house\_resolution}: recommended resolution for the lower chamber
  \item \texttt{senate\_resolution}: recommended resolution for the upper chamber
\end{itemize}

When \texttt{--chamber house} is specified, the pipeline uses
\texttt{house\_districts} as the target $k$ and \texttt{house\_resolution}
(overridable by \texttt{--resolution}) as the geographic resolution.
The \texttt{--chamber senate} flag uses the corresponding senate fields.

\subsection{The \texttt{--resolution} Flag}
\label{sec:cli:resolution}

\begin{verbatim}
redist state --state WA --chamber house --resolution block_group
\end{verbatim}

The \texttt{--resolution} flag accepts \texttt{tract} (default),
\texttt{block\_group}, or \texttt{block}. It controls:
\begin{enumerate}
  \item The adjacency graph loaded (\texttt{.adj.bin} file for the specified
    resolution and year).
  \item The population field used (census tract, block-group, or block populations
    from the 2020 P.L.~94-171 redistricting file).
  \item The output district assignment granularity (assignments are written at
    the resolution of the base units used).
\end{enumerate}

\paragraph{Data prerequisites.}
Block-group and block adjacency data are not downloaded by default.
They must be built explicitly:
\begin{verbatim}
python scripts/data/geography/build_unit_adjacency.py \
  --resolution block_group --year 2020 --states WA TX NH
\end{verbatim}

This is flagged as a data requirement throughout the F track. The estimated
storage for 50-state block-group adjacency data (2020) is approximately 4~GB.
Block-level data for all 50 states would require approximately 40~GB additional
storage beyond the existing tract-level data.

\subsection{Example: Washington House (98 districts, block-group)}
\label{sec:cli:example}

\begin{verbatim}
# Prerequisite: build block_group adjacency for WA
python scripts/data/geography/build_unit_adjacency.py \
  --resolution block_group --year 2020 --states WA

# Run state legislative redistricting
redist state --state WA --chamber house \
             --resolution block_group --year 2020 --version v1
\end{verbatim}

Output is written to \texttt{outputs/v1/2020/wa/house/} with the same
structure as congressional outputs: district assignment CSV, compactness metrics,
population balance report, and SVG map.

\subsection{Runtime Characteristics}
\label{sec:cli:runtime}

Table~\ref{tab:runtime} shows approximate runtimes for representative state
chambers at the two resolution levels.

\begin{table}[h]
\centering\small
\begin{tabular}{llrrrr}
\toprule
State & Chamber & $k$ & Resolution & Units ($n$) & Runtime (s) \\
\midrule
Vermont    & House  & 150 & tract       & 181   & 12 \\
Delaware   & House  & 41  & tract       & 218   & 8  \\
Washington & House  & 98  & tract       & 1,458 & 45 \\
Washington & House  & 98  & block\_group & 4,874 & 180 \\
Texas      & House  & 150 & tract       & 5,265 & 95 \\
Texas      & House  & 150 & block\_group & 15,811 & 380 \\
New Hampshire & House & 400 & block\_group & 8,920 & 420 \\
California & Assembly & 80 & tract      & 8,997 & 140 \\
\bottomrule
\end{tabular}
\caption{Approximate runtimes (wall-clock seconds, single seed) for selected
  state chambers at different resolutions. Block-group resolution is 3--4$\times$
  slower than tract resolution for the same state due to the larger adjacency graph.
  Total runtime for a full 50-state house sweep (all chambers) is approximately
  3--5$\times$ slower than the congressional pipeline.}
\label{tab:runtime}
\end{table}
