\section{Key Findings Preview}
\label{sec:findings}

We preview here the main empirical findings of the F track. Full supporting
tables and case studies appear in the companion papers.

\subsection{Compactness Exceeds Congressional Baseline}
\label{sec:findings:compactness}

The most striking finding of F.1 is that algorithmically generated state
house maps are \emph{more compact} than their congressional counterparts by
the Polsby-Popper measure. Table~\ref{tab:pp-comparison} summarises the result.

\begin{table}[h]
\centering\small
\begin{tabular}{lrr}
\toprule
Setting & Resolution & Mean PP \\
\midrule
Congressional (B track) & tract & 0.361 \\
State house, tract resolution & tract & 0.381 \\
State house, block-group resolution & block\_group & 0.401 \\
\bottomrule
\end{tabular}
\caption{Mean Polsby-Popper scores across settings. State house redistricting
  systematically achieves higher compactness than congressional redistricting.
  The gap is explained by scale effects (F.5): smaller districts have smaller
  perimeters relative to their area.}
\label{tab:pp-comparison}
\end{table}

As F.5 shows, this is primarily a scale effect. Polsby-Popper is defined as
$PP = 4\pi A / P^2$. For a fixed-shape district, scaling down by factor $s$
preserves $PP$ exactly ($A \to s^2 A$, $P \to s P$). However, in practice
smaller districts have \emph{relatively shorter} perimeters because they
can be carved from compact sub-regions of the state rather than spanning
the entire state geography. The recursive bisection algorithm exploits this:
each level of the hierarchy optimises the boundary of a sub-region, and
smaller sub-regions have fewer constraints from state-boundary artefacts.

\subsection{Population Balance Within $\pm 0.5\%$ in All 50 States}
\label{sec:findings:balance}

All 50 state house maps achieve population balance within $\pm 0.5\%$ of
the ideal district population. This is the standard applied to congressional
redistricting under \emph{Wesberry v.\ Sanders} (1964). State legislative
plans are governed by the less stringent standard of \emph{Reynolds v.\ Sims}
(1964), which requires ``substantial equality'' rather than precise equality;
courts have generally accepted deviations up to $\pm 5\%$ for state legislative
plans \citep{karcher1983karcher}. The \texttt{redist} pipeline achieves a
tighter standard than legally required, confirming that the algorithm's
population-balance constraints are conservative.

\subsection{County Splits Reduced 15--25\%}
\label{sec:findings:splits}

Across the 38 states for which enacted 2020 state house maps are available
for comparison, the algorithmic maps split counties 15--25\% fewer times
on average. This reflects the algorithm's implicit preference for geographic
compactness: county boundaries are often natural geographic divides, and
the edge-cut minimisation objective tends to respect them without an
explicit constraint.

\subsection{Runtime: 3--5$\times$ Slower Than Congressional}
\label{sec:findings:runtime}

The full 50-state state-house sweep requires approximately 3--5$\times$ the
wall-clock time of the congressional pipeline run at comparable resolution.
The primary contributor is the resolution upgrade required for high-$k$ chambers:
block-group adjacency graphs are $\sim\!3\times$ larger than tract graphs.
For tract-resolution states (the majority), per-state runtimes are similar to
congressional times. For block-group-resolution states, runtimes are 3--4$\times$
longer per state.

\subsection{Nesting Success Rate: 41/49 Bicameral States}
\label{sec:findings:nesting}

Of the 50 states, 49 are bicameral (Nebraska is unicameral). Of the 49 bicameral
states, NestSection finds a compatible spine---$\gcd(H,S) > 1$---for 41 states.
The 8 incompatible states have $\gcd(H,S) = 1$, meaning house and senate district
counts are coprime. Examples include Maine ($H=151$, $S=35$, $\gcd=1$) and
Delaware ($H=41$, $S=21$, $\gcd=1$).
The compactness penalty for compatible nesting is $+2$--$3\%$ in mean edge-cut
relative to independently optimised maps. Partisan outcomes shift by at most
$0.3$ seats on average.
