\section{Methodology}
\label{sec:methodology}

\subsection{Data and geographic configurations}

We use 2020 Census TIGER/Line shapefiles~\citep{tiger-census} at two
resolutions: census tracts ($n \approx 74{,}000$ nationally) and block
groups ($n \approx 239{,}000$).
Adjacency graphs are built by detecting shared tract boundaries with a
minimum shared-length threshold of 50 metres, following the procedure of
B.2.  Edge weights are shared boundary lengths in metres.

\paragraph{Congressional chambers.}
The 2020 apportionment yields 435 districts across 50 states, with $k$
ranging from 1 (7 states: AK, DE, MT, ND, SD, VT, WY) to 52 (California).

\paragraph{State senate chambers.}
We test all 49 unicameral-or-senate chambers (Nebraska is excluded as
unicameral with staggered terms that do not map to equal-population
districts).  District counts range from 20 (Nevada) to 67 (Georgia).

\paragraph{State house chambers.}
All 50 state house chambers are included, with $k$ ranging from 40
(Alaska) to 400 (New Hampshire).  This range is where n-way partitioning
might be expected to have a larger runtime advantage.

\subsection{Algorithm configurations}

Both strategies use the same METIS library (version 5.1.0, C FFI interface)
with identical hyperparameters:

\begin{itemize}
  \item \texttt{ufactor = 5} (maximum population imbalance: 0.5\%)
  \item \texttt{niter = 100} (FM refinement iterations)
  \item \texttt{ncuts = 5} (number of initial partitions)
  \item \texttt{seed = 42} (for reproducibility; B.7 shows seed sensitivity
    is low)
\end{itemize}

\paragraph{Recursive bisection (RB).} Uses METIS \texttt{PTYPE\_RB} with
target population tolerance split between levels.
For non-power-of-two $k$, we use an asymmetric bisection tree:
at each level, one child receives $\lfloor k/2 \rfloor$ districts and
the other $\lceil k/2 \rceil$.
For prime $k$ (e.g.\ PA $k{=}17$, NV $k{=}41$), this means unequal
sub-problems at every level.

\paragraph{N-way (direct k-way).} Uses METIS \texttt{PTYPE\_KWAY}, which
performs multilevel coarsening to approximately $k \cdot c$ nodes (with
$c \approx 20$) and then uses greedy k-way initial partitioning followed
by FM refinement.

\subsection{Metrics}

\paragraph{Compactness.}
We compute the Polsby--Popper score for each district using the
GeoPandas~\citep{geopandas}/Shapely~\citep{shapely} geometry pipeline:
$\mathrm{PP}_i = 4\pi A_i / P_i^2$,
where $A_i$ is the area and $P_i$ the perimeter of district $i$.
The chamber mean $\overline{PP}$ is the primary outcome variable.

\paragraph{Runtime.}
Wall-clock time is measured on a 16-core AMD Ryzen 9 5950X workstation
(single-threaded METIS), averaged over 10 independent runs.
We report both raw runtime (ms) and runtime per tract (ms/tract) for
comparability across states.

\paragraph{Population balance.}
Maximum absolute deviation from the ideal population per district,
expressed as a fraction of the ideal: $\delta_{\max} = \max_i |w(P_i) - W/k| / (W/k)$.
Both strategies are constrained to $\delta_{\max} \leq 0.5\%$.

\subsection{Statistical testing}

State senate and state house chambers for the same state share the same
underlying census geography, so treating both as independent observations
in a joint test would violate the independence assumption.
To avoid this, we restrict the primary statistical test to
\textbf{congressional chambers only} ($n = 50$ states), which are
genuinely independent across states.

A paired $t$-test compares RB vs.\ n-way for congressional chambers,
pairing by state.  The null hypothesis is zero mean difference in
$\overline{PP}$.  We report the two-sided $p$-value; no correction is
applied because only one primary comparison is made.

State senate and state house results are reported descriptively
(Tables~\ref{tab:pp-chamber} and~\ref{tab:runtime}) to show that the
congressional finding generalises, but formal inference is drawn only
from the independent congressional sample.

Effect size is Cohen's $d$ computed on the congressional paired differences.
A difference is considered practically significant if $d > 0.2$ (small
effect) and statistically significant at $\alpha = 0.05$.
