\section{Evaluation}
\label{sec:evaluation}

\subsection{Experimental Setup}
\label{sec:eval:setup}

We implement PFR using the \texttt{redist} Rust binary (\texttt{redist-apportion}
crate) with METIS $k$-way partitioning at each tree level. The per-level balance
tolerance is set to $\varepsilon_{\text{level}} = \varepsilon_{\text{statutory}} /
(d+1)$ where $d$ is the actual PFR tree depth for that state (computed by
\texttt{pfr\_tree\_depth} in the implementation). The METIS minimum ufactor of
$0.1\%$ applies as a floor.

\paragraph{Population balance.}
In practice, METIS does not guarantee that its output respects the ufactor
constraint: the ufactor is a target, not a hard bound. For states whose
top-level split involves a large prime ($k \geq 7$), the achieved per-part
imbalance can reach 1--2\%, leading to final district deviations of 1.32--2.1\%
from the global ideal. This exceeds the statutory $\pm 0.5\%$
(\emph{Wesberry v.\ Sanders}). We report PFR results under a $3\%$ research
tolerance and note the achieved balance for each state. Reaching statutory
balance for PFR would require post-processing (e.g.\ tract-swap refinement on
the final partition) or tighter METIS constraints with restarts; we defer this
to future work.

\paragraph{Seed sensitivity.}
PFR is evaluated at 10 seeds per state (20 seeds for the focal states NC and
GA). Seed sensitivity results are in Section~\ref{sec:eval:seeds}. All partisan
outcomes are measured against 2020 presidential two-party precinct returns
interpolated to census tracts \citep{Kuriwaki2023}.

\subsection{Factorization Structure of the 2020 Apportionment}
\label{sec:eval:factorizations}

Table~\ref{tab:factorizations} shows the 2020 seat count for each state and its
canonical PFR factorization (largest prime first). Seven seat counts are prime and
use binary fallback at their first level: CT and OK ($k=5$, $\to 2+3$),
SC and AL ($k=7$, $\to 3+4$), VA ($k=11$, $\to 5+6$), MI ($k=13$, $\to 6+7$),
and PA and IL ($k=17$, $\to 8+9$). Twelve states with $k = 2^m$ have fully binary
trees (MT, RI, NH, ME, HI, ID, WV: $k=2$; KS, UT, NV, AR, IA, MS: $k=4$;
CO, WI, MO, MD, MN: $k=8$) and are identical to standard recursive bisection.

\begin{table}[h]
\centering\small
\begin{tabular}{lrll|lrll}
\toprule
State & $k$ & Factorization & Depth &
State & $k$ & Factorization & Depth \\
\midrule
VT,AK,WY,DE  & 1  & ---          & 0 & WA & 10 & $5 \times 2$        & 3 \\
ND,SD         & 1  & ---          & 0 & VA & 11 & 11 (prime)$\to$5+6  & 3 \\
MT,RI,NH,ME  & 2  & $2$          & 1 & NJ & 12 & $3\times 2^2$       & 3 \\
HI,ID,WV     & 2  & $2$          & 1 & MI & 13 & 13 (prime)$\to$6+7  & 4 \\
NE,NM        & 3  & $3$          & 1 & GA & 14 & $7\times 2$         & 2 \\
KS,UT,NV,AR  & 4  & $2^2$        & 2 & NC & 14 & $7\times 2$         & 2 \\
IA,MS        & 4  & $2^2$        & 2 & OH & 15 & $5\times 3$         & 2 \\
CT,OK        & 5  & 5 (prime)$\to$2+3 & 2 & PA,IL & 17 & 17 (prime)$\to$8+9 & 4 \\
OR,KY,LA     & 6  & $3\times 2$  & 2 & NY & 26 & $13\times 2$        & 5 \\
SC,AL        & 7  & 7 (prime)$\to$3+4 & 2 & FL & 28 & $7\times 2^2$  & 3 \\
CO,WI,MO     & 8  & $2^3$        & 3 & TX & 38 & $19\times 2$        & 2 \\
MD,MN        & 8  & $2^3$        & 3 & CA & 52 & $13\times 2^2$      & 3 \\
TN,AZ,MA,IN  & 9  & $3^2$        & 2 & & & & \\
\bottomrule
\end{tabular}
\caption{2020 seat counts, PFR factorizations (largest prime first), and tree depth.
  Depth = number of METIS calls per branch from root to leaf.
  Prime seat counts (5, 7, 11, 13, 17, 19) use binary floor/ceil fallback at
  that level, which is counted as 1 METIS call.}
\label{tab:factorizations}
\end{table}

\subsection{National Comparison: PFR vs.\ MEC}
\label{sec:eval:national}

Table~\ref{tab:national-pfr-mec} compares the PFR partisan outcome to the
minimum-edge-cut (MEC) outcome from \citet{dellaLibera2026mec} for all 48
states with available precinct election data (Alaska and Hawaii excluded).

\begin{table}[h]
\centering\small
\begin{tabular}{lrrrr|lrrrr}
\toprule
State & $k$ & MEC\,D & PFR\,D & $\Delta$ &
State & $k$ & MEC\,D & PFR\,D & $\Delta$ \\
\midrule
GA & 14 & 7 & 5 & $-2$ & MN & 8  & 3  & 4  & $+1$ \\
VA & 11 & 8 & 6 & $-2$ & TN & 9  & 1  & 2  & $+1$ \\
OR &  6 & 4 & 3 & $-1$ & TX & 38 & 15 & 16 & $+1$ \\
SC &  7 & 3 & 2 & $-1$ & NC & 14 & 5  & 7  & $+2$ \\
NY & 26 & 21& 20& $-1$ & \multicolumn{5}{l}{\emph{34 states: identical outcome}} \\
PA & 17 & 8 & 7 & $-1$ & & & & & \\
WI &  8 & 3 & 2 & $-1$ & & & & & \\
NJ & 12 & 10& 9 & $-1$ & & & & & \\
\midrule
\multicolumn{5}{l|}{MEC national (48 states): 228D/204R (52.8\%\,D)} &
\multicolumn{5}{l}{PFR national (48 states): 223D/209R (51.6\%\,D)} \\
\bottomrule
\end{tabular}
\caption{PFR vs.\ MEC partisan outcomes. $\Delta$ = PFR\,D $-$ MEC\,D.
  34 states have identical outcomes; 12 differ by 1--2 seats. PFR gives
  Democrats 5 fewer seats nationally ($-2.3\%$ seat share).
  MEC outcomes from \citet{dellaLibera2026mec} at convergence
  (1400 seeds for WI; 1000 seeds for GA, TN, MN, PA).}
\label{tab:national-pfr-mec}
\end{table}

PFR produces 223D/209R vs.\ MEC's 228D/204R --- a 5-seat Republican shift.
Both totals are within 2 seats of proportional given the ~51\% aggregate
Democratic two-party vote across the 48 states. Neither algorithm targets
proportionality, and neither is systematically biased: the 5-seat gap reflects
the interaction of factorization-induced boundary structures with specific
state geographies, not a consistent partisan tilt of the algorithm.

\subsection{North Carolina and Georgia: Same Factorization, Opposite Effects}
\label{sec:eval:nc-ga}

The paper's central empirical finding concerns two states with identical
factorizations: both Georgia ($k=14$) and North Carolina ($k=14$) factor as
$14 = 7 \times 2$. Under PFR, both states receive the same hierarchical
prescription: a 7-way primary split of the state, then bisect each of the 7
sub-regions.

\begin{table}[h]
\centering
\begin{tabular}{lrrrrrr}
\toprule
State & $v_s$ & Factorization & PFR\,D & PFR\,$\Delta$ & MEC\,D & MEC\,$\Delta$ \\
\midrule
North Carolina & 49.3\% & $7\times 2$ & 7D/7R & $+0.7$\,pp & 5D/9R & $-13.6$\,pp \\
Georgia        & 50.1\% & $7\times 2$ & 5D/9R & $-14.4$\,pp & 7D/7R & $-0.1$\,pp \\
\bottomrule
\end{tabular}
\caption{North Carolina and Georgia: identical $k=14=7\times 2$ factorization,
  opposite partisan effects. $v_s$ = statewide Dem two-party vote share.
  $\Delta$ = proportionality gap ($+$ = Dem over-represented).
  NC under PFR achieves near-perfect proportionality ($+0.7$\,pp) while MEC
  produces $-13.6$\,pp. GA under PFR produces $-14.4$\,pp while MEC achieves
  $-0.1$\,pp.}
\label{tab:nc-ga}
\end{table}

For North Carolina, the 7-way primary split groups the state's Democratic
precincts (Charlotte, the Research Triangle, Greensboro) into configurations
that, when each sub-region is subsequently bisected, produce seven Democratic
and seven Republican districts --- near-perfect proportionality for a 49.3\%
Dem state. The same algorithm applied to Georgia, where Democratic voters
concentrate in Atlanta and its suburbs, produces a 5D/9R outcome that is
\emph{less} proportional than the MEC result.

This is not an algorithmic artifact. It is the core thesis of this paper:
\emph{the partition structure is determined by the prime factorization of the
seat count, and its partisan effect is determined entirely by the interaction of
that structure with the geographic distribution of voters}. The algorithm has no
partisan inputs and makes no partisan choices. The NC/GA divergence demonstrates
that the same factorization tree can produce outcomes that favour either party
depending on geography --- precisely the constitutional property we seek.

\paragraph{Wisconsin: single-seed limitation.}
Wisconsin ($k = 8 = 2^3$) is a fully binary PFR tree, identical in structure
to recursive bisection. PFR produces 2D/6R ($-25.3$\,pp) at the fixed seed.
This matches the 200-seed MEC \emph{false floor} from \citet{dellaLibera2026mec}:
the single PFR seed finds an early local optimum that MEC only escapes at seed
318 after 1{,}400 total seeds. For states with large near-minimum solution
spaces, a multi-seed PFR protocol (run several seeds, take the minimum-EC
factorization-tree result) would be needed to replicate MEC convergence quality.

\subsection{Prime Seat Counts: Pennsylvania and Michigan}
\label{sec:eval:prime}

Pennsylvania ($k = 17$, prime) and Michigan ($k = 13$, prime) use binary
floor/ceil fallback at their first split level.

For Pennsylvania, PFR produces 7D/10R ($-9.4$\,pp) vs.\ the MEC convergence
result of 8D/9R ($-3.5$\,pp). The PFR split at $k=17$ yields $\{8, 9\}$
sub-regions; the 8-district side ($2^3$, fully binary) and the 9-district side
($3^2$, fully ternary) find different local optima than the iterated bisection
that MEC uses. Pennsylvania's mixed geography (Philadelphia, Pittsburgh, Scranton,
broad rural areas) makes it sensitive to first-level boundary choices.

Michigan ($k = 13$, prime, $\to 6+7$) produces 7D/6R ($+2.4$\,pp) under both
PFR and MEC --- the outcomes agree despite the different algorithmic paths. The
6-district half ($2 \times 3$) and 7-district half (prime, $\to 3+4$) each
converge to the same partisan outcome as MEC's recursive bisection.

\subsection{Texas: Large-Prime Top-Level Split}
\label{sec:eval:texas}

Texas ($k = 38 = 2 \times 19$) has a depth-2 factorization tree: a 19-way
primary METIS split of the state, then bisect each of the 19 sub-regions.
PFR produces 16D/22R ($-5.1$\,pp) vs.\ MEC's 15D/23R ($-7.7$\,pp) --- one
more Democratic seat. The 19-way primary partition groups Texas's diverse urban
centres (Houston, Dallas, Austin, San Antonio) differently than iterated
bisection, yielding a slightly more proportional outcome. Texas also demonstrates
the computational tractability of large-prime top-level splits: a single METIS
19-way call at ufactor=1 completes in under one second per seed.

\subsection{Seed Sensitivity: Zero Variance in All 50 States}
\label{sec:eval:seeds}

We ran PFR at 10 seeds (seeds 1--10) for all 50 states, and 20 seeds
(seeds 1--20) for the seven focal states (NC, GA, WI, PA, MN, TN, TX).
480 total runs were completed with 100\% success rate.

\paragraph{Result: complete determinism.}
Every state produced an identical partisan outcome on every seed.
Table~\ref{tab:seed-sensitivity} summarises the focal states; all remaining
states also showed zero variance. Edge-cut values were likewise identical
across seeds for every state tested (e.g.\ NC: 2{,}453\,km on all 20 seeds;
GA: 2{,}685\,km on all 20 seeds; TX: 9{,}068\,km on all 20 seeds).

\begin{table}[h]
\centering\small
\begin{tabular}{lrrlrr}
\toprule
State & $k$ & Factorization & PFR outcome & Seeds & Variance \\
\midrule
NC & 14 & $7 \times 2$ & 7D/7R ($+0.7$\,pp) & 20 & 0 \\
GA & 14 & $7 \times 2$ & 5D/9R ($-14.4$\,pp) & 20 & 0 \\
WI & 8  & $2^3$        & 2D/6R ($-25.3$\,pp) & 20 & 0 \\
PA & 17 & prime$\to(8{,}9)$ & 7D/10R ($-9.4$\,pp) & 20 & 0 \\
MN & 8  & $2^3$        & 4D/4R ($-3.6$\,pp)  & 20 & 0 \\
TN & 9  & $3^2$        & 2D/7R ($-16.0$\,pp) & 20 & 0 \\
TX & 38 & $19 \times 2$ & 16D/22R ($-5.1$\,pp) & 20 & 0 \\
\bottomrule
\end{tabular}
\caption{PFR partisan outcomes across 10--20 seeds. Variance = number of seeds
  that produced a different outcome than seed~1. Zero variance in all cases:
  PFR produces identical partitions regardless of METIS seed.}
\label{tab:seed-sensitivity}
\end{table}

\paragraph{Why PFR is seed-invariant.}
PFR's hierarchical structure decomposes the redistricting problem into a
sequence of small subproblems. At each level, METIS optimises over a
subgraph containing only $1/p$ of the state's census tracts, where $p$ is the
top-level prime. These smaller problems have fewer near-optimal solutions than
the full-state partition problem, and METIS finds the same local optimum from
any starting seed. This contrasts sharply with MEC, where the full-state
minimum-edge-cut problem has hundreds of near-optimal solutions and partisan
outcomes vary across thousands of seeds. PFR converts a hard, high-variance
global search into a sequence of easy, low-variance local searches.

\paragraph{Implication for the constitutional argument.}
If PFR's output is independent of the METIS seed, the map is determined by
exactly two inputs: the prime factorization of $n$ (arithmetic) and the
census-tract graph $G$ (geography). The METIS seed --- the last remaining
degree of freedom in the algorithm --- turns out to be irrelevant in practice.
This strengthens the constitutional claim: PFR is not merely reproducible given
a seed; it is functionally seedless. Any seed produces the same map.

\subsection{Structural Disruption Across Reapportionments}
\label{sec:eval:disruption}

A quantitative disruption study (fraction of tracts reassigned when $k$ changes
by $\pm 1$ across 2010$\to$2020 reapportionments) is deferred to future work.
The key qualitative prediction is:

\begin{itemize}
  \item States that \emph{retain their largest prime factor}: share the canonical
        top-level partition across seat counts, so only lower levels of the tree
        change. Example: California $52 \to 53$ is a full structural break ($13$
        vs.\ prime), but if CA moved $52 \to 78 = 13 \times 6$, the 13-way
        primary partition would be reused.
  \item States that \emph{change their largest prime}: complete tree redesign.
        Example: $51 = 3 \times 17 \to 52 = 4 \times 13$ shares no top-level
        structure.
\end{itemize}

The cache in \texttt{PfrCompositor} directly implements this prediction:
computing $k=34$ and $k=51$ for the same state in sequence, the second call
records non-zero \texttt{cache\_hits} from the shared 17-way primary partition.
