%% 05-verification.tex — How to verify any result

\section{Verification Protocol}
\label{sec:verification}

\subsection*{Verifying Any Single Claim}

Every quantitative claim in every portfolio paper is linked to a specific
\texttt{redist} command that produced it.
To verify a claim:

\begin{enumerate}[leftmargin=*, itemsep=4pt]
  \item Find the claim in the paper (e.g., ``+22\% compactness improvement'').
  \item Locate the associated command in the paper's supplementary materials
        or in the paper's directory under \texttt{research/B.2+edge-weighted-bisection/}.
  \item Run the command with the same parameters.
  \item Compare the output to the published hash using \texttt{redist label-verify}.
\end{enumerate}

\subsection*{The \texttt{redist label-verify} Command}

\begin{verbatim}
redist label-verify <version>
\end{verbatim}

\noindent This command:
\begin{itemize}[itemsep=2pt]
  \item Reads the config file for \texttt{<version>} and computes its SHA-256.
  \item Reads all input Census files and computes a combined build hash.
  \item Reads all district assignment output files and computes the analysis hash.
  \item Reads the summary statistics and map files and computes the report hash.
  \item Compares all four hashes to the values published in the portfolio.
\end{itemize}

\noindent If all four hashes match, the run is verified.
If any hash differs, the command reports which step failed and which files differ.

\subsection*{The SHA Chain in Detail}

\paragraph{Config hash.}
The config hash covers the following parameters:
population tolerance ($\pm$0.5\%), METIS random seed (42), edge-weight scaling
factor ($\alpha = 5.0$), and the number of METIS refinement iterations ($T = 600$).
Changing any parameter changes the config hash and propagates through the chain.

\paragraph{Build hash.}
The build hash is the SHA-256 of the concatenated SHA-256 values of all
input PL 94-171 files and TIGER/Line shapefiles, in FIPS order.
This confirms that no input data was altered.

\paragraph{Analysis hash.}
The analysis hash covers the district assignment CSV for every state,
in FIPS order within each state.
Two runs on the same machine with the same inputs produce identical
analysis hashes.
Two runs on different machines of the same architecture (x86-64) also match.
Cross-architecture results (x86-64 vs.\ arm64) may differ by $\leq 0.001$~pp
due to floating-point ordering differences; the portfolio publishes separate
reference hashes for each architecture.
Results that are byte-identical on the same platform
(x86-64 Linux/Windows/macOS with the same METIS vendored build) are
fully verified by \texttt{redist label-verify}.

\paragraph{Report hash.}
The report hash covers the summary statistics CSV (compactness, population
balance, seat counts) and the map SVG files.
This confirms that post-processing and visualization did not modify
the underlying data.

\subsection*{Confirming METIS Determinism}

METIS with a fixed random seed (\texttt{-seed 42}) is deterministic
on a given architecture (x86-64 or arm64), but may produce different
results across architectures due to floating-point rounding.
The portfolio provides separate expected hashes for x86-64 and arm64.
Users on other architectures should contact the author for guidance.
