\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}[proposition]
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}

\title{AreaSection: Simultaneous Population and Land-Area Balance\\
       in Minimum-Edge-Cut Redistricting (Revision 1)}

\author{Gio Della-Libera}

\date{Draft --- \today}

\begin{document}

\maketitle

\begin{abstract}
Standard minimum-edge-cut redistricting (and its ratio-optimal extension
GeoSection, B.8) treats land area as irrelevant to bisection: the algorithm
cares only about minimising boundary length subject to population balance.
This produces a systematic pattern --- \emph{urban peeling} --- in which the
first bisection level always separates the densest metropolitan core from the
rest of the state, because a compact urban boundary is cheapest.
Urban peeling is politically contentious: it concentrates Democratic voters
into a single compact district at every recursion level, producing implicitly
packed maps even with zero partisan input.

We introduce \textbf{AreaSection}, a dual-constraint extension that adds a
second METIS balance constraint requiring each half to receive approximately
50\,\% of the state's land area ($\pm 10$\,\% swing allowed).
The area constraint is implemented via METIS's \texttt{ncon=2} multi-constraint
partitioning with per-constraint imbalance vectors (\texttt{ubvec}).
Ratio selection uses the same isoperimetric normalisation as GeoSection
($\mathrm{EC}/\sqrt{\min(i,\,k-i)}$) to prevent the caterpillar pathology.

A Lorenz-curve pre-filter identifies ratios where simultaneous
population-and-area balance is geometrically infeasible and skips them,
preventing wasted METIS calls.

Applying AreaSection to all 50 states with 50 seeds per ratio,
44 states converge cleanly.
The top-level natural ratio shifts dramatically versus GeoSection:
Illinois moves from 1:17 to 8:9, New York from 1:16 to 13:13, Florida from
1:27 to 14:14.
States with genuinely concentrated populations (Georgia: 3:11, Massachusetts: 2:7)
retain asymmetric splits because the area constraint cannot be satisfied at
more balanced ratios while maintaining contiguous regions.

A head-to-head comparison across 34 states (all states with complete data
for both algorithms) shows that AreaSection and GeoSection produce
\emph{identical seat counts} in 25 of 34 states (76\,\%).
In the 9 states that differ, 5 favour GeoSection by 1 seat and 4 favour
AreaSection by 1 seat (net: $-1$ Democratic seat under AreaSection).
There is no systematic partisan direction.
Within the states that match, AreaSection produces a cleaner competitiveness
profile: GeoSection's urban-peeling strategy creates knife-edge suburban
districts near 49\,\% Democratic vote share as a geometric byproduct;
AreaSection eliminates these.
AreaSection is a geometric-regularity mechanism, not a partisan mechanism:
it changes the bisection tree structure without systematically changing partisan seat allocations.
All 8 competitive states (46--55\,\% Democratic vote) show Republican
over-representation under both algorithms (mean $-6.2$\,pp gap), confirming
that geographic voter sorting — not the bisection algorithm — drives the
proportionality deficit.
\end{abstract}

\input{sections/01-introduction}
\input{sections/02-related-work}
\input{sections/03-algorithm}
\input{sections/04-evaluation}
\input{sections/05-discussion}
\input{sections/06-conclusion}

\bibliographystyle{plainnat}
\bibliography{references,../../references}

\end{document}
