\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}[proposition]
\newtheorem{remark}{Remark}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{rotating}
\usepackage{pdflscape}

%% Legal case emphasis
\newcommand{\case}[1]{\textit{#1}}
\newcommand{\callais}{\case{Louisiana v.\ Callais}}
\newcommand{\wesberry}{\case{Wesberry v.\ Sanders}}
\newcommand{\shaw}{\case{Shaw v.\ Reno}}
\newcommand{\miller}{\case{Miller v.\ Johnson}}
\newcommand{\rucho}{\case{Rucho v.\ Common Cause}}
\newcommand{\lwvpa}{\case{League of Women Voters v.\ Commonwealth of Pennsylvania}}
\newcommand{\allenmilligan}{\case{Allen v.\ Milligan}}

%% Note marker for estimated values
\newcommand{\est}{$^{\dagger}$}
\newcommand{\pending}{$^{\ddagger}$}

\title{\textbf{A Complete Algorithmic Toolbox for Congressional Redistricting:\\
       Legal Requirements, Algorithm Components, and Callais-Compliant\\
       Design for the Post-2026 Redistricting Cycle}\\[0.5em]
       \large B.0 Series Synthesis}

\author{%
  Gio Della-Libera\\
  \textit{Apportionment Research Group}\\
  \small\texttt{giodl@microsoft.com}
}

\date{May 2026}

\begin{document}

\maketitle

\begin{abstract}
Congressional redistricting must satisfy eleven distinct legal requirements
spanning constitutional population equality, geographic compactness, county
preservation, the Voting Rights Act, and the racial-predominance bar set by
\shaw{} (1993).
The 2026 Supreme Court decision in \callais{} adds a critical constraint:
race-conscious and partisan signals must be kept \emph{disentangled} in any
redistricting algorithm---they cannot be combined in a single optimisation run.

We present a unified algorithmic toolbox organised along three axes:
(A) vertex weights governing what METIS balances,
(B) edge management controlling compactness and subdivision preservation, and
(C) division strategy determining how the bisection tree is structured.
Fifteen algorithm components developed across the B-series papers (B.1--B.15)
map to these axes, and each legal requirement maps to at least one component.

The toolbox is structured so that each mode uses exactly one signal type.
\textbf{VRASection} (B.14) addresses Gingles Prong~1 using geographic alignment
alone---no racial target, no racial count.
\textbf{ProportionalSection} (B.12, planned) addresses proportionality using
partisan vote share alone.
A \texttt{callais\_preflight} gate enforces mutual exclusion: the two signals
cannot be combined in one run.
This is not a limitation but the legally required architecture under \callais{}.

We present a bakeoff of eight algorithm combinations across four legally
contested states (Wisconsin 8D, North Carolina 14D, Georgia 14D,
Alabama 7D), measuring edge cut, population balance, D-seat count,
majority-minority district count, natural bisection ratio, and county split count.
The results confirm that geographic voter sorting---not the bisection algorithm---
drives partisan outcomes in all four states, and that VRASection produces the
minority-opportunity district patterns required by \allenmilligan{} (2023)
without any racial targeting.

A court can audit exactly which signals were used in any plan by reading the
\texttt{whatif-manifest v1} JSON sidecar produced by the \texttt{redist} CLI.
Every run is parameter-reproducible and hash-chain verifiable.
\end{abstract}

\tableofcontents
\newpage

\input{sections/01-introduction}
\input{sections/02-legal-requirements}
\input{sections/03-algorithm-components}
\input{sections/04-requirements-matrix}
\input{sections/05-bakeoff}
\input{sections/06-callais-compliance}
\input{sections/07-discussion}
\input{sections/08-conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
