\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}[proposition]
\newtheorem{remark}{Remark}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{array}
\usepackage{microtype}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{footmisc}

%% Law-review style: section headings in small caps
\titleformat{\section}{\normalfont\large\scshape\centering}{}{0em}{}
\titleformat{\subsection}{\normalfont\normalsize\bfseries}{}{0em}{}
\titleformat{\subsubsection}{\normalfont\normalsize\itshape}{}{0em}{}

%% Spacing
\setstretch{1.5}

%% Legal case macros
\newcommand{\case}[1]{\textit{#1}}
\newcommand{\callais}{\case{Louisiana v.\ Callais}}
\newcommand{\wesberry}{\case{Wesberry v.\ Sanders}}
\newcommand{\karcher}{\case{Karcher v.\ Daggett}}
\newcommand{\shaw}{\case{Shaw v.\ Reno}}
\newcommand{\miller}{\case{Miller v.\ Johnson}}
\newcommand{\rucho}{\case{Rucho v.\ Common Cause}}
\newcommand{\lwvpa}{\case{League of Women Voters v.\ Commonwealth of Pennsylvania}}
\newcommand{\allenmilligan}{\case{Allen v.\ Milligan}}
\newcommand{\gingles}{\case{Thornburg v.\ Gingles}}
\newcommand{\harper}{\case{Harper v.\ Hall}}
\newcommand{\harkenrider}{\case{Harkenrider v.\ Hochul}}
\newcommand{\bushvvera}{\case{Bush v.\ Vera}}

%% Statute macro
\newcommand{\dia}{\textsc{Districting Integrity Act}}
\newcommand{\hh}{\textsc{Huntington-Hill}}
\newcommand{\ar}{\textsc{ApportionRegions}}
\newcommand{\usca}{\textsection}

%% Proposed statute box
\usepackage{mdframed}
\newenvironment{statute}{%
  \begin{mdframed}[linewidth=0.8pt, innerleftmargin=12pt, innerrightmargin=12pt,
                   innertopmargin=8pt, innerbottommargin=8pt,
                   backgroundcolor=gray!8]
  \small\setstretch{1.3}%
}{%
  \end{mdframed}%
}

%% Note markers
\newcommand{\est}{$^{\dagger}$}
\newcommand{\pending}{$^{\ddagger}$}

%% -----------------------------------------------------------------------

\title{%
  \textbf{The Districting Integrity Act}\\[0.4em]
  \large Huntington-Hill as the Constitutional Standard for\\
  Intrastate Congressional Apportionment\\[0.8em]
  \normalsize\textit{B.02 --- Advocacy Paper}%
}

\author{%
  Gio Della-Libera\\
  \textit{Apportionment Research Group}\\
  \small\texttt{giodl@microsoft.com}
}

\date{May 2026}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Every ten years, Congress apportions 435 seats among the fifty states using
the Huntington-Hill priority rule---an algorithm whose constitutional
pedigree is uncontested.
Yet when each state then divides its seats into districts, no federal standard
governs the method.
The result is partisan gerrymandering, litigation that consumes a decade, and
maps that fail to represent anyone's constituents well.

This Article proposes a fix grounded entirely in first principles that both
parties have already endorsed.
The argument is four steps.
\emph{First} (\wesberry{}, 1964): congressional districts must have equal
population.
\emph{Second} (Article~I, \S2): the interstate allocation method is
Huntington-Hill.
\emph{Third} (both parties, on record): the map should not be drawn by
politicians with a personal stake in the outcome.
\emph{Fourth} (logic): if Huntington-Hill is the right method for the
interstate problem---allocating seats to states---and the within-state problem
is structurally identical---allocating districts to geographic regions---then
Huntington-Hill should apply within states too.

The legislative vehicle is simple.
We call it the \dia{}.
Its operative text is one sentence, proposed as 2~U.S.C.~\usca{}~2a(c) [new]:

\begin{quote}
  \itshape
  Each State shall apportion its congressional districts by applying the
  Huntington-Hill priority rule to its census tracts, using as input only the
  prime factorization of the State's apportioned seat count and the federal
  TIGER/Line adjacency graph for the census year in which the apportionment
  was performed, with a seed determined by the formula
  \texttt{SHA-256(census\_release\_id \textbar\textbar{} `DIA\_SEED\_V1')}.
\end{quote}

This is not a new constitutional theory.
It is the existing constitutional standard applied at the next level down.
It satisfies Republican concerns about compact, non-racially-targeted maps.
It satisfies Democratic concerns about equal population and VRA compliance.
It satisfies the courts, which need not develop any new doctrine.
And it satisfies \callais{} (2026), whose disentanglement requirement fits
the algorithm's architecture exactly.

The empirical case is equally strong.
Algorithm selection \emph{itself} is a political choice: in Wisconsin,
switching from ApportionRegions to GeoSection shifts the partisan outcome by
more than twelve percentage points---purely by changing which algorithm draws
the map, with no political inputs in either algorithm.
The only way to remove politics from the process is to mandate the algorithm
that is prescribed by first principles, not selected by practitioners.
\hh{} is that algorithm.
\ar{} is its intrastate extension.
The \dia{} is the law.
\end{abstract}

\tableofcontents
\newpage

\input{sections/01-introduction}
\input{sections/02-first-principles}
\input{sections/03-why-apportionregions}
\input{sections/04-bipartisan-case}
\input{sections/05-implementation}
\input{sections/06-conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
