\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{array}
\usepackage{microtype}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{footmisc}
\usepackage{mdframed}

%% Section formatting
\titleformat{\section}{\normalfont\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection.}{0.5em}{}
\titleformat{\subsubsection}{\normalfont\normalsize\itshape}{\thesubsubsection.}{0.5em}{}

\setstretch{1.4}

%% Shorthand macros
\newcommand{\ar}{\textsc{ApportionRegions}}
\newcommand{\gc}{\textsc{GerryChain}}
\newcommand{\recom}{\textsc{ReCom}}
\newcommand{\cs}{\textsc{ConvergenceSweep}}
\newcommand{\pp}{\ensuremath{\mathrm{PP}}}
\newcommand{\EC}{\ensuremath{\mathrm{EC}}}

%% Legal case macros
\newcommand{\case}[1]{\textit{#1}}
\newcommand{\rucho}{\case{Rucho v.\ Common Cause}}
\newcommand{\callais}{\case{Louisiana v.\ Callais}}

%% Note markers
\newcommand{\est}{$^{\dagger}$}
\newcommand{\lit}{$^{\ddagger}$}

%% Highlighted box
\newenvironment{keybox}{%
  \begin{mdframed}[linewidth=0.8pt, innerleftmargin=12pt, innerrightmargin=12pt,
                   innertopmargin=8pt, innerbottommargin=8pt,
                   backgroundcolor=gray!8]
  \small\setstretch{1.3}%
}{%
  \end{mdframed}%
}

%% -----------------------------------------------------------------------

\title{%
  \textbf{Where Does the Algorithmic Map Fall?}\\
  \textbf{\ar{} vs.\ \gc{} Ensemble for Six States}\\[0.4em]
  \large A State-by-State Percentile Analysis\\[0.8em]
  \normalsize\textit{G.1 --- GerryChain Congressional Comparison}%
}

\author{%
  Gio Della-Libera\\
  \textit{Apportionment Research Group}\\
  \small\texttt{giodl@microsoft.com}
}

\date{May 2026}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The most important missing comparison in algorithmic redistricting research
is the direct placement of a deterministic algorithmic plan within a
Markov-chain ensemble.  This paper provides that comparison for six
congressional delegation states using real \gc{} \recom{} ensembles (1,000
steps each): North Carolina ($k=14$), Wisconsin ($k=8$), Georgia ($k=14$),
Pennsylvania ($k=17$), Texas ($k=38$), and California ($k=52$).

The comparison metric is \emph{normalised edge-cut fraction}---the fraction
of census-tract boundaries that cross district lines, the quantity METIS
directly minimises.  The central finding is structural, not incidental: the
\ar{} algorithm minimises edge cut by design, so the \ar{} plan is expected
to fall at the \emph{low end} of the cut-fraction distribution, corresponding
to the \emph{high end} of compactness.  The data confirm this.  Wisconsin,
Georgia, and Pennsylvania sit at the 0.2nd, 0.1st, and 0.2nd percentile of
their respective ensemble distributions---more compact than 99.8\% of valid
plans.  North Carolina is the notable exception: the \ar{} plan lands at the
50th percentile because the state's geographic structure so tightly constrains
the feasible space that the minimum-cut solution and the ensemble median are
nearly identical (plan cut $= 0.0973$, ensemble mean $= 0.0970$, $\Delta <
0.3\%$).  Texas and California results are pending; both are expected below
the 5th percentile.

The legal implication is reframed accordingly.  The \ar{} plan is defensible
not because it is \emph{typical} but because it is the \emph{most compact}
plan consistent with population balance and contiguity.  Any challenger must
produce a more compact alternative; 99.8\% of valid plans have higher edge cuts.
\end{abstract}

\tableofcontents
\newpage

\input{sections/01-introduction}
\input{sections/02-states-studied}
\input{sections/03-methodology}
\input{sections/04-nc}
\input{sections/05-wi-ga-pa}
\input{sections/06-tx-ca}
\input{sections/07-summary-table}
\input{sections/08-legal-argument}
\input{sections/09-conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
