\section{Implementation}
\label{sec:implementation}

\subsection{Crate Architecture}

The \re{} crate is structured as three source files within the
\texttt{redist-ensemble} module of the \texttt{redist} Rust workspace:

\begin{description}
  \item[\texttt{spanning.rs}] Wilson's loop-erased random walk.
    Exports \texttt{wilson\_ust(graph, rng)} returning an adjacency-list
    spanning tree. Uses \texttt{SmallRng} for per-step randomness (see
    Section~\ref{ssec:rng}).

  \item[\texttt{recom.rs}] The \recom{} proposal.
    Exports \texttt{recom\_step(partition, graph, pop, tol, rng)},
    which selects a pair, calls \texttt{wilson\_ust}, enumerates balanced
    cuts, and returns the updated partition.
    Handles tree-resample-on-failure and pair-reselection logic.

  \item[\texttt{chain.rs}] Parallel runner.
    Exports \texttt{run\_ensemble(config)} which drives $C$ independent
    chains via Rayon's \texttt{par\_iter}, collecting per-step statistics
    and computing convergence diagnostics.
\end{description}

\subsection{Graph Representation}

The census-tract adjacency graph is stored as a \texttt{Vec<Vec<usize>>}
adjacency list, loaded from the \texttt{.adj.bin} binary format shared
with \texttt{redist-data}.
Population data is a \texttt{Vec<u64>} aligned to the same vertex
indices.
Subgraphs for individual district pairs are constructed on demand as
\texttt{Vec<Vec<usize>>} slices by remapping vertex indices to the
compact range $\{0,\ldots,m-1\}$.
This subgraph construction cost is $O(m)$ per step and does not allocate
on the heap for graphs with $m \leq 512$ due to a stack-based small-buffer
optimization.

\subsection{Random Number Generation}
\label{ssec:rng}

Each chain uses a \texttt{SmallRng} instance from the \texttt{rand} crate,
seeded from the SHA-256-derived per-chain seed.
\texttt{SmallRng} is a 128-bit xoshiro128++ generator: its state is small
(16 bytes), its throughput is high ($\sim$4 billion integers per second
on modern hardware), and it passes the BigCrush statistical test suite.
For the random walk within Wilson's algorithm, the dominant operation is
selecting a uniformly random neighbor of the current vertex.
We avoid modulo bias by using Lemire's fast divisor trick~\citep{lemire2019fast}
as implemented in \texttt{rand::Rng::gen\_range}.

\subsection{Balance Check}

The balance check for a candidate spanning tree $T$ of the merged region
is implemented as a single depth-first traversal.
We root $T$ at vertex 0 of the local indexing and compute subtree
population sums in $O(m)$ time using a post-order traversal.
For each edge $(u, \mathit{parent}(u))$, the subtree rooted at $u$ has
population $\mathit{pop}[u]$, and the complementary subtree has
population $P_R - \mathit{pop}[u]$, where $P_R$ is the total population
of the merged region.
The edge is \emph{balanced} if both subtrees satisfy
$|\mathit{pop}/\bar{p} - 1| \leq \varepsilon$.
Balanced edges are collected into a \texttt{SmallVec<[usize; 8]>} to
avoid heap allocation for the common case where the number of balanced
cuts is small.

\subsection{Serde Output}

Step-level statistics---district population variances, normalized cut
fraction, Democratic seat share (if election data is loaded)---are
serialized to newline-delimited JSON using \texttt{serde\_json}.
Chain-level diagnostics ($\rhat$, $\ess$, $\ham(1)$) are written at
chain completion.
The output format is designed to be streamed directly into the
\texttt{redist-report} pipeline for court-submission documents.

\subsection{Texas and Bipartition Failures}

Texas ($n = 4{,}866$ tracts, $k = 38$ districts) presents a known
challenge: the state's shape and high $k$ mean that a randomly merged
pair of adjacent districts often contains a region with no balanced
bipartition of any random spanning tree.
\gc{} is documented to fail on Texas when run without a warmup from a
known valid starting plan~\citep{gerrychain2021}.

\re{} handles this via the pair-reselection mechanism in
Algorithm~\ref{alg:recom}: after 10 consecutive spanning tree resample
failures for a given pair, a new adjacent pair is selected.
In practice, for Texas, pair reselection triggers frequently during the
first 50--100 steps from a cold start, then subsides as the chain moves
away from the starting plan into regions of higher feasibility density.
This behavior is not a Rust-specific limitation; it reflects the intrinsic
geometry of the Texas plan space.
Rust's advantage is that the 10-resample budget is exhausted in
microseconds rather than the seconds it would take in Python, so pair
reselection adds negligible overhead.

\paragraph{Stationarity under pair reselection.}
Pair reselection preserves the \recom{} stationary distribution under
mild conditions: when the pair-selection probability is symmetric and
the pair is rejected only when no balanced cut exists (not based on the
cut quality).
In our implementation, pair reselection triggers only on balance failure
--- i.e., when all 10 sampled spanning trees for a given pair admit no
balanced bipartition --- matching \gc{}'s documented behavior.
Under this protocol, the reselection step is equivalent to a proposal
rejection that discards infeasible pair proposals uniformly, which
preserves detailed balance if the pair-selection distribution is
symmetric over adjacent district pairs.
A formal ergodicity analysis is deferred to future work; the planned
empirical G-track replication (Section~\ref{sec:conclusion}) will compare
ensemble distributions produced by \re{} and \gc{} to confirm that any
distributional difference is below statistical resolution.
