¥section{Performance Analysis}
¥label{sec:performance}

¥subsection{GerryChain Baseline Measurements}

We measured ¥gc{} throughput by running 1{,}000 ¥recom{} steps for
three states and recording wall-clock time.
These are direct empirical measurements from the ¥gc{} Python package
(version 0.3.2, NumPy 1.24, Python 3.10) run on an Intel Core i7-11800H
(2.3‾GHz base, 4.6‾GHz boost, 24‾MB L3 cache) under Windows‾11.
The starting plan for each state is a cold start from a seed-0 random
valid plan.
No warm-up steps were excluded; all 1{,}000 steps count toward the
wall-clock total.
¥gc{} was run without pair reselection (the standard configuration):

¥begin{center}
¥begin{tabular}{lrrrl}
¥toprule
State & $k$ & $n_{¥text{tracts}}$ & 1K steps (sec) & Steps/sec ¥¥
¥midrule
North Carolina & 14 & 2{,}672 & 47 & $¥approx 21$ ¥¥
Wisconsin      &  8 & 1{,}922 & 28 & $¥approx 36$ ¥¥
Georgia        & 14 & 1{,}978 & 30 & $¥approx 33$ ¥¥
¥bottomrule
¥end{tabular}
¥end{center}

The variation in throughput across states reflects differences in both
$n$ (tract count) and $k$ (district count).
Wisconsin is faster than North Carolina despite having fewer districts
($k=8$ vs $k=14$) because its lower tract count reduces the merged-region
size; Georgia achieves similar throughput to Wisconsin despite $k=14$
because its tracts are on average smaller in the adjacency graph.

¥subsection{Rust Target Estimates}

The theoretical throughput of ¥re{} follows from Wilson's complexity
analysis.
For a merged region of size $m = 2n/k$:

¥begin{enumerate}
  ¥item ¥textbf{Wilson's walk}: $O(m¥log m)$ expected operations.
    At $m ¥approx 191$ (NC average: $2 ¥times 2{,}672 / 14$),
    Wilson requires approximately $191 ¥times ¥log(191) ¥approx 191 ¥times 7.5 ¥approx 1{,}433$
    random walk steps.
  ¥item ¥textbf{Effective operations per walk step}: Neighbor selection
    (1 uniform draw), loop detection (hash set lookup), and
    path extension (pointer write)---approximately 8 instructions at the
    clock level.
  ¥item ¥textbf{Total operations per ¥recom{} step}: $¥approx 191 ¥times 8 = 1{,}528$
    effective operations.
  ¥item ¥textbf{Effective clock rate}: Modern CPUs execute at
    $¥sim$4¥,GHz but MCMC-adjacent code (random number generation,
    irregular memory access) achieves roughly 1¥,GHz effective throughput
    on graph-walk workloads of this size.
  ¥item ¥textbf{Resulting time per step}: $1{,}528 ¥,/¥, 10^9 ¥approx 1.5¥,¥mu$s.
  ¥item ¥textbf{Steps per second}: $¥approx 660{,}000$.
¥end{enumerate}

Applying a conservative $13¥times$ overhead factor for Rust runtime costs
(subgraph construction, adjacency list traversal, SmallVec allocation,
serde bookkeeping) yields a practical target of ¥textbf{50{,}000 steps/sec}.
This is still a $2{,}350¥times$ speedup over ¥gc{} for North Carolina.

¥begin{remark}
The overhead factor of $13¥times$ is deliberately conservative.
Our METIS pipeline achieved a $213¥times$ speedup over Python for a
structurally similar graph-partitioning workload‾¥citep{dellaLibera2026b1}.
The overhead factor for ¥re{} is higher because Wilson's random walk has
more irregular memory access patterns than METIS's multilevel contraction,
but $50{,}000$¥,steps/sec should be achievable without architecture-specific
optimization.
¥end{remark}

¥paragraph{Overhead sensitivity.}
The $13¥times$ overhead factor is an estimate; Table‾¥ref{tab:overhead}
shows how the projected throughput and speedup change under three scenarios.

¥begin{table}[h]
¥centering
¥caption{Throughput sensitivity to overhead factor (NC baseline: 21 steps/sec).}
¥label{tab:overhead}
¥begin{tabular}{lrrr}
¥toprule
Scenario & Overhead factor & Steps/sec & Speedup vs.¥ ¥gc{} (NC) ¥¥
¥midrule
Optimistic  & $6.5¥times$  & $¥sim$100{,}000 & $¥sim$4{,}700$¥times$ ¥¥
Central     & $13¥times$   & $¥sim$50{,}000  & $¥sim$2{,}350$¥times$ ¥¥
Conservative & $26¥times$  & $¥sim$25{,}000  & $¥sim$1{,}200$¥times$ ¥¥
¥bottomrule
¥end{tabular}
¥par¥vspace{0.5ex}
¥small All figures estimated; Phase‾2 ¥texttt{criterion.rs} benchmarks will determine the
actual overhead factor and narrow the range.
¥end{table}

¥subsection{Projected Performance Table}

¥begin{table}[h]
¥centering
¥caption{%
  Performance comparison: ¥gc{} measured vs.¥ ¥re{} estimated.
  Rust estimates based on Wilson's $O((n/k)¥log(n/k))$ analysis at
  50{,}000¥,steps/sec (conservative).
  PA and TX estimated from $n$-scaling of the NC/WI/GA measurements.
  All Rust figures require validation against ¥texttt{criterion.rs}
  (Phase‾2).
}
¥label{tab:performance}
¥begin{tabular}{lrrrrrr}
¥toprule
¥multirow{2}{*}{State} & ¥multirow{2}{*}{$k$} &
  ¥multirow{2}{*}{$n$} &
  ¥multicolumn{2}{c}{¥gc{} (measured)} &
  ¥multicolumn{2}{c}{¥re{} (estimated¥est)} ¥¥
¥cmidrule(lr){4-5}¥cmidrule(lr){6-7}
& & & 1K steps (s) & Steps/s & 10K steps (s) & Speedup ¥¥
¥midrule
NC & 14 & 2{,}672 & 47 & 21 & 0.20 & 2{,}350$¥times$ ¥¥
WI &  8 & 1{,}922 & 28 & 36 & 0.12 & 2{,}330$¥times$ ¥¥
GA & 14 & 1{,}978 & 30 & 33 & 0.13 & 2{,}310$¥times$ ¥¥
PA & 17 & 3{,}218 & $¥sim$30¥est & $¥sim$33 & 0.18 & 2{,}200$¥times$ ¥¥
TX & 38 & 4{,}866 & failure¥bench & --- & 0.31 & ---¥bench ¥¥
CA & 52 & 8{,}057 & failure¥bench & --- & 0.52 & ---¥bench ¥¥
¥bottomrule
¥end{tabular}
¥par¥vspace{0.5ex}
¥small
¥est¥,Rust estimates require validation with ¥texttt{criterion.rs}; see Section‾¥ref{ssec:phase2}.¥¥
¥bench¥,¥gc{} ¥emph{without} pair reselection fails on TX ($k=38$) and CA ($k=52$) from cold start
due to bipartition infeasibility; ¥gc{} ¥emph{with} pair reselection and ¥re{} both handle these cases
via pair reselection (Section‾¥ref{ssec:tx}).
¥gc{}'s bipartition failures on TX $k=38$ are combinatorial, not language-specific.
Both GerryChain-with-pair-reselection and ¥re{} succeed.
The advantage of ¥re{} is throughput, not correctness.
¥end{table}

¥subsection{Texas and California}
¥label{ssec:tx}

The bipartition infeasibility problem in Texas is a combinatorial property
of the TX plan space, not a language-specific limitation.
¥gc{} without pair reselection is documented to fail on Texas from cold
start‾¥citep{gerrychain2021}; ¥gc{} with pair reselection would also
handle these states, albeit more slowly.
¥re{}'s advantage is that pair reselection at microsecond speed (rather
than second speed in Python) reduces warm-up wall time without altering
the algorithm's combinatorial behavior.
The performance estimate in Table‾¥ref{tab:performance} assumes the pair
reselection overhead is absorbed within the 50{,}000¥,steps/sec budget.

Per-step cost is $O((n/k)¥log(n/k))$, so the relevant quantity is the
average merged-region size $m ¥approx 2n/k$.
For CA ($k=52$, $n=8{,}057$): $m ¥approx 310$.
For PA ($k=17$, $n=3{,}218$): $m ¥approx 379$.
PA has the largest average merged region among the six G-track states
and therefore the highest per-step cost; California's per-step cost
is lower than PA's despite its larger tract count, because its high $k$
reduces the merged-region size.
¥emph{The actual per-step cost ordering across all six states is deferred
to Phase‾2 empirical benchmarks.}
The estimated 0.52 seconds for 10{,}000 CA steps and 0.18 seconds for
10{,}000 PA steps (Table‾¥ref{tab:performance}) reflect the theoretical
ordering; Phase‾2 criterion.rs results will confirm or correct the
estimates.

¥subsection{Phase 2: Criterion.rs Benchmarks}
¥label{ssec:phase2}

¥emph{All Rust throughput figures in this paper are estimates derived from
complexity analysis and are deferred to Phase‾2 empirical validation.}
Phase‾2 will validate these estimates using ¥texttt{criterion.rs}, Rust's
standard micro-benchmarking harness, with the following protocol:

¥paragraph{Hardware.}
Benchmarks will be run on the same Intel Core i7-11800H workstation used
for the ¥gc{} baseline measurements in Section‾¥ref{sec:performance}
(2.3‾GHz base, 4.6‾GHz boost, 24‾MB L3 cache, Windows‾11, single-threaded
to isolate per-step cost).
The CPU model, clock speed, and OS will be reported with results.

¥paragraph{GerryChain comparison.}
The Phase‾2 comparison will use ¥gc{} version‾0.3.2, NumPy‾1.24,
Python‾3.10, starting from a cold-start seed-0 random valid plan for
each state (matching the Section‾¥ref{sec:performance} baseline).
Wall-clock time will be measured for 1{,}000 steps after a 50-step
warm-up (the warm-up steps are excluded from the throughput count).

¥paragraph{Acceptance criterion.}
The Phase‾2 report will confirm the 50{,}000¥,steps/sec estimate if the
¥texttt{criterion.rs} measurement for NC exceeds 30{,}000¥,steps/sec
(i.e., within 40¥% of target).
If the measurement falls below 30{,}000¥,steps/sec, the abstract and
Table‾¥ref{tab:performance} speedup figures will be revised downward.
Results will be reported as 95¥% confidence intervals from
¥texttt{criterion.rs}, not point estimates.

¥paragraph{Granularities.}
Benchmarks will be run at the following levels:
¥begin{itemize}
  ¥item ¥textbf{Wilson UST}: Time to generate one uniform random spanning
    tree for a merged region of size $m$, as a function of $m$.
  ¥item ¥textbf{Balance check}: Time for the full edge enumeration pass.
  ¥item ¥textbf{Full step}: End-to-end ¥recom{} step time including
    subgraph construction.
  ¥item ¥textbf{Chain throughput}: Steps/sec for a 1{,}000-step chain,
    compared directly against ¥gc{} wall-clock time on the same hardware.
¥end{itemize}

Results will be reported in a companion note to this paper.
