% Section 3: Methodology (800 words)

\subsection{Data and Graph Construction}

We construct a national redistricting graph from 2020 Census data. The analysis uses 84,414 census tracts covering all 50 states, obtained from the U.S. Census Bureau TIGER/Line shapefiles (2020 vintage). Each tract serves as a graph node weighted by its total population from the 2020 Decennial Census. We exclude Puerto Rico and other territories, focusing on the 435 House districts allocated to states.

Adjacency relationships are determined using rook contiguity (shared boundaries) from tract geometries. The national graph contains 266,466 edges representing adjacent tract pairs. Crucially, we identify 4,544 \textit{cross-state edges}---adjacencies between tracts in different states---which are typically prohibited in conventional redistricting but treated as valid connections in our national optimization. These cross-state edges occur at all state boundaries and represent potential locations where districts could span state lines.

The resulting graph has weighted nodes (tract populations) and unweighted edges (unit adjacency). Total U.S. population in 2020 is 331,108,434, implying a target district population of 761,169 residents ($ 331,108,434 \div 435 $).

\subsection{National Recursive Bisection Algorithm}

We apply the METIS graph partitioning algorithm \citep{karypis1998fast} via recursive bisection to partition the national graph into 435 districts. METIS is a multilevel $k$-way partitioning method that minimizes edge cuts---the number of edges connecting vertices in different partitions---while balancing vertex weights across partitions.

The algorithm operates in three phases: \textbf{(1) Coarsening:} successively merge vertices to create smaller graphs while preserving structure; \textbf{(2) Initial Partitioning:} partition the coarsest graph using spectral or greedy methods; \textbf{(3) Uncoarsening:} project the partition back to the original graph, applying Kernighan-Lin refinement at each level to improve edge cut.

We configure the algorithm with:
\begin{itemize}
    \item \textbf{Population balance:} $ \pm 0.5\% $ tolerance (ufactor = 5), matching the constraint used in state-based redistricting
    \item \textbf{Objective:} Minimize edge cut (equivalent to maximizing within-district contiguity)
    \item \textbf{Iterations:} 100 refinement iterations per bisection level
    \item \textbf{Seed:} Fixed random seed (42) for reproducibility
\end{itemize}

The algorithm does not distinguish cross-state edges from within-state edges; both have unit weight in the baseline configuration. This design choice allows METIS to freely cross state boundaries when doing so improves the edge cut objective.

\subsection{Boundary Crossing Penalty Ablation}

To explore the trade-off between compactness and boundary crossing, we conduct an ablation study varying the penalty $ \beta $ for crossing state lines. For each cross-state edge $ (i,j) $, we multiply its weight by $ (1 + \beta) $:

\begin{equation}
w_{ij} = \begin{cases}
1 & \text{if tracts } i, j \text{ in same state} \\
1 + \beta & \text{if tracts } i, j \text{ in different states}
\end{cases}
\end{equation}

We test $ \beta \in \{0.0, 0.5, 1.0, 2.0, 5.0, 10.0\} $. Higher $ \beta $ values increase the cost of cutting cross-state edges, incentivizing the algorithm to avoid boundary crossings. At $ \beta = 0 $, no penalty is applied (baseline national optimization). At $ \beta = \infty $, cross-state edges are effectively prohibited, recovering state-based redistricting.

\subsection{Compactness Measurement}

We measure district compactness using the Polsby-Popper (PP) score \citep{polsby1991third}, computed as:

\begin{equation}
\text{PP} = \frac{4\pi \times \text{Area}}{\text{Perimeter}^2}
\end{equation}

PP scores range from 0 (least compact) to 1 (perfect circle). We compute PP scores for all 435 districts in both national and state-based configurations, projecting geometries to the Albers Equal Area Conic projection (ESRI:102003) for accurate area and perimeter measurements.

District geometries are constructed by dissolving census tract geometries within each assigned district. We use the GeoPandas library \citep{geopandas} for spatial operations and Matplotlib for visualization.

\subsection{Comparison Baseline}

We compare national optimization results to state-based redistricting, where each state independently partitions its tracts into its allocated districts. The state-based configuration uses identical algorithm parameters (METIS recursive bisection, $ \pm 0.5\% $ tolerance, 100 iterations), differing only in treating each state as a separate graph. This baseline represents the current constitutional structure and provides a reference for quantifying the geometric cost of federalism.

State-based baseline compactness statistics (mean PP = 0.461) are taken from prior analyses using the same methodology, ensuring consistent measurement across configurations.

\subsection{Cross-State District Analysis}

We identify cross-state districts by checking whether any census tract pair $ (i,j) $ in the same district belongs to different states. For each cross-state district, we compute:
\begin{itemize}
    \item \textbf{Primary state:} the state contributing the most population
    \item \textbf{Primary state fraction:} the proportion of district population from the primary state
    \item \textbf{State count:} the number of states contributing tracts
\end{itemize}

This analysis quantifies the extent of fractional representation implied by national optimization and identifies which state pairs most frequently share districts.
