\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}[proposition]
\newtheorem{remark}{Remark}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{xcolor}
\usepackage{textcomp}

\title{B.12: The Proportionality Compromise:\\
Dual-Constrained Bisection for Partisan-Fair Redistricting}

\author{Gio Della-Libera}

\date{Draft --- \today}

\begin{document}

\maketitle

\begin{abstract}
Minimum-edge-cut redistricting algorithms are politically neutral by design:
they minimise boundary length subject to population balance, with no reference
to partisan data.
Yet neutral algorithms produce systematically disproportionate outcomes in
states with geographically concentrated partisan populations --- the Rodden
effect --- because Democratic voters are packed in dense urban cores while
Republican voters are dispersed.
This paper asks: what is the minimum partisan information required to
\emph{guarantee} proportional outcomes, and how far do geography-only
algorithms fall short?

We introduce a dual-constrained bisection framework that extends minimum-edge-cut
redistricting to $\mathtt{ncon}=2$ with vertex weights
$[\text{population},\, D_{\text{votes}}]$.
The core contribution is a closed-form formula for the $\mathtt{tpwgts}$
parameters that guarantee partisan-proportional district allocation:
\[
  \mathtt{tpwgts} = \left[\frac{k_D}{k},\; 1 - \frac{k_R}{2dk},\;
                          \frac{k_R}{k},\; \frac{k_R}{2dk}\right]
\]
where $d$ is the statewide Democratic vote share, $k$ is the total district
count, and $k_D$, $k_R$ are the proportional seat allocations.
This formula gives the R-bloc exactly the minimum Democratic voter share
needed for Republicans to win each district, while concentrating remaining
Democratic votes in the D-bloc.

We introduce the \emph{partisan signal strength} $\sigma$, the distance between
the proportional target and the neutral equal-distribution target.
The central empirical finding is a \textbf{proportionality paradox}: all 8
competitive states (46--55\,\% Democratic) have $\sigma < 0.05$ --- the
proportional target is essentially identical to the neutral target.
Georgia ($\sigma = 0.001$) and Wisconsin ($\sigma = 0.003$) require shifting
less than 0.3\,\% of Democratic votes from the neutral distribution to achieve
proportionality.
Yet both states show $-14$\,pp proportionality gaps under geography-only algorithms.
The gap is not caused by incorrect targeting: it is caused by the Rodden effect
making contiguous partitions geometrically incapable of reaching even the
near-neutral proportional target.

A \emph{partisan Lorenz curve} --- sorting tracts by Democratic density ---
determines when proportional packing is geographically feasible under the
contiguity constraint.
We apply this framework to all 50 states, measuring the partisan signal strength
and its relationship to the proportionality gap.

This paper does not advocate for partisan redistricting.
The federal \S104(e) prohibition on partisan inputs for congressional
districts reflects a substantive policy choice.
What this analysis shows is that the gap between geographic neutrality and
partisan proportionality is a fact about geography --- not a flaw in the
algorithm --- and that closing the gap has a precise, measurable cost in
partisan information.
\end{abstract}

\input{sections/01-introduction}
\input{sections/02-related-work}
\input{sections/03-framework}
\input{sections/04-theory}
\input{sections/05-evaluation}
\input{sections/06-compromise}
\input{sections/07-legal}
\input{sections/08-conclusion}
\input{sections/appendix-a}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
