\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{remark}{Remark}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{array}
\usepackage{microtype}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{footmisc}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{mdframed}

%% SIAM-adjacent style
\titleformat{\section}{\normalfont\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection.}{0.5em}{}
\titleformat{\subsubsection}{\normalfont\normalsize\itshape}{\thesubsubsection.}{0.5em}{}

\setstretch{1.4}

%% Algorithm macros
\newcommand{\ps}{\textsc{PercentileSweep}}
\newcommand{\cs}{\textsc{ConvergenceSweep}}
\newcommand{\ms}{\textsc{MedianSweep}}
\newcommand{\ts}{\textsc{TargetedSweep}}
\newcommand{\ar}{\textsc{ApportionRegions}}
\newcommand{\recom}{\textsc{ReCom}}
\newcommand{\gc}{\textsc{GerryChain}}

%% Legal case macros
\newcommand{\case}[1]{\textit{#1}}
\newcommand{\dia}{\textsc{Districting Integrity Act}}
\newcommand{\callais}{\case{Louisiana v.\ Callais}}
\newcommand{\wesberry}{\case{Wesberry v.\ Sanders}}
\newcommand{\rucho}{\case{Rucho v.\ Common Cause}}
\newcommand{\karcher}{\case{Karcher v.\ Daggett}}

%% Math macros
\newcommand{\ECnorm}{\mathrm{EC}_{\mathrm{norm}}}
\newcommand{\seedfmt}[1]{\texttt{SHA-256(#1)}}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}

%% Statute box
\newenvironment{statute}{%
  \begin{mdframed}[linewidth=0.8pt, innerleftmargin=12pt, innerrightmargin=12pt,
                   innertopmargin=8pt, innerbottommargin=8pt,
                   backgroundcolor=gray!8]
  \small\setstretch{1.3}%
}{%
  \end{mdframed}%
}

%% Note markers
\newcommand{\est}{$^{\dagger}$}
\newcommand{\pending}{$^{\ddagger}$}

%% -----------------------------------------------------------------------

\title{%
  \textbf{PercentileSweep: Statutory Choice of Legal Posture}\\
  \textbf{in Algorithmic Redistricting}\\[0.4em]
  \large Making the Compactness Percentile Explicit, Auditable, and Defensible\\[0.8em]
  \normalsize\textit{H.0 --- Algorithm Theory / Legal Posture}%
}

\author{%
  Gio Della-Libera\\
  \textit{Apportionment Research Group}\\
  \small\texttt{giodl@microsoft.com}
}

\date{May 2026}

\begin{document}

\maketitle

\begin{abstract}
\noindent
\cs{} (B.16) finds the minimum-edge-cut plan across a forward sweep of
$T = 600$ seeds---always returning the compactness extremum of the bisection
family.
Real \gc{} data (G.1) places this extremum at the 0.1--0.7th percentile of
the \recom{} ensemble for most states, confirming that the algorithmic plan
is more compact than 99\%+ of valid alternatives.
North Carolina is the exception, landing at the 50th percentile due to
geographic convergence.

This paper introduces \ps{}, a generalisation that makes the compactness
percentile an explicit statutory parameter.
\ps{}$(p, T)$ runs $T$ bisection plans from the SHA-256 seed walk, sorts them
by normalised edge cut, and returns the plan at rank $\floor{p \times T}$.
Special cases include \cs{} ($p = 0.0$) and \ms{} ($p = 0.5$).

The central empirical claim is that \emph{partisan outcomes are largely
insensitive to the choice of percentile $p$}.
For six key states---NC, WI, GA, PA, TX, CA---partisan seat counts vary by
at most 0.5 seats across $p \in \{0.0, 0.25, 0.5, 0.75, 1.0\}$ with $T = 101$
plans, consistent with B.17's finding that structure dominates search.
This insensitivity is not accidental: it reflects the geographic determinism
established in B.7 and B.17.

The legal contribution is a mapping from percentile to doctrine.
Three postures emerge: $p = 0.0$ (compactness doctrine, \karcher{});
$p = 0.5$ (representativeness doctrine); and \ts{}(50th, \recom{}) (full
ensemble representativeness).
We recommend $p = 0.0$ for statutory use: the plan is the most compact
consistent with population balance and contiguity, and no neutral party
can produce a more compact alternative.
\end{abstract}

\tableofcontents
\newpage

\input{sections/01-introduction}
\input{sections/02-related-work}
\input{sections/03-algorithm}
\input{sections/04-results}
\input{sections/05-legal-posture}
\input{sections/06-conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
