\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{remark}{Remark}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{array}
\usepackage{microtype}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{lmodern}
\usepackage[T1]{fontenc}

%% SIAM-adjacent style: numbered sections, upright headings
\titleformat{\section}{\normalfont\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection.}{0.5em}{}
\titleformat{\subsubsection}{\normalfont\normalsize\itshape}{\thesubsubsection.}{0.5em}{}

%% Spacing
\setstretch{1.4}

%% Algorithm macros
\newcommand{\bfsg}{\textsc{BfsGrowth}}
\newcommand{\cvd}{\textsc{CVD}}
\newcommand{\metis}{\textsc{Metis}}
\newcommand{\ec}{\ensuremath{\mathrm{EC}}}
\newcommand{\pp}{\ensuremath{\mathrm{PP}}}
\newcommand{\sa}{\textsc{SimulatedAnnealing}}
\newcommand{\be}{\textsc{BisectionEnsemble}}
\newcommand{\ar}{\textsc{ApportionRegions}}
\newcommand{\geosec}{\textsc{GeoSection}}
\newcommand{\areasec}{\textsc{AreaSection}}
\newcommand{\recom}{\textsc{ReCom}}
\newcommand{\cs}{\textsc{ConvergenceSweep}}

%% Math shorthands
\newcommand{\ECnorm}{\mathrm{EC}_{\mathrm{norm}}}
\newcommand{\EC}{\mathrm{EC}}
\newcommand{\PPmean}{\overline{\mathrm{PP}}}

%% -----------------------------------------------------------------------

\title{%
  \textbf{BFS Region-Growing: Greedy Geographic District Packing}\\
  \textbf{from $k$-Farthest Seeds}\\[0.8em]
  \normalsize\textit{B.23 --- Algorithm Design / Geometric Structure}%
}

\author{%
  Gio Della-Libera\\
  \textit{Apportionment Research Group}\\
  \small\texttt{giodl@microsoft.com}
}

\date{May 2026}

\begin{document}

\maketitle

\begin{abstract}
\noindent
We introduce \bfsg{}, the simplest redistricting algorithm: place $k$ seeds
at maximally spread locations via BFS-farthest selection, then repeatedly
assign each unassigned tract to the adjacent district with the greatest
population deficit.
The algorithm requires only the adjacency graph and population weights---no
temperature schedule, no iteration count, no distance metric.
A single pass of the priority queue produces a contiguous assignment;
contiguity is guaranteed by induction on the BFS frontier.
Post-hoc boundary-swap rebalancing (200 iterations) brings all districts
within the standard $\pm 0.5\%$ tolerance.

Total runtime is $O(n \log n)$ per bisection node, matching \metis{} and
substantially faster than simulated annealing (B.19) or CVD (B.22).
Empirical comparison on NC ($k=14$), WI ($k=8$), and TX ($k=38$) shows
that \bfsg{} provides a geometric baseline that is competitive with \metis{}
on Polsby-Popper compactness (typically $5$--$12\%$ \emph{higher} PP than
\metis{}) while sacrificing edge-cut quality ($8$--$15\%$ higher EC than
\metis{}).
The relevant comparison depends on the compactness metric: for PP-based
compactness \bfsg{} is not a lower bound, for EC-based compactness it is.
The \emph{geometric floor} claim applies specifically to edge-cut compactness.

\bfsg{} is useful as: (1) a fast warm-start for MCMC chains
(\recom{}/Forest ReCom), replacing \metis{} when a geographically
structured initial plan is preferred; (2) the ``simplest geographic
partition'' baseline for evaluating how much optimisation algorithms
add beyond basic geographic proximity; and (3) a legal-argumentation
tool demonstrating that the submitted plan outperforms the most
primitive contiguous partition.
Implementation: \texttt{--structure bfs-growth}.
\end{abstract}

\input{sections/01-introduction}
\input{sections/02-background}
\input{sections/03-algorithm}
\input{sections/04-comparison}
\input{sections/05-warmstart}
\input{sections/06-conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
