\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{remark}{Remark}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{array}
\usepackage{microtype}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, backgrounds, calc}

%% Style
\titleformat{\section}{\normalfont\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection.}{0.5em}{}
\titleformat{\subsubsection}{\normalfont\normalsize\itshape}{\thesubsubsection.}{0.5em}{}

\setstretch{1.4}

%% Macros
\newcommand{\ms}{\textsc{MultiScale}}
\newcommand{\recom}{\textsc{ReCom}}
\newcommand{\ec}{\ensuremath{\mathrm{EC}}}
\newcommand{\pp}{\ensuremath{\mathrm{PP}}}
\newcommand{\sburst}{\textsc{ShortBurst}}
\newcommand{\sburstf}{\textsc{ShortBurstForest}}
\newcommand{\be}{\textsc{BisectionEnsemble}}
\newcommand{\fr}{\textsc{ForestReCom}}
\newcommand{\msp}{\textsc{MergeSplit}}
\newcommand{\rhat}{\ensuremath{\hat{R}}}
\newcommand{\ess}{\ensuremath{\mathrm{ESS}}}
\newcommand{\acf}{\ensuremath{\mathrm{ACF}}}
\newcommand{\geoid}{\texttt{GEOID}}
\newcommand{\flip}{\textsc{Flip}}
\newcommand{\vra}{\textsc{VraRecom}}
\newcommand{\smc}{\textsc{SMC}}

\title{%
  \textbf{A Practitioner's Comparison of Redistricting Ensemble Algorithms}\\[0.8em]
  \normalsize\textit{G.14 --- Ensemble Methods / Algorithm Comparison}%
}

\author{%
  Gio Della-Libera\\
  \textit{Apportionment Research Group}\\
  \small\texttt{giodl@microsoft.com}
}

\date{May 2026}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The redistricting ensemble literature has produced at least eight distinct
algorithms: Short-Burst MCMC, SMC, Flip, Forest ReCom, Merge-Split, Multi-scale
MCMC, ShortBurstForest, and VRA-Aware ReCom.
Each algorithm makes different trade-offs among compactness optimisation,
distributional correctness, mixing speed, and runtime.
No single algorithm dominates all metrics simultaneously, yet practitioners
often face an immediate decision: which algorithm should I run for this
redistricting task?
We systematically compare all eight algorithms on North Carolina ($k = 14$),
Wisconsin ($k = 8$), and Texas ($k = 38$), measuring four metrics: edge-cut
compactness (\ec{}), mean Polsby-Popper (\pp{}), partisan seat stability,
and wall-clock runtime.
We additionally compute lag-100 autocorrelation for the chain-based methods
to assess mixing quality.
From these results we derive a practitioner decision tree that maps
redistricting requirements to the appropriate algorithm.
The three-layer compositor (\texttt{--structure}, \texttt{--weights-override},
\texttt{--search}) enables practitioners to combine structure, edge-weight
signals, and search algorithm to address any redistricting task within a
unified CLI interface.
All results are single-run (seed 42) and should be treated as illustrative
rather than definitive.
\end{abstract}

\tableofcontents
\newpage

\input{sections/01-introduction}
\input{sections/02-setup}
\input{sections/03-compactness}
\input{sections/04-distributional}
\input{sections/05-decision-framework}
\input{sections/06-conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
