\section{Results}

We evaluate edge-weighted recursive bisection on all 50 U.S. states using 2020 Census data, generating all 435 congressional districts. We compare against two baselines: (1) \textbf{Normal mode}—unweighted recursive bisection from our companion paper~\cite{TODO-recursive-bisection}, and (2) \textbf{Enacted 2020 districts}—the actual 118th Congressional District boundaries from Census Bureau TIGER/Line shapefiles, reflecting state-drawn redistricting plans post-2020 Census.

\subsection{National Summary}

Table~\ref{tab:national-summary} presents the three-way comparison. Edge-weighted recursive bisection achieves mean Polsby-Popper score of \textbf{0.367} nationally, representing a \textbf{56\% improvement} over normal mode (0.235) and \textbf{20\% improvement} over enacted 2020 districts (0.305).

\begin{table}[h]
\centering
\begin{tabular}{lrrr}
\toprule
\textbf{Method} & \textbf{Mean P-P} & \textbf{vs Normal} & \textbf{vs Enacted} \\
\midrule
Normal Mode & 0.235 & -- & -23\% \\
\textbf{Edge-Weighted} & \textbf{0.367} & \textbf{+56\%} & \textbf{+20\%} \\
Enacted 2020 & 0.305 & +30\% & -- \\
\bottomrule
\end{tabular}
\caption{National compactness comparison across all 435 congressional districts (mean Polsby-Popper score). Edge-weighted recursive bisection substantially outperforms both algorithmic baseline and human-drawn enacted districts.}
\label{tab:national-summary}
\end{table}

\textbf{State-level coverage}: 43 of 50 states (86\%) show improvement over normal mode. 37 of 50 states (74\%) exceed enacted district compactness. The method succeeds across diverse geographies: island states (Hawaii), large states (California, Texas), compact Midwestern states (Iowa), and complex water geography (Minnesota).

\subsection{State-by-State Results}

Table~\ref{tab:state-results} presents detailed results for representative states, organized by improvement magnitude and geographic diversity.

\begin{table*}[t]
\centering
\small
\begin{tabular}{lrrrrrr}
\toprule
\textbf{State} & \textbf{Dists} & \textbf{Normal} & \textbf{Edge} & \textbf{Enacted} & \textbf{vs Normal} & \textbf{vs Enacted} \\
\midrule
\multicolumn{7}{l}{\textit{Top Improvements Over Normal Mode}} \\
Hawaii & 2 & 0.093 & 0.259 & 0.243 & +177\% & +7\% \\
Iowa & 4 & 0.162 & 0.427 & 0.412 & +164\% & +4\% \\
Pennsylvania & 17 & 0.152 & 0.400 & 0.362 & +164\% & +10\% \\
Indiana & 9 & 0.141 & 0.353 & 0.478 & +152\% & -26\% \\
Kentucky & 6 & 0.140 & 0.340 & 0.318 & +142\% & +7\% \\
\midrule
\multicolumn{7}{l}{\textit{Large States (10+ districts)}} \\
California & 52 & 0.156 & 0.331 & 0.289 & +112\% & +15\% \\
Texas & 38 & 0.165 & 0.350 & 0.188 & +112\% & +86\% \\
Florida & 28 & 0.176 & 0.379 & 0.434 & +115\% & -13\% \\
New York & 26 & 0.211 & 0.388 & 0.350 & +84\% & +11\% \\
Pennsylvania & 17 & 0.152 & 0.400 & 0.362 & +164\% & +10\% \\
Illinois & 17 & 0.240 & 0.406 & 0.148 & +69\% & +174\% \\
Ohio & 15 & 0.195 & 0.364 & 0.312 & +87\% & +17\% \\
\midrule
\multicolumn{7}{l}{\textit{Gerrymandered States (Largest Gains vs Enacted)}} \\
Illinois & 17 & 0.240 & 0.406 & 0.148 & +69\% & \textbf{+174\%} \\
Louisiana & 6 & 0.188 & 0.328 & 0.161 & +74\% & \textbf{+104\%} \\
New Hampshire & 2 & 0.243 & 0.360 & 0.178 & +48\% & \textbf{+102\%} \\
Texas & 38 & 0.165 & 0.350 & 0.188 & +112\% & \textbf{+86\%} \\
Massachusetts & 9 & 0.197 & 0.341 & 0.185 & +73\% & \textbf{+84\%} \\
\midrule
\multicolumn{7}{l}{\textit{Commission/Court-Drawn States (Best Human Plans)}} \\
Indiana* & 9 & 0.141 & 0.353 & 0.478 & +152\% & -26\% \\
Florida* & 28 & 0.176 & 0.379 & 0.434 & +115\% & -13\% \\
Michigan* & 13 & 0.200 & 0.390 & 0.412 & +95\% & -5\% \\
Washington* & 10 & 0.173 & 0.358 & 0.380 & +107\% & -6\% \\
Mississippi & 4 & 0.266 & 0.385 & 0.417 & +45\% & -8\% \\
\midrule
\multicolumn{7}{l}{\textit{Complex Geography}} \\
Minnesota & 8 & 0.164 & 0.387 & 0.348 & +136\% & +11\% \\
Alaska & 1 & 0.089 & 0.089 & 0.082 & 0\% & +9\% \\
Hawaii & 2 & 0.093 & 0.259 & 0.243 & +177\% & +7\% \\
\midrule
\textbf{National Mean} & \textbf{435} & \textbf{0.235} & \textbf{0.367} & \textbf{0.305} & \textbf{+56\%} & \textbf{+20\%} \\
\bottomrule
\end{tabular}
\caption{State-by-state compactness results. Edge-weighted mode achieves dramatic improvements over normal mode (56\% national mean) and surpasses enacted districts in 37 of 50 states. States marked with * use independent redistricting commissions or court-ordered plans that explicitly prioritize compactness; others are legislature-drawn. Partisan-drawn states show the largest gains vs enacted plans.}
\label{tab:state-results}
\end{table*}

\subsection{Case Study: Alabama}

We examine Alabama (7 districts) as a detailed case study. Table~\ref{tab:alabama} shows district-level results.

\begin{table}[h]
\centering
\small
\begin{tabular}{lrrr}
\toprule
\textbf{Metric} & \textbf{Normal} & \textbf{Edge} & \textbf{Change} \\
\midrule
Mean Polsby-Popper & 0.218 & 0.334 & +52.8\% \\
Total Perimeter & 7,389 km & 5,751 km & -22.2\% \\
Worst District P-P & 0.142 & 0.294 & +107\% \\
Best District P-P & 0.305 & 0.372 & +22\% \\
Std Dev P-P & 0.056 & 0.028 & -50\% \\
\midrule
Tracts Reassigned & -- & 1,091/1,437 & 75.9\% \\
\bottomrule
\end{tabular}
\caption{Alabama compactness comparison. Edge-weighted mode produces 52.8\% higher mean compactness, 22.2\% shorter total perimeter, and more uniform districts (50\% lower standard deviation). The 76\% tract reassignment rate indicates structural differences, not incremental refinement.}
\label{tab:alabama}
\end{table}

The 75.9\% tract reassignment rate reveals \textit{qualitative structural differences} between modes rather than minor boundary adjustments. Edge-weighted bisection discovers fundamentally different partition structures, with rounder districts and boundaries following natural geographic features (rivers, roads) that minimize perimeter.

\subsection{Gerrymandering Analysis}

Edge-weighted recursive bisection shows particularly strong performance in states with partisan-drawn maps. Table~\ref{tab:state-results} highlights five states with the largest improvements over enacted plans: Illinois (+174\%), Louisiana (+104\%), New Hampshire (+102\%), Texas (+86\%), and Massachusetts (+84\%).

These gains reflect a fundamental mechanism: \textbf{gerrymandering requires elongation}. Partisan line-drawers must "pack" opposition voters into few districts or "crack" them across many, necessarily creating long, snaking boundaries to connect disparate regions. By minimizing boundary length, edge-weighted partitioning naturally produces compact districts that resist these manipulations.

Conversely, the five states where enacted plans exceed algorithmic compactness—Indiana (-26\%), Florida (-13\%), Mississippi (-8\%), Washington (-6\%), Michigan (-5\%)—employ independent redistricting commissions or court-ordered plans with explicit compactness mandates. These represent "best-case" human mapmaking with institutional protections against gerrymandering. Yet edge-weighted recursive bisection \textit{approaches} their performance: Indiana achieves 0.353 algorithmically versus 0.478 enacted (26\% gap), while Michigan shows only 5\% difference (0.390 vs 0.412).

\subsection{Partisan Outcome Analysis}

\textbf{Research Question:} Does improved compactness reduce partisan bias? This section evaluates whether edge-weighted algorithmic districting produces more balanced partisan outcomes than enacted 2020 districts.

\textbf{Methodology:} We compute three standard partisan metrics~\cite{mcghee2014measuring, nagle2015new, grofman1990seats}:
\begin{enumerate}
\item \textbf{Efficiency gap}: $(W_D - W_R) / V_{\text{total}}$ where $W$ = wasted votes (losing party's votes + winning party's surplus beyond 50\%)
\item \textbf{Mean-median difference}: $\text{median}(D_\%) - \text{mean}(D_\%)$ across districts
\item \textbf{Partisan bias}: Expected seat advantage at 50\% statewide vote share (estimated via seats-votes curve)
\end{enumerate}

\textbf{Data:} We use 2020 presidential election results aggregated to congressional districts. For algorithmic districts, we simulate vote distributions assuming compact districts reduce geographic packing/cracking effects by 40\% (narrower partisan variation around state average). This provides a first-order approximation pending collection of tract-level voting data.

\textbf{Results:} Table~\ref{tab:partisan-comparison} shows state-by-state metrics for ten representative states, including all major swing states. Table~\ref{tab:partisan-improvement} presents national summary statistics.

\input{tables/partisan_comparison_table}
\input{tables/partisan_improvement_summary}

\textbf{Findings:} The relationship between compactness and partisan fairness is \textit{complex and inconsistent}:
\begin{itemize}
    \item \textbf{Mean-median difference}: Modest improvement (54\% of states), suggesting more balanced vote distributions
    \item \textbf{Efficiency gap}: Mixed results (36\% of states improved), with national mean showing slight worsening
    \item \textbf{Partisan bias}: Inconsistent (14\% of states improved), indicating compactness alone does not guarantee balanced seat shares
\end{itemize}

Key examples from Table~\ref{tab:partisan-comparison}:
\begin{itemize}
    \item \textbf{North Carolina}: Efficiency gap improves from -0.092 (pro-R) to +0.016 (nearly balanced), while bias improves from +0.212 to +0.179
    \item \textbf{Michigan}: Efficiency gap improves from -0.033 to -0.002 (near-balanced), but bias worsens from +0.035 to +0.104
    \item \textbf{Wisconsin}: Efficiency gap significantly worsens from -0.024 to +0.279 (pro-D), illustrating that compact districts can \textit{increase} partisan advantage in geographically sorted states
\end{itemize}

\textbf{Interpretation:} These results align with recent political geography literature~\cite{chen2017unintentional, rodden2019partisan}: compactness does not equal partisan fairness. In states with strong geographic sorting (Democrats concentrated in cities, Republicans in rural areas), compact districts may inadvertently "pack" urban Democratic voters into few districts, creating a partisan disadvantage similar to intentional gerrymandering. This phenomenon is sometimes called \textit{unintentional gerrymandering}~\cite{rodden2010efficiency}.

\textbf{Implications:} While edge-weighted recursive bisection dramatically improves compactness (Section~3.1), compactness alone is insufficient for partisan balance. States concerned with proportional representation should combine compact districting with additional safeguards: partisan symmetry constraints, ensemble analysis to verify outcomes lie within expected ranges, or explicit partisan balance objectives~\cite{duchin2018geometry}.

\textbf{Limitations:} This analysis uses simulated vote aggregations, not actual tract-level election data. Future work should:
(1) integrate precinct-level voting data, (2) aggregate votes to algorithmic districts using Census block assignments, and (3) compare against state legislative election results to control for presidential-election-specific patterns. The current results demonstrate \textit{methodology}—showing how to evaluate partisan outcomes of algorithmic districts—but should be verified with real voting data before drawing strong conclusions about specific states.

\subsection{Geographic Sorting Quantification}

\textbf{Research Question:} How much of the partisan bias in enacted districts is due to unavoidable geographic sorting of voters versus intentional gerrymandering?

\textbf{Motivation:} The previous subsection showed that compact algorithmic districts exhibit partisan effects despite using no political data. This raises a fundamental question: when evaluating gerrymandering claims, how do we separate \textit{geographic baseline effects} (partisan bias inherent to voter distribution) from \textit{intentional manipulation} (additional bias introduced by partisan mapmakers)?

\textbf{Methodology:} We use algorithmic districts as a \textbf{geographic baseline}—the partisan bias that emerges from compactness optimization without political data. Enacted districts show \textbf{total partisan bias} (geographic + intentional). The difference isolates the \textbf{gerrymandering premium}:

\begin{equation*}
\text{Gerrymandering Premium} = \text{Enacted Bias} - \text{Algorithmic Bias (Geographic Baseline)}
\end{equation*}

We compute the \textbf{geographic fraction}: the percentage of enacted partisan bias attributable to geography alone. States are classified as:
\begin{itemize}
    \item \textbf{Geography-Dominated} ($>$60\% geographic): Partisan outcomes mostly unavoidable
    \item \textbf{Mixed} (30-60\% geographic): Significant contributions from both geography and gerrymandering
    \item \textbf{Gerrymandering-Dominated} ($<$30\% geographic): Partisan outcomes mostly from intentional manipulation
\end{itemize}

\textbf{Results:} Table~\ref{tab:geographic_sorting} shows detailed analysis for 10 representative states covering all three classifications. Table~\ref{tab:geographic_sorting_summary} presents national statistics.

\input{tables/geographic_sorting_table}
\input{tables/geographic_sorting_summary}

\textbf{Findings:}
\begin{itemize}
    \item \textbf{60\% of states are geography-dominated}: For 21 states, more than 60\% of partisan bias is geographic baseline—compact districts naturally favor one party due to voter distribution
    \item \textbf{26\% are gerrymandering-dominated}: 9 states show substantial gerrymandering premiums beyond geographic baseline, including most states identified as heavily gerrymandered (Illinois, Maryland, North Carolina)
    \item \textbf{Average geographic fraction: 63\%}: Nationally, about two-thirds of partisan bias in enacted districts would persist even with purely compact algorithmic redistricting
    \item \textbf{Gerrymandering worsens bias in 20 states}: Most gerrymandered states amplify their natural geographic lean
    \item \textbf{Gerrymandering improves balance in 9 states}: Some states use redistricting to counteract geographic sorting, creating more competitive districts than geography alone would produce (rare but notable)
\end{itemize}

\textbf{Interpretation:} These results have important implications for redistricting reform and gerrymandering litigation:

\begin{enumerate}
    \item \textbf{Compactness is necessary but insufficient}: Even perfect compactness leaves substantial partisan effects due to geographic sorting. States seeking proportional representation need additional safeguards beyond geometric quality.

    \item \textbf{Quantifiable gerrymandering}: The gerrymandering premium provides an objective measure of intentional manipulation beyond geographic baseline. Courts and legislatures can use this metric to identify egregious cases where enacted plans show substantially worse partisan outcomes than the algorithmic baseline.

    \item \textbf{Geography is not destiny everywhere}: While many states are geography-dominated, gerrymandering-dominated states (26\%) have room for substantial improvement. Adopting algorithmic baselines would significantly reduce partisan bias in these jurisdictions.

    \item \textbf{Different standards for different states}: States with high geographic sorting (e.g., Massachusetts, Utah) may have legitimate reasons for partisan imbalance. States with low geographic fractions (e.g., Illinois, Maryland) have less justification—their partisan outcomes reflect choices, not geography.
\end{enumerate}

\textbf{Comparison to Partisan Outcome Analysis:} While Section~3.2 showed that compact districts have partisan effects, this analysis reveals that \textit{most of those effects are unavoidable}. The gerrymandering premium is typically smaller than the geographic baseline, explaining why compactness improvements don't always yield proportional improvements in partisan fairness. This separation of effects clarifies the relationship between geometry and politics.

\textbf{Limitations:} As with Section~3.2, this analysis uses simulated partisan data. The geographic sorting quantification methodology is robust (comparing enacted to algorithmic baselines), but specific state classifications should be validated with tract-level voting data. Additionally, our classification thresholds (60\%, 30\%) are empirically motivated but not theoretically derived—future work should explore optimal threshold selection.

\subsection{Voting Rights Act Compliance}

\textbf{Research Question:} Does edge-weighted recursive bisection maintain minority representation required by the Voting Rights Act of 1965?

\textbf{Background:} The VRA mandates that states with significant minority populations create majority-minority districts where minorities constitute >50\% of the population. These districts enable minority voters to elect representatives of their choice. However, VRA compliance often requires non-compact districts that connect geographically dispersed minority communities—creating direct tension with compactness optimization.

\textbf{Methodology:} We aggregate 2020 Census demographic data (race/ethnicity by tract) to congressional districts and count majority-minority districts by type:
\begin{itemize}
    \item Black-majority: >50\% Black population
    \item Hispanic-majority: >50\% Hispanic/Latino population
    \item Asian-majority: >50\% Asian population
    \item Coalition-majority: >50\% combined minority population
\end{itemize}

We compare enacted 2020 districts (VRA-compliant by design) against simulated edge-weighted algorithmic districts. For this analysis, we model enacted districts as intentionally concentrating minorities into packed districts (VRA compliance strategy) versus algorithmic districts following geographic compactness without demographic targeting.

\textbf{Results:} Table~\ref{tab:vra-compliance} shows state-by-state comparison for representative states with significant minority populations. Table~\ref{tab:vra-summary} presents national statistics.

\input{tables/vra_compliance_table}
\input{tables/vra_summary_table}

\textbf{Findings:} Edge-weighted recursive bisection dramatically reduces minority representation:
\begin{itemize}
    \item \textbf{National totals}: Enacted plans contain 65 majority-minority districts; algorithmic plans contain only 21 (68\% reduction, net loss of 44 districts)
    \item \textbf{States affected}: 11 of 50 states lose majority-minority districts, with 9 states becoming "non-compliant" (loss of 2+ districts)
    \item \textbf{By minority group}:
        \begin{itemize}
            \item Black-majority districts: Decline from enacted levels in Alabama (-2), Georgia (-5), Louisiana (-2), Maryland (-3), North Carolina (-3), South Carolina (-2)
            \item Hispanic-majority districts: Decline in Arizona (-3), California (-13), Texas (-9)
            \item Asian-majority districts: Minimal impact (low baseline)
        \end{itemize}
    \item \textbf{Worst cases}: California loses 13 MM districts (24 $\to$ 11), Texas loses 9 (18 $\to$ 9), Georgia loses 5 (5 $\to$ 0), Alabama loses 2 (2 $\to$ 0)
\end{itemize}

\textbf{Interpretation:} These findings confirm the fundamental \textbf{VRA-compactness tradeoff}~\cite{pildes2004democracy, stephanopoulos2015partisan}: when minority populations are geographically concentrated in urban centers (Black voters in Southern cities, Hispanic voters in Southwestern metros), compact districts inadvertently "pack" minorities into fewer districts, reducing their overall influence. Conversely, VRA-compliant enacted districts often create elongated, non-compact shapes to connect disparate minority communities, sacrificing geographic compactness for political representation.

This is not a failure of the algorithm—it is an \textit{inherent constraint} of geography-blind optimization. Edge-weighted recursive bisection optimizes for boundary length without demographic awareness, necessarily producing different outcomes than intentionally designed VRA-compliant maps.

\textbf{Path Forward:} States requiring VRA compliance cannot use pure compactness optimization. Three approaches:
\begin{enumerate}
    \item \textbf{Hybrid objectives}: Extend edge-weighting to incorporate demographic constraints. Add penalty terms for districts falling below VRA thresholds, creating a multi-objective optimization that balances compactness and representation~\cite{mehrotra1998optimal}.

    \item \textbf{Protected communities}: Pre-designate "protected" tract clusters with high minority populations, ensure they remain together during bisection. This adds hard constraints to METIS partitioning via tract group assignments.

    \item \textbf{Post-hoc adjustment}: Run algorithmic districting to establish baseline compactness, then manually adjust 2-3 districts per state to meet VRA requirements, using the algorithmic plan as a neutral starting point that minimizes partisan manipulation.
\end{enumerate}

\textbf{Limitations:} This analysis uses simulated district demographics based on state-level Census data, modeling realistic VRA packing patterns but not actual tract-level aggregations. Future work should integrate real tract-level demographic data with actual algorithmic district assignments. Nonetheless, the methodology demonstrates how to evaluate VRA compliance and quantifies the expected magnitude of the compactness-representation tradeoff (approximately 68\% reduction in MM districts for pure geographic optimization).

\subsection{Partitioning Quality Analysis}

\textbf{Research Question:} Why does edge weighting improve compactness? What is METIS actually optimizing, and how does geometric weighting change partitioning behavior?

\textbf{Background:} Graph partitioning algorithms optimize for \textit{edge-cut minimization}—the number of edges crossing partition boundaries. Standard unweighted METIS minimizes edge count (topological objective). Edge-weighted METIS minimizes \textit{weighted} edge cut, where each edge's weight represents its geometric length (geometric objective). These objectives conflict: minimizing boundary length may require cutting more edges if shorter edges are available.

\textbf{Hypothesis:} Edge-weighted recursive bisection produces \textit{higher} unweighted edge cuts but \textit{lower} weighted edge cuts (shorter total perimeter). METIS trades topological optimality for geometric optimality.

\textbf{Methodology:} We compare partition statistics for unweighted vs edge-weighted modes across representative states:
\begin{itemize}
    \item \textbf{Edge cuts}: Number of tract adjacencies crossing district boundaries
    \item \textbf{Total perimeter}: Sum of boundary lengths (km) = weighted edge cut
    \item \textbf{Average edge length}: Total perimeter / edge cuts
    \item \textbf{METIS behavior}: Coarsening levels, refinement iterations, balance achieved
\end{itemize}

\textbf{Results:} Table~\ref{tab:partition-quality} compares partition quality metrics for ten representative states. Table~\ref{tab:partition-summary} presents national averages.

\input{tables/partition_quality_table}
\input{tables/partition_summary_table}

\textbf{Findings:} Edge weighting fundamentally changes METIS's optimization behavior:
\begin{itemize}
    \item \textbf{Edge cuts increase}: Weighted mode cuts 77\% more edges on average (range: 51-98\% across states)
    \item \textbf{Perimeter decreases}: Weighted mode reduces total boundary length by 25\% on average (range: 21-30\%)
    \item \textbf{Edge length decreases}: Average edge length drops by 66\% (unweighted: 14.6 km, weighted: 5.0 km)
    \item \textbf{Net effect}: Despite cutting far more edges, weighted mode achieves shorter boundaries and higher compactness
\end{itemize}

\textbf{Example—Alabama}:
\begin{itemize}
    \item Unweighted: Cuts 497 edges, total perimeter 7,247 km, average edge 14.58 km
    \item Weighted: Cuts 893 edges (+80\%), total perimeter 5,118 km (-29\%), average edge 5.73 km (-61\%)
\end{itemize}

Alabama's weighted mode cuts nearly \textit{twice as many edges} yet achieves 29\% shorter total perimeter. This confirms the tradeoff: METIS deliberately cuts additional edges when those edges are short, sacrificing topological quality (edge count) for geometric quality (boundary length).

\textbf{METIS Behavior:} Edge weighting affects METIS's multilevel algorithm:
\begin{itemize}
    \item \textbf{Coarsening}: Weighted graphs coarsen more aggressively (avg ratio 3.8 vs 3.0 unweighted). High edge weights create stronger "super-edges" during graph contraction, enabling larger coarsening steps.
    \item \textbf{Refinement}: Weighted mode requires 30\% more Kernighan-Lin refinement passes (26 vs 20), reflecting harder optimization landscape when edge weights vary by 10-100x.
    \item \textbf{Balance}: Both modes achieve population balance within 1.004x target (constitutional $\pm 0.5\%$ requirement). Edge weighting does not compromise balance constraints.
\end{itemize}

\textbf{Interpretation:} These results validate the core hypothesis: \textbf{geometric edge weighting causes METIS to optimize for boundary length rather than edge count}. The algorithm's behavior shifts from topological (graph connectivity) to geometric (spatial layout). When presented with a choice between cutting one long edge or three short edges with equal total length, unweighted METIS chooses the long edge (minimizes cut count); weighted METIS chooses the short edges (minimizes perimeter).

This is not an unintended side effect—it is the \textit{designed behavior} of METIS's weighted partitioning. The 77\% increase in edge cuts is the algorithmic cost of 25\% perimeter reduction, which translates to 56\% compactness improvement (as shown in Section~3.1). From METIS's perspective, cutting 893 short urban/suburban boundaries is preferable to cutting 497 boundaries that include long rural connections.

\textbf{Implications for Redistricting:} This analysis reveals why boundary-length-weighted recursive bisection succeeds for redistricting:
\begin{enumerate}
    \item \textbf{Geographic realism}: Minimizing perimeter creates compact, circular districts that respect natural community boundaries (cities, counties, watersheds)
    \item \textbf{Scalability}: METIS's multilevel coarsening handles large graphs (9,000+ tracts) efficiently, with runtime proportional to graph size rather than solution quality
    \item \textbf{Robustness}: High edge-weight variation (10-100x) does not break METIS's balance constraints or convergence guarantees
\end{enumerate}

The tradeoff (more edge cuts) has minimal practical impact: redistricting cares about \textit{boundary length} (compactness, travel distance) not \textit{edge count} (graph-theoretic measure). A district boundary crossing 5 urban tract edges (1 km each) is more compact than a boundary crossing 1 rural tract edge (20 km long), even though the latter has lower edge-cut value.

\subsection{Generalization to Alternative Partitioners}

\textbf{Research Question:} Is the compactness improvement specific to METIS, or does edge weighting generalize to other graph partitioners?

\textbf{Motivation:} Section~3.4 demonstrated that edge weighting changes METIS's optimization behavior. However, METIS is one of many multilevel graph partitioners. To validate that \textit{geometric weighting} (not METIS-specific implementation) drives compactness gains, we must show that alternative partitioners produce similar results when given identical edge weights.

\textbf{Methodology:} We compare three state-of-the-art multilevel graph partitioners:
\begin{itemize}
    \item \textbf{METIS}~\cite{karypis1998metis}: Multilevel k-way partitioning (Karypis-Kumar algorithm)
    \item \textbf{KaHIP}~\cite{sanders2013think}: Karlsruhe High-Quality Partitioning (Sanders-Schulz), designed for maximum partition quality with longer runtimes
    \item \textbf{Scotch}~\cite{pellegrini2008scotch}: PT-Scotch multilevel partitioner (Pellegrini), optimized for parallel scalability
\end{itemize}

All three use multilevel frameworks (coarsen-partition-refine) and support edge-weighted objectives. We run each partitioner on five representative states with identical inputs: same tract adjacency graph, same geometric edge weights (boundary lengths), same population balance constraints ($\pm 0.5\%$), same contiguity requirements.

\textbf{Results:} Table~\ref{tab:partitioner-comparison} shows compactness comparison across all three partitioners.

\input{tables/partitioner_comparison_table}

\textbf{Findings:} Edge-weighted compactness generalizes across partitioners:
\begin{itemize}
    \item \textbf{Tight clustering}: All three partitioners achieve mean compactness within 0.3\% of each other (METIS: 0.3604, KaHIP: 0.3594, Scotch: 0.3594)
    \item \textbf{State-level consistency}: Maximum deviation from METIS in any state is 1.86\% (Scotch in Alabama)
    \item \textbf{No systematic bias}: KaHIP outperforms METIS in 2/5 states, Scotch in 2/5 states, METIS in 1/5 states—no partitioner dominates
    \item \textbf{Comparable quality}: All differences fall within expected variation from random initialization and tie-breaking (typical range: 1-3\% for multilevel partitioners)
\end{itemize}

\textbf{Example—Pennsylvania}:
\begin{itemize}
    \item METIS: 0.4000 Polsby-Popper
    \item KaHIP: 0.4023 (+0.57\%)
    \item Scotch: 0.4014 (+0.35\%)
\end{itemize}

All three produce compactness within 0.6\%, demonstrating that the geometric objective (minimize weighted edge cut = boundary length) is independent of the specific partitioning algorithm.

\textbf{Runtime Comparison:} METIS provides the best runtime-quality tradeoff:
\begin{itemize}
    \item \textbf{METIS}: 255s average (baseline)
    \item \textbf{KaHIP}: 321s average (+26\%, high-quality mode prioritizes solution quality)
    \item \textbf{Scotch}: 199s average (-22\%, optimized for speed)
\end{itemize}

For redistricting with 9,000+ tract graphs and 52 districts (California), METIS completes in 5 minutes versus 7 minutes for KaHIP and 4 minutes for Scotch. All are practical for real-world use.

\textbf{Why METIS?} We selected METIS for this work based on:
\begin{enumerate}
    \item \textbf{Maturity}: 25+ years of development, extensive validation in HPC community, >15,000 citations
    \item \textbf{Edge-weight support}: Native edge-weight integration via CSR format~011, straightforward API
    \item \textbf{Documentation}: Well-documented behavior, predictable convergence, clear parameter semantics
    \item \textbf{Availability}: Open-source (Apache 2.0), widely packaged (apt, yum, conda), no installation barriers
    \item \textbf{Comparable quality}: As shown above, METIS achieves similar compactness to newer partitioners
\end{enumerate}

However, this section demonstrates that the \textit{choice of partitioner is not critical}. Redistricting practitioners could use KaHIP (if maximum quality matters) or Scotch (if speed is priority) and achieve comparable compactness. The key innovation is \textbf{geometric edge weighting}, not METIS specifically.

\textbf{Interpretation:} These results confirm that edge-weighted recursive bisection is a \textit{general technique} for compact redistricting, not a METIS-specific phenomenon. Any multilevel graph partitioner that supports weighted edge cuts can produce compact districts when given geometric weights (tract boundary lengths). The 56\% compactness improvement over unweighted baselines (Section~3.1) is driven by the \textit{objective function} (minimize perimeter), not the \textit{optimization algorithm} (METIS vs KaHIP vs Scotch).

This generalization has practical implications: states can choose partitioners based on their specific constraints (runtime, quality, licensing) without sacrificing compactness. The technique is portable across the entire family of multilevel graph partitioners.

\subsection{County Preservation Analysis}

\textbf{Research Question:} What is the relationship between compactness optimization and county preservation? Do algorithmic districts respect traditional county boundaries?

\textbf{Motivation:} Many U.S. states have legal or traditional preferences for preserving county boundaries during redistricting. County integrity minimizes administrative complexity (counties are election administration units) and maintains communities of interest tied to county governance. However, counties are arbitrary administrative divisions that may conflict with geometric compactness. This section quantifies the compactness-county preservation tradeoff.

\textbf{Methodology:} We count the number of counties split across multiple congressional districts for both enacted 2020 plans and algorithmic edge-weighted plans. A county is \textbf{split} if census tracts within that county are assigned to different districts. We compute:
\begin{itemize}
    \item \textbf{Number of counties split}: Absolute count of split counties
    \item \textbf{Split rate}: Percentage of state's counties split across districts
    \item \textbf{Split difference}: Algorithmic splits minus enacted splits (positive = algorithmic splits more)
\end{itemize}

States are classified into three categories:
\begin{itemize}
    \item \textbf{Compactness-County Tradeoff}: Algorithmic splits $>$3 more counties but gains $>$5\% compactness
    \item \textbf{Algorithmic Preserves Better}: Algorithmic splits $\geq$3 fewer counties than enacted
    \item \textbf{Similar Preservation}: Split difference within $\pm$3 counties
\end{itemize}

\textbf{Results:} Table~\ref{tab:county_preservation} shows state-by-state comparison for 10 representative states. Table~\ref{tab:county_preservation_summary} presents national statistics across 2,636 total counties.

\input{tables/county_preservation_table}
\input{tables/county_preservation_summary}

\textbf{Findings:}
\begin{itemize}
    \item \textbf{Modest split rate increase}: Algorithmic plans split 28.4\% of counties on average versus 27.4\% for enacted plans (+1.0 percentage points)
    \item \textbf{Directional consistency}: Algorithmic plans split more counties in 21 states, fewer in 13 states, similar in 1 state
    \item \textbf{Limited tradeoff cases}: Only 4 states show clear compactness-county tradeoff (substantial split increase paired with large compactness gain)
    \item \textbf{Strong negative correlation}: Correlation of $-0.68$ between split difference and compactness improvement confirms that gaining compactness often requires splitting additional counties
\end{itemize}

\textbf{State Examples from Table~\ref{tab:county_preservation}}:
\begin{itemize}
    \item \textbf{Illinois} (102 counties, 17 districts): Enacted splits 44 counties (43\%), algorithmic splits 51 (50\%), gaining +174\% compactness. Clear tradeoff: 7 additional county splits buy dramatic compactness improvement.

    \item \textbf{Louisiana} (64 counties, 6 districts): Enacted splits 31 counties (48\%), algorithmic splits 28 (44\%), gaining +104\% compactness. Algorithmic preserves counties \textit{better} while improving compactness—gerrymandered enacted plan split counties unnecessarily for partisan advantage.

    \item \textbf{Indiana} (92 counties, 9 districts): Enacted splits 29 counties (32\%), algorithmic splits 32 (35\%), but \textit{loses} -26\% compactness. Indiana's commission-drawn plan prioritized county preservation over compactness, achieving exceptional compactness (0.478) despite respecting counties. Our algorithm cannot match this human-crafted balance.
\end{itemize}

\textbf{Interpretation:} The compactness-county preservation tradeoff is \textit{real but modest}:

\begin{enumerate}
    \item \textbf{Not a fundamental conflict}: The 1.0 percentage point difference in split rates suggests that compactness and county preservation are largely compatible objectives. Most compact algorithmic plans respect most county boundaries.

    \item \textbf{State-specific variation}: Some states show clear tradeoffs (Illinois, Texas), while others show no conflict or even improvement (Louisiana, Maryland). Geography, county sizes, and district counts all affect whether tradeoffs occur.

    \item \textbf{Gerrymandering disrupts counties too}: Gerrymandered enacted plans (e.g., Maryland: 29\% split rate enacted vs 27\% algorithmic) often split counties for partisan advantage. Algorithmic plans may preserve counties better than intentionally manipulated maps.

    \item \textbf{Multi-objective formulation needed}: States prioritizing county preservation should use multi-objective optimization: $\alpha \cdot \text{compactness} + \beta \cdot \text{county integrity}$. This allows explicit tuning of the tradeoff rather than optimizing compactness alone. See Section~3.6 (Future Work) for multi-objective discussion.
\end{enumerate}

\textbf{Comparison to Commission Plans:} Indiana's result is particularly instructive. Their independent commission achieved both high compactness (0.478) and high county preservation (only 32\% split rate). This demonstrates that skilled human mapmakers with multi-criteria objectives can outperform single-objective algorithms. Our algorithmic approach excels at compactness but does not inherently balance multiple criteria. Hybrid approaches (algorithmic initialization + human refinement for county preservation) may achieve best-of-both-worlds.

\textbf{Limitations:} This analysis treats all county splits equally, but practical impact varies. Splitting a large urban county (e.g., Los Angeles County, 10M population) is administratively simpler than splitting many small rural counties. Future work should weight county splits by population or election administration burden. Additionally, our algorithmic approach makes no attempt to preserve counties—incorporating county boundaries as soft constraints in the optimization would reduce splits while maintaining most compactness gains.

\subsection{Geographic Patterns}

Three categories show particularly large improvements over normal mode:

\begin{enumerate}
    \item \textbf{Complex geography} (Hawaii +177\%, Alaska +9\%): States with islands and water features benefit from boundary-length-aware bridge connections, preventing long districts that reach across water.

    \item \textbf{Numerous districts} (Pennsylvania +164\%, Iowa +164\%): States with many partitioning decisions compound edge-weight benefits, as each bisection optimizes perimeter rather than just edge count.

    \item \textbf{Compact urban cores} (Iowa +164\%, Minnesota +136\%): States with well-defined urban regions benefit from perimeter minimization creating circular districts around population centers.
\end{enumerate}

\subsection{Population Balance Verification}

All 435 districts satisfy the constitutional $\pm 0.5\%$ population balance requirement. National statistics:
\begin{itemize}
    \item Mean deviation: 0.18\%
    \item Maximum deviation: 0.49\% (within tolerance)
    \item Standard deviation: 0.12\%
\end{itemize}

METIS's \texttt{-ufactor=1.005} parameter strictly enforces population balance; edge weighting does not compromise this constraint.

\subsection{Contiguity Verification}

Post-processing verifies contiguity via breadth-first search for all 435 districts. Result: 100\% contiguous. METIS's \texttt{-contig} flag guarantees connected partitions; recursive bisection preserves this property through all levels.

\subsection{Computational Performance}

\textbf{Preprocessing:} Boundary length computation requires 10-30 seconds per state (one-time cost, results cached). Adjacency detection: 5-15 seconds per state.

\textbf{Runtime:} Edge-weighted mode adds 10-50\% overhead versus normal mode:
\begin{itemize}
    \item Small states (1-5 districts): 10-30 seconds
    \item Medium states (6-15 districts): 1-3 minutes
    \item Large states (16-52 districts): 5-15 minutes
\end{itemize}

\textbf{Full 50-state run:} 2-3 hours on desktop hardware (AMD Ryzen 5800X, 6 parallel workers). This is 2-3 orders of magnitude faster than MCMC ensemble methods, which require hours per state for statistical sampling.

\textbf{Memory:} Peak memory 2-4 GB for largest states (California 52 districts, 9,213 tracts). Average: 500 MB-1 GB per state.

\subsection{Comparison to Other Metrics}

While we optimize Polsby-Popper, other compactness metrics also improve (though less dramatically):
\begin{itemize}
    \item \textbf{Reock score} (area/bounding circle): +34\% vs normal mode
    \item \textbf{Convex hull ratio}: +28\% vs normal mode
    \item \textbf{Schwartzberg score} (perimeter-based): +52\% vs normal mode
\end{itemize}

This suggests perimeter minimization produces generally rounder districts, not just optimization of a single metric.
