\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{remark}{Remark}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{array}
\usepackage{microtype}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{lmodern}
\usepackage[T1]{fontenc}

%% Style
\titleformat{\section}{\normalfont\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection.}{0.5em}{}
\titleformat{\subsubsection}{\normalfont\normalsize\itshape}{\thesubsubsection.}{0.5em}{}

\setstretch{1.4}

%% Macros
\newcommand{\sbf}{\textsc{ShortBurstForest}}
\newcommand{\sbms}{\textsc{ShortBurstMergeSplit}}
\newcommand{\sburst}{\textsc{ShortBurst}}
\newcommand{\flip}{\textsc{Flip}}
\newcommand{\recom}{\textsc{ReCom}}
\newcommand{\frecom}{\textsc{ForestRecom}}
\newcommand{\ms}{\textsc{MergeSplit}}
\newcommand{\ec}{\ensuremath{\mathrm{EC}}}
\newcommand{\pp}{\ensuremath{\mathrm{PP}}}
\newcommand{\be}{\textsc{BisectionEnsemble}}
\newcommand{\cs}{\textsc{ConvergenceSweep}}
\newcommand{\rhat}{\ensuremath{\hat{R}}}
\newcommand{\ess}{\ensuremath{\mathrm{ESS}}}

\title{%
  \textbf{Short-Burst Optimization with Calibrated Chains}\\
  \textbf{Composing Burst Optimization and Distributional Correctness}\\[0.8em]
  \normalsize\textit{G.12 --- Ensemble Methods / Composable Search}%
}

\author{%
  Gio Della-Libera\\
  \textit{Apportionment Research Group}\\
  \small\texttt{giodl@microsoft.com}
}

\date{May 2026}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Short-Burst (G.6) is a meta-algorithm: it runs \(n_{\mathrm{bursts}}\) short
Markov chains and selects the endpoint with the best edge-cut.
The underlying chain type is a parameter.
We implement two new chain variants---\sbf{} and \sbms{}---by composing
Short-Burst with ForestRecom (G.9) and MergeSplit (G.10) chains,
respectively.
This pairing delivers two properties simultaneously: compactness optimization
(the contribution of the Short-Burst wrapper) and approximate distributional
correctness (the contribution of the ForestRecom and MergeSplit chains).
We evaluate all three Short-Burst variants---standard \recom{}-based,
\sbf{}, and \sbms{}---on North Carolina ($k=14$), Wisconsin ($k=8$),
and Texas ($k=38$) at equal compute budget (1000 total steps per run).
The key empirical finding is that standard \sburst{} achieves comparable
or slightly better final edge-cut than the calibrated variants because its
higher per-burst acceptance rate ($\approx 80\%$) explores more plans per
burst; ForestRecom and MergeSplit spend compute on the Metropolis--Hastings
ratio computation that is unnecessary for pure compactness optimization.
However, \sbf{} produces better \emph{ensemble coverage} near good plans
because each accepted step is an approximately correct sample from the
stationary distribution.
We introduce the \emph{per-burst acceptance rate} diagnostic, recorded in
the audit manifest, which quantifies the effective burst length for each
chain type.
The three variants are available via
\texttt{--search short-burst-forest} and
\texttt{--search short-burst-merge-split}.
\end{abstract}

\tableofcontents
\newpage

\input{sections/01-introduction}
\input{sections/02-background}
\input{sections/03-algorithm}
\input{sections/04-comparison}
\input{sections/05-when-to-use}
\input{sections/06-conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
