\label{sec:checklist}

Table~\ref{tab:checklist} presents the expert witness checklist, mapping
\texttt{redist} command outputs to each Gingles prong.  The checklist is
organized as a sequential workflow; outputs from earlier steps feed into
later analysis.

\begin{table}[h]
\centering
\caption{Expert Witness Checklist: \texttt{redist} Outputs for Each Gingles Prong}
\label{tab:checklist}
\begin{threeparttable}
\begin{tabular}{p{1cm}p{3.5cm}p{4cm}p{3.5cm}}
\toprule
Prong & Legal Element & \texttt{redist} Command & Output for Report \\
\midrule
\multirow{4}{*}{1} &
  Minority constitutes majority of CVAP in proposed district &
  \texttt{redist analyze --types demographic --vra-mode} &
  Minority CVAP \% per district \\
\cmidrule{2-4}
  & Geographic compactness of minority community &
  \texttt{redist analyze --types compactness} &
  Alignment score $A_d$; Polsby-Popper score \\
\cmidrule{2-4}
  & Ensemble robustness of Prong 1 finding &
  \texttt{redist analyze --ensemble-size 1000} &
  5th--95th pct alignment range \\
\cmidrule{2-4}
  & Visual illustration of district and minority heatmap &
  \texttt{redist map --overlay minority-cvap} &
  Map figure for exhibit \\
\midrule
\multirow{3}{*}{2} &
  Minority political cohesion (primary elections preferred) &
  \texttt{redist analyze --types bloc-voting} &
  $\hat{\beta}_{\text{min}}$ with 95\% CI per election \\
\cmidrule{2-4}
  & Cross-election stability of cohesion estimate &
  Multiple \texttt{--election} arguments &
  Range table across elections \\
\cmidrule{2-4}
  & Identification of minority-preferred candidate &
  Same command &
  Candidate-level vote shares \\
\midrule
\multirow{4}{*}{3} &
  Majority bloc voting (baseline, no partisan control) &
  \texttt{redist analyze --types bloc-voting} &
  $\hat{\beta}_{\text{race}}$ without control \\
\cmidrule{2-4}
  & Racial component after \textit{Callais} partisan control &
  \texttt{redist analyze --types bloc-voting --partisan-control} &
  $\hat{\beta}_{\text{race}}$ with control; HC3 SE \\
\cmidrule{2-4}
  & Holm multiple-testing correction &
  Same command &
  Holm-corrected $p$-values per election \\
\cmidrule{2-4}
  & LOO robustness &
  Same command &
  LOO stability table \\
\midrule
Supporting &
  Overall criteria compliance for proposed remedial map &
  \texttt{redist analyze --types all} &
  Full criteria-compliance report \\
\cmidrule{2-4}
  & Deposition-ready log with hash chain &
  \texttt{redist depo --verify} &
  Canonical JSONL log + manifest \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item Note: Each \texttt{redist analyze} invocation should specify
  \texttt{--state}, \texttt{--year}, and \texttt{--version} flags to
  identify the run.  The deposition log (\texttt{redist depo}) produces
  a tamper-evident hash chain for all outputs, suitable for discovery
  production.
\end{tablenotes}
\end{threeparttable}
\end{table}

\subsection{Deposition Readiness}

Expert testimony in Section 2 litigation will be subject to deposition.
The \texttt{redist depo} module produces a canonical JSONL log of all
analysis steps with a SHA-256 hash chain, enabling verification that
reported outputs match what the algorithm produced~\citep{dellaLibera2026vra}.
The log is suitable for discovery production and provides a complete
audit trail from input data to expert report figures.

\subsection{Common Expert Challenges and Responses}

Expert witnesses using this methodology should anticipate the following
challenges in deposition or cross-examination:

\begin{description}
  \item[``The alignment score threshold is arbitrary.'']
    Response: The 0.5 threshold is calibrated to the universe of districts
    that courts have accepted as satisfying Prong 1.  The ensemble analysis
    provides the full distribution of alignment scores for the proposed
    district; the threshold is a summary, not the primary finding.

  \item[``The partisan control absorbs true racial effects.'']
    Response: The partisan control uses presidential vote, which reflects
    the party alignment of the \emph{district} environment, not the racial
    composition of individual precincts.  The coefficient $\hat{\beta}_{\text{race}}$
    measures the relationship between precinct racial composition and
    majority vote share holding constant the district's partisan baseline.
    The LOO analysis confirms that the finding is not sensitive to the
    specific elections selected.

  \item[``The algorithm could have drawn a different map that doesn't satisfy Prong 1.'']
    Response: The ensemble range reported for Prong 1 documents the fraction
    of map configurations that satisfy the geographic compactness criterion.
    If the Prong 1 finding is robust across more than 90\% of configurations,
    the alternative maps that fail Prong 1 are outliers in the space of
    population-balanced, compact configurations.
\end{description}
