\section{Discussion}
\label{sec:discussion}

\subsection{AreaSection as a Geometric Regularity Mechanism}

The central empirical finding of this paper is that AreaSection changes the
\emph{structure} of redistricting outcomes without changing the \emph{outcome}
for partisan seat counts.
In Wisconsin and North Carolina --- and, as shown in \S\ref{sec:evaluation},
across all 44 successfully completed states --- AreaSection and GeoSection
produce the same partisan seat allocations.

We use the term \emph{geometric regularity} rather than \emph{geographic
fairness} to describe this contribution.
``Fairness'' in redistricting law has a specific meaning: it refers to equal
treatment under the Equal Protection Clause or to the absence of discriminatory
effect under the Voting Rights Act.
AreaSection does not alter either.
What it does is constrain the \emph{bisection tree structure}: each first-level
split respects a geographic symmetry condition (comparable land area on both
sides) that standard bisection ignores.
Whether this geometric regularity property is a legal virtue is a question for
courts and legislatures; the empirical question this paper addresses is whether
it has partisan consequences, and the answer is: no.

This finding has an important implication: the proportionality gap in competitive
states is not a product of urban peeling.
It is a product of the underlying population distribution.
A 50.3\,\% Democratic state (Wisconsin) with Republicans distributed across
a larger geographic footprint will consistently produce a 3D/5R congressional
delegation under any geographically neutral algorithm, regardless of whether
the bisection strategy is urban-peeling (GeoSection) or area-balanced
(AreaSection).

\paragraph{On competitiveness.}
AreaSection eliminates knife-edge districts: GeoSection creates two Wisconsin
districts at 49.4\,\% Democratic vote share, while AreaSection creates zero
such seats.
This finding cuts in both directions and we do not present it as a normative
benefit without qualification.

Competitive districts are widely regarded as electorally valuable --- they
produce responsive elections, prevent entrenchment, and give minority-party
voters real leverage.
If knife-edge districts disappeared, a swing toward Republicans could lock in
a 5R/3D result even more durably than a map with two 49.4\,\% seats.
Whether seat stability is preferable to seat contestability is a normative and
legal question that different democratic theories answer differently.

We report the competitiveness reduction as an empirical observation.
An algorithm designer choosing between GeoSection and AreaSection should
consider both the seat-count equivalence and the competitiveness difference;
the choice is not a priori neutral.

\paragraph{Legal theory.}
The area balance constraint has a natural legal articulation:
census tracts are administrative units whose areas reflect the physical
geography of the state, not the partisan composition of their residents.
A redistricting algorithm that targets geographic symmetry respects the
physical structure of the state in a way that population-only algorithms
do not.
This is a procedural rather than partisan claim: the constraint is
content-neutral (it treats all land area equally regardless of who lives there)
and its partisan consequences are empirically nil.

Under \emph{Rucho v. Common Cause} (2019), federal courts cannot remedy
partisan gerrymandering under the Constitution, but state courts can and
have done so \citep{lwv2018pa,harper2022nc}.
State courts reviewing redistricting algorithms may find the geographic
balance criterion independently defensible:
``we drew district boundaries to minimise total boundary length while
giving each half of the state comparable geographic footprint''
is a claim with a clear, politically neutral rationale.

\subsection{Why Seats Don't Change}
\label{sec:why-seats-stable}

The stability of seat counts across GeoSection and AreaSection reflects
a deeper mathematical structure.
In states with strong geographic partisan sorting (urban Democrats,
rural Republicans), the \emph{number} of Democratic seats is determined
by how many contiguous geographic regions have Democratic majorities,
not by where the bisection boundaries fall.
AreaSection changes where the boundaries fall, but not which geographic
regions are sufficiently Democratic to elect a Democrat.

This suggests that the partisan outcome is, in a precise sense, a property
of the population distribution, not of the bisection algorithm.
GeoSection (B.7) establishes this claim rigorously for the seed-sweep
convergence case.
AreaSection provides corroborating evidence from a different direction:
changing the geometric constraint does not change the seat count.

\subsection{Limitations}
\label{sec:limitations}

\paragraph{Large-state balance failures.}
Six states --- FL, IL, MI, NY, PA, TX --- fail the 1.5\,\% population balance
check.
These six states account for 139 of 435 congressional seats (32\,\%) and are
among the most politically consequential.

The failure mode is \emph{architectural, not incidental}.
The dual constraint at the first bisection level produces subgraph populations
that deviate from the ideal district-population target.
For example, a 3:5 top-level split that achieves only 27.5\,\% instead of
37.5\,\% of the state population leaves the 3-district subgraph with
$\sim$540K per district rather than 737K.
The recursive bisection cannot compensate: it has insufficient population.
The deviation is not a rounding issue but a consequence of METIS prioritising
population balance over area balance when both constraints conflict --- the
left subgraph gets the fraction of population that satisfies the tighter
constraint, which may not be exactly the target.

The same 5\,\% area swing sensitivity test confirms this interpretation: running
FL, IL, MI, NY, PA, TX with a 5\,\% balance tolerance (well above the
constitutional standard) still fails because the subgraph population imbalance
is architectural.
A complete fix requires one of:
(a) applying the dual constraint at \emph{every} recursive level, not just
    the first, so each subgraph also targets its own proportional area share;
(b) post-processing with a population-balancing cell-swap algorithm that
    adjusts boundaries after the tree is constructed; or
(c) using per-subgraph population balance (ideal = subgraph\_pop / k) instead
    of the national ideal.
Option (c) is simplest and would likely resolve most failures.
We defer this to future work.

\paragraph{Experimental vs constitutional balance tolerance.}
The 44-state sweep used a 1.5\,\% balance tolerance, which exceeds the
0.5\,\% standard established by \emph{Karcher v. Daggett} (1983) for
congressional redistricting.
Results at the 0.5\,\% standard are reported as a sensitivity table in
Appendix A.
Preliminary results show [TBD: insert counts when sweep completes].
States where 0.5\,\% fails but 1.5\,\% succeeds have discrete-tract rounding
as the limiting factor; this is also fixable by option (c) above.

\paragraph{Area vs population correlation.}
The area constraint is an indirect measure of geographic diversity.
In states where the area-to-population ratio is approximately uniform
(agricultural Midwest), AreaSection behaves identically to GeoSection.
The constraint only has meaningful effect in states with strong
population concentration, which are precisely the states where urban peeling
is most problematic.

\paragraph{Sensitivity to area swing parameter.}
Table~\ref{tab:swing-sensitivity} reports the natural ratio and area fraction
achieved for three representative states across three values of
$\mathtt{ubvec}[1] = $ area swing.

\begin{table}[h]
\centering
\caption{Sensitivity to area swing parameter.
Natural ratio and area fraction at the winning split for 20 seeds per ratio.
OK = within target window; VIOLATED = METIS exceeded tolerance.}
\label{tab:swing-sensitivity}
\begin{tabular}{llrrr}
\toprule
State & Swing & Natural ratio & Area & Status \\
\midrule
WI ($p^* = 0.93$) & 1.05 ($\pm 5\,\%$) & 4:4 & 50.9\,\% & OK \\
WI & 1.10 ($\pm 10\,\%$) & 4:4 & 53.9\,\% & OK \\
WI & 1.20 ($\pm 20\,\%$) & 2:6 & 41.3\,\% & VIOLATED \\
\midrule
GA ($p^* = 0.91$) & 1.05 ($\pm 5\,\%$) & 7:7 & 52.1\,\% & OK \\
GA & 1.10 ($\pm 10\,\%$) & 3:11 & 54.9\,\% & OK \\
GA & 1.20 ($\pm 20\,\%$) & 3:11 & 59.2\,\% & OK \\
\midrule
NC ($p^* = 0.75$) & 1.05 ($\pm 5\,\%$) & 6:8 & 52.2\,\% & OK \\
NC & 1.10 ($\pm 10\,\%$) & 6:8 & 52.2\,\% & OK \\
NC & 1.20 ($\pm 20\,\%$) & 7:7 & 40.8\,\% & VIOLATED \\
\bottomrule
\end{tabular}
\end{table}

The sensitivity results reveal that 1.10 is not an arbitrary choice but sits
at a \emph{regime boundary} in both directions:

\begin{itemize}
\item At 1.05 (tight): Georgia is forced to a 7:7 equal split.
  The Atlanta metropolitan area, which occupies $\sim$8\,\% of Georgia's land,
  cannot satisfy the $\pm 5\,\%$ window at any ratio below 6:8; the algorithm
  is pushed to a near-equal bisection.
  While this may be desirable for anti-peeling purposes, it may also be
  overly aggressive for states where asymmetric geography is genuine.

\item At 1.20 (loose): Wisconsin reverts to a 2:6 ratio
  (area=41.3\,\%, VIOLATED — just outside the $\pm 20\,\%$ window) because the
  Milwaukee area's compact boundary is cheaper than the equal geographic split.
  The area constraint has become too permissive to prevent peeling.

\item At 1.10 (current): both WI and GA produce non-trivial but non-extreme splits
  (4:4 and 3:11 respectively) with area fractions within tolerance.
  NC is stable at 6:8 for both 1.05 and 1.10.
\end{itemize}

The 1.10 default is the tightest value that produces non-trivial, area-compliant
splits for all three tested states.
Tighter values (below $\sim 1.07$) risk forcing unnecessarily equal splits for
geographically concentrated states; looser values (above $\sim 1.15$) risk
re-enabling peeling for uniform states.

\paragraph{Area measure.}
We use TIGER ALAND (land area only, excluding water).
ALAND + AWATER would change results for coastal and Great Lakes states
(NY, MI, WI) by attributing large water areas to coastal tracts.
ALAND is the more legally defensible choice: water is not habitable territory
and should not count toward a district's geographic footprint.

\subsection{Comparison to Polsby-Popper and Other Compactness Metrics}

Standard compactness metrics (Polsby-Popper, Reock, convex-hull ratio) evaluate
the shape of individual districts after the fact.
AreaSection is an \emph{algorithmic} constraint applied during bisection:
it shapes the structure of the bisection tree before any individual district
boundaries are drawn.

These are complementary, not competing, approaches.
AreaSection produces bisection trees that respect geographic balance;
Polsby-Popper audits verify that individual districts within those trees
are not pathologically non-compact.
A complete redistricting analysis would apply both.
