\documentclass[12pt]{article}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{array}
\usepackage{multirow}
\usepackage{enumitem}
\usepackage{threeparttable}
\usepackage{url}

\onehalfspacing

\title{From Algorithm to Map: Implementation Case Studies\\
for Three Adoption Pathways}

\author{Giovanni Della-Libera\\
\small{Working Paper}}

\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Algorithmic redistricting has three realistic adoption pathways: through an
independent commission, through ordinary statute, and through court-ordered
remedial relief.  Each pathway places different institutional actors in the
role of map adopter, raises different procedural questions, and invites
different legal challenges.  This paper develops concrete case studies for
each pathway using the \texttt{redist} algorithmic redistricting system.
Case Study 1 examines the Arizona Independent Redistricting Commission (AIRC),
which already employs politically balanced membership and community-of-interest
hearings; we analyze how the commission could integrate algorithm output into
its existing workflow while preserving its adjustment authority.  Case Study 2
examines the California Citizens Redistricting Commission (CRC), whose
statutory criteria include proportionality and county contiguity requirements
that map directly onto \texttt{redist}'s VRASection and county-sticky edge
weights.  Case Study 3 examines North Carolina post-\textit{Harper v.\ Hall},
where the state supreme court's authority to appoint a special master and order
a remedial map creates conditions under which algorithm output could constitute
the remedy absent a demonstrated constitutional defect.  A practical workflow
table identifies who runs which command, at which stage, under what oversight.
We conclude with a candid assessment of the challenges---data preparation
burden, public transparency requirements, and challenge processes---that any
adopting jurisdiction will face.
\end{abstract}

\noindent\textbf{Keywords:} redistricting, independent commission, court
remedy, special master, Arizona, California, North Carolina, implementation

\section{Introduction}
\input{sections/01-introduction}

\section{Case Study 1 --- Arizona AIRC (Commission Pathway)}
\input{sections/02-arizona}

\section{Case Study 2 --- California CRC (Statute Pathway)}
\input{sections/03-california}

\section{Case Study 3 --- North Carolina Post-\textit{Harper} (Court-Order Pathway)}
\input{sections/04-north-carolina}

\section{Practical Workflow}
\input{sections/05-workflow}

\section{Challenges and Solutions}
\input{sections/06-challenges}

\section{Conclusion}
\input{sections/07-conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
