% Methodology

This section describes our computational implementation of overlapping party-specific districts for all 50 states, the comparison baselines we use to evaluate the system, and the metrics we employ to assess proportionality, compactness, and representation quality.

\subsection{Algorithm Implementation}

We implement the overlapping party-district system using recursive bisection on census tract adjacency graphs. The complete workflow proceeds in three stages:

\textbf{Stage 1: Data preparation}

For each state, we construct a graph where nodes represent census tracts and edges connect spatially adjacent tracts \citep{author2024recursive}. Node weights encode population (from 2020 Census); edge weights reflect spatial proximity (inverse distance between tract centroids). We compute statewide vote shares using the 2020 presidential election, aggregating county-level results \citep{leip2020}. This provides party vote shares $v_D$ (Democratic), $v_R$ (Republican), and $v_O$ (other parties).

\textbf{Stage 2: Proportional seat allocation}

Using the D'Hondt method (Section 2.1), we allocate each state's seats to parties based on vote shares. For a state with $N$ seats:

\begin{equation}
s_i = \text{seats allocated to party } i \text{ via D'Hondt with threshold } \frac{1}{N+1}
\end{equation}

This produces seat counts $s_D$, $s_R$, $s_O$ such that $s_D + s_R + s_O = N$. Parties below the threshold receive $s_i = 0$.

\textbf{Stage 3: Independent party-specific redistricting}

For each party $i$ with $s_i \geq 2$ seats, we run recursive bisection to create $s_i$ districts:

\begin{algorithm}[H]
\caption{Party-Specific Recursive Bisection}
\begin{algorithmic}[1]
\State \textbf{Input:} State graph $G$, target seats $s_i$ for party $i$
\State \textbf{Output:} Set of $s_i$ districts covering entire state
\State Initialize: Single district = entire state
\While{number of districts $< s_i$}
    \State Select largest district $d$ by population
    \State Partition $d$ into $d_1$ and $d_2$ using METIS bisection \citep{karypis1998metis}
    \State Objective: minimize edge cut, balance population
\EndWhile
\State \textbf{Return:} $s_i$ districts
\end{algorithmic}
\end{algorithm}

For parties with $s_i = 1$, the entire state constitutes their single district (no subdivision). For parties with $s_i = 0$, no districts are created.

This process runs \emph{independently} for each party. Democratic districts are created without knowledge of Republican districts and vice versa. The independence ensures that one party cannot manipulate boundaries to disadvantage another party—each controls only its own map.

\textbf{Computational complexity}

For a state with $N$ seats split between $k$ parties, the algorithm performs $k$ independent recursive bisection runs. Each run takes $O(m \log m)$ time where $m$ is the number of census tracts \citep{karypis1998metis}. For the entire United States (50 states, ~74,000 tracts), the complete computation finishes in under 4 hours on a standard workstation (Intel i7, 32GB RAM).

\subsection{Comparison Baselines}

We compare our overlapping party-district system to two baselines:

\textbf{Baseline 1: Enacted Districts}

The current congressional maps used in the 2022 midterm elections. These districts were drawn by state legislatures (partisan control), independent commissions (neutral bodies), or courts (litigation outcomes). Enacted districts reflect real-world redistricting with all its political pressures, legal constraints, and strategic manipulation. They serve as the status quo comparator.

\textbf{Baseline 2: Neutral Algorithmic Districts}

Districts generated via recursive bisection without partisan input, identical to the method in \citet{author2024recursive}. These represent the ``control'' condition: purely geometric optimization with no consideration of partisan outcomes. Comparing party-specific districts to neutral algorithmic districts isolates the effect of party-based subdivision from general algorithmic properties.

\textbf{Our System: Party-Specific Overlapping Districts}

The overlapping maps produced by our three-stage algorithm. Each state has multiple overlapping district maps (one per party above the threshold). These achieve perfect proportionality by design while maintaining geometric optimization within each party's map.

\subsection{Metrics}

We evaluate the three systems across four dimensions:

\textbf{Proportionality}

We measure partisan balance using the efficiency gap \citep{stephanopoulos2015partisan}:

\begin{equation}
\text{EG} = \frac{\text{Wasted votes}_D - \text{Wasted votes}_R}{\text{Total votes}}
\end{equation}

where wasted votes are (1) votes for losing candidates and (2) votes beyond 50\% for winning candidates. An efficiency gap of 0 indicates perfect proportionality; positive values favor Democrats, negative values favor Republicans. For enacted and neutral districts, we compute EG from simulated elections. For our system, EG is guaranteed to be 0 because seats are allocated proportionally before districts are drawn.

\textbf{Compactness}

We use the Polsby-Popper score \citep{polsby1991third}:

\begin{equation}
\text{PP} = \frac{4\pi \cdot \text{Area}}{\text{Perimeter}^2}
\end{equation}

Higher PP indicates more compact districts (circle = 1.0; elongated shapes approach 0). We compute mean and median PP across all districts within each party's map and compare to enacted and neutral baselines.

\textbf{Overlap Complexity}

For overlapping districts, we measure:
\begin{itemize}
    \item Average number of districts per geographic point: $\bar{d} = \frac{1}{|P|}\sum_{p \in P} |\text{districts containing } p|$
    \item Maximum overlap depth: $\max_p |\text{districts containing } p|$
\end{itemize}

where $P$ is the set of census tracts. In a two-party system with no minor parties exceeding thresholds, $\bar{d} = 2$ (every point in exactly one Democratic and one Republican district).

\textbf{Voter Representation}

We compute the percentage of voters with at least one co-partisan representative:

\begin{equation}
\text{RepCoverage} = \frac{|\{\text{voters with co-partisan rep}\}|}{|\{\text{all voters}\}|}
\end{equation}

For enacted and neutral districts, this depends on geographic distribution: voters in minority-party districts have no co-partisan representation. For our system, RepCoverage = 100\% by construction, as every party-affiliated voter has a representative from their party (assuming their party exceeded the threshold).

\textbf{Data sources}

All metrics are computed using publicly available data: Census Bureau population and geography (2020 decennial census), Dave Leip's Atlas of U.S. Presidential Elections (2020 county-level results), and TIGER/Line shapefiles for spatial boundaries. Enacted district shapefiles come from state redistricting authorities and are available via the Census Bureau's redistricting data program.

This methodology enables a controlled comparison: the same underlying geography, population distribution, and partisan vote shares are used across all three systems. Differences in outcomes reflect the districting system itself, not data artifacts or measurement choices.
