% Data and Methodology

\subsection{Census Data}

We use decennial census data from three redistricting cycles:

\begin{table}[h]
\centering
\begin{tabular}{lrrr}
\toprule
\textbf{Year} & \textbf{Census Tracts} & \textbf{Total Population} & \textbf{Districts} \\
\midrule
2000 & 66,304 & 281,421,906 & 435 \\
2010 & 74,002 & 308,745,538 & 435 \\
2020 & 85,331 & 331,449,281 & 435 \\
\bottomrule
\end{tabular}
\caption{Census data summary across three decades}
\end{table}

Census tracts provide the finest-grained units with stable boundaries suitable for temporal comparison. Population data comes from the decennial census PL 94-171 redistricting files.

\textbf{Apportionment}: Congressional seats are allocated among states using the Huntington-Hill method:
\begin{itemize}
    \item 2000: Baseline allocation (435 seats across 50 states)
    \item 2010: 8 states gained, 10 states lost seats
    \item 2020: 6 states gained, 7 states lost seats
\end{itemize}

\subsection{Algorithmic Redistricting}

We apply recursive bisection using METIS 5.1.0 with \textit{identical parameters} across all three years:

\begin{itemize}
    \item \textbf{Algorithm}: Recursive bisection (graph partitioning)
    \item \textbf{Population balance}: $\pm0.5\%$ of target
    \item \textbf{Edge weights}: $\alpha = 5.0$ for minority-minority edges (maintaining communities of interest)
    \item \textbf{Contiguity}: Enforced via graph connectivity
    \item \textbf{No partisan data}: Algorithm has no access to voter registration or election results
\end{itemize}

\textbf{Why consistency matters}: Using identical parameters isolates temporal effects (population change, geographic shifts) from methodological variations. Any differences in outcomes across years reflect genuine demographic trends, not algorithmic artifacts.

\subsection{Metrics}

\subsubsection{Compactness}

We measure compactness using the Polsby-Popper score:
\[
PP = \frac{4\pi \cdot \text{Area}}{\text{Perimeter}^2}
\]
PP ranges from 0 (infinitely elongated) to 1 (perfect circle). We compute PP for all 435 districts in each year.

\subsubsection{Geographic Stability}

To measure boundary stability across census cycles, we compute Intersection-over-Union (IoU):
\[
\text{IoU}(D_t, D_{t+10}) = \frac{|D_t \cap D_{t+10}|}{|D_t \cup D_{t+10}|}
\]
where $D_t$ is a district in year $t$ and $D_{t+10}$ is the geographically closest district 10 years later. IoU ranges from 0 (no overlap) to 1 (identical boundaries).

\subsubsection{Enacted Comparison}

We compare algorithmic districts to enacted districts using shapefiles from Dave's Redistricting App (DRA), which provides standardized enacted district boundaries for all three census cycles.

\subsection{Statistical Analysis}

We perform paired t-tests to assess compactness changes within states across years, ANOVA to compare commission vs non-commission states, and correlation analysis to examine relationships between population growth and district characteristics.
