\section{Formal Properties of the Flip Chain}
\label{sec:formal}

\subsection{Algorithm Specification}

Let $G = (V, E)$ be the census-tract adjacency graph for a state, with
$|V| = n$ tracts.
A redistricting plan $\pi : V \to [k]$ is valid if every district is
connected and population-balanced (within $\epsilon = 0.005$ of the target).
Let $P_{\mathrm{ideal}} = \sum_{v} \mathrm{pop}(v) / k$.

The deterministic seed schedule for the \flip{} chain uses the following
derivation:
\[
  \mathrm{chain\_seed}(b, 0)
  = \mathrm{SHA256}\!\left(\texttt{"FLIP\_CHAIN\_"} \,\|\, 0^{\mathrm{le64}}
    \,\|\, \texttt{"\_"} \,\|\, b^{\mathrm{le64}}\right)[0{:}8]
    \;\text{as } u64,
\]
where $b$ is the base seed, $\cdot^{\mathrm{le64}}$ denotes little-endian
8-byte encoding, and $[0{:}8]$ takes the first 8 bytes.

\begin{algorithm}
\caption{\flip{} Chain}
\label{alg:flip}
\begin{algorithmic}[1]
\Require Initial plan $\pi_0$, steps $F$, percentile $p \in [0,1]$,
         base seed $b$
\Ensure  Selected plan $\pi^*$
\State $\pi \gets \pi_0$
\State $\mathrm{visited} \gets [(\ec(\pi_0),\, \pi_0)]$
\State $\mathrm{rng} \gets \mathrm{SmallRng::seed\_from\_u64}(\mathrm{chain\_seed}(b, 0))$
\For{$i \gets 1$ \textbf{to} $F$}
  \State $\mathcal{B} \gets \{v \in V : \exists\, u \in N(v),\; \pi(u) \neq \pi(v)\}$
         \Comment{boundary tracts}
  \If{$\mathcal{B} = \emptyset$} \textbf{break} \EndIf
  \State $t \gets \mathrm{rng.choose}(\mathcal{B})$
         \Comment{uniform sample from boundary tracts}
  \State $\mathcal{D}_t \gets \{\pi(u) : u \in N(t),\; \pi(u) \neq \pi(t)\}$
         \Comment{adjacent districts}
  \State $d^* \gets \mathrm{rng.choose}(\mathcal{D}_t)$
         \Comment{proposed target district}
  \State $\pi' \gets \pi$ with $\pi'(t) \gets d^*$
  \State $A \gets \pi^{-1}(\pi(t)) \setminus \{t\}$,\quad $B \gets \pi^{-1}(d^*) \cup \{t\}$
  \If{$G[A]$ is connected
     \textbf{ and } $|P(A) - P_{\mathrm{ideal}}| \leq \epsilon \cdot P_{\mathrm{ideal}}$
     \textbf{ and } $|P(B) - P_{\mathrm{ideal}}| \leq \epsilon \cdot P_{\mathrm{ideal}}$}
    \State $\pi \gets \pi'$
    \State $\mathrm{visited}.\mathrm{push}((\ec(\pi),\, \pi))$
  \EndIf
\EndFor
\State Sort $\mathrm{visited}$ by $\ec$ ascending
\State \Return $\mathrm{visited}[\lfloor p \cdot |\mathrm{visited}|\rfloor]$
\end{algorithmic}
\end{algorithm}

Three design choices in Algorithm~\ref{alg:flip} deserve comment.
First, only the district \emph{losing} tract $t$ (the donor, $d_A =
\pi^{-1}(\pi(t))$) must be re-checked for contiguity.
Since $\pi_0$ is a valid plan, the subgraph induced by $d_A \setminus
\{t\}$ may be disconnected if $t$ was a cut vertex of $G[d_A]$; the
contiguity BFS checks this explicitly.
By contrast, the district \emph{gaining} $t$ (the recipient, $d_B =
\pi^{-1}(d^*)$) is guaranteed to remain connected: $t$ is adjacent to
some $u \in d_B$ by the boundary-tract selection rule
($d^* \in \mathcal{D}_t$), and $d_B$ is connected in the current plan,
so $d_B \cup \{t\}$ is connected without any additional check.
Second, the population check is symmetric: both the donor district (now
$n/k - 1$ tracts lighter) and the recipient district (now $n/k + 1$ tracts
heavier) must satisfy the balance tolerance.
Third, every accepted plan is stored (not just the final plan), so the
$p$-percentile selection examines the full trajectory.

\paragraph{CLI invocation.}
\begin{verbatim}
redist state --state NC --year 2020 \
    --search flip --flip-steps 10000 --percentile 0.0
\end{verbatim}

\begin{remark}[Target distribution]
\label{rem:target-dist}
The hard-constraint \flip{} chain (accepting only proposals satisfying
population balance and contiguity) is not reversible in general with
respect to the uniform distribution over valid plans: the proposal
probabilities $Q(\pi, \pi')$ and $Q(\pi', \pi)$ differ because the set
of valid flip proposals from $\pi$ need not equal the set from $\pi'$.
The visited sequence used for sensitivity analysis is a random walk
trace, not an approximate sample from the stationary distribution.
This distinguishes \flip{} from \recom{}, which is designed as a Markov
chain sampler targeting a specific distribution~\citep{deford2021}.
\end{remark}

\subsection{Time Complexity}

\begin{proposition}[$O(1)$ amortised step cost]
\label{prop:o1}
Each step of \flip{} runs in $O(\deg_{\max})$ time if the boundary tract
set $\mathcal{B}$ is maintained incrementally, and $O(n/k)$ worst-case for
the contiguity BFS.
\end{proposition}

\begin{proof}
The boundary tract set $\mathcal{B}$ changes by at most $1 + \deg(t)$
elements per accepted step: tract $t$ exits $\mathcal{B}$, and its
neighbours may enter or leave depending on whether they still border a
different district.
Hence $\mathcal{B}$ can be maintained in $O(\deg_{\max})$ amortised time
per step.
The contiguity BFS for district $A = \pi^{-1}(\pi(t)) \setminus \{t\}$
visits at most $|A| \leq n/k$ tracts.
In practice, for census-tract graphs with average degree $\approx 6$,
the BFS terminates quickly because most tracts are not on the boundary
of the district.
\end{proof}

Compare to \recom{}, which samples a random spanning tree over $A \cup B$
(a merged district of size $\approx 2n/k$) at each step, requiring
$O(n/k \log n/k)$ time~\citep{deford2021}.
\flip{} is therefore $O(\log n/k)$ times cheaper per step, and in practice
the constant factor is large enough that \flip{} runs at 10$\times$--50$\times$
the raw throughput of \recom{}.

\subsection{Acceptance Rate}

\begin{proposition}[High acceptance rate]
\label{prop:accept}
In our experiments on census-tract graphs, the \flip{} acceptance rate
is approximately $40$--$70\%$, substantially higher than \recom{} ($5$--$30\%$).
\end{proposition}

The intuition is straightforward: boundary tracts were selected precisely
because they lie at the boundary of two districts, so they tend to be close
to the average population per district.
A single-tract move therefore rarely violates the balance constraint.
Contiguity failure is equally rare: if tract $t$ is a boundary tract adjacent
to district $d^*$, the district $A$ losing $t$ typically remains connected
because $t$ was on the periphery of $A$, not a bottleneck.
The rare failure case is when $t$ is an \emph{isthmus} of $A$---the unique
tract connecting two parts of the district.

\recom{}, by contrast, proposes merging two districts and resampling a
spanning tree cut; finding a balanced spanning tree cut requires a precise
population split, which fails often for large districts~\citep{deford2021}.

\subsection{Path-Length Lower Bound}

Despite high acceptance rates, \flip{} chains require many steps to traverse
significant distances in plan space.

\begin{proposition}[Path-length lower bound on plan-space exploration]
\label{thm:slow-mix}
Let $\pi_0$ be an initial redistricting plan with $k$ districts, each
containing $m = n/k$ tracts on average.
For any target plan $\pi^*$ that differs from $\pi_0$ in $\ell$ boundary
tracts (where $\ell$ may be as large as $O(n/k)$), the expected number of
\flip{} steps to reach $\pi^*$ from $\pi_0$ is at least $\Omega(\ell)$.
\end{proposition}

\begin{proof}
Each \flip{} step changes exactly one tract assignment.
If $\pi_0$ and $\pi^*$ differ in $\ell$ tracts, any path from $\pi_0$ to
$\pi^*$ in the \flip{} transition graph must contain at least $\ell$ accepted
steps.
The expected number of steps (including rejections) to make $\ell$ accepted
moves is $\ell / p_{\mathrm{acc}} = \Omega(\ell)$.
This is a \emph{path-length} lower bound: it bounds the minimum number of
steps needed to reach a specific target plan, not the mixing time of the
chain in the $L^1$ sense.
\end{proof}

\begin{remark}
The stronger claim---that the $L^1$ mixing time of the \flip{} chain is
exponential in $n$ in the worst case---is not proved here.
Such exponential lower bounds for local Markov chains on combinatorial
structures are established in~\citet{sinclair1992} via conductance arguments,
and similar behaviour is conjectured for redistricting chains on realistic
census-tract graphs.
However, since \flip{} is not run for ensemble sampling in our framework
(we use it only for local sensitivity analysis starting from a fixed plan),
the $L^1$ mixing time is not directly relevant to our application; the
path-length bound above is the operative limit on \flip{}'s exploratory reach.
\end{remark}

\subsection{Reachability: A Strict Subset of the Plan Space}
\label{subsec:reachability}

A fundamental limitation of \flip{} for ensemble sampling is that not all
valid plans are reachable from a given starting plan via single-tract flips.

\begin{theorem}[Reachability gap]
\label{thm:reachability}
Let $\Pi$ denote the set of all valid $k$-district plans of $G$.
For any initial plan $\pi_0 \in \Pi$, the set of plans reachable from
$\pi_0$ via \flip{} moves is a strict subset of $\Pi$.
\end{theorem}

\begin{proof}[Proof sketch]
Consider a 1D path graph $G = v_1 - v_2 - \cdots - v_n$ partitioned into
$k$ contiguous intervals.
The only valid \flip{} moves shift a boundary between two adjacent intervals
by one vertex.
The plan $\pi_0$ assigning $v_1, \ldots, v_{n/2}$ to district 1 and
$v_{n/2+1}, \ldots, v_n$ to district 2 cannot be reached from the plan
$\pi^*$ assigning $v_1$ to district 2 and $v_2, \ldots, v_n$ to district 1
in one step.
In general, for planar graphs with $k \geq 3$ districts, some plans require
simultaneous multi-tract moves to maintain contiguity---for example, when
district boundaries intersect in a ``corridor'' configuration where the only
path from $\pi_0$ to $\pi^*$ passes through an invalid intermediate plan
(a disconnected district).
Such configurations are endemic in real census-tract graphs due to the
geographic complexity of state boundaries, rivers, and urban cores.

\medskip
\noindent\textbf{Planar example ($k = 3$).}
We exhibit a concrete planar instance with $k = 3$ districts.
Consider a $3 \times 3$ grid graph with tracts labelled $0$--$8$ in
row-major order (tract $i$ occupies row $\lfloor i/3 \rfloor$, column
$i \bmod 3$).
Define the initial plan $\pi_0$ as: district~1 $= \{0, 3, 6\}$ (left
column), district~2 $= \{1, 4, 7\}$ (middle column), district~3 $=
\{2, 5, 8\}$ (right column).
Now consider the target plan $\pi_1$: district~1 $= \{0, 1, 3, 4\}$,
district~2 $= \{2, 5\}$, district~3 $= \{6, 7, 8\}$.
Moving from $\pi_0$ to $\pi_1$ requires simultaneously moving tracts~1
and~4 from district~2 to district~1, and moving tract~3 from district~1
to district~2.
No single-tract flip achieves this without an invalid intermediate
state: for example, flipping tract~4 from district~2 to district~1
before flipping tract~1 leaves district~2 as $\{1, 7\}$, which is
disconnected (tract~1 and tract~7 share no edge once tract~4 is
removed).
Therefore $\pi_1 \notin \mathcal{R}(\pi_0)$ via boundary flips.
\end{proof}

This reachability gap is the decisive reason \flip{} cannot replace \recom{}
for ensemble sampling.
\recom{} merges two districts and resamples the spanning tree, which can
produce qualitatively different district configurations in a single step.
\flip{} cannot bridge topological obstacles that require simultaneous
multi-tract moves.
