\documentclass[sigconf,review]{acmart}

\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{threeparttable}

\title{Edge-Weighted Recursive Bisection for Compact Congressional Districts}

\author{Giovanni Della-Libera}

\begin{document}

\begin{abstract}
Congressional redistricting requires balancing population equality with geographic compactness, but traditional graph partitioning approaches treat all adjacencies equally, ignoring the physical geometry of boundaries. We introduce \textbf{edge-weighted recursive bisection}, where edge weights represent boundary lengths between census tracts, enabling METIS to minimize total district perimeters during recursive partitioning. This geometric weighting produces a 56\% improvement in compactness (Polsby-Popper) over unweighted baselines and exceeds enacted 2020 districts by 20\% across all 50 states (435 districts). Our approach handles complex geographic features including water crossings, islands, and point adjacencies, while maintaining strict population balance ($\pm 0.5\%$) and contiguity guarantees. Critically, edge weighting uses no political data---compactness emerges from geometric optimization without partisan input. However, we show through partisan metrics analysis that geographic compactness does not guarantee partisan fairness: algorithmically compact districts exhibit partisan effects due to geographic sorting of voters, though typically smaller than enacted gerrymandered plans. Results show particularly strong improvements in gerrymandered states (Illinois +174\%, Louisiana +104\%), demonstrating that boundary length minimization produces compact districts that resist intentional elongation. This work advances graph partitioning for redistricting by incorporating geometric information at national scale.
\end{abstract}

\keywords{redistricting, graph partitioning, METIS, compactness, edge weighting}

\maketitle

\input{sections/introduction}
\input{sections/related-work}
\input{sections/methodology}
\input{sections/results}
\input{sections/discussion}
\input{sections/conclusion}

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
