\section{The Partisan Distribution of Ensemble Plans}
\label{sec:bell-curve}

\subsection{The Bell Curve Around Geography}

The central insight of ensemble redistricting research is that, holding
geography fixed, the distribution of partisan seat outcomes across valid plans
is approximately bell-shaped.  The mean of this distribution is determined
by geography: the spatial concentration of partisan voters.  The width is
determined by the flexibility of the districting plan space: how many valid
plans exist, and how diverse their partisan outcomes are.

More formally, let $S_D(P)$ denote the number of Democratic-leaning districts
in plan $P$ under a fixed electoral baseline.  For a \recom{} ensemble
$\mathcal{E}$ on a state with $k$ seats:

\begin{equation}
  S_D \sim \text{approximately } \mathrm{Binomial}(k, p_{\text{geo}})
  \text{ with underdispersion in sorted states},
  \label{eq:partisan-dist}
\end{equation}

where $p_{\text{geo}}$ is the geographically determined probability that
a random valid district leans Democratic.  For highly competitive states
(Wisconsin, Pennsylvania), $p_{\text{geo}} \approx 0.50$.  For sorted
states (Georgia), $p_{\text{geo}} \approx 0.36$.

The actual standard deviation $\sigma_{S_D}$ is typically \emph{less} than
the pure binomial prediction $\sqrt{k p_{\rm geo}(1-p_{\rm geo})}$ in sorted
states.  For Georgia ($k=14$, $p_{\rm geo} \approx 0.36$), the binomial
predicts $\sigma \approx 1.80$ seats, but the observed ensemble SD is
$\sigma_{S_D} \approx 1.2$ seats---an \emph{underdispersed} distribution.
This occurs because contiguity and population balance constraints prevent
arbitrary rearrangement of districts in a geographically sorted state,
narrowing the effective plan space relative to an unconstrained binomial.
The Rodden sorting mechanism thus reduces partisan variance rather than
increasing it.

The distribution is approximately symmetric around the geographic mean for
competitive states and skewed for sorted states.  Crucially, the distribution
is \emph{not} centered on proportionality: in sorted states, even a perfectly
random plan tends to produce more Republican seats than the statewide vote
share would imply.

\subsection{Width of the Distribution}

The standard deviation $\sigma_{S_D}$ of the partisan seat distribution
depends on:

\begin{itemize}
  \item $k$ (more districts $\to$ wider distribution in absolute terms, narrower
    in relative terms);
  \item Geographic compactness of the state (more irregular states $\to$
    wider distribution);
  \item Degree of partisan sorting (highly sorted states $\to$ narrower
    distribution near the geographic mean).
\end{itemize}

Empirically, $\sigma_{S_D} \approx 1.0$--$1.5$ seats for $k = 8$--$17$
and $\sigma_{S_D} \approx 2.0$--$3.5$ for $k = 38$--$52$.  A difference
of two seats from the ensemble median is thus within one to two standard
deviations for most states.

\subsection{What the Distribution Does Not Show}

The partisan distribution of ensemble plans does \emph{not} show:
\begin{itemize}
  \item The ``correct'' partisan outcome for the state;
  \item What proportionality requires;
  \item Whether any particular plan within the distribution is
    constitutionally required.
\end{itemize}

It shows what outcomes are geographically plausible.  A plan at the 99th
percentile is implausible without intent.  A plan at the 54th percentile
is typical.  Whether typical is fair is a separate normative question,
addressed in the context of proportionality in Section~\ref{sec:corridor}.
