\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}[proposition]
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}

\title{Preserving Political Subdivisions in Algorithmic Redistricting:\\
       Governmental Hierarchy as Edge-Weight Signal}

\author{Gio Della-Libera}

\date{Draft --- \today}

\begin{document}

\maketitle

\begin{abstract}
Most state redistricting statutes list ``preserving political subdivisions''
as an explicit criterion alongside compactness and population equality.
Yet no algorithmic redistricting system operationalizes this criterion
as a formal, tunable parameter.
We introduce \emph{subdivision-respecting redistricting}: a single real-valued
stickiness parameter $\alpha$ that makes intra-county edges more expensive
to cut in the METIS graph bisection framework, with a natural extension to
any level of the governmental hierarchy (MCDs, places, voting precincts).
When $\alpha = 0$ the algorithm reduces to standard minimum-edge-cut bisection;
as $\alpha \to \infty$ the algorithm only splits a county when population
balance requires it.
County affiliation is freely available from Census TIGER tract GEOIDs ---
no additional data download.
We prove that the optimal weighted min-cut partition has weakly fewer county
splits as $\alpha$ increases and characterise the Pareto frontier empirically
via 50-state sweep plus 25-seed ensemble analysis of seven focal states.
At $\alpha = 5$, county splits fall by 45\% across 44 states (14.3 $\to$ 7.8 per state)
at a 2.5$\times$ increase in total boundary length.
The Pareto frontier has a characteristic knee near $\alpha \approx 2$: most
split reduction occurs cheaply; beyond $\alpha = 5$ only structurally unavoidable
splits remain.
A 25-seed ensemble across seven focal states confirms partisan neutrality:
mean Democratic seat change $= +0.18$ (not significant, $p > 0.05$ for all states).
The stickiness parameter is a single published number verifiable from public TIGER data.
\end{abstract}

\input{sections/01-introduction}
\input{sections/02-related-work}
\input{sections/03-methodology}
\input{sections/04-evaluation}
\input{sections/05-discussion}
\input{sections/06-conclusion}

\bibliographystyle{plainnat}
\bibliography{references,../../references}

\end{document}
