\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}[proposition]
\newtheorem{remark}{Remark}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{array}
\usepackage{listings}

\title{StabilitySection: Cross-Census Stability of\\
       GeoSection-Optimal Redistricting Maps}

\author{Gio Della-Libera}

\date{Draft --- \today}

\begin{document}

\maketitle

\begin{abstract}
A redistricting algorithm is \emph{census-stable} for a state if it produces
the same natural partition---the same first-level split ratio, the same
bisection tree structure, and the same partisan seat distribution---across
multiple decennial census cycles.
We introduce the \textbf{Census Stability Score} (CSS), a composite metric
measuring GeoSection's cross-census robustness for each state, and present the
StabilitySection framework for computing it.

Cross-census stability has a precise legal significance that within-year seed
stability (B.7) cannot provide.
If GeoSection produces the same optimal map for a state using 2000, 2010,
and 2020 population data---twenty years of demographic change---then the map is
\emph{not a product of any particular census cycle's partisan distribution}.
It is the geographic structure of the state, expressed consistently across
three independent decennial snapshots.
A challenger cannot argue the map was drawn to exploit the current partisan
distribution: the current census year is merely one of three that converged
to the same answer.

We present complete 2020 GeoSection results for 44 multi-district states
and the CSS framework with full methodology.
For 2000 and 2010, the cross-year sweep is in progress; we report
preliminary ratio-stability findings for the available subset of states and
provide \textbf{[TBD: 2010 sweep pending]} placeholders for the
three-census CSS.
The 2000 vs.\ 2020 comparison---a twenty-year span---is already a strong
empirical anchor.
States expected to have CSS $\geq 0.90$: the Midwest row states (IA, KS, NE,
MO, IN) and New England (ME, NH, VT).
States expected to be structurally volatile: the Sun Belt growth corridor
(TX, AZ, GA, FL, NV), where suburban expansion has reorganised the
population-geography relationship between census cycles.
\end{abstract}

\input{sections/01-introduction}
\input{sections/02-related-work}
\input{sections/03-methodology}
\input{sections/04-evaluation}
\input{sections/05-discussion}
\input{sections/06-conclusion}

\bibliographystyle{plainnat}
\bibliography{references,../../references}

\end{document}
