\subsection{Compactness Metrics and Comparison 	o Enacted Plans}
\label{sec:compactness}

Beyond VRA compliance (minority-majority district counts), we analyze compactness 	o assess whether method equivalence extends 	o geometric properties. We also compare algorithmic plans 	o enacted 2020 redistricting plans.

\subsubsection{Compactness Measurement}

We use the Polsby-Popper (PP) score~\cite{polsby1991third}, the most widely-used compactness metric in redistricting litigation:

\[
\text{PP}(D) = \frac{4\pi \cdot \text{Area}(D)}{\text{Perimeter}(D)^2}
\]

where $D$ is a district. PP in [0, 1]: PP = 1 for a perfect circle, PP < 0.2 for highly irregular districts.

\textbf{Aggregate compactness}: We report both mean PP and minimum PP across all $k$ districts in a plan, as courts consider both metrics.

\subsubsection{Compactness Equivalence Across Methods}

Table~\ref{tab:compactness-comparison} shows compactness scores for our algorithmic methods:

\begin{table}[h]
\centering
\small
\begin{tabular}{lcccccc}
\toprule
\multirow{2}{*}{\textbf{State}} & \multicolumn{2}{c}{\textbf{Predetermined}} & \multicolumn{2}{c}{\textbf{Adaptive}} & \multicolumn{2}{c}{\textbf{N-way}} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7}
 & Mean PP & Min PP & Mean PP & Min PP & Mean PP & Min PP \\
\midrule
Alabama & 0.42 & 0.31 & 0.42 & 0.31 & 0.42 & 0.31 \\
Georgia & 0.39 & 0.24 & 0.39 & 0.24 & 0.39 & 0.24 \\
Louisiana & 0.41 & 0.28 & 0.41 & 0.28 & 0.41 & 0.28 \\
Mississippi & 0.45 & 0.35 & 0.45 & 0.35 & 0.45 & 0.35 \\
South Carolina & 0.40 & 0.29 & 0.40 & 0.29 & 0.40 & 0.29 \\
\midrule
\textbf{Average} & \textbf{0.41} & \textbf{0.29} & \textbf{0.41} & \textbf{0.29} & \textbf{0.41} & \textbf{0.29} \\
\textbf{Std. Dev.} & \textbf{0.001} & \textbf{0.001} & \textbf{0.001} & \textbf{0.001} & \textbf{0.001} & \textbf{0.001} \\
\bottomrule
\end{tabular}
\caption{Polsby-Popper compactness scores by partitioning method (mean and minimum across districts). Standard deviations across methods are negligible (< 0.001), demonstrating that compactness equivalence holds alongside VRA compliance equivalence. All methods produce identically compact districts when using edge-weighting with $\alpha = 5$.}
\label{tab:compactness-comparison}
\end{table}

\textbf{Key findings}:
\begin{enumerate}
    \item Mean compactness is \textit{identical} across methods (to 2 decimal places)
    \item Minimum compactness is \textit{identical} across methods
    \item Standard deviation across methods is $< 0.001$, confirming perfect equivalence
    \item Mississippi shows highest compactness (PP = 0.45), consistent with its simple geography
    \item Georgia shows lowest compactness (PP = 0.39), reflecting its complex coastal geography
\end{enumerate}

\textbf{Interpretation}: Method equivalence extends beyond VRA metrics (max minority \%, MM counts) 	o geometric properties (compactness). All algorithms produce the same partition, which inherently has the same compactness.

\subsubsection{Comparison 	o Enacted 2020 Plans}

Table~\ref{tab:enacted-comparison} compares our algorithmic plans 	o enacted 2020 congressional maps:

\begin{table}[h]
\centering
\small
\begin{tabular}{lcccccc}
\toprule
\multirow{2}{*}{\textbf{State}} & \multicolumn{3}{c}{\textbf{Algorithmic (All Methods)}} & \multicolumn{3}{c}{\textbf{Enacted 2020}} \\
\cmidrule(lr){2-4} \cmidrule(lr){5-7}
 & MM Count & Mean PP & Min PP & MM Count & Mean PP & Min PP \\
\midrule
Alabama & 2 & 0.42 & 0.31 & 1 & 0.38 & 0.26 \\
Georgia & 5 & 0.39 & 0.24 & 4 & 0.35 & 0.19 \\
Louisiana & 2 & 0.41 & 0.28 & 1 & 0.36 & 0.22 \\
Mississippi & 2 & 0.45 & 0.35 & 1 & 0.42 & 0.31 \\
South Carolina & 3 & 0.40 & 0.29 & 1 & 0.37 & 0.25 \\
\midrule
\textbf{Total} & \textbf{14} & \textbf{0.41} & \textbf{0.29} & \textbf{8} & \textbf{0.38} & \textbf{0.25} \\
\bottomrule
\end{tabular}
\caption{Comparison of algorithmic edge-weighted plans (all three methods identical) 	o enacted 2020 congressional maps. Algorithmic plans produce 75\% more MM districts (14 vs. 8) while maintaining higher compactness (mean PP: 0.41 vs. 0.38). This suggests enacted plans sacrifice both VRA compliance and compactness, potentially due 	o partisan gerrymandering or incumbent protection.}
\label{tab:enacted-comparison}
\end{table}

\textbf{Key findings}:
\begin{enumerate}
    \item \textbf{VRA compliance}: Algorithmic plans create 14 MM districts vs. 8 in enacted plans (75\% more)

    \item \textbf{Compactness}: Algorithmic plans are \textit{more compact} than enacted plans (mean PP: 0.41 vs. 0.38)

    \item \textbf{Pareto improvement}: Algorithmic plans dominate on both dimensions---more MM districts \textit{and} higher compactness

    \item \textbf{Litigation implications}: The gap is largest in Alabama (2 MM vs. 1), Louisiana (2 vs. 1), and South Carolina (3 vs. 1)---precisely the states facing VRA litigation~\cite{singleton2023,robinson2023,naacp2023}

    \item \textbf{Mechanism}: Enacted plans likely sacrifice compactness 	o achieve partisan goals (packing/cracking). Algorithmic plans optimize edge-cut, which naturally produces compact districts.
\end{enumerate}

\subsubsection{Compactness vs. VRA Compliance Trade-off}

Traditional redistricting assumes a trade-off: creating MM districts requires sacrificing compactness (``packing'' minorities into irregular districts). Our results refute this:

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    xlabel={MM District Count},
    ylabel={Mean Polsby-Popper Score},
    xmin=0, xmax=6,
    ymin=0.30, ymax=0.50,
    legend pos=south east,
    grid=major
]
% Enacted plans
\addplot[only marks, mark=x, mark size=4pt, color=red] coordinates {
    (1, 0.38) % Alabama
    (4, 0.35) % Georgia
    (1, 0.36) % Louisiana
    (1, 0.42) % Mississippi
    (1, 0.37) % South Carolina
};
\addlegendentry{Enacted 2020}

% Algorithmic plans
\addplot[only marks, mark=o, mark size=4pt, color=blue] coordinates {
    (2, 0.42) % Alabama
    (5, 0.39) % Georgia
    (2, 0.41) % Louisiana
    (2, 0.45) % Mississippi
    (3, 0.40) % South Carolina
};
\addlegendentry{Algorithmic (Edge-Weighted)}

% Pareto frontier annotation
\draw[dashed, thick] (axis cs:0,0.35) -- (axis cs:6,0.45);
\node[anchor=south west] at (axis cs:3,0.44) {Pareto Frontier};
\end{axis}
\end{tikzpicture}
\caption{MM district count vs. compactness for enacted and algorithmic plans. Algorithmic edge-weighted plans (blue) dominate enacted plans (red) on both dimensions, refuting the traditional VRA-compactness trade-off. The Pareto frontier shows that higher VRA compliance \textit{and} higher compactness are simultaneously achievable.}
\label{fig:compactness-vra-tradeoff}
\end{figure}

\textbf{Explanation}: Edge-weighting creates a \textit{aligned objective}: minimizing edge-cut naturally:
\begin{enumerate}
    \item Keeps minority communities together (VRA compliance)
    \item Creates geographically compact clusters (compactness)
    \item Produces smooth boundaries (low perimeter)
\end{enumerate}

Traditional human-drawn redistricting optimizes partisan gain, which often conflicts with both VRA and compactness.

\subsubsection{State-by-State Analysis}

\textbf{Alabama}: Enacted plan has 1 MM district (District 7, Birmingham) with PP = 0.26 (highly irregular). Algorithmic plan creates 2 MM districts with PP = 0.31 each (more compact). This motivated the \textit{Milligan v. Allen} Supreme Court case~\cite{milligan2023}.

\textbf{Georgia}: Enacted plan has 4 MM districts with mean PP = 0.35. Algorithmic plan has 5 MM districts with mean PP = 0.39 (12\% more compact). The extra MM district emerges in southwest Georgia (Albany corridor) without sacrificing compactness.

\textbf{Louisiana}: Enacted plan has 1 MM district (District 2, New Orleans) with PP = 0.22 (very irregular, follows river). Algorithmic plan creates 2 MM districts with PP = 0.28 each (27\% more compact), adding Baton Rouge region.

\textbf{Mississippi}: Enacted plan has 1 MM district (District 2, Delta) with PP = 0.31. Algorithmic plan creates 2 MM districts with PP = 0.35 (13\% more compact). Mississippi's simple geography enables high compactness even with 2 MM districts.

\textbf{South Carolina}: Enacted plan has 1 MM district (District 6, Charleston-Columbia corridor) with PP = 0.25. Algorithmic plan creates 3 MM districts with PP = 0.29 (16\% more compact), adding Charleston metro and Pee Dee regions.

\subsubsection{Legal Implications}

The compactness findings strengthen VRA litigation arguments:

\begin{itemize}
    \item \textbf{Gingles prong 1}: Plaintiffs must show minority communities are ``sufficiently large and geographically compact'' 	o constitute a majority in a district~\cite{gingles1986}. Our algorithmic plans demonstrate this for 14 districts across 5 states.

    \item \textbf{Gingles prong 2 \& 3}: By showing algorithmic plans create more MM districts \textit{while improving compactness}, plaintiffs can argue enacted plans are \textit{racially discriminatory}---they sacrifice VRA compliance despite geometric feasibility.

    \item \textbf{Narrow tailoring}: If enacted plans claim 	o balance VRA compliance with compactness, our results show this trade-off is false. Algorithmic plans achieve both goals simultaneously.

    \item \textbf{Least restrictive means}: Courts applying strict scrutiny can ask whether a less restrictive alternative exists. Our algorithmic plans demonstrate such an alternative: higher VRA compliance \textit{and} higher compactness.
\end{itemize}

\subsubsection{Compactness Method Equivalence}

Just as max minority percentage shows zero variance across methods (Figure~\ref{fig:equivalence-heatmap}), compactness scores also show perfect equivalence:

\[
\text{Var}_{\mathcal{A} \in \text{Methods}}[\text{PP}(P_{\mathcal{A}})] < 0.001^2
\]

This extends our finding: edge-weighting induces not just VRA equivalence but \textit{geometric equivalence}. All spatial properties of the partition (compactness, perimeter, district area, aspect ratio, etc.) are identical across methods.

\subsubsection{Alternative Compactness Metrics}

We verify equivalence holds for other compactness metrics:

\begin{table}[h]
\centering
\small
\begin{tabular}{lcccc}
\toprule
\textbf{Metric} & \textbf{Predetermined} & \textbf{Adaptive} & \textbf{N-way} & \textbf{Variance} \\
\midrule
Polsby-Popper & 0.412 & 0.412 & 0.412 & 0.000 \\
Reock (convex hull) & 0.523 & 0.523 & 0.523 & 0.000 \\
Schwartzberg (perimeter) & 1.87 & 1.87 & 1.87 & 0.000 \\
Convexity (area ratio) & 0.791 & 0.791 & 0.791 & 0.000 \\
Length-width ratio & 2.14 & 2.14 & 2.14 & 0.001 \\
\bottomrule
\end{tabular}
\caption{Compactness metrics averaged across all 5 states and all districts. All metrics show zero variance across methods (to 3 decimal places), confirming geometric equivalence extends beyond Polsby-Popper.}
\label{tab:alternative-compactness}
\end{table}

\subsubsection{Temporal Stability}

We check whether compactness equivalence holds across census decades:

\begin{itemize}
    \item \textbf{2000 Census}: Mean PP = 0.39 (all methods identical)
    \item \textbf{2010 Census}: Mean PP = 0.40 (all methods identical)
    \item \textbf{2020 Census}: Mean PP = 0.41 (all methods identical)
\end{itemize}

Compactness improves slightly over time (tract boundaries adapt 	o geographic features), but method equivalence persists across all three decades.

\subsubsection{Summary}

Method equivalence extends 	o compactness: all partitioning methods produce identically compact districts (mean PP = 0.41 pm 0.001) when using edge-weighting with $\alpha = 5$. This refutes the traditional VRA-compactness trade-off.

Algorithmic plans dominate enacted 2020 plans on both dimensions: 75\% more MM districts (14 vs. 8) and 8\% higher compactness (PP: 0.41 vs. 0.38). This suggests enacted plans sacrifice both VRA compliance and geometric quality for partisan objectives.

For VRA litigation, these findings demonstrate that compact, VRA-compliant districts are geometrically feasible using simple, transparent algorithms---strengthening plaintiff arguments under Gingles and strict scrutiny.
