\documentclass[11pt,twocolumn]{article}

% Packages
\usepackage[margin=0.75in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{url}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}

% Formatting
\setlength{\columnseprule}{0.4pt}
\captionsetup{font=small,labelfont=bf}

% Title and Author
\title{\textbf{Algorithmic Congressional Redistricting via\\Edge-Weighted Recursive Bisection}}

\author{
Giovanni Della-Libera\\
\textit{Working Paper}\\
January 2026
}

\date{}

\begin{document}

\maketitle

\begin{abstract}
Congressional redistricting in the United States faces persistent challenges from gerrymandering and public distrust. While the Huntington-Hill method resolved seat apportionment disputes through mathematical objectivity, boundary drawing remains subject to partisan manipulation. We present an algorithmic approach using edge-weighted recursive bisection—a graph partitioning method that directly minimizes district perimeter by incorporating actual boundary lengths as edge weights. Applied to all 435 congressional districts across 50 states using 2020 Census data, our approach achieves a mean Polsby-Popper compactness score of 0.367—a 56\% improvement over unweighted baseline recursive bisection (0.235) and 20\% better than enacted 2020 congressional districts (0.305). Cross-census validation using 2010 data demonstrates algorithmic consistency (0.320 vs. 0.353, only 10.3\% variation) despite vastly different political environments, providing strong evidence that geographic factors, not political opportunities, drive algorithm performance. The 50\% reduction in the gap between algorithmic and enacted districts from 2010 to 2020 quantifies the effectiveness of redistricting reforms. We demonstrate that 37 of 50 states exceed enacted district compactness, with particularly strong performance in partisan-drawn states (Illinois +174\%, Louisiana +104\%, New Hampshire +102\%). Our results show that algorithmic redistricting using edge-weighted graph partitioning can match or exceed human mapmaking in compactness while providing algorithmic transparency, reproducible baselines for measuring gerrymandering, and eliminating opportunities for partisan manipulation.
\end{abstract}

% Sections
\input{sections/01_introduction}
\input{sections/02_background}
\input{sections/03_methodology}
\input{sections/04_implementation}
\input{sections/05_results}
\input{sections/06_analysis}
\input{sections/07_discussion}
\input{sections/08_conclusion}

\section*{Acknowledgments}
We thank the U.S. Census Bureau for providing comprehensive redistricting data, the MIT Election Data + Science Lab for election results, and the Census Bureau's TIGER/Line program for enacted district boundaries. This research was conducted using the METIS graph partitioning library developed by George Karypis and Vipin Kumar.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
