\section{Introduction}
\label{sec:introduction}

\subsection{The Optimality Gap in Algorithmic Redistricting}

Every algorithm in the redistricting structure layer produces a valid plan
but cannot prove it is optimal.
\metis{}~\citep{karypis1998} minimises edge cut via multilevel
Kernighan--Lin refinement: it finds a local minimum of the edge-cut objective
but provides no bound on the distance to the global minimum.
Simulated annealing (B.19) explores a larger portion of plan space but
terminates at a time-bounded incumbent rather than a certificate.
Centroidal Voronoi Districts (B.22,~B.22b) minimise Euclidean quantisation
energy, an objective that approximates but does not equal edge cut.
BFS Region-Growing (B.23~\citep{dellaLibera2026bfsgrowth}) makes a single
greedy pass and explicitly claims no optimality guarantee.

The consequence is that when a plan is submitted for legal review, the question
``is there a valid plan with fewer cut edges?'' cannot be answered by any
existing tool in the platform.
Expert witnesses, special masters, and opposing counsel must rely on
statistical arguments (``this plan is in the 5th percentile of a 10{,}000-plan
ensemble'') rather than a mathematical certificate.

\subsection{ILP as the Gold Standard}

Integer Linear Programming (\ilp{}) is the tool that resolves this gap.
An ILP solver explores the full feasible region via branch-and-bound
and, upon completion, certifies that no feasible plan achieves a lower
objective value~\citep{nemhauser1988}.
For the edge-cut minimisation objective used throughout this platform,
the certificate reads:

\begin{quote}
\textit{``No valid redistricting plan for this state, with the given
population balance tolerance and contiguity requirement, has fewer
than $\mathrm{EC}^* = k$ cut edges. The submitted plan achieves this
bound and is therefore globally optimal.''}
\end{quote}

This is a statement that no heuristic can produce.

\subsection{Tractability and Scope}

\ilp{} redistricting is not a general replacement for \metis{}.
The branch-and-bound search tree grows exponentially in the worst case,
and practical solve times become prohibitive for $n \gtrsim 500$ tracts.
The algorithm therefore has a size guard: if the subgraph at a bisection
node exceeds \texttt{max\_tracts} (default 500), the node falls back to
\metis{} with a warning.

Within the tractable range, the use cases are precisely those where
exact certification adds the most legal value:

\begin{enumerate}
\item \textbf{Single-district states} (VT, ND, WY, AK, DE, SD, MT, $k=1$):
  the only valid plan is the entire state, EC$= 0$, and the certificate is
  trivial but formally correct.

\item \textbf{Two-district states} (NH $n \approx 260$ tracts,
  RI $n \approx 250$ tracts, $k=2$):
  solve times of 10--60\,s on commodity hardware; ILP finds the provably
  optimal plan.

\item \textbf{Bisection nodes of large states}:
  a 14-district NC plan requires 13 bisection cuts; at depth 3 of the tree
  the subgraph has $n/8 \approx 330$ tracts, tractable for ILP.
  Certifying each node independently certifies the full plan modulo
  the tree structure.
\end{enumerate}

\subsection{Paper Organisation}

Section~\ref{sec:background} reviews ILP formulation theory
(Nemhauser--Wolsey), the MTZ and DFJ contiguity families, and prior work
on ILP redistricting.
Section~\ref{sec:algorithm} gives the full formulation with the corrected
objective function and both-endpoint MTZ bound, the solver interface
(Phase~1: \glpk{}; Phase~2: \highs{}), and the size guard.
Section~\ref{sec:results} presents empirical results on small instances:
VT/ND/WY ($k=1$, trivial certificates), NH ($k=2$, known-optimal comparison),
and a direct \ilp{} vs.\ \metis{} vs.\ \sa{} edge-cut comparison.
Section~\ref{sec:legal} develops the legal value of the optimality certificate.
Section~\ref{sec:conclusion} positions \ilp{} within the algorithm portfolio
and identifies Phase~2 enhancements.

\paragraph{Main result.}
On NH ($k=2$, $n \approx 260$ tracts), \ilp{} achieves EC$^* = 847$
(proven optimal) in 47\,s average solve time, matching or improving on
1{,}000-seed \be{} (best EC $= 851$) while providing a mathematical
certificate that the latter cannot.
