\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}[proposition]
\newtheorem{lemma}{Lemma}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}

\title{F.3: Resolution Selection for State Legislative Redistricting:\\
       When Census Tracts Are Too Coarse}

\author{Gio Della-Libera}

\date{Draft --- \today}

\begin{document}

\maketitle

\begin{abstract}
Census tracts, the standard geographic unit for congressional redistricting,
contain 1,200--8,000 residents each and are well-matched to districts of
700,000+ residents. State legislative districts, by contrast, may contain
as few as 3,500 residents (New Hampshire House), making the average tract
contain \emph{more} than one district. Redistricting at tract resolution
in these settings is geometrically degenerate: districts must be composed
of fractional tracts, population balance constraints cannot be met, and
compactness scores are artificially inflated by the coarseness of the
boundary set.
We introduce the resolution-selection rule $k/n_{\mathrm{tracts}} > 0.05$:
when the district-to-tract ratio exceeds $5\%$, we shift from census-tract
to census block-group resolution. Block groups (600--3,000 residents each)
restore the invariant that each district contains at least 20 base units,
enabling meaningful compactness optimisation and sub-$0.5\%$ population balance.
We validate this threshold empirically on Washington (98 districts),
Texas (150 districts), and California (80 districts), showing that
block-group resolution improves mean Polsby-Popper from $0.368$ to $0.398$
and reduces population deviation from $0.8\%$ to $0.4\%$ for the Washington
House. Runtime at block-group resolution is 3--4$\times$ slower than at
tract resolution due to the larger adjacency graph.
We supply a resolution recommendation table for all 50 states and all
lower chambers, and discuss the MAUP implications of resolution choice
for partisan and compactness outcomes.
\end{abstract}

\input{sections/01-introduction}
\input{sections/02-resolution-problem}
\input{sections/03-rule-of-thumb}
\input{sections/04-empirical-comparison}
\input{sections/05-maup-analysis}
\input{sections/06-runtime-analysis}
\input{sections/07-recommendation-table}
\input{sections/08-conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
