\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}[proposition]
\newtheorem{remark}{Remark}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algpseudocode}

\title{VRASection: Geographic Alignment Score for\\
       Minority Opportunity District Bisection}

\author{Gio Della-Libera}

\date{Draft --- \today}

\begin{document}

\maketitle

\begin{abstract}
Congressional redistricting algorithms that attempt to satisfy the Voting
Rights Act (\S 2) by adding minority population as an explicit METIS
constraint (``MetisVra'') fail in 3 of 5 majority-minority states due to
numerical scale mismatch: minority VAP fractions (0.0--1.0) and census
population counts (10,000--200,000) differ by 60--800$\times$, producing
infeasible or degenerate partitions \citep{b3paper}.

We introduce \textbf{VRASection}, which abandons explicit minority constraints
and instead exploits the geographic structure of \emph{Gingles Prong 1}:
legally cognizable minority communities are, by definition, sufficiently
large and geographically compact.
This geographic concentration is already encoded in the census-tract adjacency
graph; VRASection makes it measurable through a \emph{geographic alignment score}
$A(\text{split}) = |\text{MVAP}_\text{frac}(L) - \text{MVAP}_\text{frac}(R)|$
layered on top of GeoSection's isoperimetric ratio scan \citep{b8paper}.

The modified selection objective is:
\[
  \text{score}(\text{ratio}) = \mathrm{EC}_{\text{norm}} - w_\text{vra}
  \cdot A \cdot \max(\mathrm{EC}_{\text{norm}},\, 1)
\]
where $w_\text{vra} = 0.40$ by default.
At this weight, compactness is weighted 50\,\% more heavily than alignment,
ensuring geographic compactness remains the predominant criterion
under the \emph{Shaw v.\ Reno} / \emph{Miller v.\ Johnson} racial-predominance tests.
The alignment score uses only the spatial distribution of minority VAP ---
not racial identity itself --- satisfying Gingles Prong 1 without constituting
racial targeting.

Applied to Alabama (7 districts, 27\,\% Black VAP), VRASection selects
a $2:5$ first-level bisection, separating the Black Belt region from the
northern and coastal areas, compared to GeoSection's $1:6$ urban peel
(Birmingham isolation).
The $2:5$ split more effectively concentrates minority population on one
side for downstream minority-opportunity-district formation.
Full comparison runs across 6 historically significant VRA states are
pending; Alabama's $2:5$ result provides the lead case study.

VRASection requires no user-specified minority-fraction targets.
It passes the MetisVra failure modes by design: there is no second METIS
constraint and no scale mismatch.
We analyse the algorithm's legal robustness, sensitivity to $w_\text{vra}$,
and relationship to the \emph{Callais v.\ Landry} (2025) evidence standard.
\end{abstract}

\input{sections/01-introduction}
\input{sections/02-related-work}
\input{sections/03-algorithm}
\input{sections/04-evaluation}
\input{sections/05-discussion}
\input{sections/06-conclusion}

\bibliographystyle{plainnat}
\bibliography{references,../../references}

\end{document}
