\section{BFS Growth as MCMC Warm-Start}
\label{sec:warmstart}

\subsection{Motivation}

MCMC redistricting chains---\recom{}, Forest ReCom, and their variants---
require an initial feasible plan from which the chain begins.
The quality of this warm-start affects two aspects of chain behaviour:
\begin{enumerate}
\item \textbf{Burn-in length.} If the initial plan is far from the
  chain's stationary distribution, many steps are needed before the
  chain's samples are representative. This means results from early
  steps must be discarded as burn-in.
\item \textbf{Autocorrelation.} If the initial plan is atypical
  (unusually high or low EC, unusual partisan structure), the chain
  may exhibit high autocorrelation for longer, reducing the effective
  sample size.
\end{enumerate}

The default warm-start in the B-series platform is a single \metis{} call.
\metis{} produces a low-EC plan that may be an outlier in the ensemble
distribution---most valid redistricting plans are not minimum-EC plans.
\bfsg{} produces a higher-EC, more geographically typical plan that may
lie closer to the centre of the ensemble distribution.

\subsection{Theoretical Argument}

Let $\pi$ denote the stationary distribution of the \recom{} chain over
valid redistricting plans of a given state.
The typical plan under $\pi$ is characterised by the ensemble's modal
EC and PP values, which are substantially higher than the METIS minimum
and somewhat below the CVD or BFS Growth values.

If the initial plan $x_0$ has EC close to the ensemble mode, the chain
needs fewer steps to mix because each ReCom merge-split proposal is more
likely to produce a plan similar to $x_0$---and similarly valued by $\pi$.
If $x_0$ is a METIS minimum-EC outlier, early proposals that increase EC
(toward the mode) are accepted, but the chain moves slowly because it
starts at a corner of the distribution.

\bfsg{} plans have higher EC and more geographically typical boundaries
than METIS plans.
If the ensemble mode's EC is between the METIS minimum and the BFS Growth
value, then the BFS Growth warm-start is closer to the mode in EC-space,
potentially reducing burn-in.

\subsection{Empirical Results}

We ran Forest \recom{} for 500 steps on NC $k=14$ (2020, geographic
weights) starting from two warm-starts:
\begin{itemize}
  \item[(a)] Standard \metis{} plan ($\EC \approx 3{,}124$, seed $s=42$).
  \item[(b)] \bfsg{} plan ($\EC \approx 3{,}418$, seed $s=42$).
\end{itemize}

\paragraph{NC ($k=14$) results.}
The \bfsg{} warm-start begins with higher EC but reaches the chain's
stationary EC distribution after approximately 80 more steps than the
\metis{} warm-start, which starts closer to the chain's modal EC.
The autocorrelation of the EC statistic at lag 50 is $\rho_{50} = 0.44$
for the \bfsg{} warm-start versus $\rho_{50} = 0.38$ for the \metis{}
warm-start.
This difference is consistent with \bfsg{} not improving convergence speed
in this NC configuration.

The ensemble EC mode for NC 2020 under geographic weights is approximately
$2{,}400$--$2{,}600$ (from B.7 500-seed sweep data~\citep{b7paper}).
\metis{} single-seed EC is $2{,}891$ (Table~\ref{tab:main-comparison}),
above the mode, while \bfsg{} EC of $3{,}418$ is further above it.
Neither warm-start is at the ensemble mode; \metis{} is closer.

\paragraph{Summary.}
\metis{} remains the recommended default warm-start for NC.
\bfsg{} may be preferable when plan diversity is more valuable than fast
convergence---for example, when running short chains to sample from a wide
geographic spread of initial conditions rather than to converge to the
stationary distribution.

\subsection{Autocorrelation Decay}

Let $f: \mathcal{P} \to \mathbb{R}$ be a plan statistic (e.g., EC, mean PP,
partisan seat count).
The autocorrelation at lag $\ell$ is:
\[
  \rho_f(\ell) = \frac{\mathrm{Cov}(f(x_t),\; f(x_{t+\ell}))}
                      {\mathrm{Var}(f(x_t))},
\]
where $(x_t)$ is the Markov chain.
The \emph{integrated autocorrelation time} $\tau_f = \sum_{\ell=1}^\infty \rho_f(\ell)$
quantifies how many steps are needed for effectively independent samples.

A smaller $\tau_f$ from a BFS Growth warm-start relative to a METIS
warm-start would indicate faster mixing when starting from the BFS plan.
The NC experiment (Section above) tested this with the EC statistic:
the lag-50 autocorrelation was $0.44$ (BFS Growth) vs.\ $0.38$ (METIS),
showing that BFS Growth does not reduce autocorrelation relative to METIS
on NC.
The theoretical condition for BFS Growth to win is:

\begin{quote}
\textit{For states where the BFS Growth plan's EC is closer to the
ensemble mode than the METIS plan's EC, the ReCom chain warm-started
from BFS Growth has smaller $\tau_f$ for EC-correlated statistics.}
\end{quote}

This condition does not hold for NC: the METIS plan ($\EC \approx 3{,}124$)
is closer to the ensemble mode ($2{,}400$--$2{,}600$) than the BFS Growth
plan ($\EC \approx 3{,}418$).
Future work may identify states or weight configurations where the
condition is met and BFS Growth warm-starts accelerate convergence.

\subsection{Practical Recommendation}

Given the current analysis, the practical warm-start recommendation is:

\begin{itemize}
\item \textbf{Default: \metis{} warm-start.}
  \metis{} is better understood, and its EC properties are well-documented.
  Use \metis{} as the default warm-start until BFS Growth warm-starts
  are empirically validated.

\item \textbf{Alternative: \bfsg{} warm-start for geographic analysis.}
  When the research question concerns geographic compactness or the
  spatial structure of valid plans, a BFS Growth warm-start provides a
  more geographically typical initial plan.
  The BFS Growth plan's higher EC means the chain starts with more
  boundary-swap flexibility, potentially exploring a wider range of
  plans early.

\item \textbf{Multi-start: both.}
  For rigorous ensemble analysis, run two parallel chains from both
  warm-starts and compare their convergence diagnostics.
  Agreement between the two chains' stationary-phase distributions
  confirms mixing; disagreement signals that longer chains are needed.
\end{itemize}

\subsection{Implementation}

To use \bfsg{} as a warm-start for a \recom{} ensemble:

\begin{verbatim}
# Step 1: Generate BFS Growth initial plan
redist state --state NC --year 2020 \
  --structure bfs-growth --search single

# Step 2: Use the output plan as ReCom warm-start
redist ensemble --state NC --year 2020 \
  --initial-plan runs/nc_bfs/2020/nc_plan.json \
  --chain recom --steps 10000
\end{verbatim}

The \texttt{--initial-plan} flag accepts any previously generated plan
in the platform's JSON format, including BFS Growth plans.
The audit chain (Section~3 of the spec) records the \texttt{base\_seed}
and \texttt{bfs\_seed} in \texttt{runs/{label}/{year}/index.json},
enabling full reproducibility of the warm-start.
