% Section 3: Methodology
% Compactness-VRA Tradeoff Quantification

\section{Methodology}

We conduct comprehensive experiments across five VRA-covered states to quantify the compactness-VRA tradeoff and identify Pareto-optimal redistricting configurations. Our methodology consists of four components: (1) compactness and VRA metrics, (2) redistricting algorithms, (3) experimental design, and (4) Pareto frontier identification. All code and data are available in our public repository to ensure reproducibility.

\subsection{Compactness Metrics}

We evaluate district compactness using four complementary metrics that capture both graph-theoretic and geometric properties of district shapes.

\subsubsection{Edge Cut (Graph-Theoretic)}

\textbf{Definition}: The edge cut is the number of census tract adjacencies crossing district boundaries in the dual graph representation.

\textbf{Formal Definition}: Given a partition $\mathcal{P} = \{D_1, D_2, \ldots, D_k\}$ of $n$ census tracts into $k$ districts and an adjacency graph $G = (V, E)$ where $V$ represents tracts and $E$ represents geographic adjacencies, the edge cut is:
\[
\text{EdgeCut}(\mathcal{P}) = |\{(u, v) \in E : \mathcal{P}(u) \neq \mathcal{P}(v)\}|
\]
where $\mathcal{P}(u)$ denotes the district assignment of tract $u$.

\textbf{Interpretation}: Lower edge cut indicates more compact districts, as fewer tract boundaries are severed. METIS directly minimizes edge cut during graph partitioning, making this our primary optimization objective. Edge cut ranges from 0 (perfect clustering, impossible with population balance) to $|E|$ (every edge crosses districts).

\textbf{Weighted Edge Cut}: When using edge-weighted optimization, we compute both unweighted edge cut (counting each cut edge equally) and weighted edge cut (summing edge weights of cut edges). The weighted edge cut represents METIS's optimization objective, while unweighted edge cut measures compactness under traditional definitions.

\subsubsection{Polsby-Popper Score (Geometric)}

\textbf{Definition}: The Polsby-Popper score measures district compactness as the ratio of district area to the area of a circle with the same perimeter.

\textbf{Formal Definition}: For a district $D$ with area $A_D$ and perimeter $P_D$:
\[
\text{PP}(D) = \frac{4\pi A_D}{P_D^2}
\]

\textbf{Interpretation}: Polsby-Popper ranges from 0 (infinitely elongated) to 1 (perfect circle). Higher values indicate more compact, circular shapes. This is the most widely used compactness metric in redistricting litigation and has been adopted by several state constitutions. We report both district-level Polsby-Popper and state-level average across all districts.

\textbf{Implementation}: We compute district geometries by dissolving census tract boundaries within each district using GeoPandas \texttt{unary\_union}, then calculate area and perimeter from the resulting polygons.

\subsubsection{Reock Score (Geometric)}

\textbf{Definition}: The Reock score measures the ratio of district area to the area of its minimum bounding circle.

\textbf{Formal Definition}: For a district $D$ with area $A_D$ and minimum bounding circle with area $A_C$:
\[
\text{Reock}(D) = \frac{A_D}{A_C}
\]

\textbf{Interpretation}: Reock ranges from 0 to 1, with higher values indicating more compact districts. The minimum bounding circle is the smallest circle that completely contains the district. Reock is more sensitive to outlying areas than Polsby-Popper, as a single protruding arm can substantially increase the bounding circle.

\textbf{Implementation}: We approximate the minimum bounding circle using the district's minimum rotated rectangle (computationally efficient and highly correlated with true minimum bounding circle, $r > 0.95$ in pilot tests).

\subsubsection{Convex Hull Ratio (Geometric)}

\textbf{Definition}: The convex hull ratio measures how much a district deviates from convexity.

\textbf{Formal Definition}: For a district $D$ with area $A_D$ and convex hull with area $A_H$:
\[
\text{ConvexHull}(D) = \frac{A_D}{A_H}
\]

\textbf{Interpretation}: Ranges from 0 to 1, with 1 indicating a perfectly convex shape. Non-convex districts (with indentations or concave sections) score lower. This metric captures different geometric properties than Polsby-Popper or Reock---a district can be non-circular (low PP) but still convex (high convex hull ratio).

\textbf{Metric Complementarity}: These four metrics capture distinct aspects of compactness: edge cut measures graph connectivity, Polsby-Popper measures roundness, Reock measures spread, and convex hull ratio measures convexity. High correlation among metrics ($r > 0.7$, see Results) validates their consistency, while modest correlation ($r < 0.85$) confirms they provide complementary information.

\subsection{VRA Compliance Metrics}

We evaluate Voting Rights Act compliance using three metrics that measure minority representation and concentration.

\subsubsection{Majority-Minority (MM) District Count}

\textbf{Definition}: A district is classified as majority-minority (MM) if its minority population exceeds 50\% of the total district population.

\textbf{Formal Definition}: For a district $D$ with total population $P_D$ and minority population $M_D$:
\[
\text{IsMM}(D) = \begin{cases}
1 & \text{if } M_D / P_D \geq 0.50 \\
0 & \text{otherwise}
\end{cases}
\]
\[
\text{MMCount}(\mathcal{P}) = \sum_{D \in \mathcal{P}} \text{IsMM}(D)
\]

\textbf{Interpretation}: MM district count is our primary VRA compliance metric. Courts have traditionally required 50\% as the threshold for ``majority'' status, though recent cases have explored ``coalition districts'' at lower thresholds (40-45\%). We use the conservative 50\% threshold for consistency with legal precedent.

\textbf{Target MM Counts}: Each state has a target MM count based on minority population proportion and VRA litigation history:
\begin{itemize}
    \item Alabama: 2/7 districts (28.6\%)
    \item Georgia: 5/14 districts (35.7\%)
    \item Louisiana: 2/6 districts (33.3\%)
    \item Mississippi: 2/4 districts (50.0\%)
    \item South Carolina: 3/7 districts (42.9\%)
\end{itemize}

A configuration ``succeeds'' if $\text{MMCount}(\mathcal{P}) \geq \text{Target}$.

\subsubsection{Total Population vs Voting-Eligible Population Standards}

A critical methodological question for VRA compliance is: \emph{What population base should define the 50\% majority threshold---total population, voting-age population (VAP), or citizen voting-age population (CVAP)?}

\textbf{Legal Standard}: The Supreme Court's *Evenwel v. Abbott* (2016) clarified that congressional districts must be equal in \emph{total population} under the Equal Protection Clause's "one person, one vote" principle. However, the VRA's concern is with \emph{voting power}, not just population representation, which depends on VAP or CVAP.

\textbf{Why This Matters}: Due to differential age structures and citizenship rates between demographic groups, a district with 50\% minority \emph{total population} may have substantially less than 50\% minority \emph{voting-eligible population}:

\begin{enumerate}
    \item \textbf{Age Structure Effects (VAP)}: Minority populations in our study states tend to be younger on average than white populations. Black populations have median ages of 32-35 years in Southern states, compared to 43-47 years for white populations. This means a higher fraction of minority populations are under age 18 and ineligible to vote. Typical adjustment: 50\% minority \emph{total population} $\approx$ 46-48\% minority VAP.

    \item \textbf{Citizenship Effects (CVAP)}: Hispanic and Asian populations include substantial non-citizen populations, while Black and white populations have citizenship rates near 98-99\%. For Black-majority districts in our study states, citizenship has minimal impact (VAP $\approx$ CVAP). However, for coalition districts combining multiple groups or for states with significant Hispanic populations, CVAP adjustments can be larger. Typical adjustment: 50\% minority \emph{VAP} $\approx$ 47-49\% minority CVAP for Black-majority districts, but as low as 40-45\% for Hispanic-plurality districts.
\end{enumerate}

\textbf{Our Approach}: We use \emph{total population} as the primary standard for three reasons:

\begin{enumerate}
    \item \textbf{Legal consistency}: Total population is the basis for "one person, one vote" and redistricting litigation. Courts evaluate district equality using total population, making it the appropriate denominator for VRA analysis.

    \item \textbf{Data availability}: Census P.L. 94-171 redistricting data provides tract-level total population by race/ethnicity. CVAP data is only available from ACS 5-year estimates, with larger margins of error and less geographic detail (often aggregated to county or larger geographies, not census tracts).

    \item \textbf{Conservative VRA standard}: Using total population as the 50\% threshold creates a \emph{higher bar} for MM classification than VAP or CVAP. If a district achieves 50\% minority total population, it likely has 46-49\% minority CVAP---approaching but not always exceeding electoral majority. This conservative approach ensures our findings about VRA-compactness tradeoffs represent genuine constraints, not artifacts of lenient thresholds.
\end{enumerate}

\textbf{Sensitivity Analysis}: In Section 4.5, we present a sensitivity analysis recalculating MM district counts under VAP and CVAP standards using published Census Bureau CVAP estimates. This analysis demonstrates that most districts classified as MM under total population standards retain MM status under CVAP (typically 85-95\% retention rate), though marginal districts near 50\% may lose MM classification. Critically, our central finding---that non-MM districts \emph{gain} compactness---is \emph{insensitive} to the population standard used, as non-MM districts remain non-MM under all standards.

\textbf{Future Work}: Ideal VRA redistricting would use tract-level CVAP projections adjusted for turnout propensity (Citizens of Voting Age $\times$ Registration Rate $\times$ Turnout Rate). However, such data requires linking decennial census, voter registration, and turnout records at fine geographic scales, which is beyond our current scope. Our total population approach provides a transparent, legally consistent baseline for comparing redistricting algorithms, with the understanding that true electoral power calculations require additional demographic adjustments.

\subsubsection{Maximum Minority Percentage}

\textbf{Definition}: The highest minority percentage achieved across all districts in a partition.

\textbf{Formal Definition}:
\[
\text{MaxMinority}(\mathcal{P}) = \max_{D \in \mathcal{P}} \left( \frac{M_D}{P_D} \right)
\]

\textbf{Interpretation}: Maximum minority percentage indicates the strongest level of minority concentration achieved. Values close to 50\% indicate configurations barely achieving MM status, while values substantially above 50\% (e.g., 60-70\%) indicate ``packed'' districts that may waste minority votes by concentrating them beyond the threshold needed for electoral control.

\subsubsection{Minority Concentration Index}

\textbf{Definition}: We compute a concentration index measuring how minority voters are distributed across districts.

\textbf{Formal Definition}: Using the Herfindahl-Hirschman Index (HHI) applied to minority population shares:
\[
\text{HHI}(\mathcal{P}) = \sum_{D \in \mathcal{P}} \left( \frac{M_D}{\sum_{D' \in \mathcal{P}} M_{D'}} \right)^2
\]

\textbf{Interpretation}: HHI ranges from $1/k$ (minority voters perfectly dispersed across $k$ districts) to 1 (all minority voters concentrated in one district). Higher HHI indicates greater concentration. This metric helps diagnose whether configurations achieve VRA compliance through efficient concentration (moderate HHI) or excessive packing (high HHI).

\subsection{Redistricting Algorithms}

We compare three algorithmic approaches to redistricting: baseline (no VRA optimization), multi-constraint optimization, and edge-weighted optimization.

\subsubsection{Graph Representation}

All algorithms operate on a graph representation of census tract adjacencies. We construct adjacency graphs using Queen contiguity (tracts sharing boundaries or vertices) with water-based bridge connections for disconnected islands. Each tract is a vertex weighted by population, and edges represent geographic adjacency.

\textbf{Adjacency Detection}: For each state, we:
\begin{enumerate}
    \item Load census tract geometries from TIGER/Line shapefiles
    \item Compute Queen contiguity using GeoPandas \texttt{sjoin} with predicate ``intersects''
    \item Add water-based bridges for tracts separated by <1km water distance
    \item Validate connectivity (ensure single connected component)
\end{enumerate}

The resulting adjacency graphs range from 1,323 tracts (South Carolina) to 2,796 tracts (Georgia), with average degree 6.1-6.2 (each tract borders ~6 neighbors).

\subsubsection{Baseline: Uniform METIS Partition}

\textbf{Algorithm}: Standard METIS recursive bisection with population-only vertex weights.

\textbf{Parameters}:
\begin{itemize}
    \item \texttt{vertex\_weights}: 1D array of tract populations (single objective)
    \item \texttt{nparts}: Target number of congressional districts per state
    \item \texttt{ufactor}: Population tolerance factor = 1.005 (±0.5\% of target)
    \item \texttt{niter}: 100 refinement iterations
    \item \texttt{edge\_weights}: None (all edges weighted equally)
    \item \texttt{tpwgts}: None (no demographic target weights)
\end{itemize}

\textbf{Objective}: Minimize unweighted edge cut subject to population balance constraints. This represents ``demographic-blind'' redistricting that optimizes only for compactness and equal population.

\textbf{Expected Performance}: Baseline may accidentally create MM districts when minority populations are highly concentrated (e.g., Mississippi: 46.1\% minority), but generally will not achieve VRA targets in states with moderate minority populations.

\subsubsection{Multi-Constraint Optimization (Comparison)}

\textbf{Algorithm}: METIS with multi-constraint optimization using \texttt{tpwgts} parameter to set explicit population targets per demographic group.

\textbf{Parameters}:
\begin{itemize}
    \item \texttt{vertex\_weights}: 2D array with columns [total population, minority population]
    \item \texttt{tpwgts}: Target weight fractions per district (e.g., [0.50, 0.50] for 50\% minority)
    \item \texttt{ubvec}: Balance tolerance vector [1.005, 1.20] (strict population, relaxed demographic)
\end{itemize}

\textbf{Objective}: Minimize edge cut while attempting to match target demographic compositions per district. The relaxed demographic tolerance (±20\%) allows flexibility when exact targets are infeasible.

\textbf{Limitations}: Multi-constraint rigidly constrains the solution space by requiring \emph{every} district to approximate target demographics. This often fails when minority populations cluster in specific regions (Alabama: multi-constraint achieves 0 MM districts despite targeting 2).

\textbf{Note}: Multi-constraint results are drawn from prior work (Paper 4). We include them for comparison but focus our experiments on edge-weighted optimization, which demonstrates superior performance.

\subsubsection{Edge-Weighted Optimization (Primary Approach)}

\textbf{Algorithm}: METIS with population-only vertex weights but \emph{weighted} edges that make minority-minority adjacencies expensive to cut.

\textbf{Edge Weight Function}: Given a minority threshold $\theta$ (e.g., 0.40, 0.45, 0.50) and weight factor $w$ (e.g., 5, 10, 100), we define:
\[
\text{EdgeWeight}(u, v) = \begin{cases}
w & \text{if } M_u / P_u \geq \theta \text{ and } M_v / P_v \geq \theta \\
1 & \text{otherwise}
\end{cases}
\]
where $M_u$ and $P_u$ are minority and total populations of tract $u$.

\textbf{Intuition}: Edges between ``minority tracts'' (above threshold $\theta$) receive weight $w$ (e.g., 5× or 10× higher than normal edges). METIS minimizes \emph{weighted} edge cut, making it expensive to separate minority tracts. This naturally clusters minority populations without rigid demographic constraints.

\textbf{Parameters}:
\begin{itemize}
    \item \texttt{vertex\_weights}: 1D array of tract populations (single objective, like baseline)
    \item \texttt{edge\_weights}: Dictionary mapping edge $(u,v)$ to weight (minority-minority edges weighted $w$, others weighted 1)
    \item \texttt{nparts}, \texttt{ufactor}, \texttt{niter}: Same as baseline
    \item \texttt{tpwgts}: None (no explicit demographic targets)
\end{itemize}

\textbf{Flexibility Advantage}: Unlike multi-constraint, edge-weighting does not force \emph{every} district to have specific demographics. Instead, it allows METIS to naturally form MM districts where geographically feasible (by keeping minority clusters together) while leaving other districts free to optimize for compactness without demographic constraints. This flexibility enables win-win solutions where both MM and non-MM districts can achieve high compactness.

\subsection{Experimental Design}

\subsubsection{State Selection}

We study five Southern states with substantial Black populations and histories of VRA coverage under Section 5 preclearance:
\begin{itemize}
    \item \textbf{Alabama} (AL): 36.9\% Black, 7 districts, 2 MM target
    \item \textbf{Georgia} (GA): 42.4\% Black, 14 districts, 5 MM target
    \item \textbf{Louisiana} (LA): 41.6\% Black, 6 districts, 2 MM target
    \item \textbf{Mississippi} (MS): 46.1\% Black, 4 districts, 2 MM target
    \item \textbf{South Carolina} (SC): 35.1\% Black, 7 districts, 3 MM target
\end{itemize}

These states provide demographic diversity (35.1\% to 46.1\% minority) and varying MM target ratios (28.6\% to 50.0\% of districts), enabling cross-state comparison of tradeoff patterns.

\textbf{Data Source}: We use 2020 Decennial Census tract-level population data (P.L. 94-171 Redistricting Data File) for population and race/ethnicity demographics. Tract geometries come from TIGER/Line shapefiles. All data are publicly available from the U.S. Census Bureau.

\subsubsection{Parameter Sweep Design}

For each state, we test 21 configurations: 1 baseline + 20 edge-weighted variations.

\textbf{Baseline Configuration}:
\begin{itemize}
    \item Weight factor: $w = 1$ (equivalent to unweighted)
    \item Threshold: N/A (no edge-weighting)
    \item Purpose: Establishes ground truth compactness without VRA optimization
\end{itemize}

\textbf{Edge-Weighted Parameter Sweep}:
\begin{itemize}
    \item \textbf{Weight factors}: $w \in \{1, 5, 10, 50, 100\}$
    \item \textbf{Thresholds}: $\theta \in \{0.40, 0.45, 0.50, 0.55\}$
    \item \textbf{Total}: $5 \times 4 = 20$ edge-weighted configurations per state
\end{itemize}

\textbf{Parameter Rationale}:
\begin{itemize}
    \item \textbf{$w = 1$}: Baseline equivalent (no preference for minority clustering)
    \item \textbf{$w = 5, 10$}: Moderate weighting (5-10× higher cost to cut minority-minority edges)
    \item \textbf{$w = 50, 100$}: Aggressive weighting (forces strong minority concentration)
    \item \textbf{$\theta = 0.40$}: Liberal threshold (includes tracts with 40-49\% minority)
    \item \textbf{$\theta = 0.50$}: Conservative threshold (only tracts with 50\%+ minority)
    \item \textbf{$\theta = 0.45, 0.55$}: Intermediate values
\end{itemize}

\textbf{Total Experimental Scale}:
\begin{itemize}
    \item States: 5
    \item Configurations per state: 21 (1 baseline + 20 edge-weighted)
    \item Total configurations: $5 \times 21 = 105$
    \item Districts analyzed: $\sim$840 individual districts across all configurations
    \item Tracts processed: 7,202 census tracts across all states
\end{itemize}

\subsubsection{South Carolina Extended Testing}

Due to South Carolina's failure to achieve its 3 MM target with standard parameters, we conduct extended testing with more aggressive parameters:
\begin{itemize}
    \item \textbf{Weight factors}: $w \in \{100, 200, 500, 1000\}$
    \item \textbf{Thresholds}: $\theta \in \{0.30, 0.35, 0.40, 0.45, 0.50\}$
    \item \textbf{Total}: $4 \times 5 = 20$ additional configurations
\end{itemize}

This extended testing probes the limits of edge-weighted optimization and establishes the geographic feasibility threshold.

\subsubsection{Metric Collection}

For each configuration, we collect:
\begin{enumerate}
    \item \textbf{State-level metrics}: Edge cut (unweighted, weighted), average Polsby-Popper, average Reock, average convex hull ratio, MM count, maximum minority percentage
    \item \textbf{District-level metrics}: Per-district Polsby-Popper, Reock, convex hull ratio, minority percentage, MM status, internal edge count, tract count
    \item \textbf{Configuration metadata}: State, weight factor, threshold, success status (MM count $\geq$ target)
\end{enumerate}

All metrics are computed using custom Python scripts leveraging GeoPandas (geometric operations), SciPy (graph operations), and NumPy (statistical calculations).

\subsection{Ensemble Validation via ReCom}

A critical question for any redistricting algorithm is whether its solutions represent genuinely optimal configurations or merely algorithm-specific artifacts. The computational social science standard for validation is \textbf{ensemble comparison}: comparing deterministic algorithm outputs to distributions generated by Markov Chain Monte Carlo (MCMC) methods that explore the full space of valid redistricting plans.

\subsubsection{Why Ensemble Comparison Matters}

\textbf{Limitation of Deterministic Algorithms}: Our edge-weighted METIS approach produces a single partition for each parameter configuration. While this partition minimizes weighted edge cut subject to constraints, we cannot know \emph{a priori} whether it represents a typical solution, an outlier, or a genuinely optimal point in the vast space of valid redistricting plans.

\textbf{Ensemble Methods as Gold Standard}: Recent computational redistricting literature (DeFord et al. 2021; Fifield et al. 2020; Mattingly \& Vaughn 2014) advocates ensemble methods that generate thousands of valid plans via random sampling. By examining the \emph{distribution} of compactness and VRA compliance across ensembles, we can assess whether our METIS solutions:
\begin{enumerate}
    \item Are typical of random exploration (not special)
    \item Are outliers in favorable directions (genuinely superior)
    \item Lie on the Pareto frontier (cannot be improved on both objectives simultaneously)
\end{enumerate}

\subsubsection{ReCom Algorithm}

We use the \textbf{Recombination (ReCom)} algorithm (DeFord et al. 2021), a Markov chain that samples valid redistricting plans by:

\begin{enumerate}
    \item \textbf{Select random edge}: Choose an edge between two districts at random
    \item \textbf{Merge districts}: Temporarily merge the two districts into a single region
    \item \textbf{Re-bisect region}: Use recursive bisection to split the merged region back into two districts with equal population
    \item \textbf{Accept new partition}: The new partition becomes the next state in the chain
\end{enumerate}

ReCom satisfies \textbf{detailed balance} (reversibility) and \textbf{irreducibility} (can reach all valid partitions), ensuring it explores the space of valid redistricting plans uniformly (up to population constraints).

\textbf{Key Property}: ReCom is ``VRA-neutral''---it does not optimize for minority representation. Plans with high MM district counts arise purely from random demographic-geographic alignments, not intentional clustering. This makes ReCom ensembles an ideal baseline for assessing whether edge-weighted METIS achieves VRA compliance through genuine optimization or mere luck.

\subsubsection{Ensemble Generation Protocol}

For Alabama (our flagship case study), we generate a 10,000-plan ReCom ensemble:

\begin{itemize}
    \item \textbf{Starting partition}: Baseline METIS partition (ensures valid initial state)
    \item \textbf{Chain length}: 10,000 steps after 1,000-step burn-in (discard initial states to reduce dependence on starting partition)
    \item \textbf{Constraints}: Contiguity (all districts connected), population balance (±0.5\%), 7 districts
    \item \textbf{No VRA optimization}: ReCom proposals are purely random (no demographic weighting)
\end{itemize}

For each plan in the ensemble, we compute:
\begin{itemize}
    \item Edge cut (unweighted)
    \item Mean Polsby-Popper score
    \item MM district count (50\% threshold)
    \item Maximum minority percentage
\end{itemize}

\subsubsection{Comparison Metrics}

We assess our METIS solutions relative to the ensemble distribution using:

\begin{enumerate}
    \item \textbf{Percentile rankings}: What fraction of ensemble plans have worse compactness / fewer MM districts than our solution?
    \item \textbf{Domination analysis}: How many ensemble plans \emph{dominate} our solution (better or equal on both objectives with strict inequality on at least one)?
    \item \textbf{Pareto frontier distance}: Is our solution on the ensemble Pareto frontier, or is it dominated by typical ReCom exploration?
\end{enumerate}

\textbf{Interpretation}:
\begin{itemize}
    \item If our edge-weighted solution is dominated by $>10\%$ of ensemble: Algorithm performs poorly, typical random exploration does better
    \item If dominated by $1-10\%$ of ensemble: Good solution, but room for improvement
    \item If dominated by $<1\%$ of ensemble: Near Pareto-optimal, validates algorithm effectiveness
\end{itemize}

\subsubsection{Implementation}

We implement ensemble generation using the GerryChain library (Metric Geometry and Gerrymandering Group, 2019), a Python package specifically designed for MCMC redistricting. GerryChain provides:
\begin{itemize}
    \item Pre-implemented ReCom proposal function
    \item Built-in contiguity and population balance constraints
    \item Efficient updaters for compactness metrics
    \item Trace logging for convergence diagnostics
\end{itemize}

Ensemble generation for Alabama (7 districts, 1,186 tracts, 10,000 plans) requires approximately 2-3 hours on a standard workstation (single-core execution). For computational feasibility, we focus ensemble validation on Alabama as our representative case, with the understanding that patterns observed in Alabama likely generalize to other states given similar demographic geographies.

\subsection{Pareto Frontier Identification}

\subsubsection{Pareto Optimality Definition}

A configuration $c$ is \textbf{Pareto-optimal} if no other feasible configuration $c'$ dominates it. Configuration $c'$ dominates $c$ if:
\begin{enumerate}
    \item $\text{MMCount}(c') \geq \text{MMCount}(c)$ (equal or better VRA compliance), AND
    \item $\text{EdgeCut}(c') \leq \text{EdgeCut}(c)$ (equal or better compactness), AND
    \item At least one inequality is strict (better on at least one objective).
\end{enumerate}

Formally:
\[
c' \succ c \iff \left( \text{MM}(c') \geq \text{MM}(c) \land \text{EC}(c') \leq \text{EC}(c) \right) \land \left( \text{MM}(c') > \text{MM}(c) \lor \text{EC}(c') < \text{EC}(c) \right)
\]

A configuration is Pareto-optimal if no configuration dominates it:
\[
c \in \text{Pareto}(\mathcal{C}) \iff \nexists c' \in \mathcal{C} : c' \succ c
\]

\subsubsection{Frontier Construction Algorithm}

For each state, we construct the Pareto frontier using the following algorithm:

\begin{algorithm}[H]
\caption{Pareto Frontier Identification}
\begin{algorithmic}[1]
\STATE \textbf{Input}: Set of configurations $\mathcal{C} = \{c_1, c_2, \ldots, c_n\}$
\STATE \textbf{Output}: Pareto-optimal subset $\mathcal{P} \subseteq \mathcal{C}$
\STATE Initialize $\mathcal{P} \gets \emptyset$
\FOR{each configuration $c \in \mathcal{C}$}
    \STATE $\text{dominated} \gets \text{False}$
    \FOR{each other configuration $c' \in \mathcal{C} \setminus \{c\}$}
        \IF{$c' \succ c$} \COMMENT{$c'$ dominates $c$}
            \STATE $\text{dominated} \gets \text{True}$
            \STATE \textbf{break}
        \ENDIF
    \ENDFOR
    \IF{$\text{dominated} = \text{False}$}
        \STATE $\mathcal{P} \gets \mathcal{P} \cup \{c\}$
    \ENDIF
\ENDFOR
\RETURN $\mathcal{P}$
\end{algorithmic}
\end{algorithm}

\textbf{Complexity}: $O(n^2)$ for $n$ configurations, computationally trivial for our scale ($n = 21$ per state).

\subsubsection{Multi-Objective Visualization}

We visualize Pareto frontiers as scatter plots with:
\begin{itemize}
    \item \textbf{X-axis}: Edge cut (compactness, lower is better)
    \item \textbf{Y-axis}: MM district count (VRA compliance, higher is better)
    \item \textbf{Colors}: Gradient by maximum minority percentage (diagnostic)
    \item \textbf{Markers}: Pareto-optimal points highlighted with distinct markers (stars) and colors (red)
\end{itemize}

This visualization makes dominated configurations (below and to the right of the frontier) immediately apparent.

\subsubsection{Tradeoff Slope Quantification}

For each consecutive pair of Pareto-optimal points $(c_i, c_{i+1})$ sorted by increasing MM count, we compute the marginal tradeoff slope:
\[
\text{Slope}_{i \to i+1} = \frac{\text{EdgeCut}(c_{i+1}) - \text{EdgeCut}(c_i)}{\text{MMCount}(c_{i+1}) - \text{MMCount}(c_i)}
\]

\textbf{Interpretation}:
\begin{itemize}
    \item Negative slope: Gaining MM districts \emph{improves} compactness (win-win region)
    \item Zero slope: Gaining MM districts has no compactness cost (neutral)
    \item Positive slope: Gaining MM districts costs compactness (traditional tradeoff)
    \item Steep positive slope: Marginal MM districts are very expensive
\end{itemize}

\subsection{Statistical Analysis}

\subsubsection{Metric Correlations}

We compute Pearson correlation coefficients among compactness metrics (edge cut, Polsby-Popper, Reock, convex hull) using all 105 configurations. This assesses consistency across geometric definitions of compactness.

Expected correlations:
\begin{itemize}
    \item Edge cut vs. geometric metrics: Negative correlation ($r \approx -0.6$ to $-0.7$), as lower edge cut generally implies more compact geometric shapes.
    \item Geometric metrics inter-correlation: Positive correlation ($r \approx 0.7$ to $0.8$), as they measure related shape properties.
\end{itemize}

\subsubsection{Significance Testing}

For key claims (e.g., ``non-MM districts improve''), we conduct paired t-tests comparing:
\begin{itemize}
    \item Baseline vs. best edge-weighted configuration (within-state)
    \item MM vs. non-MM district compactness (within-configuration)
\end{itemize}

We report $p$-values and effect sizes (Cohen's $d$) for transparency.

\subsubsection{Robustness Checks}

METIS includes stochastic elements (initial partition randomization). We test robustness by:
\begin{enumerate}
    \item Running selected configurations with 10 different random seeds
    \item Computing coefficient of variation (CV) for key metrics
    \item Reporting mean and standard deviation for configurations with high variation
\end{enumerate}

Preliminary tests show low variation (CV $< 5\%$ for edge cut, $< 3\%$ for MM count) due to METIS's deterministic refinement and fixed random seeds, validating single-run results.

\subsection{Computational Infrastructure}

Practical applicability of our approach depends on computational feasibility. We provide theoretical complexity analysis, empirical runtime measurements, and scalability projections to assess whether edge-weighted METIS optimization is viable for real-world redistricting.

\subsubsection{Theoretical Complexity}

METIS multilevel graph partitioning has near-linear time complexity \citep{karypis1998}. The algorithm consists of three phases:

\textbf{1. Coarsening}: Successively collapse adjacent vertices to create smaller graphs. Complexity: $O(|E|)$ where $|E|$ is the number of edges.

\textbf{2. Initial Partitioning}: Partition the coarsest graph using recursive bisection or k-way partitioning. Complexity: $O(|V|\log|V|)$ where $|V|$ is the number of vertices.

\textbf{3. Refinement}: Project the partition back through uncoarsening levels, applying local refinement at each level (Kernighan-Lin or boundary refinement). Complexity: $O(|E|)$ per level, $O(|E|\log|V|)$ total.

Overall complexity for a single bisection: $O(|E| + |V|\log|V|)$. For $k$ districts requiring $\log_2(k)$ recursive bisection levels, total complexity is $O(k(|E| + |V|\log|V|))$, which is near-linear in practice since $k$ is small (4-14 districts) and $\log|V|$ grows slowly.

\textbf{Edge-Weighting Overhead}: Computing demographic edge weights requires $O(|E|)$ operations to evaluate minority percentages for adjacent tract pairs. This does not change asymptotic complexity---edge-weighting is absorbed into the $O(|E|)$ term. Empirically, edge-weighting adds $\sim$12\% overhead (see Table~\ref{tab:computational_complexity}).

\subsubsection{Hardware and Software}

\textbf{Hardware}: Experiments run on a workstation with Intel i9-12900K (16 cores, 3.2 GHz base / 5.2 GHz boost), 64GB DDR4-3200 RAM, Windows 11. METIS uses 4 parallel workers for multilevel refinement.

\textbf{Software Stack}:
\begin{itemize}
    \item METIS 5.1.0 (graph partitioning) via command-line \texttt{gpmetis} executable
    \item Python 3.13 (data processing, metric computation, workflow orchestration)
    \item GeoPandas 0.14.2 (geometric operations, spatial joins)
    \item NumPy 1.26 / SciPy 1.11 (numerical computations, statistical tests)
    \item Matplotlib 3.8 / Seaborn 0.13 (visualization, Pareto frontiers)
    \item GerryChain 0.2.22 (ensemble validation, ReCom implementation)
\end{itemize}

\subsubsection{Runtime Performance}

Table~\ref{tab:computational_complexity} reports runtime estimates for baseline (uniform edge weights) vs edge-weighted METIS optimization across our five study states.

\input{results/computational_complexity/timing_table.tex}

\textbf{Key Findings}:
\begin{itemize}
    \item \textbf{Baseline mean runtime}: 0.42 seconds per state (range: 0.15s for Mississippi to 0.98s for Georgia)
    \item \textbf{Edge-weighted mean runtime}: 0.47 seconds per state (12\% overhead)
    \item \textbf{Overhead factor}: 1.12× on average (edge-weighting adds minimal computational cost)
    \item \textbf{Scaling with graph size}: Runtime scales near-linearly from 880 tracts (Mississippi) to 1,960 tracts (Georgia), consistent with $O(|E| + |V|\log|V|)$ complexity
\end{itemize}

\textbf{Full Parameter Sweep}: Our experimental design tests 105 configurations (21 per state × 5 states). Total runtime: $105 \times 0.47s \approx 50$ seconds $<$ 1 minute. Post-processing (metric computation, visualization) adds $\sim$5-10 minutes per state, yielding $<$1 hour total experimental cost.

\subsubsection{Scalability to Block-Level Resolution}

Census block-level data provides finer geographic resolution (50,000 blocks per state vs 1,400 tracts) but increases computational cost. Extrapolating from tract-level performance:

\textbf{Scaling Factor}: Block-level graphs have $\sim$35× more vertices. With $O(|E| + |V|\log|V|)$ complexity, expected runtime scaling is:
\[
\text{Runtime}_{\text{block}} \approx \text{Runtime}_{\text{tract}} \times 35 \times \frac{\log(50{,}000)}{\log(1{,}400)} \approx \text{Runtime}_{\text{tract}} \times 50
\]

\textbf{Extrapolated Block-Level Runtime}: $0.42s \times 50 \approx 21$ seconds per state (baseline), $0.47s \times 50 \approx 24$ seconds per state (edge-weighted). Full parameter sweep: $105 \times 24s \approx 42$ minutes.

\textbf{Implication}: Block-level optimization remains computationally feasible ($<$1 hour for full sweep), enabling future work to refine district boundaries beyond tract resolution.

\subsubsection{Comparison: Baseline vs Edge-Weighted}

Edge-weighting introduces two computational steps:
\begin{enumerate}
    \item \textbf{Edge weight computation}: For each edge $(u, v)$, compute minority percentages $p_u, p_v$ and assign weight $w = \beta$ if $\min(p_u, p_v) \geq \tau$, otherwise $w = 1$. Complexity: $O(|E|)$.
    \item \textbf{Weighted METIS partitioning}: METIS processes weighted edges natively (no algorithmic change). Complexity: $O(|E| + |V|\log|V|)$, same as baseline.
\end{enumerate}

Empirical overhead is 12\%, confirming that edge-weight computation is negligible relative to METIS partitioning. The $O(|E|)$ edge-weighting cost is absorbed into the $O(|E| + |V|\log|V|)$ METIS complexity.

\textbf{Conclusion}: Edge-weighted optimization is computationally inexpensive. The 12\% overhead is negligible compared to the substantial VRA compliance benefits demonstrated in our results (Alabama: 0 MM → 2 MM with 3.2\% compactness improvement).

\subsubsection{Data Availability and Reproducibility}

All code, data, and results are available at [repository URL to be added upon publication]. This includes:
\begin{itemize}
    \item Census tract shapefiles (2020 TIGER/Line) and demographic data (Redistricting Data PL 94-171)
    \item Adjacency graphs in coordinate format (COO)
    \item Python scripts for redistricting, metric computation, and visualization
    \item METIS configuration files (ufactor, niter, objtype parameters)
    \item Raw results (CSV files with all 105 configurations, district-level metrics)
    \item Ensemble validation datasets (10,000 ReCom plans for Alabama)
    \item Statistical analysis scripts (bootstrap, permutation tests, effect sizes)
\end{itemize}

Computational experiments are fully reproducible given these materials. Runtime may vary depending on hardware specifications (reported runtimes assume Intel i9-12900K or equivalent).

\subsection{Limitations and Assumptions}

\subsubsection{Single Minority Group}

Our experiments focus on Black-White demographics in Southern states. We do not model:
\begin{itemize}
    \item Multi-group scenarios (Black + Latino, Asian, etc.)
    \item Coalition districts requiring joint representation
    \item Within-group heterogeneity (immigrant vs. native-born)
\end{itemize}

This simplification aligns with VRA litigation in our study states but limits generalization to more diverse states.

\subsubsection{Fixed District Counts}

We hold the number of congressional districts fixed per state (determined by Census apportionment). Varying $k$ could alter Pareto frontiers, but this is not policy-relevant as district counts are constitutionally determined.

\subsubsection{Geographic Resolution}

We use census tracts (~1,500 tracts per state) as atomic units. Finer-resolution block-level data (~50,000 blocks per state) could enable more precise district boundaries but would increase computational costs 30-fold. Tract-level analysis is standard in redistricting research and litigation.

\subsubsection{Population Data Only}

We consider only residential population, ignoring:
\begin{itemize}
    \item Voter registration rates (minorities may have lower registration)
    \item Turnout rates (minorities may have lower turnout)
    \item Age distributions (minorities may have younger populations with fewer eligible voters)
\end{itemize}

These factors affect \emph{electoral} effectiveness but are beyond demographic redistricting optimization. Courts have historically used total population (not voting-eligible population) for VRA analysis.

\subsubsection{Compactness as Sole Competing Objective}

We focus on the VRA-compactness tradeoff. Redistricting involves additional objectives:
\begin{itemize}
    \item Communities of interest (cities, counties)
    \item Partisan fairness (efficiency gap, proportional representation)
    \item Incumbent protection
\end{itemize}

Multi-objective extensions are possible but beyond our current scope.
