\section{Conclusion}

Recursive graph bisection demonstrates that algorithmic objectivity can extend from congressional apportionment to redistricting. Applied at national scale across three census decades, the method produces 1,305 constitutionally compliant districts while structurally unable to access partisan data. Five findings establish feasibility: technical requirements met, VRA compliance exceeding human maps, a 42\% demographic threshold determining minority representation feasibility, partisan patterns reflecting geography rather than manipulation, and temporal stability across census cycles.

The paradigm shift moves from asking ``How can we prevent gerrymandering?'' to ``What if the algorithm cannot see party?'' The impossibility defense---stronger than intent-based tests---offers courts a judicially manageable standard. Legislatures and commissions gain transparent alternatives implementable without constitutional amendment. Open-source release planned for the 2030 census cycle will enable independent verification and broader adoption.

Gerrymandering persists not because solutions are unknown but because reform requires removing discretion from those who benefit from its exercise. Mathematical governance succeeded for apportionment because Congress, in a rare moment of institutional humility, accepted that formulas could allocate House seats more legitimately than politicians. The same path remains available for redistricting.

Eighty years after mathematical governance resolved congressional apportionment, algorithmic redistricting offers a similar path toward restoring electoral legitimacy---not through optimal outcomes, which mathematics cannot guarantee, but through structural objectivity, which it can.
