\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{algorithm}
\usepackage{algorithmic}

\title{N-way vs Recursive Bisection for VRA-Compliant Redistricting:\\
A Comprehensive 50-State Comparison}
\author{Giovanni Della-Libera}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Graph partitioning algorithms offer two fundamental approaches to congressional redistricting: n-way partitioning (direct division into $k$ districts) and recursive bisection (hierarchical binary splitting). This paper presents the first comprehensive empirical comparison of these methods for Voting Rights Act (VRA) compliance using edge-weighted graph partitioning across all 50 states.

We conduct ablation studies testing five weight factors (1$\times$--100$\times$) and four minority thresholds (40\%--55\%) for both methods, totaling 1,760 redistricting runs across 44 multi-district states. Our key finding challenges conventional assumptions: \textbf{both methods achieve statistically equivalent success rates} (n-way: 47.5\%, recursive: 48.3\%, $p=0.634$), with negligible effect size (Cohen's $d=-0.018$).

However, methods exhibit distinct characteristics. N-way partitioning offers \textbf{67.5\% faster runtime} (3.68s vs 11.33s average), making it preferable for rapid prototyping and large-scale analysis. Recursive bisection achieves a \textbf{higher performance ceiling} with optimal parameters (56.8\% vs 52.3\% best configuration), benefiting careful parameter tuning. Methods show different parameter sensitivities—n-way prefers lower thresholds (40\%), recursive prefers higher thresholds (50\%)—and state-specific advantages, with n-way winning in Virginia (+35\%) and recursive in Connecticut (+45\%).

This equivalence finding has important implications for redistricting reform: practitioners can confidently choose either method based on computational constraints (speed vs optimization) rather than effectiveness concerns. Both approaches produce comparable VRA compliance when properly configured with edge-weighting and total minority coalitions.
\end{abstract}

\section{Introduction}
\input{sections/01_introduction}

\section{Background: Graph Partitioning Methods}
\input{sections/02_background}

\section{Theoretical Analysis}
\input{sections/03_theory}

\section{Experimental Design}
\input{sections/04_methodology}

\section{Results}
\input{sections/05_results}

\section{Discussion}
\input{sections/06_discussion}

\section{Conclusion}
\input{sections/07_conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
