\section{Related Work}
\label{sec:related}

\paragraph{Algorithmic redistricting.}
The METIS multilevel graph partitioning framework~\citep{Karypis1998} has become
the standard tool for compactness-optimizing redistricting~\citep{Browdy1990,
Liu2016}.
Edge-weighted METIS (B.2 in this series) replaces uniform edge weights with
TIGER boundary lengths, aligning the edge-cut objective with geographic compactness
measured by perimeter.
The present paper extends this by adding a second edge-weight signal: county
membership.

\paragraph{Subdivision preservation in law.}
The legal requirement to preserve political subdivisions dates to the
Supreme Court's reapportionment decisions of the 1960s.
\emph{Reynolds v.\ Sims} (1964) established equal population as the primary
federal criterion for state legislative districts.
\emph{Rucho v.\ Common Cause} (2019) held that federal courts may not hear
partisan gerrymandering claims, but left open state-court challenges based on
traditional redistricting criteria including subdivision preservation.
In \emph{Harper v.\ Hall} (2022), the North Carolina Supreme Court struck down
a congressional map that split counties unnecessarily.
This paper provides the first formal algorithmic operationalization of this
statutory criterion.

\paragraph{County splits in practice.}
The Princeton Gerrymandering Project and the Redistricting Data Hub track county
splits as a standard fairness metric.
\citet{Chen2013} showed that randomly sampled compact plans split fewer counties
than enacted maps in several states, suggesting enacted maps are gerrymandered
even by this apolitical measure.
We contribute the first large-scale comparison of how an algorithmic parameter
($\alpha$) shifts the county-splits distribution while holding everything else
constant.

\paragraph{Composition with other criteria.}
Recent work has studied how multiple redistricting criteria interact.
\citet{Duchin2020} showed that compactness and VRA compliance can conflict.
Our compositional framework (Section~\ref{sec:method}) allows $\alpha$ to be
combined with minority-weighting~\citep[B.2]{DL2026b} or GeoSection
splits~\citep[B.8]{DL2026c} without coupling the implementations.
