\section{Evaluation}
\label{sec:eval}

\subsection{Setup}

We ran all 50 states using the 2020 Census at congressional district count,
seed $= 1$, comparing $\alpha \in \{0, 5\}$.
For each state we recorded:
(i) total edge cut (km, from the manifest),
(ii) county splits (number of counties whose tracts appear in more than one district,
computed from TIGER GEOID prefixes),
(iii) Democratic seat count and proportionality gap (pp).

One-district states (VT, AK, WY, DE, ND, SD) have no splits or cuts by definition
and are excluded from aggregate statistics; the 44 multi-district states are reported.

\textbf{Limitation:} All results in this section use a single seed ($= 1$).
METIS is stochastic; single-seed outcomes can differ substantially from the
converged minimum found over an ensemble.
The county-split figures are more stable across seeds than partisan outcomes,
but readers should interpret all per-state values as illustrative rather than
converged.
Section~\ref{sec:discussion} discusses the implications.

\subsection{County Splits}

Table~\ref{tab:splits} shows the complete 50-state results.
At $\alpha = 5$, county splits fall from an average of 14.3 to 7.8 per state,
a 45\% reduction.
Of the 44 multi-district states:
\begin{itemize}
  \item \textbf{37 states} see strictly fewer splits ($\alpha = 5$ better)
  \item \textbf{6 states} are unchanged (HI, MT, NV, WV, MD, WI share this property)
  \item \textbf{1 state} (NE, 3 districts) sees one additional split due to population balance requirements forcing a cross-county cut regardless of $\alpha$
\end{itemize}

The largest absolute reductions are in large states with many counties:
Texas ($-24$), Virginia ($-20$), Georgia ($-19$), North Carolina ($-16$),
and Michigan ($-15$).

\begin{table}[t]
\centering
\caption{County splits and edge cut: $\alpha = 0$ vs.\ $\alpha = 5$, seed $= 1$.
States ordered by number of districts. Splits$_0$ = county splits at $\alpha=0$;
Splits$_5$ = county splits at $\alpha=5$.}
\label{tab:splits}
\small
\begin{tabular}{lrrrrrr}
\toprule
State & Districts & Splits$_0$ & Splits$_5$ & $\Delta$ & EC$_0$ (km) & EC$_5$ (km) \\
\midrule
MT  & 2  & 2  & 2  & 0    & 753   & 1393  \\
RI  & 2  & 2  & 2  & 0    & 51    & 161   \\
ME  & 2  & 3  & 2  & $-1$ & 228   & 608   \\
NH  & 2  & 3  & 2  & $-1$ & 128   & 511   \\
ID  & 2  & 3  & 2  & $-1$ & 381   & 730   \\
WV  & 2  & 2  & 2  & 0    & 349   & 583   \\
HI  & 2  & 2  & 2  & 0    & 79    & 435   \\
NE  & 3  & 3  & 4  & $+1$ & 469   & 1552  \\
NM  & 3  & 4  & 2  & $-2$ & 921   & 1803  \\
KS  & 4  & 6  & 3  & $-3$ & 924   & 2087  \\
UT  & 4  & 5  & 4  & $-1$ & 642   & 2039  \\
NV  & 4  & 2  & 2  & 0    & 389   & 3342  \\
AR  & 4  & 9  & 2  & $-7$ & 1130  & 1982  \\
IA  & 4  & 9  & 3  & $-6$ & 1187  & 1687  \\
MS  & 4  & 7  & 3  & $-4$ & 931   & 1758  \\
CT  & 5  & 5  & 4  & $-1$ & 413   & 1278  \\
OK  & 5  & 12 & 6  & $-6$ & 1234  & 2620  \\
OR  & 6  & 8  & 5  & $-3$ & 1521  & 3314  \\
KY  & 6  & 13 & 5  & $-8$ & 1339  & 2637  \\
LA  & 6  & 12 & 5  & $-7$ & 1357  & 3165  \\
SC  & 7  & 11 & 6  & $-5$ & 1428  & 3407  \\
AL  & 7  & 15 & 7  & $-8$ & 1714  & 3377  \\
CO  & 8  & 13 & 6  & $-7$ & 1483  & 3396  \\
WI  & 8  & 15 & 8  & $-7$ & 1683  & 3869  \\
MD  & 8  & 8  & 8  & 0    & 690   & 2395  \\
MO  & 8  & 12 & 6  & $-6$ & 1720  & 3372  \\
MN  & 8  & 16 & 6  & $-10$& 1573  & 3222  \\
TN  & 9  & 19 & 10 & $-9$ & 1796  & 4503  \\
AZ  & 9  & 7  & 3  & $-4$ & 1920  & 5670  \\
MA  & 9  & 8  & 7  & $-1$ & 700   & 2837  \\
IN  & 9  & 20 & 9  & $-11$& 1614  & 3248  \\
WA  & 10 & 13 & 7  & $-6$ & 2216  & 6316  \\
VA  & 11 & 33 & 13 & $-20$& 1824  & 4639  \\
NJ  & 12 & 15 & 13 & $-2$ & 780   & 2843  \\
MI  & 13 & 27 & 12 & $-15$& 2541  & 5292  \\
GA  & 14 & 33 & 14 & $-19$& 2823  & 5887  \\
NC  & 14 & 30 & 14 & $-16$& 2595  & 6416  \\
OH  & 15 & 28 & 15 & $-13$& 2565  & 5374  \\
PA  & 17 & 31 & 18 & $-13$& 2558  & 6630  \\
IL  & 17 & 21 & 12 & $-9$ & 2396  & 6393  \\
NY  & 26 & 31 & 17 & $-14$& 2372  & 6231  \\
FL  & 28 & 27 & 19 & $-8$ & 3699  & 10592 \\
TX  & 38 & 51 & 27 & $-24$& 8448  & 19305 \\
CA  & 52 & 34 & 25 & $-9$ & 7544  & 25192 \\
\midrule
\textbf{Mean} & & \textbf{14.3} & \textbf{7.8} & \textbf{$-6.5$} &
  \textbf{1662} & \textbf{4184} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Edge-Cut Cost}

The price of fewer splits is longer cuts.
At $\alpha = 5$, average edge cut increases from 1,662 km to 4,184 km, a factor
of 2.5.
The extra boundary length corresponds to following existing administrative
boundaries rather than choosing the shortest available cut.
Importantly, this cost scales with county geometry rather than with $k$: states
where county lines happen to be near the natural bisection boundaries (e.g.,
Iowa, Arkansas) show smaller relative cost increases.

\subsection{Ensemble County-Split Results}
\label{sec:ensemble}

Table~\ref{tab:ensemble} reports mean county splits and edge cut over 25 seeds
per state at $\alpha \in \{0, 5\}$ for seven focal states.
The split reduction is consistent and substantial across seeds: the range bars
confirm that no seed at $\alpha = 5$ produces more splits than the worst seed
at $\alpha = 0$ for any of the six large focal states.

\begin{table}[t]
\centering
\caption{Multi-seed ensemble: county splits and edge cut over 25 seeds.
Format: mean (min--max). EC ratio = EC$_5$ / EC$_0$ (normalised to $\alpha=0$ baseline).}
\label{tab:ensemble}
\small
\begin{tabular}{lrrrrrr}
\toprule
State & $k$ & Splits$_0$ & Splits$_5$ & $\Delta$ splits & EC ratio \\
\midrule
GA & 14 & 33.7 (24--38) & 12.3 (8--16)  & $-21.4$ & 2.1$\times$ \\
MI & 13 & 22.8 (18--27) & 11.6 (9--14)  & $-11.2$ & 2.3$\times$ \\
NC & 14 & 29.6 (24--35) & 13.1 (9--17)  & $-16.5$ & 2.5$\times$ \\
PA & 17 & 31.2 (28--37) & 17.1 (14--21) & $-14.1$ & 2.5$\times$ \\
TX & 38 & 49.3 (42--58) & 24.0 (22--29) & $-25.3$ & 2.3$\times$ \\
VA & 11 & 29.4 (24--37) & 12.2 (10--18) & $-17.2$ & 2.1$\times$ \\
CA & 52 & 31.6 (31--34) & 24.8 (23--26) & $-6.8$  & 3.2$\times$ \\
\midrule
\textbf{Mean} & & \textbf{32.5} & \textbf{16.4} & \textbf{$-16.1$} & \textbf{2.4$\times$} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Partisan Effects (Ensemble)}
\label{sec:partisan}

Table~\ref{tab:partisan} reports mean Democratic seats over 25 seeds at
$\alpha \in \{0, 5\}$ for the seven focal states.
The ensemble analysis resolves the seed-sensitivity problem that affected the
single-seed pilot: Georgia's mean shifts from 5.6D (not 4D as in the single-seed
result) to 6.0D --- a difference of 0.4 seats across 14 districts.

\begin{table}[t]
\centering
\caption{Ensemble partisan outcomes: mean D seats over 25 seeds, $\alpha \in \{0,5\}$.}
\label{tab:partisan}
\small
\begin{tabular}{lrrrrr}
\toprule
State & D vote (\%) & D seats$_0$ (mean) & D seats$_5$ (mean) & $\Delta$ & $p$-value \\
\midrule
GA & 48.5 & 5.6 & 6.0 & $+0.4$ & 0.18 \\
MI & 48.9 & 6.2 & 6.2 & $0.0$  & 1.00 \\
NC & 49.3 & 5.7 & 5.6 & $-0.1$ & 0.81 \\
PA & 50.6 & 7.3 & 7.6 & $+0.3$ & 0.35 \\
TX & 46.8 & 18.5 & 18.6 & $+0.1$ & 0.85 \\
VA & 54.2 & 7.1 & 7.5 & $+0.4$ & 0.22 \\
CA & 60.3 & 45.8 & 46.3 & $+0.5$ & 0.09 \\
\midrule
\textbf{Mean} & & & & \textbf{$+0.18$} & \\
\bottomrule
\end{tabular}
\end{table}

The mean D seat change is $+0.18$ across 7 focal states covering 149 congressional
districts --- statistically and substantively indistinguishable from zero.
No individual state change reaches $p < 0.05$ under a two-sided Wilcoxon signed-rank
test across the 25-seed ensemble.
The null hypothesis that $\alpha$ does not affect expected partisan outcomes
cannot be rejected.\footnote{$p$-values are from two-sided Wilcoxon signed-rank
tests comparing per-seed D-seat counts at $\alpha=0$ vs.\ $\alpha=5$ within
each state. Independence across seeds is violated in principle since the same
adjacency graph is used; we regard these as directional indicators rather than
exact significance statements.}
