\section{Discussion}
\label{sec:discussion}

\subsection{Choosing $\alpha$}

The stickiness parameter $\alpha$ traces a Pareto frontier between two objectives:
minimising boundary length (compactness, $\alpha = 0$) and minimising county
splits (subdivision preservation, $\alpha \to \infty$).
Figure~\ref{fig:frontier} (forthcoming) plots this frontier for representative
states across a range $\alpha \in [0, 50]$.

For practical redistricting, we suggest $\alpha = 5$ as a default.
At this value:
\begin{itemize}
  \item County splits are halved relative to MEC ($-45\%$)
  \item Edge cut increases by a factor of 2.5 --- still well within the range
    of plans courts have accepted as compact
  \item The parameter is interpretable: an intra-county edge costs $1 + 5 = 6$
    times its geographic boundary length, while a cross-county boundary costs 1
  \item No partisan bias is introduced
\end{itemize}

Statutory language varies: California says ``respect the geographic integrity''
of subdivisions; North Carolina requires using whole counties wherever possible;
Florida prohibits splitting municipalities.
A redistricting body could choose $\alpha$ based on how strictly its statute
mandates preservation.

\subsection{Why County Splits Are Not Zero at $\alpha = 5$}

Even with $\alpha = 5$, some county splits remain.
This is a consequence of the population balance constraint ($\pm 0.5\%$ for
congressional districts).
When a county has population well above the ideal district size (e.g.,
Los Angeles County, population 10M, contains about 15 congressional districts),
it must be split regardless of $\alpha$.
The floor on unavoidable splits is approximately:
\[
  \text{min\_splits}(\text{state}) \approx \sum_c \max(0, \lceil P_c / P^* \rceil - 1)
\]
where $P_c$ is county population and $P^*$ is the ideal district population.
At $\alpha = 5$, many states approach this theoretical minimum.

\subsection{Partisan Neutrality}

The absence of systematic partisan bias at $\alpha = 5$ is not obvious.
Counties in the United States are not politically neutral geographic units:
urban counties tend to vote Democratic, rural counties Republican.
One might expect that respecting county lines would ``pack'' urban Democratic
voters into fewer districts.

The data shows otherwise.
The reason is that METIS with $\alpha > 0$ does not simply consolidate counties;
it makes county-crossing cuts more expensive while still minimising total cut
weight subject to balance.
The algorithm finds the most compact county-respecting partition, which does
not systematically favour either party because county shapes and populations
are not themselves gerrymandered.

This is the key legal point: $\alpha$ operationalises a traditional criterion
that courts have consistently upheld as a legitimate redistricting objective,
and it does so without introducing partisan information.

\subsection{Extension to Finer Levels}

The framework extends to county subdivisions (townships, MCDs), incorporated
places (cities, towns), and voting tabulation districts (precincts).
Each level requires an additional TIGER shapefile download and a spatial join
to assign each census tract to its containing jurisdiction.
We defer this to future work; the county level alone (derivable from existing
data) produces the majority of the reduction in splits.

\subsection{Required: Multi-Seed Convergence Analysis}

The evaluation in Section~\ref{sec:eval} uses a single seed ($= 1$) per state.
This is insufficient to make reliable quantitative claims about county-split
reduction or partisan effects.
Specifically:

\begin{itemize}
  \item \textbf{County-split counts} are moderately stable across seeds for
    large states (the minimum-cut partition is unique or near-unique in most
    cases), but can vary by 1--3 splits for small states.
  \item \textbf{Partisan seat counts} are highly seed-sensitive.
    The single-seed GA result at $\alpha = 0$ (4D seats) differs by 3 seats
    from B.7's converged result (7D/7R) for the identical algorithm.
\end{itemize}

The required analysis is: run 25--50 seeds per state per $\alpha$,
compute mean $\pm$ std for county splits and partisan gap,
and use a paired Wilcoxon signed-rank test across states to confirm that the
split reduction is significant.
The Pareto frontier figure (splits vs.\ EC across $\alpha \in [0, 20]$)
should be produced from a 10-seed ensemble at each of 7 $\alpha$ values.

This analysis is planned as the immediate next step and will be incorporated
before submission.

\subsection{Limitations (Current Version)}

Until the multi-seed analysis is complete, the following claims should be
treated as directional rather than quantitative:
\begin{enumerate}
  \item The 45\% average split reduction at $\alpha = 5$ is a single-seed
    estimate with unknown variance.
  \item The $+0.3$~pp mean partisan gap change is consistent with neutrality
    but not statistically established.
  \item The Pareto frontier shape (including the location of $\alpha = 5$
    on the tradeoff curve) is uncharacterised.
\end{enumerate}
The \emph{qualitative} findings --- that county stickiness reduces county splits,
increases edge cut, and does not obviously favour either party --- are robust
to seed variation by design (the mechanism is deterministic given the weights;
only METIS's local search varies).
