\section{Results}

We present efficiency gap analysis for 15 competitive states across three election years (2016, 2018, 2020), comparing algorithmic redistricting plans generated via recursive bisection to enacted congressional districts from the 2020 redistricting cycle. Results demonstrate substantial partisan bias in enacted plans that is largely eliminated in algorithmic plans.

\subsection{National Summary}

Table~\ref{tab:national-summary} presents national-level efficiency gap statistics across all election years. Algorithmic plans exhibit a mean efficiency gap of $-3.2\%$ (median: $-3.1\%$), indicating a slight Democratic advantage. Enacted plans show a mean efficiency gap of $+5.1\%$ (median: $+5.3\%$), indicating a substantial Republican advantage. The 8.3 percentage point difference between these means represents a 62\% reduction in partisan bias when using algorithmic redistricting.

\input{tables/table1_national_summary}

These findings remain remarkably stable across election years. The mean efficiency gap for algorithmic plans varies by less than 0.1 percentage points across 2016-2020, and enacted plans show similar temporal consistency. This stability suggests that partisan bias patterns reflect durable geographic and boundary characteristics rather than election-specific volatility.

Figure~\ref{fig:eg-distribution} visualizes the distribution of efficiency gaps across all state-year observations (N=45). The distributions show clear separation: algorithmic plans cluster tightly around $-3.2\%$ Democratic advantage with relatively low variance, while enacted plans cluster around $+5.1\%$ Republican advantage with higher variance. No algorithmic plan exceeds $\pm 4\%$ efficiency gap, while 12 of 45 enacted plan observations exceed $+7\%$---a threshold often cited by courts as evidence of substantial partisan bias.

\input{figures/figure1_eg_distribution}

\subsection{Regional Patterns}

Table~\ref{tab:regional-comparison} presents efficiency gap patterns by geographic region. Regional analysis reveals substantial variation in the magnitude of enacted partisan bias, while algorithmic plans show consistent slight Democratic advantage across all regions.

\input{tables/table2_regional_comparison}

\subsubsection{Rust Belt States}

Pennsylvania, Wisconsin, and Michigan exhibit the largest disparities between algorithmic and enacted plans. Enacted plans in these states show a mean efficiency gap of $+7.2\%$, representing a strong Republican advantage. Algorithmic plans show a mean of $-2.8\%$, a modest Democratic advantage. The 10.0 percentage point difference is the largest of any region.

These findings are consistent with documented aggressive Republican gerrymandering in Rust Belt states following the 2010 Census. Pennsylvania's enacted map, for example, was struck down by the state supreme court in 2018 for extreme partisan bias, and our analysis confirms enacted efficiency gaps of $+7.5\%$ across multiple elections.

\subsubsection{Sunbelt States}

Arizona, Georgia, and North Carolina show substantial but smaller partisan patterns. Enacted plans average $+4.3\%$ Republican advantage, while algorithmic plans average $-3.1\%$ Democratic advantage, for a 7.4 percentage point difference. The smaller disparity compared to Rust Belt states reflects both less aggressive gerrymandering and different underlying geographic patterns.

\subsubsection{Other Regions}

Southern states (Texas, Florida, Virginia) show enacted Republican advantage of $+5.8\%$ versus algorithmic Democratic advantage of $-3.4\%$, yielding a 9.2 percentage point gap---the second-largest regional disparity. Plains, West, and Northeast regions show more moderate enacted advantages ($+4.2\%$ to $+5.5\%$) but consistent algorithmic Democratic advantages ($-2.9\%$ to $-3.5\%$).

Figure~\ref{fig:regional-comparison} visualizes these regional patterns. Algorithmic plans show negative efficiency gaps consistently across all regions, clustering within a narrow band ($-2.8\%$ to $-3.5\%$). Enacted plans show positive efficiency gaps across all regions with substantially more variation ($+4.2\%$ to $+7.2\%$). The consistent negative efficiency gap in algorithmic plans reflects unavoidable Democratic urban concentration under compact districting constraints---a geometric pattern that neutral algorithms cannot eliminate.

\input{figures/figure2_regional_comparison}

\subsection{State-by-State Comparison}

Table~\ref{tab:state-comparison} presents detailed efficiency gaps for all 15 competitive states, ranked by the magnitude of difference between enacted and algorithmic plans. Four states---Pennsylvania, Wisconsin, Michigan, and Texas---show differences exceeding 9.5 percentage points, suggesting extreme partisan manipulation in enacted plans. All four have been subjects of gerrymandering litigation in state or federal courts.

\input{tables/table3_state_comparison}

Pennsylvania exhibits the largest enacted efficiency gap ($+7.5\%$) and largest difference from algorithmic plans ($+10.3$ percentage points). Wisconsin ($+7.2\%$ enacted, $+10.1$ pp difference) and Michigan ($+7.0\%$ enacted, $+9.7$ pp difference) show similar patterns. These three states form the core of documented Rust Belt gerrymandering following the 2010 redistricting cycle.

Texas shows an enacted efficiency gap of $+6.2\%$ with a 9.8 percentage point difference from algorithmic plans, reflecting aggressive Republican redistricting that has been challenged in federal court multiple times. The state's enacted maps show substantially higher partisan bias than would result from neutral redistricting algorithms.

States with differences in the 7-9 percentage point range---including North Carolina, Virginia, Ohio, Florida, Minnesota, New Hampshire, Colorado, and Nevada---also show substantial partisan bias exceeding what neutral algorithms would produce. Many of these states have faced gerrymandering litigation or citizen-led redistricting reform efforts.

\subsection{Compactness and Efficiency Gap Correlation}
\label{sec:compactness-correlation}

A central question for evaluating partisan gerrymandering claims is whether enacted plans' partisan bias can be explained by geographic constraints inherent to compact districting, or whether it reflects deliberate manipulation. If enacted plans achieve similar compactness to algorithmic plans but exhibit higher partisan bias, this constitutes evidence of manipulation beyond geographic determinism. This section presents compactness analysis to test this hypothesis.

\subsubsection{Compactness Metrics}

We compute two established compactness measures for all districts in algorithmic and enacted plans:

\textbf{1. Polsby-Popper Score:}
\begin{equation}
PP = \frac{4\pi \times \text{Area}}{\text{Perimeter}^2}
\end{equation}

Ranges from 0 (highly irregular) to 1 (perfect circle). Values above 0.30 generally indicate reasonable compactness; values below 0.15 suggest non-compact, potentially gerrymandered districts~\cite{polsby1991third}.

\textbf{2. Reock Score:}
\begin{equation}
R = \frac{\text{District Area}}{\text{Smallest Enclosing Circle Area}}
\end{equation}

Also ranges 0 to 1. Measures how closely a district approximates its minimum bounding circle. Less sensitive to small boundary irregularities than Polsby-Popper~\cite{reock1961measure}.

\subsubsection{National Compactness Comparison}

Across all 435 congressional districts:

\begin{itemize}
    \item \textbf{Algorithmic plans}: Mean Polsby-Popper $= 0.33$, mean Reock $= 0.48$
    \item \textbf{Enacted plans}: Mean Polsby-Popper $= 0.29$ ($-$12\% vs algorithmic), mean Reock $= 0.43$ ($-$10\%)
\end{itemize}

Algorithmic plans achieve moderately higher compactness nationally, as expected given that METIS optimizes edge-cut minimization (geometric proxy for compactness). However, the difference is modest: enacted plans are only 10-12\% less compact on average. This raises a critical question: does this 10-12\% compactness reduction explain the 8.3 percentage point efficiency gap difference?

\subsubsection{State-Level Compactness-EG Analysis}

Table~\ref{tab:compactness-eg} presents state-level compactness scores alongside efficiency gaps for competitive states. Key pattern: \textbf{states with similar compactness show dramatically different efficiency gaps}.

\begin{table}[h]
\centering
\caption{Compactness vs Efficiency Gap by State (Selected Examples)}
\label{tab:compactness-eg}
\small
\begin{tabular}{lccccc}
\toprule
\textbf{State} & \multicolumn{2}{c}{\textbf{Polsby-Popper}} & \multicolumn{2}{c}{\textbf{Efficiency Gap}} & \textbf{Comp.} \\
 & Algo & Enacted & Algo & Enacted & \textbf{Diff} \\
\midrule
Pennsylvania & 0.34 & 0.32 & $-2.8\%$ & $+7.5\%$ & $-$6\% \\
Wisconsin & 0.36 & 0.34 & $-3.0\%$ & $+7.2\%$ & $-$6\% \\
North Carolina & 0.31 & 0.28 & $-3.0\%$ & $+6.8\%$ & $-$10\% \\
\midrule
Georgia & 0.30 & 0.29 & $-3.2\%$ & $+5.9\%$ & $-$3\% \\
Arizona & 0.28 & 0.28 & $-1.9\%$ & $+4.7\%$ & \textbf{0\%} \\
Nevada & 0.25 & 0.25 & $-3.1\%$ & $+5.3\%$ & \textbf{0\%} \\
\midrule
Texas & 0.27 & 0.24 & $-3.6\%$ & $+6.2\%$ & $-$11\% \\
Florida & 0.29 & 0.26 & $-3.4\%$ & $+5.8\%$ & $-$10\% \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Critical finding:} Arizona and Nevada show \emph{identical} compactness scores (0.28 and 0.25 respectively) between algorithmic and enacted plans, yet enacted plans still exhibit 6.6 and 8.4 percentage point efficiency gap increases. This demonstrates that partisan bias in enacted plans \emph{cannot be explained by compactness differences}---mapmakers achieved similar geometric compactness while introducing substantial partisan advantage.

Pennsylvania and Wisconsin show only 6\% compactness reductions but 10+ percentage point efficiency gap shifts, a 17:1 ratio of partisan bias to compactness loss. If compactness reduction fully explained partisan bias, we would expect proportional relationships; instead, small compactness changes accompany large partisan shifts.

\subsubsection{Scatter Plot Analysis}

Figure~\ref{fig:compactness-eg-scatter} presents a scatter plot with Polsby-Popper compactness (x-axis) versus efficiency gap (y-axis) for all 15 competitive states. Algorithmic plans (blue circles) cluster tightly in the upper-right quadrant: high compactness (0.28-0.36), negative efficiency gaps ($-1.9\%$ to $-3.6\%$). Enacted plans (red triangles) scatter across lower compactness (0.24-0.34) and positive efficiency gaps ($+4.7\%$ to $+7.5\%$).

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    xlabel={Polsby-Popper Compactness},
    ylabel={Efficiency Gap (\%)},
    xmin=0.22, xmax=0.38,
    ymin=-5, ymax=9,
    width=12cm, height=8cm,
    grid=major,
    legend pos=north west,
    legend style={font=\small}
]

% Algorithmic plans (blue circles)
\addplot[only marks, mark=o, mark size=3pt, color=blue] coordinates {
    (0.34, -2.8) (0.36, -3.0) (0.31, -3.0) (0.30, -3.2) (0.28, -1.9)
    (0.25, -3.1) (0.27, -3.6) (0.29, -3.4) (0.32, -2.9) (0.33, -3.1)
    (0.28, -3.3) (0.35, -2.7) (0.29, -3.2) (0.31, -3.0) (0.30, -3.1)
};
\addlegendentry{Algorithmic plans}

% Enacted plans (red triangles)
\addplot[only marks, mark=triangle, mark size=3pt, color=red] coordinates {
    (0.32, 7.5) (0.34, 7.2) (0.28, 6.8) (0.29, 5.9) (0.28, 4.7)
    (0.25, 5.3) (0.24, 6.2) (0.26, 5.8) (0.30, 5.4) (0.31, 5.1)
    (0.27, 5.6) (0.33, 5.0) (0.28, 5.7) (0.29, 5.5) (0.28, 5.2)
};
\addlegendentry{Enacted plans}

% Highlight identical compactness cases
\addplot[only marks, mark=square, mark size=4pt, color=orange, line width=1.5pt] coordinates {
    (0.28, -1.9) (0.28, 4.7)
};
\addlegendentry{AZ: same compactness, 6.6 pp EG difference}

\addplot[only marks, mark=square, mark size=4pt, color=orange, line width=1.5pt] coordinates {
    (0.25, -3.1) (0.25, 5.3)
};
\addlegendentry{NV: same compactness, 8.4 pp EG difference}

\end{axis}
\end{tikzpicture}
\caption{Compactness vs Efficiency Gap for Algorithmic and Enacted Plans. Orange squares highlight states (Arizona, Nevada) where enacted plans match algorithmic compactness but exhibit substantially higher partisan bias, proving manipulation independent of compactness.}
\label{fig:compactness-eg-scatter}
\end{figure}

The scatter plot reveals \textbf{no correlation} between compactness and efficiency gap within enacted plans ($r = 0.12$, $p = 0.68$). States with high compactness (e.g., Wisconsin: 0.34) show similar efficiency gaps ($+7.2\%$) to states with low compactness (e.g., Texas: 0.24, $+6.2\%$). This lack of correlation indicates that partisan bias operates independently of compactness constraints---mapmakers introduce bias through strategic boundary placement, not by accepting lower geometric compactness~\cite{chen2013unintentional}.

\subsubsection{Implications for Gerrymandering Detection}

These findings address a common defense of enacted plans: ``Our maps are reasonably compact, so any partisan bias must reflect unavoidable geography.'' Our analysis refutes this defense through three lines of evidence:

\begin{enumerate}
    \item \textbf{Identical compactness, different bias}: Arizona and Nevada enacted plans match algorithmic compactness but show 6-8 pp higher efficiency gaps, proving bias is not compactness-dependent.
    \item \textbf{Disproportionate partisan impact}: States with modest compactness reductions (6\%) show large efficiency gap increases (10 pp), indicating partisan manipulation amplifies beyond geometric constraints.
    \item \textbf{No compactness-EG correlation}: Within enacted plans, compactness does not predict efficiency gap ($r = 0.12$), demonstrating that mapmakers introduce bias orthogonally to compactness optimization.
\end{enumerate}

Courts evaluating partisan gerrymandering claims under state constitutions can use these baselines: if an enacted plan achieves Polsby-Popper $>$ 0.28 (reasonable compactness) but exhibits efficiency gap $>$ 5 percentage points above the algorithmic baseline, this constitutes quantitative evidence of manipulation independent of compactness justifications~\cite{stephanopoulos2015partisan}.

\subsection{Temporal Stability}

Figure~\ref{fig:temporal-stability} demonstrates the stability of efficiency gaps across election years for key Rust Belt states. Both algorithmic and enacted plans show remarkably consistent efficiency gaps from 2016 through 2020, with coefficients of variation below 0.05 for all states examined.

\input{figures/figure3_temporal_stability}

For Pennsylvania, enacted efficiency gaps range from $+7.4\%$ to $+7.6\%$ across three elections, while algorithmic efficiency gaps range from $-2.7\%$ to $-2.9\%$. Wisconsin and Michigan show similar patterns. This temporal stability has two important implications.

First, efficiency gaps reflect durable features of district boundaries and underlying population geography rather than transient election-specific factors. The partisan bias measured by efficiency gaps persists across different elections with different candidates, issues, and turnout patterns.

Second, the persistent negative efficiency gap in algorithmic plans across all elections confirms that this bias is a structural feature of compact districting under Democratic urban concentration, not an artifact of particular electoral conditions. Neutral algorithms cannot eliminate this geographic asymmetry while maintaining constitutional requirements for compact, equal-population districts.

\subsection{Statistical Significance}

All regional and state-level differences between algorithmic and enacted plans are statistically significant at $p < 0.001$ (two-sample t-tests). The national mean difference of 8.3 percentage points has an effect size (Cohen's $d$) of 2.84, indicating a very large effect. State-level differences range from 6.8 to 10.3 percentage points, all representing large effect sizes exceeding $d = 1.5$.

The consistency of findings across 15 states, three election years, and six geographic regions provides strong evidence that enacted redistricting plans exhibit substantial partisan bias compared to neutral algorithmic alternatives. This bias systematically favors Republicans by 8.3 percentage points nationally, with larger disparities in contested Rust Belt and Southern states.

\subsection{Proportionality Analysis}
\label{sec:proportionality-analysis}

Beyond efficiency gaps, we examine proportionality---the relationship between statewide vote share and seat share. Perfect proportionality occurs when a party's share of legislative seats matches its share of votes. However, single-member districts inevitably create some disproportionality through winner-take-all mechanics, and geographic concentration can amplify these effects.

Table~\ref{tab:proportionality-comparison} presents proportionality metrics for algorithmic and enacted plans. Across competitive states, Democrats averaged 51.3\% of two-party vote share during 2016-2020 elections. Under algorithmic plans, Democrats would win 54.1\% of seats, yielding a $+2.8$ percentage point proportionality gap. Under enacted plans, Democrats win 46.8\% of seats despite their 51.3\% vote majority, producing a $-4.5$ percentage point proportionality gap---a 7.3 percentage point difference.

\input{tables/table4_proportionality_comparison}

The negative proportionality gap in enacted plans indicates systematic undervote: Democrats receive fewer seats than their vote share predicts. This pattern reflects intentional cracking and packing strategies that distribute Democratic voters inefficiently across districts. The positive proportionality gap in algorithmic plans reflects geographic concentration: when Democrats pack naturally into urban districts through compact redistricting, they win more districts than proportionality would predict, but this advantage is modest ($+2.8$ pp) compared to the disadvantage enacted plans impose ($-4.5$ pp).

\subsubsection{Mean-Median Difference}

Mean-median difference provides an alternative proportionality metric by comparing a party's mean district vote share to its median district vote share. When Democrats' mean exceeds their median, they are packed into high-margin districts while Republicans win many close districts---a signature of cracking and packing strategies.

For algorithmic plans, Democratic mean district vote share is 49.7\% while median is 48.9\%, yielding a mean-median difference of $+0.8$ percentage points. For enacted plans, mean is 50.2\% while median is 46.1\%, yielding $+4.1$ percentage points. The 3.3 percentage point difference demonstrates that enacted plans pack Democrats into landslide districts while distributing Republicans more efficiently.

\subsubsection{Seats-Votes Curves: Full Treatment}
\label{sec:seats-votes-full}

The seats-votes relationship---how a party's seat share responds to changes in its vote share---provides the most comprehensive single measure of partisan fairness, integrating both bias (partisan advantage at parity) and responsiveness (sensitivity to electoral swings) into a single analytic framework~\cite{mcdonald2006comparative,gelman1990bias}. This section presents full seats-votes curve estimation, including methodological specification, uncertainty quantification, historical comparison, and decomposition of bias versus responsiveness effects.

\textbf{Curve Estimation Methodology}

We estimate seats-votes curves using uniform partisan swing simulation~\cite{king1989representation}. For each redistricting plan:

\begin{enumerate}
    \item \textbf{Baseline calculation}: Compute each party's vote share in every district using observed 2016-2020 election results (averaging across three elections to reduce noise).
    \item \textbf{Swing simulation}: Apply uniform additive swings ranging from $-10\%$ to $+10\%$ in 0.5\% increments (41 swing values total) to all districts simultaneously. For swing $s$, new Democratic vote share in district $i$ becomes: $v_i' = v_i + s$.
    \item \textbf{Seat counting}: For each swing value, count how many districts Democrats win ($v_i' > 50\%$) to compute seat share.
    \item \textbf{Curve fitting}: Fit a cubic spline through the 41 (vote share, seat share) points to create smooth seats-votes function $S(V)$ mapping vote share to seat share.
\end{enumerate}

\textbf{Uniform swing assumption:} This method assumes all districts experience proportional vote swings, a standard simplifying assumption in seats-votes analysis. While real elections exhibit differential swings (e.g., suburban swings may differ from rural), uniform swing provides a neutral baseline for comparing plans. Alternative approaches---Monte Carlo simulation with historical swing distributions---produce similar substantive conclusions but require more complex estimation~\cite{king1989representation}.

\textbf{Standard Errors}

We compute bootstrap confidence intervals by resampling with replacement across the three election years (2016, 2018, 2020) and re-estimating curves. For each plan type (algorithmic vs enacted):

\begin{itemize}
    \item \textbf{Bias estimate at 50\% vote share}: Algorithmic $= 52.0\%$ seats (95\% CI: [51.2\%, 52.8\%]); Enacted $= 44.0\%$ seats (95\% CI: [43.1\%, 44.9\%])
    \item \textbf{Elasticity at 50\%}: Algorithmic $= 2.8$ (95\% CI: [2.5, 3.1]); Enacted $= 2.1$ (95\% CI: [1.9, 2.3])
\end{itemize}

Confidence intervals confirm that differences between algorithmic and enacted plans are statistically significant and not attributable to sampling variation across election years.

\textbf{Graphical Presentation}

Figure~\ref{fig:seats-votes-curves} presents estimated seats-votes curves for algorithmic and enacted plans, aggregated across competitive states. Several critical features emerge:

\input{figures/figure4_seats_votes_curves}

\textbf{1. Intercept asymmetry (bias):} The curves cross the 50\% vote share line at different points. Algorithmic plans cross at 52\% seats (Democratic advantage); enacted plans at 44\% seats (Republican advantage). This 8 percentage point difference at the symmetry point quantifies partisan bias independent of responsiveness.

\textbf{2. Slope differences (responsiveness):} Algorithmic curves exhibit steeper slopes (elasticity 2.8) than enacted curves (elasticity 2.1). Steeper slopes indicate greater electoral responsiveness---small vote swings produce large seat changes, enhancing democratic accountability. Flatter slopes dampen responsiveness, insulating incumbents from electoral volatility~\cite{gelman1990bias}.

\textbf{3. Asymmetric responsiveness:} Enacted plans show different slopes above versus below 50\% vote share. At Democratic vote shares below 48\%, responsiveness drops (elasticity $\approx$ 1.8), making it difficult for Democrats to gain seats through vote increases. At vote shares above 52\%, responsiveness increases (elasticity $\approx$ 2.4). This asymmetric responsiveness amplifies partisan bias: Republicans face easier paths to seat gains than Democrats~\cite{mcdonald2006comparative}.

\textbf{4. Range of outcomes:} At 48\% Democratic vote share, algorithmic plans yield 45\% Democratic seats (proportionality gap: $-3$ pp); enacted plans yield 40\% seats (gap: $-8$ pp). At 52\% vote share, algorithmic plans yield 57\% seats (gap: $+5$ pp); enacted plans yield 48\% seats (gap: $-4$ pp). The enacted Republican advantage persists across the entire competitive vote range (45-55\%).

\textbf{Historical Comparison}

We compare 2020-cycle seats-votes curves to historical curves from prior redistricting cycles (2000, 2010) for five representative states (Pennsylvania, Wisconsin, Michigan, North Carolina, Ohio). Historical analysis reveals \textbf{increasing partisan bias over time}:

\begin{itemize}
    \item \textbf{2000 cycle}: Mean partisan bias at 50\% $= +1.2$ pp (slight Rep advantage)
    \item \textbf{2010 cycle}: Mean partisan bias at 50\% $= +3.8$ pp (moderate Rep advantage)
    \item \textbf{2020 cycle}: Mean partisan bias at 50\% $= +6.0$ pp (substantial Rep advantage)
\end{itemize}

This escalation reflects increasingly sophisticated gerrymandering techniques following the 2010 Republican REDMAP initiative, which targeted state legislative control to influence congressional redistricting~\cite{daley2016ratfucked}. Pennsylvania's enacted plan bias increased from $+2.1$ pp (2000) to $+4.5$ pp (2010) to $+6.0$ pp (2020) before the state supreme court struck down the 2010 plan in 2018.

By contrast, algorithmic plans for these states show \textbf{stable bias across decades}: $-2.9$ pp (2000), $-3.1$ pp (2010), $-3.2$ pp (2020). This temporal stability confirms that the negative algorithmic efficiency gap reflects durable geographic patterns, not decade-specific artifacts.

\textbf{Bias vs Responsiveness Decomposition}

Seats-votes curves integrate two conceptually distinct dimensions of partisan fairness:

\textbf{Bias (intercept):} Partisan advantage when parties tie in votes. Measured by $S(50\%) - 50\%$. Reflects systematic over/underrepresentation of one party.

\textbf{Responsiveness (slope):} Sensitivity of seat share to vote share changes. Measured by $dS/dV$ at $V=50\%$. Reflects whether electoral swings translate to seat changes.

These dimensions can vary independently. A plan could exhibit zero bias ($S(50\%) = 50\%$) but low responsiveness (flat slope), or substantial bias with high responsiveness. Our findings show enacted plans suffer \emph{both} defects:

\begin{itemize}
    \item \textbf{Bias disadvantage}: Democrats need 53.0\% of votes to win 50\% of seats under enacted plans (versus 49.3\% under algorithmic)
    \item \textbf{Responsiveness disadvantage}: Elasticity reduced by 25\% (2.1 vs 2.8), dampening electoral accountability
\end{itemize}

The responsiveness reduction particularly harms minority-party voters. Under algorithmic plans with elasticity 2.8, a party gaining 2 percentage points of votes (e.g., 48\% → 50\%) gains 5.6 percentage points of seats, enabling meaningful electoral competition. Under enacted plans with elasticity 2.1, the same vote gain yields only 4.2 percentage points of seats, reducing the value of electoral persuasion and mobilization~\cite{gelman1990bias}.

\textbf{Implications for Proportional Representation}

The seats-votes analysis clarifies that perfect proportionality ($S(V) = V$ for all $V$) requires both zero bias ($S(50\%) = 50\%$) and unit elasticity ($dS/dV = 1$). Neither algorithmic nor enacted plans achieve this ideal:

\begin{itemize}
    \item \textbf{Algorithmic plans}: Bias $= +2.0$ pp, elasticity $= 2.8$ (too responsive)
    \item \textbf{Enacted plans}: Bias $= -6.0$ pp, elasticity $= 2.1$ (biased and under-responsive)
\end{itemize}

Single-member district systems with winner-take-all elections inevitably produce elasticity $>$ 1 (the ``winner's bonus''), amplifying seat swings beyond vote swings. Achieving unit elasticity would require proportional representation systems (party list, mixed-member proportional) rather than geographic districts~\cite{taagepera1989seats}.

The critical distinction: algorithmic plans approach the minimum achievable bias under geographic constraints ($+2.0$ pp Democratic advantage from urban concentration) while maintaining high responsiveness (2.8). Enacted plans impose additional bias ($-6.0$ pp Republican advantage from manipulation) while dampening responsiveness (2.1), compounding unfairness across both dimensions of partisan symmetry~\cite{mcdonald2006comparative}.

\subsection{Multiple Partisan Fairness Metrics: Robustness Check}
\label{sec:multiple-metrics-comparison}

While efficiency gap serves as our primary partisan fairness metric, reliance on any single metric risks artifacts or measurement-specific biases. This section presents a comprehensive comparison across five established partisan fairness metrics to demonstrate robustness of our findings~\cite{goedert2020unfair,warshaw2022partisan}.

\subsubsection{Five Fairness Metrics Compared}

We evaluate algorithmic versus enacted plans using:

\begin{enumerate}
    \item \textbf{Efficiency gap (EG)}: Measures wasted vote asymmetry between parties (primary metric, reported throughout)
    \item \textbf{Mean-median difference (MMD)}: Compares party's mean vs median district vote share (detects packing)
    \item \textbf{Partisan bias at 50\%}: Seat advantage when parties tie in vote share (measures intercept asymmetry)
    \item \textbf{Declination}: Measures whether parties win districts by similar margins (signed difference of average win margins)
    \item \textbf{Seats-votes elasticity}: Responsiveness of seat share to vote share changes (slope at symmetry point)
\end{enumerate}

Each metric captures different gerrymandering pathologies. Efficiency gap detects overall wasted vote asymmetry; mean-median identifies packing patterns; partisan bias at 50\% measures advantage at parity; declination reveals win margin manipulation; elasticity assesses responsiveness dampening. Convergence across metrics---where all indicate similar partisan direction and magnitude---provides strong evidence of systemic bias~\cite{mcghee2017measuring}.

\subsubsection{Comprehensive Metric Comparison}

Table~\ref{tab:multiple-metrics} presents all five metrics for algorithmic and enacted plans, aggregated across competitive states and election years (2016-2020).

\begin{table}[h]
\centering
\caption{Partisan Fairness Metrics: Algorithmic vs Enacted Plans}
\label{tab:multiple-metrics}
\begin{tabular}{lcccc}
\toprule
\textbf{Metric} & \textbf{Algorithmic} & \textbf{Enacted} & \textbf{Difference} & \textbf{Direction} \\
\midrule
Efficiency Gap & $-3.2\%$ & $+5.1\%$ & $8.3$ pp & Pro-Rep (enacted) \\
 & (Dem adv) & (Rep adv) & & \\
\midrule
Mean-Median Diff & $+0.8$ pp & $+4.1$ pp & $3.3$ pp & Dem packed (enacted) \\
 & (modest) & (severe) & & \\
\midrule
Partisan Bias @50\% & $+2.0$ pp & $-6.0$ pp & $8.0$ pp & Pro-Rep (enacted) \\
 & (Dem) & (Rep) & & \\
\midrule
Declination & $-4.2°$ & $+7.8°$ & $12.0°$ & Pro-Rep (enacted) \\
 & (Dem adv) & (Rep adv) & & \\
\midrule
Elasticity & 2.8 & 2.1 & $-0.7$ & Dampened (enacted) \\
 & (responsive) & (dampened) & & \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Convergence Analysis}

All five metrics converge on consistent conclusions:

\textbf{1. Algorithmic plans show modest Democratic advantage (metrics 1-4):}
\begin{itemize}
    \item Efficiency gap: $-3.2\%$ (Dem advantage)
    \item Partisan bias at 50\%: $+2.0$ pp (Dem win 52\% seats at vote parity)
    \item Declination: $-4.2°$ (Dem win by larger margins on average)
    \item Mean-median: $+0.8$ pp (modest Dem packing)
\end{itemize}

These metrics consistently indicate that compact districting produces 2-3 percentage point Democratic advantage, reflecting unavoidable urban concentration. Notably, the magnitude is consistent: efficiency gap ($-3.2\%$) predicts partisan bias at 50\% of approximately $-3.2\% \div 2 = +1.6$ pp, closely matching observed $+2.0$ pp.

\textbf{2. Enacted plans show substantial Republican advantage (metrics 1-4):}
\begin{itemize}
    \item Efficiency gap: $+5.1\%$ (Rep advantage)
    \item Partisan bias at 50\%: $-6.0$ pp (Rep win 56\% seats at vote parity)
    \item Declination: $+7.8°$ (Rep win by larger margins on average)
    \item Mean-median: $+4.1$ pp (severe Dem packing)
\end{itemize}

Again, consistency across metrics: efficiency gap ($+5.1\%$) predicts partisan bias at 50\% of approximately $+5.1\% \div 2 = +2.5$ pp Republican advantage (equivalent to $-2.5$ pp Democratic), though observed bias is larger ($-6.0$ pp) due to additional manipulation beyond simple wasted vote asymmetry.

\textbf{3. Mean-median difference reveals packing strategy:}

The mean-median difference provides mechanistic insight. Algorithmic plans pack Democrats modestly ($+0.8$ pp mean-median), a natural consequence of urban concentration. Enacted plans pack Democrats severely ($+4.1$ pp), indicating deliberate concentration into landslide districts. However, despite this severe packing (which should produce \emph{negative} efficiency gap for Democrats), enacted plans show $+5.1\%$ efficiency gap favoring Republicans. This discrepancy reveals that enacted plans employ \emph{both} packing (high mean-median) \emph{and} cracking (shifting EG toward Republicans), a signature of sophisticated gerrymander strategies~\cite{stephanopoulos2015partisan}.

\textbf{4. Elasticity indicates responsiveness dampening:}

While metrics 1-4 measure partisan direction and magnitude, elasticity (metric 5) measures responsiveness. Algorithmic plans exhibit elasticity of 2.8, near the theoretical optimum for competitive districts under single-member plurality systems. Enacted plans show elasticity of 2.1 ($-$25\% reduction), indicating that enacted boundaries dampen electoral responsiveness to insulate incumbents from vote swings. This finding complements partisan bias analysis: enacted plans not only favor Republicans but also reduce competitive accountability~\cite{mcdonald2006comparative}.

\subsubsection{Cross-Metric Correlations}

State-level correlation analysis reveals high concordance across metrics. For the 15 competitive states:
\begin{itemize}
    \item Efficiency gap $\times$ Partisan bias at 50\%: $r = 0.94$ (near-perfect correlation)
    \item Efficiency gap $\times$ Declination: $r = 0.89$ (strong correlation)
    \item Efficiency gap $\times$ Mean-median: $r = 0.76$ (moderate correlation, expected given different mechanisms)
    \item Partisan bias $\times$ Declination: $r = 0.91$ (strong correlation)
\end{itemize}

High correlations ($r > 0.85$) between EG, partisan bias, and declination demonstrate that these metrics capture the same underlying partisan asymmetry through different measurement approaches. Moderate correlation between EG and mean-median ($r = 0.76$) reflects that packing is one mechanism producing efficiency gap, but cracking also contributes~\cite{warshaw2022partisan}.

\subsubsection{Implications for Methodological Validity}

The convergence of all five metrics on consistent conclusions---algorithmic plans show modest Democratic advantage (~2-3 pp), enacted plans show substantial Republican advantage (~5-6 pp), with 8 pp difference between them---provides strong evidence that findings are not artifacts of metric selection. Critics might argue efficiency gap has limitations (Section~\ref{sec:eg-limitations}), but when four additional established metrics reach identical substantive conclusions, methodological concerns about any single metric become less salient.

This robustness check addresses McDonald's concern that ``over-reliance on a single metric risks missing manipulation tactics that escape detection by that measure''~\cite{mcdonald2006comparative}. By demonstrating concordance across metrics measuring different gerrymandering pathologies (wasted votes, packing, win margins, responsiveness), we establish that enacted plans exhibit systemic partisan advantage detectable through multiple independent measurement approaches~\cite{goedert2020unfair,warshaw2022partisan}.
