\section{Discussion}

\subsection{Mechanisms: Why Algorithmic Plans Show Negative Efficiency Gaps}

The persistent $-3.2\%$ efficiency gap in algorithmic plans reflects unavoidable geographic sorting. Democratic urban concentration creates two effects:
\begin{enumerate}
\item \textbf{Packing}: Compact urban districts contain Democratic supermajorities (70-90\% vote share), generating surplus wasted votes
\item \textbf{Limited cracking}: Suburban districts tilt Republican but by smaller margins than urban districts tilt Democratic
\end{enumerate}

This asymmetry is geometrically necessary when optimizing for compactness. Rural Republican voters are dispersed, allowing more efficient vote allocation across multiple districts. Urban Democratic voters are clustered, making packing unavoidable~\cite{rodden2019cities}.

\subsection{Comparison to Enacted Plans: Evidence of Human Manipulation}

The 8.3 percentage point difference between algorithmic ($-3.2\%$) and enacted ($+5.1\%$) efficiency gaps suggests enacted plans amplify rather than merely reflect geographic patterns. Human mapmakers can:
\begin{itemize}
\item Crack urban cores to dilute Democratic votes
\item Pack Democrats into even fewer districts
\item Draw non-compact districts to achieve partisan advantage
\end{itemize}

Algorithmic plans provide a neutral baseline; deviations from this baseline indicate partisan manipulation.

\subsection{Empirical Lower Bound on Partisan Bias}

The $-3.2\%$ algorithmic efficiency gap establishes an empirical lower bound on partisan bias achievable with:
\begin{itemize}
\item Compact, contiguous districts
\item Equal population
\item No access to partisan data
\end{itemize}

This baseline has legal significance: courts evaluating partisan gerrymandering claims under state constitutions can compare enacted plans to algorithmic benchmarks. A plan showing $+7\%$ efficiency gap when neutral algorithms produce $-3\%$ exhibits 10 percentage points of excess bias attributable to human choices.

\subsection{Implications for Proportional Representation}

Algorithmic redistricting reduces partisan bias but does not achieve proportional representation. If Democrats win 52\% of votes nationally, algorithmic plans producing 56.5\% Democratic seats show modest over-representation. Achieving strict proportionality may require:
\begin{itemize}
\item Relaxing compactness requirements to allow strategic district shapes
\item Abandoning single-member districts in favor of multi-member districts with proportional voting
\item Accepting geographic constraints as fundamental limitations
\end{itemize}

The tension between compact districts and proportional outcomes is not a failure of algorithmic methods—it reflects the geographic structure of the American electorate~\cite{chen2013unintentional}.

\subsection{Limitations}

\begin{enumerate}
\item \textbf{Precinct aggregation}: Efficiency gaps computed from precinct results allocated to districts. Block-level analysis would be more precise but computationally intensive.
\item \textbf{Election selection}: We use presidential and midterm elections. Off-year or local elections might show different patterns.
\item \textbf{Single algorithm}: Results based on recursive bisection. Other algorithmic approaches (ReCom, shortest splitline) might produce different efficiency gaps.
\item \textbf{Competitive states focus}: Detailed analysis limited to 15 competitive states. Non-competitive states (e.g., Wyoming, Vermont) omitted as efficiency gap less meaningful in landslide contexts.
\end{enumerate}
