\documentclass[12pt]{article}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{array}
\usepackage{multirow}
\usepackage{enumitem}
\usepackage{threeparttable}

\onehalfspacing

\title{Do Algorithmic Districts Produce Competitive Elections?\\
Evidence from All 50 States}

\author{Giovanni Della-Libera\\
\small{Working Paper}}

\date{\today}

\begin{document}

\maketitle

\begin{abstract}
A central argument for algorithmic redistricting is that geometrically neutral
maps produce more competitive congressional elections than politically drawn
ones.  We test this claim by applying 2020 presidential vote returns to
algorithmically generated and enacted congressional districts across all
50 states.  Using a margin-of-victory threshold of 10 percentage points in the
two-party presidential vote as the definition of a competitive district
(and 5 points for \emph{swing} districts), we find that algorithmic maps
produce approximately 85 competitive districts compared with 65 under enacted
plans---a 31\% increase.  Among competitive algorithmic districts, the partisan
split is 43 Democratic-leaning to 42 Republican-leaning, indicating near
symmetry.  Crucially, this competitiveness is a \emph{byproduct} of geometric
neutrality, not a design target: the algorithm optimizes population equality
and compactness while blind to election data.  We further show that
competitive algorithmic districts remain competitive under plausible national
vote-share swings of up to 4 percentage points, suggesting durability beyond
a single election cycle.  State heterogeneity is substantial: states that gain
the most competitive districts under algorithmic maps tend to be geographically
diverse swing states where enacted maps concentrate partisans into safe seats.
These findings support the democratic-responsiveness argument for algorithmic
redistricting without overstating the algorithm's role.
\end{abstract}

\noindent\textbf{Keywords:} redistricting, electoral competition, algorithmic
maps, competitive districts, partisan symmetry, swing districts

\section{Introduction}
\input{sections/01-introduction}

\section{Defining Competition}
\input{sections/02-defining-competition}

\section{Methodology}
\input{sections/03-methodology}

\section{Results: Competitive Districts---Algorithmic vs.\ Enacted}
\input{sections/04-results}

\section{Partisan Asymmetry Among Competitive Districts}
\input{sections/05-partisan-asymmetry}

\section{Longitudinal Projection}
\input{sections/06-longitudinal}

\section{State Heterogeneity}
\input{sections/07-state-heterogeneity}

\section{Conclusion}
\input{sections/08-conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
