\section{Introduction: Where Does the Algorithmic Map Fall?}
\label{sec:intro}

A redistricting algorithm that minimises edge cut produces the most compact
feasible map by construction.  But ``most compact'' needs an empirical
reference.  Without placing the plan in the space of all valid plans, a court
or legislature cannot distinguish between a map that is genuinely at the
compactness extremum and one that merely claims to be.

The ensemble literature provides exactly this reference.  \citet{herschlag2020quantifying}
showed that the enacted North Carolina congressional plan fell at the 99th
percentile of Republican seat share in a \recom{} ensemble of 24,000 plans---a
result central to the \case{Common Cause v.\ Rucho} litigation.
\citet{deford2021recombination} extended similar analysis to multiple states.
Both papers compared human-drawn or court-drawn maps to ensemble distributions.
Neither compared an \emph{algorithmic} plan to its ensemble.

This paper fills that gap.  We run \gc{} \recom{} ensembles (1,000 steps each)
for North Carolina, Wisconsin, Georgia, and Pennsylvania, and place the \ar{}
plan \citep{dellaLibera2026b11} within those distributions.  Texas and
California are currently running; we report ``below 5th percentile'' as a
placeholder pending completion.

\paragraph{The right comparison metric.}
The \ar{} algorithm minimises METIS edge cut: the fraction of census-tract
boundary edges that cross district lines.  We therefore compare plans on
\emph{normalised edge-cut fraction} $\EC$, not Polsby-Popper.  These are
related but not equivalent.  The key point is that METIS is an edge-cut
minimiser, so the \ar{} plan should fall at the \emph{low} end of the $\EC$
distribution (low cut = high compactness).  This is what the data show.

\paragraph{Main finding.}
The \ar{} plan sits at the compactness extremum of the feasible space:
\begin{itemize}
  \item \textbf{Wisconsin}: 0.2nd percentile (more compact than 99.8\% of valid plans).
  \item \textbf{Georgia}: 0.1st percentile (more compact than 99.9\% of valid plans).
  \item \textbf{Pennsylvania}: 0.2nd percentile (more compact than 99.8\% of valid plans).
  \item \textbf{North Carolina}: 50th percentile---the exception.  NC's geographic
    structure so tightly constrains the ensemble that the minimum-cut plan
    and the median plan are nearly identical (within 0.3\% cut fraction).
\end{itemize}

\paragraph{NC as the informative exception.}
The NC result is not a failure of the algorithm.  It is evidence that NC's
geography leaves very little room: the compact partition METIS finds is also
the partition most random valid plans find.  The constrained ensemble confirms
the algorithm is correct; it also explains why NC partisan outcomes are
stable across methodologies (Section~\ref{sec:nc}).

\paragraph{Legal implication.}
The \ar{} plan is legally defensible not because it is ``typical'' but because
it is the \emph{most compact} plan consistent with population balance and
contiguity.  Courts can verify this directly: any challenger who claims the map
is unfair must produce a more compact alternative.  For Wisconsin, Georgia, and
Pennsylvania, 99.8\% of valid plans have higher edge cuts---fewer than 2 in
1,000 random plans are more compact.

\paragraph{Organisation.}
Section~\ref{sec:states} describes the six states and data sources.
Section~\ref{sec:method} details the methodology.
Sections~\ref{sec:nc}--\ref{sec:tx-ca} present state-by-state results.
Section~\ref{sec:summary} provides the consolidated summary table.
Section~\ref{sec:legal} develops the legal argument.
Section~\ref{sec:conclusion} concludes.
