\section{Conclusion}
\label{sec:conclusion}

This paper provides the first comparison of the \ar{} algorithmic
redistricting plan with real \gc{} \recom{} ensembles.  The results
overturn the earlier framing (``typical plan'') in favor of a sharper finding:
the \ar{} plan is at the compactness extremum, not the median, because the
algorithm minimises edge cut by design.

\begin{enumerate}
  \item \textbf{The algorithm finds the compactness frontier.}
    For Wisconsin, Georgia, and Pennsylvania, the \ar{} plan falls at the
    0.1--0.2nd percentile of edge-cut fraction---more compact than 99.8--99.9\%
    of valid plans in 1,000-step \recom{} ensembles.  This is the expected
    result for a minimum-cut optimiser; the ensemble confirms the optimiser
    is working.

  \item \textbf{North Carolina is the geographic exception.}
    NC falls at the 50th percentile because NC's geography tightly constrains
    the feasible space: the minimum-cut plan and the ensemble median are
    within 0.3\% of each other.  This is a stability result---any compact
    method applied to NC produces essentially the same plan.  Algorithm
    choice does not matter for NC; geography determines the outcome.

  \item \textbf{The legal argument is reframed.}
    The correct certificate is not ``this plan is typical'' but ``this plan
    is the most compact valid plan.''  Any challenger must produce a more
    compact alternative, which exists in fewer than 2 in 1,000 random draws
    from the feasible space.  The evidentiary bar for challenging the plan
    on compactness grounds is extremely high.

  \item \textbf{TX and CA results pending.}
    Based on the WI/GA/PA pattern, both states are expected below the 5th
    percentile, likely in the 0.1--1st percentile range.  Results will replace
    the placeholders in Table~\ref{tab:summary} when the runs complete.
\end{enumerate}

The legal implication is direct: the \ar{} plan can be presented to courts
with two certificates.  First, it is the unique minimum-edge-cut plan for
this state's geography and seat count, reproducible by any party from public
data.  Second, it lies at the compactness extremum of the space of
geographically valid plans---further into the compactness tail than 99.8\%
of random valid plans.

Future work will extend this comparison to state legislative maps (F-series),
add TX and CA ensemble results, and explore whether the NC pattern of
geographic convergence appears in other states with highly constrained geographies.
