\section{The ConvergenceSweep Bridge}
\label{sec:bridge}

\subsection{Two Notions of Convergence}

The term ``convergence'' is used differently in the ensemble and algorithmic
settings.  For ensemble methods, convergence refers to the Markov chain
approaching its stationary distribution.  For \cs{} \citep{dellaLibera2026b16},
convergence refers to the optimisation search reaching the minimum edge-cut
map for a given state and seat count.

These are distinct phenomena, but they interact.  The key question is:

\begin{keybox}
At the threshold $T = 600$, \cs{} produces the minimum-edge-cut (\ar{}) plan.
This plan is a \emph{specific point} in the plan space that the \recom{} chain
also explores.  Where does this specific point fall in the \recom{} distribution?
\end{keybox}

\subsection{Characterising the Local Optimum}

The \ar{} plan is a local optimum of the METIS edge-cut objective with
respect to small perturbations.  Under \recom{} moves, the plan will not
remain in its basin of attraction for long---a single recombination move
typically changes $5$--$15\%$ of district assignments and can escape local
optima.  This means the \ar{} plan is \emph{not} the typical plan in the
\recom{} ensemble.

However, the \ar{} plan's \emph{compactness} (as measured by Polsby-Popper
score) is empirically high: B.11 shows a mean $\bar\pp \approx 0.38$--$0.45$
across states, versus \recom{} ensemble means of $0.28$--$0.35$.  This
places the \ar{} plan typically at the $65$th--$75$th percentile of the
\recom{} compactness distribution---more compact than average but not extreme.

\subsection{The Minimum-EC Plan in the ReCom Distribution}

Let $P^*_{\EC}$ denote the minimum-edge-cut plan found by \cs{}.  The
edge-cut objective $\EC(P)$ is negatively correlated with Polsby-Popper
compactness (lower edge cut implies fewer shared-boundary census tracts
across district lines, which correlates with more cohesive district shapes).
Within the \recom{} ensemble, plans with lower $\EC$ are rarer, because
\recom{} does not target $\EC$ minimisation.

Empirically, we estimate $P^*_{\EC}$ falls at approximately the
\begin{itemize}
  \item $65$th--$75$th percentile of the $\bar\pp$ distribution,
  \item $45$th--$60$th percentile of the $S_D$ (partisan seats) distribution,
  \item $50$th--$65$th percentile of the $\mathrm{MM}$ distribution,
\end{itemize}
with state-specific variation characterised in G.1--G.3.  The partisan
percentile near the median is the central finding: the \ar{} plan's
partisanship is close to what geography determines, not to what a partisan
mapmaker would achieve.

\subsection{Diagnostic Consistency}

The $T = 600$ threshold for \cs{} was certified in B.16 using empirical data
from B.7 (50-state seed sweep).  The same data supports ensemble diagnostic
thresholds: if the seed sweep finds no improvement after 600 seeds, the
plan is stable under seed perturbation.  This is analogous to---but distinct
from---an ensemble chain achieving \ess{} $> 400$.

ConvergenceSweep $T = 600$ and ensemble \ess{} are independent certificates
of robustness: \ess{} measures MCMC mixing in the ensemble space; $T = 600$
measures monotonic convergence in the seed-quality space.  Both certify
stability but via orthogonal mechanisms.  They are not numerically comparable
and should not be conflated.  Together they provide complete coverage of the
two-method framework: \ess{} certifies that the ensemble distribution estimate
is stable under additional draws; $T = 600$ certifies that the deterministic
search has found the best available plan under additional seeds.
