\section{Implementation Notes}
\label{sec:implementation}

\subsection{Rayon Parallelism}

The key observation enabling parallelism is that particle proposals at each
stage are \emph{independent}: the spanning tree $T_i^{(t)}$ and cut edge
$e_i^{(t)}$ for particle $i$ depend only on particle $i$'s current state
and its deterministic per-particle seed.
No cross-particle communication is required until the ESS check (which reads
all weights) and the resampling step (which redistributes all particles).

The implementation uses Rayon's \texttt{par\_iter\_mut()} at the stage loop
level:
\begin{verbatim}
particles
    .par_iter_mut()
    .enumerate()
    .for_each(|(i, particle)| {
        let seed = particle_seed(base_seed, stage, i as u64);
        let mut rng = SmallRng::seed_from_u64(seed);
        propose_district(particle, &adj, &pop, tolerance, &mut rng,
                         &mut log_weights[i]);
    });
\end{verbatim}
Each thread independently runs the Wilson UST and valid-cut enumeration for
its assigned particle.
The speedup is approximately linear in the number of cores up to the
memory-bandwidth limit; on an 8-core workstation, we observe approximately
$6\times$ speedup relative to the single-threaded implementation.

\subsection{Kahan Summation for Weight Normalisation}

The final weight normalisation computes $\mathbf{w} = \mathrm{softmax}(\log
\tilde{\mathbf{w}})$ using Kahan compensated summation~\citep{delmoral2004}.
The standard implementation $w_i = \exp(\log \tilde{w}_i) / \sum_j
\exp(\log \tilde{w}_j)$ is numerically unstable when the range of log-weights
is large (which occurs for large $k$ when particles have very different weight
histories).

The Kahan implementation first shifts by the maximum log-weight:
\[
  \log M = \max_i \log \tilde{w}_i, \qquad
  w_i = \frac{\exp(\log \tilde{w}_i - \log M)}
             {\sum_j \exp(\log \tilde{w}_j - \log M)},
\]
then applies Kahan compensated summation to the denominator.
This bounds the relative error of the denominator at $O(\varepsilon_{\rm
machine})$ regardless of $N$, compared to $O(N \varepsilon_{\rm machine})$
for naive summation.

\subsection{Memory Layout}

District assignments are stored as \texttt{Vec<u8>} (one byte per tract,
value is district index $\in [k]$, with 0 reserved for unassigned).
This supports $k \leq 255$ districts, which covers all US congressional
district configurations (the maximum is California at $k = 52$).

For $N = 20{,}000$ particles on Texas ($n \approx 5{,}200$ tracts, $k = 38$
districts), the total memory for district assignments is
$20{,}000 \times 5{,}200 = 104{,}000{,}000$ bytes $\approx 100$ MB.
With the additional overhead of adjacency graphs, population vectors, and
RNG state, the total memory footprint is under 1 GB on a standard workstation.

\subsection{Particle Count Recommendations}

The particle count $N$ controls the Monte Carlo error of the \smc{} estimator.
The standard error of $\hat{f}_N$ scales as $1/\sqrt{\ess}$, where
$\ess \geq \tau \cdot N$ by the resampling rule.
For $\tau = 0.5$ (the default), we have $\ess \geq N/2$ at all times.

Based on our synthetic experiments (L1 test suite, 48 tests on graphs with
$k \leq 14$) and the NC L2 test, we recommend:

\begin{enumerate}
  \item $N = 1{,}000$ for states with $k \leq 5$ (AK, DE, MT, ND, SD, VT, WY).
        The \ess{} degradation is modest at small $k$; $N = 1{,}000$ gives
        $\mathrm{SE}(\hat{f}) \lesssim 0.03$ for typical district-level statistics.
  \item $N = 5{,}000$ for states with $6 \leq k \leq 20$ (NC, WI, and
        most mid-size states).
        This gives $\ess \geq 2{,}500$ and $\mathrm{SE}(\hat{f}) \lesssim 0.014$.
  \item $N = 20{,}000$ for states with $k \geq 30$ (CA, TX, NY, FL).
        The higher particle count is needed to maintain $\ess \geq 10{,}000$
        against the increased weight variance at large $k$.
\end{enumerate}

In our synthetic experiments on the L1 test suite, \smc{} with $N = 1{,}000$
on graphs with $k = 5$ produces estimated seat fractions within $0.02$ of the
analytic result (where computable by exhaustive enumeration on small instances).

\subsection{SmallRng (xoshiro128++) and Domain Separation}

Per-particle and per-resample RNG states use \texttt{SmallRng} (the
\texttt{xoshiro128++} algorithm in the \texttt{rand} crate), which passes the
PractRand statistical test suite and provides $2^{128}$ period --- sufficient
for any redistricting application.

The SHA-256 domain separation (the prefixes
\texttt{"SMC\_PARTICLE\_"} and \texttt{"SMC\_RESAMPLE\_"}) ensures that
per-particle and per-resample seeds are drawn from disjoint regions of the
seed space.
Without domain separation, a seed collision between particle $i$ at stage $t$
and a resample event at round $r$ would cause correlated randomness; with
domain separation, such collisions are computationally infeasible.

\subsection{NC 2020 L2 Test}

The \texttt{redist-smc} L2 test (\texttt{\#[ignore]}, run manually) exercises
the full SMC algorithm on North Carolina 2020 census data with $N = 200$
particles (for speed) and base seed 42.
In our experiments, the run completes in approximately 40 seconds on an 8-core
workstation and produces a valid weighted ensemble.
The \ess{} after the final stage is approximately $140$--$180$ (out of 200
particles), confirming that weight collapse is not severe for the NC 2020
graph at $N = 200$.

All quantitative claims in this section are from single-run experiments and
should not be over-interpreted.
Replication with larger $N$ and multiple seeds is deferred to Phase 2.
