\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}[proposition]
\newtheorem{remark}{Remark}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algpseudocode}

\title{GeoSection: Isoperimetrically-Normalised Ratio-Optimal\\
       Bisection for Congressional Redistricting}

\author{Gio Della-Libera}

\date{Draft --- \today}

\begin{document}

\maketitle

\begin{abstract}
Minimum-edge-cut recursive bisection produces compact congressional district maps
by repeatedly splitting the state's census-tract adjacency graph into
population-balanced halves.
Standard recursive bisection always uses the near-equal $\lfloor k/2\rfloor:\lceil k/2\rceil$ ratio.
We show this hides a pathology: when a state contains a compact metropolitan
core, the unnormalised edge-cut always favours the $1:(k-1)$ ratio
--- carving off one city district and leaving the rest of the state
to be bisected at the next level.
Iterated, this produces the \emph{caterpillar}: a linear chain of peeled
urban districts that looks like deliberate partisan packing even with
zero partisan input.

We introduce \textbf{GeoSection}, which adds two innovations to standard
recursive bisection.
First, at every bisection level, GeoSection tries all feasible split ratios
$1:(k-1)$ through $\lfloor k/2\rfloor:\lceil k/2\rceil$ with $N$ seeds each
and selects the ratio minimising the \emph{isoperimetrically normalised}
edge-cut $\mathrm{EC}/\sqrt{\min(i,\,k-i)}$.
The $\sqrt{\cdot}$ denominator is the isoperimetric correction: for a convex region,
the minimum boundary length to enclose fraction $i/k$ of the population
scales as $\sqrt{\min(i,k-i)/k}$, so dividing by $\sqrt{\min(i,k-i)}$
makes all ratios compete on equal footing.
Second, each subregion after a split recomputes its own principal axis
(PCA of tract centroids) and applies a directional penalty $\lambda$
to discourage cuts parallel to the new subregion's minor axis,
producing straighter district boundaries.

Applied to 45 states (2020 Census, 50 seeds per ratio), GeoSection
selects near-equal ratios for 28 states and genuinely asymmetric ratios
for 17 states where urban concentration is geographically real.
Wisconsin: $1:7$ (Milwaukee is genuinely compact, 85\,km boundary).
North Carolina: $6:8$ east/west split (Charlotte/Raleigh are not
compact enough to win under normalisation).
Tennessee: $1:8$ (Nashville/Memphis justify a peel).

A comparison against standard minimum-edge-cut bisection (B.7 baseline)
shows that GeoSection improves the proportionality gap by 1--2 seats in
states with strong geographic concentration and is neutral in states
with uniform density.
For North Carolina, GeoSection's $6:8$ split produces 5\,D/9\,R
($-13.6$\,pp proportionality gap); standard bisection produces 5\,D/9\,R
($-13.6$\,pp) as well.
Geographic voter sorting dominates the outcome under both algorithms.
\end{abstract}

\input{sections/01-introduction}
\input{sections/02-related-work}
\input{sections/03-algorithm}
\input{sections/04-evaluation}
\input{sections/05-discussion}
\input{sections/06-conclusion}

\bibliographystyle{plainnat}
\bibliography{references,../../references}

\end{document}
